---
title: "mfrmr Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{mfrmr Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
is_cran_check <- !isTRUE(as.logical(Sys.getenv("NOT_CRAN", "false")))
vignette_artifact <- function(name) {
  path <- system.file(
    "extdata", "vignette-artifacts", name,
    package = "mfrmr",
    mustWork = TRUE
  )
  read.csv(
    path,
    stringsAsFactors = FALSE,
    check.names = FALSE,
    na.strings = character(),
    colClasses = "character"
  )
}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  eval = !is_cran_check
)
```

## Quick start

This example estimates person abilities while accounting for rater severity
and criterion difficulty. A *facet* is a source of variation in scores; here,
`Rater` and `Criterion` are facets, and individual raters and criteria are their
*levels*. The data are synthetic, with one row per rating and scores from 1 to 4.

`head(toy)` shows the first six rows. The quoted column names in `fit_mfrm()`
are case-sensitive; `Study` and `Group` are extra labels unused by this model.
`MML` selects marginal maximum likelihood; `RSM` selects a rating-scale model
with shared category thresholds (the transitions between adjacent scores).

```{r quick-start}
# Load the package
library(mfrmr)

# Load example ratings and look at the first six rows
toy <- load_mfrmr_data("example_operational")
head(toy)

# Fit the model
fit <- fit_mfrm(
  data = toy,
  person = "Person",
  facets = c("Rater", "Criterion"),
  score = "Score",
  method = "MML",
  model = "RSM"
)

# Plot the results (Wright map)
plot(fit)

# Save the summary, then display its tables
results <- summary(fit)
results$person_overview # One row summarizing person ability estimates
results$facet_overview  # One row per facet: number of levels, mean, SD, range

# Check the interpretation status and recommended next step
results$decision
```

`<-` saves an object without printing it. Here `toy` holds the data, `fit`
holds the model, and `results` holds its summary. `$` selects a named part:
`results$person_overview` displays just that table. Enter `results` to print
the full summary.

The Wright map displays the estimates in logits, the model's measurement units,
rather than the original 1-to-4 scores. With this example's default orientation,
higher person estimates mean higher ability; higher rater estimates mean
stricter ratings, and higher criterion estimates mean greater difficulty.

The overview tables describe distributions: `person_overview` has one row for
all 48 persons, and `facet_overview` has one row for each facet. `Mean`/`MeanEstimate`
is the average, `SD`/`SDEstimate` is the spread, and `Min`/`Max` or
`MinEstimate`/`MaxEstimate` give the endpoints. Rater and criterion means are
constrained to zero here; their SDs and ranges show differences among levels.

### Inspect individual estimates

```{r quick-start-estimates}
estimates <- as.data.frame(fit)
head(subset(estimates, Facet == "Person")) # First six persons
subset(estimates, Facet == "Rater")       # All raters
subset(estimates, Facet == "Criterion")   # All criteria
```

`Facet` identifies the type of estimate, `Level` identifies the person, rater,
or criterion, and `Estimate` is its value in logits.

### Check what needs review

Read `results$decision`, especially `Why` and `NextAction`, before interpreting
or reporting estimates. `FormalInference = "No"` in this first summary can
mean that precision has not yet been reviewed; it does not necessarily mean
that fitting failed. The default summary does not compute diagnostics.

```{r quick-start-diagnostics}
diagnostics <- diagnose_mfrm(fit)
diagnostic_summary <- summary(diagnostics)
diagnostic_summary$decision
```

For your own data, follow [Use your own CSV](#use-your-own-csv) below.
For reporting, `res <- mfrm_results(fit, diagnostics = diagnostics)`
builds a comprehensive object that reuses the checks above. Pass `res` to
`mfrm_report()` or `export_mfrm_results()`; `results` remains the basic summary.

## Use your own CSV

The goal here is to estimate person ability while accounting for rater severity
and criterion difficulty in your own ratings. Start with the same row layout
as the packaged example: **one score from one rater for one person on one
criterion**. For example:

| Person | Rater | Criterion | Score |
| --- | --- | --- | --- |
| 001 | R1 | Content | 3 |
| 001 | R1 | Style | 2 |
| 001 | R2 | Content | 4 |
| 001 | R2 | Style | 3 |

The repeated `001` is intentional: all four scores belong to the same person.
Keep IDs consistent across rows. These four rows explain the layout; use the
complete set of ratings for estimation. `Score` contains the ordered integer
categories from the rubric, not totals or averages across raters. This example
uses two non-person facets; choose facets that represent your design.

### 1. Read and inspect the file

You can practice the full workflow without supplying a file. This block writes
the packaged synthetic ratings to a new temporary CSV:

```{r csv-practice-file}
library(mfrmr)
csv_path <- tempfile(fileext = ".csv")
write.csv(load_mfrmr_data("example_operational"), csv_path,
          row.names = FALSE, na = "")
```

For your own data, export your rating sheet as **CSV UTF-8**, with column names
in the first row. Skip the practice block above and select your file instead:

```{r csv-select-file, eval = FALSE}
library(mfrmr)
csv_path <- file.choose()
```

For a reusable script, replace `file.choose()` with a quoted path such as
`"data/ratings.csv"`, relative to the folder shown by `getwd()`.
Both routes now use the same import and analysis code:

```{r csv-read}
ratings <- read.csv(
  csv_path,
  colClasses = "character",
  na.strings = "",
  check.names = FALSE,
  fileEncoding = "UTF-8-BOM"
)
# Treat the documented missing-score marker only in the score column
ratings <- recode_missing_codes(ratings, columns = "Score", codes = "NA")
head(ratings)
names(ratings)
table(ratings$Score, useNA = "ifany")
```

Reading columns as text preserves distinct IDs such as `001`, `1`, and `NA`;
`mfrmr` converts numeric score strings such as `"3"` for estimation.
`check.names = FALSE` preserves the headers. Empty cells are missing in every
column; the literal marker `NA` is recoded only in `Score`. Use the
missing-score markers declared for your data. Zero is a score when the rubric
includes zero, and must not be used to fill unassigned rating cells.

The quoted column names in the next two steps must match `names(ratings)`,
including capitalization and spaces. You do not have to rename your file's
columns. For headers `Student`, `Judge`, `Task`, and `Rating`, change the
arguments in **both** `describe_mfrm_data()` and `fit_mfrm()` as follows:

| Argument | Four-column example | Your alternative headers |
| --- | --- | --- |
| `person` | `"Person"` | `"Student"` |
| `facets` | `c("Rater", "Criterion")` | `c("Judge", "Task")` |
| `score` | `"Score"` | `"Rating"` |

Also use `columns = "Rating"` in the recoding call and inspect `ratings$Rating`
instead of `ratings$Score`. Extra columns such
as the practice file's `Study` and `Group` are unused by this model.

### 2. Review the rows and the rubric

Set the minimum and maximum from the **rubric**, not the observed minimum and
maximum in this sample. The code below describes a 1-to-4 rubric. Keep the same
values in the data review and the fit. `keep_original = TRUE` preserves the
intended category structure, including categories with no observations.

```{r csv-data-review}
data_review <- describe_mfrm_data(
  data = ratings,
  person = "Person",
  facets = c("Rater", "Criterion"),
  score = "Score",
  rating_min = 1,
  rating_max = 4,
  keep_original = TRUE
)
data_review$row_retention
data_review$missing_by_column
data_review$score_distribution
data_review$design_connectivity
```

Read these four tables in order:

- `row_retention`: compare input and retained `Rows`; inspect any `DroppedRows`.
  A missing score or required ID removes that rating row, not automatically
  the person's other ratings. The package does not fill missing ratings.
- `missing_by_column`: locate missing input values. Non-numeric score text
  can also cause row loss; inspect warnings and `data_review$preparation_notes`
  when the retained count is unexpected.
- `score_distribution`: `RawN` counts the retained ratings in each category.
  A zero count in an internal category, such as 3 on a 1-to-4 scale, prevents
  fitting with `keep_original = TRUE`. Review the data and rubric before
  changing categories; increasing optimizer iterations cannot supply missing
  category information.
- `design_connectivity`: `Components = 1` means a facet's levels are connected
  through shared persons. More than one component needs design review before
  comparing levels across components. Connectivity alone does not establish
  full model identification.

For the practice CSV, all 282 input rows are retained, all four categories have
observations, and each facet has one connected component. Six planned ratings
are absent from the file; column missing counts cannot detect absent rows.
If you have a planned assignment roster, pass it as `expected_design` to
`describe_mfrm_data()` to assess these omissions. The applied example later
in this vignette shows how. Without a roster, structural missingness is
reported as not assessed; a complete person-by-rater-by-criterion crossing is
not assumed.

### 3. Fit and read the results

After resolving the data-review findings, fit the same rows, columns, and
rubric. This example uses an RSM with shared category thresholds; the choice
should reflect the scoring design. Diagnostics then supply evidence for the
interpretation decision.

```{r csv-fit}
csv_fit <- fit_mfrm(
  data = ratings,
  person = "Person",
  facets = c("Rater", "Criterion"),
  score = "Score",
  rating_min = 1,
  rating_max = 4,
  keep_original = TRUE,
  method = "MML",
  model = "RSM"
)
csv_diagnostics <- diagnose_mfrm(csv_fit)
csv_results <- summary(csv_fit, diagnostics = csv_diagnostics)
csv_results$decision
```

Read `Why` and `NextAction` before using the estimates. A `FormalInference`
value of `"No"` means that the stated obstacle to formal use still needs
attention. Diagnostic plots can help investigate it. Even a supported precision
decision does not establish the validity of the assessment or answer the
study's substantive question.

```{r csv-results}
plot(csv_fit)
csv_estimates <- as.data.frame(csv_fit)
head(subset(csv_estimates, Facet == "Person"))
subset(csv_estimates, Facet == "Rater")
subset(csv_estimates, Facet == "Criterion")
```

`Level` identifies the person, rater, or criterion; `Estimate` is the fitted
value in logits. With these defaults, higher person values mean higher
ability, higher rater values mean stricter ratings, and higher criterion values
mean greater difficulty. The rater and criterion estimates are centered at
zero within each facet. In the practice file, R06 is about +0.41 and R01 about
-0.61: R06 is estimated to rate more strictly after accounting for person
ability and criterion difficulty. This difference alone does not establish
statistical significance.

With custom facet names, change the `Facet` filters too, for example from
`"Rater"` to `"Judge"`; person rows retain `Facet == "Person"` even when the
input ID column is named `Student`. To inspect all persons, remove `head()`.
The overview tables in `csv_results` summarize distributions; the rows in
`csv_estimates` give individual estimates.

### If a check stops you

| What you see | What to check or change |
| --- | --- |
| The CSV is not found | Use `file.choose()` or check the quoted path relative to `getwd()`. An `.xlsx` workbook must first be exported as CSV. |
| All fields appear in one column, or the header is garbled | Check the separator and encoding. Use `sep = ";"` for a semicolon-delimited file; re-export as CSV UTF-8 if text is unreadable. |
| `Column(s) not found in data` | Compare `names(ratings)` with the names in `person`, `facets`, and `score`; check spelling, spaces, and case. |
| Non-numeric or fractional scores | Check `table(ratings$Score, useNA = "ifany")`. Map rubric labels explicitly to their ordered integer codes. Do not round totals or averaged ratings to make them fit. |
| Scores outside the supplied range, such as 99 | Check whether the value is an error or a documented missing code. Correct the value or recode the declared missing marker; keep the range tied to the rubric. |
| Fewer retained rows than expected | Inspect `data_review$missing_by_column` and `data_review$preparation_notes`. Correct unintended missing IDs or score values, then rerun the review and fit. |
| Duplicated person-by-facet cells | Inspect `data_review$duplicate_cell_detail`. Correct accidental duplicates. If these are distinct tasks or occasions, include the corresponding facet; do not automatically delete legitimate ratings. |
| A category has no observations, or the design is disconnected | Review category coding and assignments. Do not invent ratings or change optimizer settings to resolve missing information. |

If your data documentation defines `99` and `.` as missing **scores**, recode
only that column before step 2. This preserves an ID such as `99`:

```{r csv-missing-codes, eval = FALSE}
ratings <- recode_missing_codes(
  ratings, columns = "Score", codes = c("99", ".")
)
table(ratings$Score, useNA = "ifany")
```

Use only the codes declared for your data, and change `"Score"` if your score
column has another name. Then repeat steps 2 and 3 with the recoded data.

### If criteria occupy separate columns

For a sheet with one row per person-rater pair and separate criterion scores,
reshape the criterion columns into rating rows:

```{r csv-wide-to-long}
wide <- data.frame(
  Person = c("001", "001"),
  Rater = c("R1", "R2"),
  Content = c(3, 4),
  Style = c(2, 3)
)
ratings_long <- tidyr::pivot_longer(
  wide,
  cols = c("Content", "Style"),
  names_to = "Criterion",
  values_to = "Score"
)
ratings_long
```

The output is the four-row layout at the start of this section. For your full
sheet, read the CSV into `wide`, select its actual criterion columns in `cols`,
and use `ratings <- ratings_long` before step 2. Keep the complete dataset;
these four demonstration rows are not sufficient for the intended analysis.

## Continue with diagnostics and reporting

The following sections reuse the model and diagnostics from **Quick start**.
They ask three questions: are the estimates interpretable, what needs closer
inspection, and what information belongs in a report? For your own analysis,
use `csv_fit` and `csv_diagnostics` in these same helper calls and review your
own assignment roster if one exists.

### 1. Check the data and estimation record

```{r toy-setup}
fit_toy <- fit
diag_toy <- diagnostics

# The packaged roster declares which ratings were planned
data("mfrmr_example_operational_design", package = "mfrmr")
data_review_toy <- describe_mfrm_data(
  data = toy,
  person = "Person",
  facets = c("Rater", "Criterion"),
  score = "Score",
  rating_min = 1,
  rating_max = 4,
  keep_original = TRUE,
  expected_design = mfrmr_example_operational_design
)
data_summary_toy <- summary(data_review_toy)
data_summary_toy$structural_missingness
data_summary_toy$design_connectivity

fit_summary_toy <- summary(fit_toy)
fit_summary_toy$overview[, c("Model", "Method", "N", "Persons", "Converged")]
fit_summary_toy$decision
```

The example has 282 observed ratings from 48 persons, six raters, and three
criteria. All observed rows are retained. Six additional ratings were planned
but are absent; they are visible only because the roster was supplied.
Distinguish omitted planned ratings from input rows excluded during cleaning.
The package does not assume that every rater scored every person.

Record the estimator, model, score coding, constraints, and numerical settings.
For this RSM, MML uses a standard-normal person distribution; reported person
scores are EAP estimates, and non-person facets are fixed effects centered
within facet. RSM shares category thresholds. Choose PCM with an explicit
`step_facet` when the rubric and question call for separate thresholds.

The default MML grid is 31 quadrature points. A comparison that could change
the study's conclusion needs a prespecified denser common-grid sensitivity
check with `mml_quadrature_sensitivity()`. A temporary 7-point grid is only a
computational screening run. `maxit` is a computational ceiling, not a
convergence criterion; inspect `ConvergenceStatus` and `Numerical` before
increasing it. Keep the same data, model, method, and planned controls when
investigating an iteration limit. Choose JML for its fixed-person estimand
when appropriate, not as a faster substitute for MML.

```{r toy-setup-artifacts, echo = FALSE, eval = is_cran_check}
vignette_artifact("workflow_fit_decision.csv")
vignette_artifact("workflow_fit_overview.csv")
```

### 2. Read precision and model fit before interpreting differences

```{r precision-and-fit}
diagnostic_summary_toy <- summary(diag_toy)
diagnostic_summary_toy$decision
diagnostic_summary_toy$overall_fit

precision_toy <- precision_review_report(fit_toy, diagnostics = diag_toy)
precision_toy$profile[, c(
  "PrecisionTier", "SupportsFormalInference", "PersonSEBasis", "NonPersonSEBasis"
)]
precision_toy$checks

# View the fitted locations with their uncertainty
plot(fit_toy, diagnostics = diag_toy, show_ci = TRUE)
```

The fit-only summary reports `FormalInference = "No"` until precision evidence
is supplied. Read the diagnostic decision's `Why` and `NextAction` together
with the precision checks. A `model_based` label alone does not override a
failed fit-readiness component or justify a particular substantive claim.

Infit and Outfit describe departures from the fitted response model; their
reference value is 1. A global average near 1 can coexist with local problems.
In this example the MML person uncertainty is posterior SD, while the
non-person SEs use observed information. Report the basis and interval method
when presenting uncertainty; do not call every interval a posterior quantile
interval or treat a point difference as a significance test.

```{r diagnostic-overview-artifact, echo = FALSE, eval = is_cran_check}
vignette_artifact("workflow_diagnostic_overview.csv")
```

### 3. Identify what needs closer inspection

**Are the rating categories working as intended?** Examine category use and
the transitions between scores on the declared rubric:

```{r category-review}
scale_toy <- rating_scale_table(fit_toy, diagnostics = diag_toy)
scale_toy$category_table[, c("Category", "Count", "AvgPersonMeasure", "Infit", "Outfit")]
scale_toy$threshold_table
plot(scale_toy)
```

The bars show observed category counts; the line shows fitted expected counts.
Look for rarely used categories, changes in average person measure across
categories, and threshold order. Explain the observed pattern in terms of the
rubric. A warning is a reason to investigate, not an automatic instruction to
merge categories.

**Which ratings differ most from model expectations?** This example uses an
explicit exploratory rule and shows up to ten cases:

```{r diagnostics-reporting}
unexpected_toy <- unexpected_response_table(
  fit_toy,
  diagnostics = diag_toy,
  abs_z_min = 1.5,
  prob_max = 0.4,
  top_n = 10
)
unexpected_toy$summary
unexpected_toy$table[, c("Person", "Rater", "Criterion", "Observed", "Expected", "StdResidual")]
plot(unexpected_toy)
```

Read the rule in `$summary` with the residuals. The displayed cases are a
selected preview; `top_n` limits the table, while the summary counts all
flagged ratings. Here this broad rule flags 141 of 282 ratings (50%), although
only ten cases are displayed. Report the rule with the count and percentage.
Inspect the original work and scoring context before changing a
rating or excluding a person. A flag does not identify its cause.

The default diagnostics retain the residual/EAP path and the latent-integrated
marginal screening path. The latter includes category and pairwise agreement
gaps. Keep their bases separate when reporting:

```{r diagnostic-bases}
diagnostic_summary_toy$diagnostic_basis[, c("DiagnosticPath", "Status", "ReportingUse")]
plot_marginal_fit(fit_toy, diagnostics = diag_toy)
```

Follow a pairwise warning with `plot_marginal_pairwise()`. Add residual PCA
only when residual structure is relevant to the question, using
`diagnose_mfrm(fit_toy, residual_pca = "both")`. Report the mode and limits of
the screen; residual PCA alone does not establish unidimensionality. See
`vignette("mfrmr-mml-and-marginal-fit", package = "mfrmr")` and
`vignette("mfrmr-visual-diagnostics", package = "mfrmr")` for these follow-ups.

**What should I discuss with a rater?** Start from the coverage and category
checks above, then view severity and its uncertainty alongside the screening
results. Reuse the same fit and diagnostics:

```{r rater-feedback, fig.alt = "Rater severity estimates with 95% normal intervals on the fitted logit scale. Positive estimates indicate stricter ratings relative to the fitted reference."}
plot_rater_severity_profile(
  fit_toy, diagnostics = diag_toy, facet = "Rater", show_bands = FALSE
)
rater_review <- facet_quality_dashboard(
  fit_toy, diagnostics = diag_toy, facet = "Rater"
)
knitr::kable(
  rater_review$detail[, c("Level", "N", "Estimate", "SE", "Infit", "Outfit")],
  digits = 3, row.names = FALSE, caption = "Rater estimates and fit"
)
knitr::kable(
  rater_review$detail[, c("Level", "MissingMetrics", "SeverityFlag", "MisfitFlag")],
  row.names = FALSE, caption = "Availability and screening flags"
)
rater_review$settings
writeLines(strwrap(rater_review$notes, width = 72))
```

A stricter rater is not necessarily inconsistent or incorrect. Individual
interval overlap is not a test of the difference between two raters. Read
`MissingMetrics` before interpreting flags: zero observed flags is not a
complete pass when diagnostics are unavailable. A `REVIEW ONLY` label retains
the fit's restrictions; the plot does not override them.

In this synthetic example, none of the six raters has a severity or misfit
flag under the displayed settings. The earlier response-level screen still
selected 141 ratings: the two screens assess different units and use different
rules. Neither result establishes that every rating is appropriate.

Use the selected unexpected ratings to discuss the rubric and scoring context,
not to automatically exclude a rater. Limited overlap may call for more shared
ratings before comparing raters. A change after training alone does not
establish a training effect. For your own analysis, use `csv_fit` and
`csv_diagnostics`, and replace `"Rater"` with your actual facet name.
Keep the whole dashboard with `saveRDS()` to retain its settings and notes;
`export_mfrm_bundle(..., include = c("dashboard", "html"))` includes them
with the exported tables. See `?facet_quality_dashboard` for that export route.

### 4. Assemble the evidence for a report

```{r primary-route}
res_toy <- mfrm_results(fit_toy, diagnostics = diag_toy, include = "publication")
report_toy <- mfrm_report(res_toy, style = "apa")
report_toy$first_screen[, c("Area", "Status", "MainIssue", "NextAction")]

checklist_toy <- reporting_checklist(fit_toy, diagnostics = diag_toy)
subset(checklist_toy$checklist, !DraftReady,
       c("Section", "Item", "NextAction"))
```

`DraftReady` describes available drafting material with caveats. An unrequested
bias or linking analysis may be irrelevant to the question; do not run every
helper merely to turn every flag `TRUE`. The report's gaps concern the supplied
analysis objects. They cannot confirm recruitment, rater training, ethics,
the study rationale, or whether the prose answers the research question.

For a journal article, the companion
`vignette("mfrmr-reporting-and-apa", package = "mfrmr")` provides a
**Manuscript coverage map**: what to report, where the numerical evidence is,
and what the author must supply. It also shows estimates with uncertainty,
reliability and agreement with distinct interpretations, and a worked
question-to-result explanation.

```{r report-overview-artifact, echo = FALSE, eval = is_cran_check}
vignette_artifact("workflow_next_actions.csv")
```

### 5. Display a table and save the analysis

```{r report-table}
measurements_toy <- fit_measures_table(fit_toy, diagnostics = diag_toy)
rater_table_toy <- subset(
  measurements_toy$table, Facet == "Rater",
  c("Level", "N", "Measure", "SE", "CI_Lower", "CI_Upper", "Infit", "Outfit")
)
apa_table(rater_table_toy, digits = 3,
          caption = "Rater severity and fit",
          note = "N counts rating rows. Rater effects are centered within facet and measured in logits; higher values indicate stricter ratings. Intervals are 95% normal approximations from MML observed-information SEs. Infit and Outfit have reference value 1.")
```

Report the score scale and direction with the table. For person results,
state the EAP/posterior-SD basis; for rater results, explain that variation in
severity is different from inconsistent scoring. Add a sentence answering
the study's question instead of repeating every cell of the table.

```{r report-export}
# A new temporary directory keeps this synthetic example repeatable
export_dir <- tempfile("mfrmr-workflow-export-")
export_toy <- export_mfrm_results(
  res_toy,
  output_dir = export_dir,
  include = c("default", "report"),
  acknowledge_sensitive = TRUE
)
export_preview <- head(export_toy$written_files[, c("Component", "Format", "Path")])
export_preview$Path <- basename(export_preview$Path)
knitr::kable(export_preview, row.names = FALSE,
             caption = "First six exported files (filenames only)")
```

This archive retains the collected results. `acknowledge_sensitive = TRUE` is used
because the example contains synthetic data; it does not remove identifiers
or local paths from real analyses. Review the files before sharing them.
The table previews six filenames; `export_toy$written_files` retains the full
list, and its `Path` column locates the generated files. Temporary
files are for practice; choose a study directory for an archive you will keep.
With a precomputed fit as input, this archive's replay code assumes that
`fit` and `diagnostics` already exist. Keep your original data-to-fit script.
For a fit replay including its input CSV, and for saving a manuscript table,
figure, and their notes, follow Section 5 of
`vignette("mfrmr-reporting-and-apa", package = "mfrmr")`.

```{r export-artifact, echo = FALSE, eval = is_cran_check}
vignette_artifact("workflow_export_files.csv")
```

### Further analyses should answer a stated question

| Question | Follow-up |
| --- | --- |
| Do category thresholds need to differ by criterion? | `compare_mfrm()` and `mml_quadrature_sensitivity()`; fit candidates to the same observations and review numerical sensitivity. |
| Do specific rater-by-criterion or group contrasts depart from the model? | `estimate_bias()` or `analyze_dff()`; specify the contrasts, screening rules, and multiplicity plan. See the reporting vignette and `mfrmr-linking-and-dff`. |
| Would the conclusions change under a bounded GPCM? | `gpcm_capability_matrix()` and `mfrmr-gpcm-scope`; explain what discrimination reweighting means for the score interpretation. |
| Are forms or waves on a comparable scale? | `mfrmr-linking-and-dff` and `mfrmr-portable-calibration`; common labels alone do not establish linking. |
| How does a proposed rating design perform under stated assumptions? | `build_mfrm_sim_spec()`, `evaluate_mfrm_recovery()`, and `assess_mfrm_recovery()`; report generating conditions, repetitions, failures, and Monte Carlo uncertainty. |
| How should new persons be scored under an existing calibration? | `predict_mfrm_units()` and `sample_mfrm_plausible_values()`; reuse an eligible fitted calibration and report the conditioning assumptions. |

Simulation examples with one or two repetitions test the computational setup;
they do not estimate operating characteristics adequately for a paper. A
simulation's assumptions and results also do not replace evidence from the
observed assessment. Keep these analyses in the report only when their role
in answering the study question is clear.

## An observed writing-assessment example

The optional `sirt` package supplies `data.ratings1`, observed German-writing
ratings from a 2009 Austrian educational standards survey. Its 274 rows contain
135 students, seven observed raters and five rubric criteria scored 0 to 3.
The [data documentation](https://alexanderrobitzsch.r-universe.dev/sirt/doc/manual.html#data.ratings)
describes the source. This dataset is separate from mfrmr's synthetic examples.

The practical question is which differences in severity, category use or
consistency merit discussion with a rater. The example below uses a PCM so
criteria can have different category thresholds. It assumes a common writing
scale and conditionally independent ratings; inspecting the output does not
establish those assumptions. Repeated rubric ratings of one performance may
still share influences that this model does not represent.

Run this optional example after installing `sirt`. It is not run when building
this vignette. Display aliases replace source identifiers in the example's
tables; aliases do not make a rating archive anonymous.

```{r observed-writing-example, eval = FALSE}
library(mfrmr)
data("data.ratings1", package = "sirt")
wide <- data.ratings1
persons <- match(wide$idstud, unique(wide$idstud))
raters <- match(wide$rater, unique(wide$rater))
writing <- do.call(rbind, lapply(paste0("k", 1:5), function(criterion) {
  data.frame(
    Person = sprintf("P%03d", persons), Rater = paste0("R", raters),
    Criterion = criterion, Score = wide[[criterion]]
  )
}))
review <- describe_mfrm_data(
  writing, person = "Person", facets = c("Rater", "Criterion"), score = "Score",
  rating_min = 0, rating_max = 3
)
review$row_retention
review$design_connectivity

writing_fit <- fit_mfrm(
  writing, "Person", c("Rater", "Criterion"), "Score",
  method = "MML", model = "PCM", step_facet = "Criterion",
  rating_min = 0, rating_max = 3, quad_points = 61, maxit = 400
)
writing_diagnostics <- diagnose_mfrm(writing_fit, residual_pca = "none")
feedback <- facet_quality_dashboard(
  writing_fit, diagnostics = writing_diagnostics, facet = "Rater"
)
feedback$detail[, c("Level", "N", "Estimate", "SE", "Infit", "Outfit", "MisfitFlag")]
feedback$settings
plot_rater_severity_profile(
  writing_fit, diagnostics = writing_diagnostics, facet = "Rater", show_bands = FALSE
)
saveRDS(feedback, "writing-feedback.rds")
archive <- export_mfrm_bundle(
  writing_fit, diagnostics = writing_diagnostics, facet = "Rater",
  output_dir = "writing-analysis", data = writing, acknowledge_sensitive = TRUE
)
```

Reshaping gives 1,370 criterion ratings; the 274 original rows are not the
number of students. Each rater has 185 to 205 criterion ratings here. The
example produces a low-Infit flag for R7: unusually predictable ratings also
trigger this screen. This is a prompt to inspect rubric use and individual
ratings, not evidence that R7 is inaccurate or should be removed. Severity
and consistency answer different questions, and this dataset supplies no
known true rater quality against which to validate either interpretation.

The quadrature setting belongs to this example; use
`mml_quadrature_sensitivity()` when numerical sensitivity matters to a
comparison. Preserve the dashboard's notes, review rating-category use and
common-person coverage, and use the actual scripts or rubric descriptors in
feedback discussions. A different-looking result alone does not establish
that a rater needs training or that training would improve the scores.
