---
title: "mfrmr Reporting and APA"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{mfrmr Reporting and APA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
is_cran_check <- !isTRUE(as.logical(Sys.getenv("NOT_CRAN", "false")))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  eval = !is_cran_check
)
```

A useful report explains the question, why the scoring design and model can
answer it, what the results show, and how far the answer applies. This guide
connects those steps to actual `mfrmr` tables and figures. The
[Manuscript coverage map](#manuscript-coverage-map) also identifies information
that the researcher must provide outside the fitted model.

The running question is: **after accounting for person ability and criterion
difficulty, how much do the raters differ in severity?** The synthetic data
illustrate the reporting workflow; they are not evidence about a real
assessment or population. For loading your own CSV, see
`vignette("mfrmr-workflow", package = "mfrmr")`.

## 1. Fit once and retain the analysis record

```{r setup}
# Load the package and synthetic ratings
library(mfrmr)
toy <- load_mfrmr_data("example_operational")
head(toy)

# Estimate person ability, rater severity, and criterion difficulty
fit <- fit_mfrm(
  data = toy,
  person = "Person",
  facets = c("Rater", "Criterion"),
  score = "Score",
  method = "MML",
  model = "RSM"
)

# Reuse these diagnostics in the tables and report below
diag <- diagnose_mfrm(fit, diagnostic_mode = "both", residual_pca = "none")
results <- summary(fit, diagnostics = diag)
```

One row is one rating. The rubric uses **1 to 4**, and all four categories
occur. The example has 48 persons, six raters, and three criteria, with 282
observed ratings. Here `N` is the number of rating rows, not the number of
independent participants. Repeated ratings from the same person share that
person's latent ability in the model.

MML means marginal maximum likelihood; RSM is a rating-scale model with
shared category thresholds. In this default fit, the person distribution is
fixed to standard normal, person scores are expected a posteriori (EAP)
estimates, and rater and criterion effects are fixed parameters centered
within each facet. Higher rater estimates mean stricter ratings, in logits.
In a real study, explain why this population assumption and shared-threshold
rubric match the question. Record any different estimator, constraints,
anchors, weights, interactions, or score orientation.

```{r analysis-record}
results$overview[, c("Model", "Method", "N", "Persons", "Converged")]
results$settings_overview[, c(
  "QuadPoints", "RatingMin", "RatingMax", "RatingRangeSource", "NoncenterFacet"
)]
results$row_retention
results$decision

# Record the iteration limit, integration grid, and numerical stopping details
spec <- specifications_report(fit)
spec$convergence_control
results$overview[, c("OptimizerMethod", "ConvergenceStatus", "RequestedReltol", "EffectiveReltol")]
```

The default numerical integration uses 31 quadrature points. Retain the
analysis script and recorded controls with the results; convergence alone
does not establish numerical stability or model adequacy. If the conclusions
hinge on a close model comparison, use `mml_quadrature_sensitivity()` with a
prespecified denser common grid and report whether the conclusion changes.

### Show what was planned, observed, and excluded

```{r rating-design}
data("mfrmr_example_operational_design", package = "mfrmr")
data_review <- describe_mfrm_data(
  data = toy,
  person = "Person",
  facets = c("Rater", "Criterion"),
  score = "Score",
  rating_min = 1,
  rating_max = 4,
  keep_original = TRUE,
  expected_design = mfrmr_example_operational_design
)
data_summary <- summary(data_review)
data_summary$missing
data_summary$structural_missingness
data_summary$design_connectivity

# Describe the scores and the amount of rating evidence
data_summary$score_distribution
data_summary$facet_overview
data.frame(MeanScore = mean(toy$Score), SDScore = sd(toy$Score))
table(table(toy$Person)) # Number of persons with each observed rating count
as.data.frame(results$person_overview[, c(
  "Persons", "DistributionN", "ReviewExcludedExtremeEAPs", "Mean", "SD", "MeanPosteriorSD"
)])
```

The supplied roster contains 288 planned rating cells, of which six are
absent. All 282 input rows are retained in fitting. These are different
counts: absent planned ratings are not rows discarded by `fit_mfrm()`.
Report the assignment scheme, overlap, exclusions, and reasons for missing
ratings. Without a roster, do not label every unobserved person-by-rater
combination as a missing planned rating. A connected design is relevant to
comparability, but does not establish anchor invariance or sufficient precision.

These counts describe different units. In the example, 42 persons have six
ratings and six have five; rater workloads range from 38 to 56 rating rows.
For repeated testing, state whether a person ID identifies a person, a response,
or an occasion, and how dependence across occasions is handled. Preserve the
original rubric, any scoring exceptions, and the exact recoding rule. Such
details affect what the model can answer: Goodwin (2016, Sections 3–4 and
Table 1) documents score-use contexts, rating history, exclusions, and category
recoding before reporting rater differences.

State which persons enter distribution summaries. Eckes (2005, Tables 1–2)
marks summaries restricted to nonextreme examinees. In this example the person
summary includes all 48 persons. `SD` describes variation across person EAP
estimates; `MeanPosteriorSD` describes their average within-person uncertainty.
They are not interchangeable, and neither is a rater or criterion SE.

## 2. Check uncertainty and fit before writing the answer

```{r precision}
prec <- precision_review_report(fit, diagnostics = diag)
prec$profile[, c(
  "PrecisionTier", "SupportsFormalInference", "PersonSEBasis", "NonPersonSEBasis"
)]
prec$checks

# These diagnostics address departures from the fitted response model
diag_summary <- summary(diag)
diag_summary$overall_fit
diag_summary$diagnostic_basis[, c("DiagnosticPath", "Status", "ReportingUse")]
```

Read `results$decision`, including `Why` and `NextAction`, together with these
checks. A `model_based` precision tier describes the uncertainty calculation;
it does not by itself justify every inferential claim. Retain restrictions
reported for the fit, design, or particular statistic. For this fit, person
uncertainty is posterior SD and non-person SEs use observed information.

Infit and Outfit have a reference value of 1 and summarize departures from the
model. Report the statistics and the source and purpose of any review cutoff;
a flag is not an automatic exclusion rule. Global fit can hide local problems.
The default `diagnostic_mode = "both"` retains residual/EAP diagnostics and
latent-integrated marginal screens. Their reference bases differ; inspect
`prec$fit_separation_basis` before describing MnSq, ZSTD, separation, or
reliability as if they were a single adequacy test.

When reporting the frequency of large residuals, state the rule and denominator.
For example, Eckes (2005, p. 204) reports counts at two absolute standardized-
residual cutoffs. The following shows those descriptive counts for the current
EAP-residual path:

```{r residual-counts}
z <- diag$obs$StdResidual
residual_counts <- data.frame(
  AbsoluteZCutoff = c(2, 3),
  Evaluated = sum(is.finite(z)),
  Unavailable = sum(!is.finite(z)),
  Flagged = c(sum(abs(z[is.finite(z)]) >= 2),
              sum(abs(z[is.finite(z)]) >= 3))
)
residual_counts$Percent <- 100 * residual_counts$Flagged / residual_counts$Evaluated
residual_counts
```

Use the residual basis appropriate to your analysis. These are descriptive
tail counts, not a calibrated global test or a universal rule that a certain
percentage establishes fit. The observed-category probability screen in
`unexpected_response_table()` is an additional rule unless explicitly disabled;
its default count is not the same as counting only `abs(StdResidual) >= 2`.

Keep the marginal screening results visible alongside those counts:

```{r marginal-reporting}
diag_summary$marginal_fit[, c(
  "OverallRMSD", "StepGroupsFlagged", "FacetLevelsFlagged",
  "PairwiseFlaggedLevelPairs", "ReportingUse"
)]
diag$marginal_fit$thresholds
diag$marginal_fit$pairwise$thresholds
subset(diag$marginal_fit$facet_level$summary_stats, Flagged,
       c("Facet", "Level", "GroupCount", "RMSD", "MaxAbsStdResidual"))
subset(diag$marginal_fit$pairwise$pair_stats, Flagged,
       c("Facet", "Level1", "Level2", "LevelPairCount", "ExactAgreement", "ExpectedExactAgreement"))
```

The package's default marginal rules use an absolute standardized-residual
cutoff of 2 or a root-mean-square proportion discrepancy of 0.05; pairwise
screens also use an exact/adjacent agreement-gap cutoff of 0.10. Record the
rules used in your run and inspect flagged groups with `plot_marginal_fit()`
and `plot_marginal_pairwise()`. These remain screening results; an acceptable
overall average does not resolve local warnings.

## 3. Show estimates, uncertainty, and the answer

```{r rater-estimates}
measurements <- fit_measures_table(fit, diagnostics = diag, ci_level = 0.95)
rater_table <- subset(
  measurements$table, Facet == "Rater",
  c("Level", "N", "Measure", "SE", "CI_Lower", "CI_Upper", "Infit", "Outfit")
)
rater_table <- rater_table[order(rater_table$Measure), ]

rater_apa <- apa_table(
  rater_table,
  digits = 3,
  caption = "Rater severity and fit in the synthetic MML/RSM example",
  note = paste(
    "282 ratings of 48 persons by six raters on three criteria using a 1-4 rubric.",
    "N counts rating rows. Measures and SEs are in logits; higher values mean stricter ratings.",
    "Intervals are 95% normal approximations from observed-information SEs.",
    "Rater effects are centered within facet. Infit and Outfit have reference value 1."
  )
)
rater_apa

# Display the estimates on the common logit scale
plot(fit, diagnostics = diag, show_ci = TRUE)
```

Pass the same `diag` to the table and plot so their intervals use the same
SEs. Without `diagnostics`, Wright-map intervals use exploratory
observation-table SEs, which can differ from the MML observed-information
SEs reported here. Steps on this map are point estimates; their uncertainty
is reported in the separate threshold table below.

`apa_table()` prints the table, caption, and note; its `$table` member holds
the data for further formatting. `digits` rounds all numeric columns
uniformly. Review statistic-specific rounding, especially for *p* values and
intervals, and the target journal's format before submission.

A worked explanation for this synthetic example is:

> We asked how much raters differed in severity after accounting for person
> ability and criterion difficulty. The model includes those facets so that
> raw rater averages are not the only basis for comparison. Estimated rater
> severity ranged from -0.606 logits (R01; SE = 0.224, 95% CI [-1.046, -0.166])
> to 0.412 logits (R06; SE = 0.249, 95% CI [-0.077, 0.901]), a span of about
> 1.018 logits. Thus, the fitted model describes R06 as stricter than R01 in
> these data. This span is descriptive: it is not an interval or a hypothesis
> test for the difference. The evidence does not establish why the raters
> differ or how much a particular examinee's decision would change.

The intervals compare each centered rater effect with its reference; their
overlap is not a pairwise test. A contrast needs uncertainty for that contrast,
including covariance, and a multiplicity plan if several comparisons are
made. Whether a difference matters in practice depends on the rubric, the
score use, and any decision rule. Do not convert a logit span directly into
rubric points or claim that severity differences prove biased or inconsistent
scoring. For person tables, state the EAP/posterior-SD basis and the interval
method explicitly; normal approximations are not posterior quantile intervals.

### Keep separation, agreement, and category functioning distinct

```{r reliability-and-agreement}
apa_table(
  diag_summary$reliability[, c("Facet", "Levels", "Separation", "Strata", "Reliability", "RealReliability")],
  digits = 3,
  caption = "Separation and reliability by facet",
  note = "Interpret the facet and error basis explicitly; rater separation reliability is not interrater agreement."
)

agreement <- interrater_agreement_table(fit, diagnostics = diag)
agreement$summary[, c(
  "OpportunityCount", "ExactAgreement", "ExpectedExactAgreement", "AdjacentAgreement"
)]
```

A rater separation reliability describes how distinctly rater locations are
estimated relative to their errors; a high value does not mean that raters
agree. Person separation reliability addresses differentiation among persons
under its stated error basis. Specify whether the model or misfit-adjusted
(`RealReliability`) index is reported, and the replication/error sources
relevant to the intended score use. These coefficients do not describe every
possible source of measurement error.

Also identify the formula. In `diag$reliability`, `Reliability` is the
separation-based ratio `max(V - E, 0) / V`, where `V` is the sample variance
of finite facet estimates and `E` is the mean squared model SE over available
SEs. The ratio is unavailable when `V` is zero or unavailable. `Separation`
is `sqrt(max(V - E, 0) / E)` when defined, and `Strata` is
`(4 * Separation + 1) / 3`; they are distinct indices, not observed groups.
The underlying quantities are available directly:

```{r reliability-basis}
diag$reliability[, c("Facet", "Levels", "ObservedVariance", "ModelErrorVariance", "Reliability")]
```

This is not the posterior-variance EAP reliability reported by Xiao et al.
(2026, Section 2.5, Equation 4), whose denominator combines the spread of EAP
scores with posterior variances. Nor is it the squared correlation with known
true abilities used in their simulation. Report the package's coefficient
under its own name and basis; numerical comparisons across papers require
matching definitions and scoring designs. For example, this analysis's person
separation reliability is 0.664, not a TAM EAP-reliability estimate.

Here the observed exact agreement is 54/138 = 39.1% across available paired
ratings of the same person and criterion; adjacent agreement (within one
category) is 85.5%. `ExpectedExactAgreement` is the fitted-model baseline
(35.1% here), not an independent chance-agreement correction. Report the
opportunity count and overlap pattern, not just a percentage. These are
observed-score summaries; keep them separate from the marginal diagnostic
screens above. Use `agreement$pairs` when differences among rater pairs matter.

```{r categories}
scale <- rating_scale_table(fit, diagnostics = diag)
scale$category_table[, c("Category", "Count", "AvgPersonMeasure", "Infit", "Outfit")]
scale$threshold_table
plot(scale)
```

For a threshold table with uncertainty, use the diagnostics already computed:

```{r threshold-uncertainty}
steps <- diag$parameter_uncertainty$steps
steps[, c("Step", "Estimate", "SE", "CI_Lower", "CI_Upper", "CI_Level")]
steps[, c("Step", "SE_Status", "CIEligible", "CIUse")]
```

`Step_1`, `Step_2`, and `Step_3` are the transitions 1–2, 2–3, and 3–4 here.
The table uses MML observed-information SEs and 95% normal intervals. A bare
fit's `steps` table supplies point estimates; `rating_scale_table()` does not
automatically add the uncertainty from a separately supplied diagnostics
object. Check its availability and reporting status before quoting intervals.
For PCM, retain `StepFacet` to identify the threshold family. Do not substitute
a criterion-location SE for a threshold SE, or a centered step for the combined
criterion-plus-step location. These distinctions matter when preparing tables
such as the criterion-specific thresholds and SEs in Eckes (2005, Table 5).

The bars show observed category counts; the line shows fitted expected counts.
Report category frequencies, threshold estimates and ordering, and any sparse
or unused categories. Explain what the pattern means for the rubric. These
checks can motivate rubric review; they do not automatically justify merging
categories, changing scores, or claiming unidimensionality.

### If the question concerns observed versus adjusted scores

Eckes (2005, Table 3) places observed scores, adjusted scores, logit measures,
SEs, and rating counts together. State the adjustment reference when using a
similar presentation. Here the reference is zero for the other facets:

```{r observed-and-adjusted}
fair <- fair_average_table(fit, diagnostics = diag, reference = "zero")
person_scores <- fair$raw_by_facet$Person
person_scores <- person_scores[order(person_scores$Level), ]
head(person_scores[, c(
  "Level", "TotalCount", "ObservedAverage", "FairZ", "Measure", "ModelSE"
)], 6)
```

These are the first six person IDs, selected for illustration. `FairZ` is an
expected score on the fitted 1–4 scale, at the person's estimated ability with
the other facet references set to zero. It is not a z-score. `ModelSE` here is
the posterior SD of person ability in logits, not the SE of `FairZ`. The
RSM/PCM `fair_average_table()` does not supply fair-score SEs. Its mean-reference
alternative, `FairM`, uses mean other-facet measures; for non-person rows that
also includes the mean person measure. Neither averages over the actual
assignment distribution.

A score-use claim needs the decision thresholds, rounding rule, reference
environment, uncertainty treatment, and full-sample counts of changed
decisions. The six-row preview is not that analysis. Statistical adjustment
alone does not establish fairness or authorize replacing operational scores.

## A short Methods and Results example

The following is an original write-up of this **synthetic illustration**,
using the outputs above. It is not a reconstruction of any cited study. The
research questions determine which of these supporting results belong in a
real article's main text or appendix.

**Methods.** We examined how much raters differed in severity after accounting
for person ability and criterion difficulty. The synthetic dataset represented
48 persons, six raters, and three criteria scored from 1 to 4. Each person was
assigned two raters across three criteria in an incomplete but connected
design. Of 288 planned ratings, six were deliberately omitted in the example;
all 282 observed ratings were retained. The raw rating mean was 2.38
(SD = 1.00), and individual raters contributed 38–56 ratings.

We fitted an RSM with person, rater, and criterion effects using marginal
maximum likelihood in `mfrmr` (software versions are recorded below). Shared
category steps represented a common transition structure across criteria.
The working person distribution was fixed to standard normal, person scores
were EAP estimates, and rater and criterion effects were centered within facet.
Higher rater estimates indicated stricter scoring. The fit used 31 quadrature
points, an iteration limit of 400, and the recorded L-BFGS-B optimizer; the
requested relative tolerance was 1e-9 and the effective selected-stage tolerance
was 1e-13. We examined the stored readiness and precision checks, residual/EAP
and marginal diagnostics, and category use. Rater and step intervals were 95%
normal approximations from observed-information SEs. Person uncertainty was
reported as posterior SD.

**Results.** The fit's readiness and precision checks passed. Estimated rater
severity spanned 1.018 logits, from -0.606 (R01; SE = 0.224) to 0.412 (R06;
SE = 0.249); the individual intervals are reported in the rater table above.
Thus, the fitted model described a range of rater severity after accounting
for the included facets. This descriptive span did not test a pairwise
contrast or quantify changes to assessment decisions.

Global residual-path Infit and Outfit were 0.866 and 0.857. Seven of 282
standardized residuals (2.48%) had absolute values of at least 2, and one
(0.35%) reached 3. The marginal screens flagged R06 and the R04–R05 rater pair
under the stated rules; these local signals require follow-up despite an
overall marginal RMSD of 0.0062. Categories 1–4 occurred 62, 96, 78, and 46
times. Shared step
estimates were ordered (-1.223, 0.163, and 1.060 logits; SEs = 0.169, 0.167,
and 0.177). Person and rater separation reliabilities were 0.664 and 0.677.
Observed exact agreement was 54/138 (39.1%) of available paired ratings of
the same person and criterion; agreement within one category was 85.5%.
These statistics describe distinct aspects of this fit. They do not establish
the cause of severity differences, rubric validity, or the adequacy of an
operational scoring decision. The illustration does not evaluate a different
population model, a pairwise inferential contrast, or decision accuracy.

For an empirical report, add recruitment, participant and rater characteristics,
training and scoring procedures, rubric rationale, actual missingness reasons,
and the study-level information in the map below. Report unresolved diagnostics
and relevant sensitivity results even if the optimizer converged. Never fill
unknown study details with assumptions from the synthetic example.

## 4. Review gaps and generate draft text

```{r checklist}
chk <- reporting_checklist(fit, diagnostics = diag, include_references = TRUE)
chk$section_summary
subset(chk$checklist, !DraftReady, c("Section", "Item", "NextAction"))

res <- mfrm_results(fit, diagnostics = diag, include = "publication")
report <- mfrm_report(res, style = "apa")
report$first_screen[, c("Area", "Status", "MainIssue", "NextAction")]
```

`DraftReady` means that drafting material is available with the documented
caveats. Readiness flags and `report$report_gaps` concern the supplied analysis
objects, not the completeness or validity of the research article. Check
`report$claim_readiness` for restrictions on a planned claim. Add a bias,
linking, or residual-structure analysis only if it answers a study question;
an unrequested optional analysis is not necessarily a reporting omission.

```{r apa}
apa <- build_apa_outputs(
  fit,
  diagnostics = diag,
  context = list(
    assessment = "Synthetic writing ratings",
    setting = "Illustration of the reporting workflow",
    scale_desc = "1-4 rubric scale",
    rater_facet = "Rater"
  )
)
cat(apa$report_text)
```

The draft reuses the supplied diagnostics. This example requested
`residual_pca = "none"`, so it contains no residual PCA findings. To report
PCA, first request the intended scope in `diagnose_mfrm()` and review those
results before regenerating the draft.

Verify every supplied context label against the study record. For example,
`scale_desc` is author-provided text, not a check of the observed score range.
`summary(apa)$content_checks` checks the generated content against its output
contract; it cannot verify recruitment, rater training, a causal explanation,
or whether the prose answers the question. Edit the draft to connect the
numerical findings to the question, as in the worked example above.

## Manuscript coverage map

Use this map while writing. The package supplies numerical evidence for part
of a paper; the right-hand column identifies information to add or justify.
Select items relevant to the design and intended claims. This is a practical
MFRM reporting guide, not a certification of journal compliance.

| What the reader needs | Evidence available in `mfrmr` | What the author supplies or verifies |
| --- | --- | --- |
| Problem, intended score use, and research questions | No fitted-object substitute. | Who uses the scores, for what decision, what is unresolved, and the expected contribution. Identify confirmatory and exploratory questions and any preregistration. |
| Participants and sample rationale | `data_summary$overview`, `$facet_overview`; `results$overview`. | Recruitment, inclusion criteria, relevant characteristics, setting and dates, sampling unit, sample-size rationale, and generalization target. Distinguish people from rating rows. |
| Raters, tasks, and rubric | Input facet labels and `data_summary$score_distribution`. | Rater expertise, recruitment and training, calibration, scoring instructions, task selection, category meanings, administration, blinding where relevant, and rubric rationale. For automated scoring, record the model/version, scoring date, prompts, exemplars and other scoring settings. Labels alone do not document these. |
| Assignment, overlap, and linkage | `data_summary$structural_missingness`, `$design_connectivity`, `$linkage_summary`; diagnostic subset information. | Who rated whom, crossing/nesting, planned overlap, occasions/forms, allocation process, and substantive justification for common-scale comparisons. Supply the roster and any anchor provenance. |
| Missing data, duplicates, and exclusions | `data_summary$missing`, `$duplicate_cell_summary`, `$row_retention`; `results$row_retention`. | Missing codes and reasons, planned versus absent ratings, treatment of repeated records, exclusion rules and counts at each stage, and missingness assumptions. Explain differences between preparation and fitting counts. |
| Model and estimand | `results$settings_overview`, `$population_overview`; model call and `specifications_report(fit)`. | Why RSM/PCM/GPCM and MML/JML answer the question; population distribution, threshold structure, constraints, orientation, weights, anchors, and any interactions. State what score or contrast is estimated. |
| Computation and reproducibility | `results$overview`, `$decision`; recorded estimation controls and replay/manifest export. | Software versions, optimizer and integration settings, tolerances, iteration limits, convergence problems and remedies, and relevant numerical sensitivity. Retain the exact analysis script. |
| Estimates and uncertainty | `as.data.frame(fit)`; `fit_measures_table()`; person/facet/step summaries. | Units, direction, reference constraints, SE and interval basis and level, uncertainty for planned contrasts, effect magnitude, and practical interpretation. Do not invent unavailable contrast tests. |
| Model fit and local diagnostics | `diag_summary$overall_fit`, `$diagnostic_basis`; `plot_marginal_fit()`, `unexpected_response_table()`, optional residual PCA. | Which assumptions each analysis probes, diagnostic mode, thresholds and their rationale/source, inspected flags, and follow-up decisions. Global fit does not establish validity; PCA does not prove unidimensionality. |
| Precision, separation, and agreement | `prec$profile`, `$checks`, `$fit_separation_basis`; `diag_summary$reliability`; `agreement$summary` and `$pairs`. | Interpreted score, relevant error sources/replications, index definition, adjustment and uncertainty basis, paired-rating denominator and overlap. Separate rater severity, inconsistency, agreement, and person-score precision. |
| Category functioning | `scale$category_table`, `$threshold_table`, category plots; `diag$parameter_uncertainty$steps` for step SEs, intervals, and reporting status. | Intended categories, threshold family and reference, sparse-category implications, rubric-based interpretation, and reasons and consequences of any recoding. Distinguish step uncertainty from facet-location uncertainty. |
| Comparisons, bias, DFF, or linking, when relevant | `compare_mfrm()`, `build_model_choice_review()`, `estimate_bias()`, `analyze_dff()`, or `build_linking_review()`. | Planned contrasts, same-data/comparability checks, numerical sensitivity, reference groups and anchor rationale, uncertainty, multiplicity, and departures from plan. Screening flags alone do not establish unfairness or drift absence. |
| Latent regression or simulation, when relevant | Population coefficients/coding/caveats; recovery/evaluation helpers. | Covariate coding and omissions; or generating conditions, repetitions, random seeds, failures, evaluation metrics and Monte Carlo uncertainty. Explain each analysis's role and distinguish empirical results from results under simulation assumptions. |
| Answer, limitations, and implications | Tables, plots, and generated draft text supply supporting results. | Answer each question in words, with direction, magnitude, uncertainty, conditions, alternative explanations, limitations, and implications for the intended score use. Do not infer validity or causality from fit alone. |
| Ethics and transparency | Local export bundle and software citation records cover only part of this. | Applicable ethics approval/consent, confidentiality, funding and conflicts, data/materials/code availability or access restrictions, and complete references. Review the actual files before sharing. |

Journal-level reporting guidance emphasizes linking questions, design,
analyses, and interpretation; use the applicable JARS-Quant items for your
study. The *Standards for Educational and Psychological Testing* additionally
connect documentation and precision evidence to the proposed score
interpretation and use, including the error sources represented by a
reliability coefficient (Appelbaum et al., 2018; AERA, APA, & NCME, 2014,
Chapters 2 and 7).

## 5. Save the analysis and complete the references

```{r export}
# A new temporary folder makes this example repeatable.
# For your study, choose a new permanent project folder instead.
output_dir <- tempfile("mfrmr-report-")
bundle <- export_mfrm_bundle(
  fit,
  diagnostics = diag,
  data = toy,
  output_dir = output_dir,
  prefix = "analysis01",
  include = c(
    "core_tables", "checklist", "dashboard", "apa",
    "summary_tables", "manifest", "script", "html"
  ),
  acknowledge_sensitive = TRUE
)
selected_files <- subset(
  bundle$written_files,
  Component %in% c("core_measures", "replay_script", "bundle_html")
)
data.frame(
  Component = selected_files$Component,
  File = basename(selected_files$Path)
)
```

Full paths and data-handling notes remain in `bundle$written_files`.
`data = toy` includes the original input CSV needed by the fit replay script.
Omitting `data` leaves a `your_data.csv` placeholder to replace manually.
The bundle's core step CSV contains point estimates; save the diagnostic
step table as well when reporting its SEs and intervals.

Save the selected manuscript table, its caption and note, the threshold
uncertainty table, and the draft carrying your supplied context:

```{r manuscript-files}
# Keep full precision in CSV; the displayed table is rounded to three decimals
write.csv(rater_table, file.path(output_dir, "Table1-raters.csv"), row.names = FALSE)
writeLines(capture.output(print(rater_apa)), file.path(output_dir, "Table1-raters.txt"))
write.csv(steps, file.path(output_dir, "Step-uncertainty.csv"), row.names = FALSE)
write.csv(mfrmr_example_operational_design, file.path(output_dir, "Planned-ratings.csv"), row.names = FALSE)
write.csv(data_summary$structural_missingness, file.path(output_dir, "Planned-missingness.csv"), row.names = FALSE)
writeLines(apa$report_text, file.path(output_dir, "Methods-Results-draft.txt"))

# Draw to the file device, then close it even if drawing fails
png(file.path(output_dir, "Figure1-Wright.png"), width = 1800, height = 1500, res = 200)
tryCatch(
  plot(fit, diagnostics = diag, show_ci = TRUE),
  finally = dev.off()
)
writeLines(c(
  "Figure 1. Person, facet, and step locations in the synthetic MML/RSM analysis.",
  "282 ratings; 48 persons, six raters, three criteria; rubric 1-4.",
  "The histogram shows person EAP estimates. All locations are in logits.",
  "Higher rater/criterion values mean stricter scoring/greater difficulty; effects are centered within facet.",
  "Facet whiskers are 95% normal intervals using MML observed-information SEs from the same diagnostics as Table 1.",
  "Steps are shared adjacent-category transitions shown as point estimates; see Step-uncertainty.csv for their intervals.",
  "The display does not test rater contrasts or establish unidimensionality."
), file.path(output_dir, "Figure1-Wright-caption.txt"))

# Filenames are shown here; output_dir retains the folder location.
list.files(output_dir, pattern = "^(Table1|Step-uncertainty|Planned|Figure1|Methods-Results)")
```

`Table1-raters.csv` retains unrounded values and does not contain the table caption or
note; keep the paired text file. The PNG likewise needs its accompanying
caption. The bundle generates its own generic APA draft, so retain
`Methods-Results-draft.txt` for the draft with the `context` supplied above.
Check and edit that text against the study record before manuscript use.
Use the journal's figure dimensions and resolution for the final submission.
Keep the planned-rating roster alongside its missingness summary: the observed
input CSV alone cannot identify which unobserved combinations were planned.

To refit in a new R session with the same package version, run
`source("/path/to/analysis01_replay.R", chdir = TRUE)`, replacing the path
with the exported file's actual location. `chdir = TRUE` lets the script find
the adjacent input CSV. This reruns the fit and diagnostics and writes a
`replayed_bundle` folder; it does not recreate your edited prose or selected
manuscript figures. Keep the analysis script containing those steps too.

The archive does not replace the study description or deidentify the data.
Inspect its text as well as its tables, and use a new directory for a new analysis.

```{r citations}
# Get the citation for the package version actually used, and for R
citation("mfrmr")
citation()

# Methodological reading prompts, not complete bibliography entries
chk$references

# Retain the computing environment with the analysis record
sessionInfo()
```

`chk$references` contains abbreviated citations and topics. Obtain and verify
complete bibliographic records for the methods actually used, for example in
Zotero, and cite the relevant source next to the model or diagnostic decision.
Do not cite every listed method simply because it is available. A software
citation identifies the tool; it does not replace methodological references.

For the conditional analyses in the map, runnable setups and reporting limits
are provided in `help("compare_mfrm")`, `help("estimate_bias")`, and the
`mfrmr-linking-and-dff`, `mfrmr-gpcm-scope`, and
`mfrmr-mml-and-marginal-fit` vignettes. For latent regression, see the population
model sections of `help("fit_mfrm")` and `help("mfrmr_reporting_and_apa")`.
Fit any additional model explicitly before
using it in a comparison or coefficient table.

## References

Appelbaum, M., Cooper, H., Kline, R. B., Mayo-Wilson, E., Nezu, A. M., & Rao,
S. M. (2018). Journal article reporting standards for quantitative research in
psychology: The APA Publications and Communications Board task force report.
*American Psychologist, 73*(1), 3–25.
[Article and reporting standards](https://doi.org/10.1037/amp0000191).

American Educational Research Association, American Psychological Association,
& National Council on Measurement in Education. (2014). *Standards for
educational and psychological testing*. American Educational Research
Association. [Open-access full text](https://www.testingstandards.net/uploads/7/6/6/4/76643089/standards_2014edition.pdf).

Eckes, T. (2005). Examining rater effects in TestDaF writing and speaking
performance assessments: A many-facet Rasch analysis. *Language Assessment
Quarterly, 2*(3), 197–221. [Article](https://doi.org/10.1207/s15434311laq0203_2).

Goodwin, S. (2016). A Many-Facet Rasch analysis comparing essay rater behavior
on an academic English reading/writing test used for two purposes.
*Assessing Writing, 30*, 21–31. [Article](https://doi.org/10.1016/j.asw.2016.07.004).

Xiao, X., Patz, R. J., & Wilson, M. R. (2026). Revisiting reliability with human
and machine learning raters under scoring design and rater configuration in
the many-facet Rasch model. *British Journal of Mathematical and Statistical
Psychology*. [Article](https://doi.org/10.1111/bmsp.70034).
