---
title: "mfrmr Linking and DFF"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{mfrmr Linking and DFF}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
is_cran_check <- !isTRUE(as.logical(Sys.getenv("NOT_CRAN", "false")))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  eval = !is_cran_check
)
```

This vignette covers the package-native route for:

- checking whether a design is connected enough for a common scale
- exporting anchor candidates from an existing fit
- screening differential facet functioning (DFF)
- deciding when subgroup contrasts are descriptive versus formally comparable

For a broader workflow guide, see `vignette("mfrmr-workflow", package = "mfrmr")`.
For the shorter help-page map, see
`help("mfrmr_linking_and_dff", package = "mfrmr")`.

## Minimal setup

```{r setup}
library(mfrmr)

bias_df <- load_mfrmr_data("example_bias")

# This example uses compact quadrature for a shorter runtime.
# For final DFF or linking evidence, refit with the package default or a higher
# quadrature setting and record that setting in the analysis log.
fit <- fit_mfrm(
  bias_df,
  person = "Person",
  facets = c("Rater", "Criterion"),
  score = "Score",
  method = "MML",
  model = "RSM",
  quad_points = 31
)

diag <- diagnose_mfrm(fit, residual_pca = "none")
```

## 1. Check connectedness first

Use `subset_connectivity_report()` before interpreting subgroup or cross-form
contrasts.

```{r connectivity}
sc <- subset_connectivity_report(fit, diagnostics = diag)

sc$summary[, c("Subset", "Observations", "ObservationPercent")]
plot(sc, type = "design_matrix", preset = "publication")
```

Interpretation:

- Sparse rows or columns indicate weaker design coverage.
- Weak coverage should lower confidence in subgroup comparisons.
- Use the co-observation design graph, not rater agreement,
  severity-direction, or halo response networks, to assess empirical
  connectedness. Response-network correlations do not create a common scale.

## 2. Export anchor candidates

`make_anchor_table()` mechanically exports fitted values in the direct-anchor
schema. Its output is a candidate table, not a validated anchor set.

```{r anchors}
anchors <- make_anchor_table(fit, facets = "Criterion")
head(anchors)
```

By default, `make_anchor_table()` refuses a source whose current readiness
record does not have `InferenceReady = TRUE`. Use
`readiness_policy = "review"` only to inspect review-only candidate values, not
to reuse them as anchors. For a ready reference fit, also document cross-run
element identity and invariance and confirm compatible model, score,
orientation, and population conventions. Then use `review_mfrm_anchors()` to
check table syntax and receiving-data support. That review does not certify the
substantive link.

Keep the roles distinct: common ratings or common elements supply observed
design overlap; `anchors` fix individual parameters; `group_anchors` constrain
a group mean under an externally justified target or equal-mean assumption.
Neither constraint type manufactures empirical overlap.

### Design the link, not only the percentage

No anchor percentage or count is universally adequate. In sparse designs,
link quality also depends on coverage, the distribution and redundancy of the
links, whether one articulation point or critical edge holds the network
together, the location and model fit of the linking set, and the intended use
of ranks or classifications. Inspect both Rater-centered and Person/task-
centered assignment graphs rather than treating a minimum count as proof of a
common scale. This distinction is consistent with work on sparse linking
networks by [Myford and Wolfe (2000)](https://doi.org/10.1002/j.2333-8504.2000.tb01832.x),
[Wind and Jones (2018)](https://doi.org/10.1177/0013164417703733), and
[Uto (2021)](https://doi.org/10.3758/s13428-020-01498-x).

Fixed anchor values also carry uncertainty from their source calibration.
Current mfrmr drift and chain summaries do not propagate every source-fit,
offset, and cross-fit covariance component. When results depend on the anchor
set, compare defensible alternatives and report that sensitivity; fixed-
parameter calibration error should not be mistaken for zero uncertainty
([Robitzsch, 2024](https://doi.org/10.3390/appliedmath4030063)).

## 3. Residual DFF as a screening layer

Residual DFF is the fast screening route. It is useful for triage, but it is
not automatically a logit-scale inferential contrast.

```{r dff-residual}
dff_resid <- analyze_dff(
  fit,
  diag,
  facet = "Criterion",
  group = "Group",
  data = bias_df,
  method = "residual"
)

dff_resid$summary
head(
  dff_resid$dif_table[, c("Level", "Group1", "Group2", "Classification", "ClassificationSystem")],
  8
)
plot_dif_heatmap(dff_resid)
```

Interpretation:

- Treat `residual` output as screening evidence.
- Check `ClassificationSystem` to see how the current residual screen was labeled.
- Reserve `ScaleLinkStatus` and `ContrastComparable` for refit-based contrasts.

## 4. Refit DFF when subgroup comparisons are defensible

The refit route reports logit-scale point contrasts when subgroup linking
supports a common scale. Any available SE or test statistic conditions on the
baseline anchors and omits their uncertainty and cross-refit covariance;
`FormalInferenceEligible` remains `FALSE`, even with adequate linking.
The default refit linking screen requires at least five anchored levels in
total across the non-target linking facets in each subgroup. A weak link
withholds contrast SEs and p-values. Inspect `LinkingAnchoredLevels`,
`LinkingThreshold`, and `ScaleLinkStatus`; five is a local screening threshold,
not a universal adequate anchor count or proof of invariance.

Refits currently refuse baselines with user-specified active population
models, fitted facet interactions, or group-anchor constraints because those
structures cannot yet be replayed and linked completely within subgroups.
The default GPCM-MML identification intercept is handled separately. Residual
DFF remains a screening alternative; see `?analyze_dff` for the exact scope.

```{r dff-refit}
dff_refit <- analyze_dff(
  fit,
  diag,
  facet = "Criterion",
  group = "Group",
  data = bias_df,
  method = "refit"
)

dff_refit$summary
head(
  dff_refit$dif_table[, c("Level", "Group1", "Group2", "Classification", "ContrastComparable")],
  8
)
```

## 5. Cell-level follow-up

If the level-wise screen points to a specific facet, follow up with the
interaction table and narrative report.

```{r dff-follow-up}
dit <- dif_interaction_table(
  fit,
  diag,
  facet = "Criterion",
  group = "Group",
  data = bias_df
)

head(dit$table)

dr <- dif_report(dff_resid)
cat(dr$narrative)
```

## 6. Model-estimated facet interactions

Residual bias and DFF tools screen for unusual cells after fitting the
additive model. When the interaction hypothesis is specified in advance, use
`facet_interactions` to estimate the named two-way non-person facet
interaction in the model likelihood.

```{r model-estimated-interaction, eval = FALSE}
fit_add <- fit_mfrm(
  bias_df,
  person = "Person",
  facets = c("Rater", "Criterion"),
  score = "Score",
  method = "MML",
  model = "RSM",
  quad_points = 7
)

fit_interaction <- fit_mfrm(
  bias_df,
  person = "Person",
  facets = c("Rater", "Criterion"),
  score = "Score",
  method = "MML",
  model = "RSM",
  facet_interactions = "Rater:Criterion",
  quad_points = 31
)

interaction_effect_table(fit_interaction)
compare_mfrm(Additive = fit_add, Interaction = fit_interaction, nested = TRUE)
```

Interpretation:

- Name the facet pair explicitly before fitting.
- Treat the interaction estimates as fixed-effect deviations from the
  additive MFRM under zero marginal-sum constraints.
- Inspect sparse interaction cells before reporting substantive claims.
- Keep this route separate from residual screening with `estimate_bias()`.

## 7. Multi-wave anchor review

When you work across administrations, the route usually moves from a declared
common-anchor design to anchored fitting and then to drift review. The block
below is schematic: `wave1` and `wave2` must be administrations for which the
listed facet levels have documented cross-wave identity and unchanged meaning.

Do **not** substitute `ej2021_study1` and `ej2021_study2` here. They are
independent legacy synthetic studies; reused raw labels do not identify common
persons, raters, or anchors and do not link their scales.

```{r drift-route, eval = FALSE}
declared_common_facets <- c("Criterion")

fit1 <- fit_mfrm(
  wave1, "Person", c("Rater", "Criterion"), "Score",
  method = "MML"
)

anchored <- anchor_to_baseline(
  wave2,
  fit1,
  person = "Person",
  facets = c("Rater", "Criterion"),
  score = "Score",
  anchor_facets = declared_common_facets
)

fit2 <- fit_mfrm(
  wave2, "Person", c("Rater", "Criterion"), "Score",
  method = "MML"
)
drift <- detect_anchor_drift(list(Wave1 = fit1, Wave2 = fit2))
plot_anchor_drift(drift, type = "drift", preset = "publication")
```

The current drift and chain helpers estimate one pooled offset across every
selected common element and facet. Their SE ratios do not include uncertainty
from estimating that offset or cross-fit covariance, and cumulative chain
offsets have no propagated SE. `Offset_SD` is residual spread rather than an
offset standard error. Use one coherent linking facet unless a common shift
across facet blocks is defensible, and treat flags and support counts as review
screens rather than formal tests or proof of scale equivalence.

## Recommended sequence

For a compact linking route:

1. `fit_mfrm()`
2. `diagnose_mfrm()`
3. `subset_connectivity_report()`
4. `make_anchor_table()` or `review_mfrm_anchors()`
5. `analyze_dff()`
6. `dif_report()` and `plot_dif_heatmap()`
7. `interaction_effect_table()` after a confirmatory `facet_interactions` fit
8. `anchor_to_baseline()` / `detect_anchor_drift()` when working across waves

## Related help

- `help("mfrmr_linking_and_dff", package = "mfrmr")`
- `help("subset_connectivity_report", package = "mfrmr")`
- `help("analyze_dff", package = "mfrmr")`
- `help("interaction_effect_table", package = "mfrmr")`
- `help("detect_anchor_drift", package = "mfrmr")`

## References

- Myford, C. M., & Wolfe, E. W. (2000). Strengthening the ties that bind:
  Improving the linking network in sparsely connected rating designs. *ETS
  Research Report Series*, 2000(1).
- Robitzsch, A. (2024). Bias and linking error in fixed item parameter
  calibration. *AppliedMath*, 4(3), 1181--1191.
- Uto, M. (2021). Accuracy of performance-test linking based on a many-facet
  Rasch model. *Behavior Research Methods*, 53(4), 1440--1454.
- Wind, S. A., & Jones, E. (2018). The stabilizing influences of linking set
  size and model--data fit in sparse rater-mediated assessment networks.
  *Educational and Psychological Measurement*, 78(4), 679--707.
