---
title: "Exploring rater attributes with missing data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Exploring rater attributes with missing data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
available <- requireNamespace("mice", quietly = TRUE) &&
  requireNamespace("cluster", quietly = TRUE)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = available)
```

An assessment team wants to describe its raters' backgrounds before planning
training. Do groups summarize different combinations of experience, workload,
and training? Would those descriptions change with the number of groups,
feature weights, or plausible replacements for unrecorded attributes?

This example follows that question from a feature table to a comparison of
partitions. It uses **120 fictional raters**, not observations from an assessment
program. No true groups are planted. Successful execution does not establish
that these features, imputation models, or groups are appropriate for your
raters. The groups describe backgrounds; they do not measure rater quality,
severity, or ability, and do not identify which training will be effective.

These APIs belong to development version 0.2.4.9001. Install the optional
`cluster` and `mice` packages to run the entire example.

## Choose attributes and their meaning

Use one row per rater, rather than repeating attributes for every rating.
Keep identifiers separate from features. Select variables for the question
before examining which choices produce attractive groups.

| Column | Meaning and coding | Role here |
| --- | --- | --- |
| `ExperienceYears` | Completed years of rating experience; numeric | Feature |
| `AnnualRatings` | Number of responses rated in the previous year; numeric | Feature |
| `Specialty` | Main subject area; unordered factor | Feature |
| `TrainingLevel` | Highest completed training stage; ordered factor | Feature |
| `Certified` | Holds the relevant certification; logical | Feature |
| `WorkshopHours` | Hours of workshops attended in the previous year; numeric | Auxiliary imputation predictor |
| `MentoringYears` | Years in a formal mentoring role; undefined without that role | Reviewed, then excluded from the all-rater feature set |

```{r attributes}
library(mfrmr)
set.seed(20260921)
n <- 120L
raters <- data.frame(
  Rater = sprintf("R%03d", seq_len(n)),
  ExperienceYears = sample(0:30, n, replace = TRUE),
  AnnualRatings = round(exp(rnorm(n, log(250), 0.6))),
  Specialty = factor(sample(c("Language", "Science", "Social science"),
                            n, replace = TRUE)),
  TrainingLevel = ordered(sample(c("Introductory", "Intermediate", "Advanced"),
                                 n, replace = TRUE),
                          levels = c("Introductory", "Intermediate", "Advanced")),
  Certified = sample(c(FALSE, TRUE), n, replace = TRUE),
  WorkshopHours = sample(0:80, n, replace = TRUE)
)
has_mentoring_role <- sample(c(FALSE, TRUE), n, replace = TRUE, prob = c(.7, .3))
raters$MentoringYears <- ifelse(has_mentoring_role,
  pmin(raters$ExperienceYears, sample(0:10, n, replace = TRUE)), NA_real_)

# Introduce unrecorded values only after generating the fictional attributes.
feature_names <- c("ExperienceYears", "AnnualRatings", "Specialty",
                   "TrainingLevel", "Certified")
missing_counts <- c(12L, 18L, 6L, 10L, 10L)
for (j in seq_along(feature_names)) {
  raters[[feature_names[j]]][sample.int(n, missing_counts[j])] <- NA
}
head(raters)
```

The [Gower distance calculation](https://stat.ethz.ch/R-manual/R-devel/library/cluster/html/daisy.html)
accommodates these mixed types. Numeric features use their
observed ranges. Ordered levels use ordered integer codes, so the declared
order and the spacing implicit in that coding matter. Logical features treat
both matches (`FALSE/FALSE` and `TRUE/TRUE`) alike. Certification is therefore
not coded as a special presence-only attribute. Equal feature weights are a
starting choice; correlated or redundant attributes can still receive too much
combined influence. A highly unusual numeric value can also change its
feature's range and hence other raters' distances.

## Distinguish unrecorded and undefined values

Here a missing mentoring duration means that the attribute does not apply.
It is not a duration waiting to be estimated. A duration of zero would instead
mean an applicable mentoring role with no completed years.

```{r missing-reasons}
all_features <- c(feature_names, "MentoringYears")
initial <- mfrm_features(raters, "Rater", all_features)
reasons <- initial$missing
reasons$Reason <- ifelse(reasons$Feature == "MentoringYears",
                        "Not applicable: no mentoring role", "Not recorded")
review <- mfrm_features(raters, "Rater", all_features, reasons)
summary(review)
table(review$missing$Reason)
```

The target is a description of **all raters**. Requiring a mentoring duration
would exclude nonmentors. We therefore leave it out of the selected features
and the imputation predictors, retaining the original review. A separate
analysis confined to mentors would address a different population.

```{r selected-features}
selected_reasons <- subset(review$missing, Feature %in% feature_names)
features <- mfrm_features(raters, "Rater", feature_names, selected_reasons)
features$feature_summary
table(features$row_summary$Complete)
```

`mfrm_cluster(features, k = 3)` would stop on the remaining missing values.
Explicit `missing = "omit"` is available, but selects complete raters and
changes the population being described. It does not correct missing-data bias.

## Fit an explicit imputation model

Only cells reviewed as unrecorded are eligible here. The fictional omissions
were introduced at random; in an actual program, an explanation based on the
data collection process is needed. Missingness reasons alone do not identify
a statistical missingness mechanism. Unobserved-value-dependent nonresponse
requires additional sensitivity assumptions.

The [mice documentation](https://amices.org/mice/reference/mice.html) describes
the methods and predictor controls used below. `WorkshopHours` can help predict
unrecorded attributes without becoming a clustering feature. The identifier
is neither imputed nor used to predict other variables.

```{r imputation}
eligible <- subset(features$missing, Reason == "Not recorded")
imputation_data <- raters[c("Rater", feature_names, "WorkshopHours")]
where <- matrix(FALSE, nrow(imputation_data), ncol(imputation_data),
                dimnames = dimnames(imputation_data))
where[cbind(match(eligible$ID, imputation_data$Rater),
            match(eligible$Feature, names(imputation_data)))] <- TRUE
methods <- mice::make.method(imputation_data, where = where)
methods[feature_names] <- c("pmm", "pmm", "polyreg", "polr", "logreg")
methods["Rater"] <- ""
predictors <- mice::make.predictorMatrix(imputation_data)
predictors[, "Rater"] <- 0
predictors["Rater", ] <- 0
model <- mice::mice(imputation_data, m = 5, maxit = 5,
  method = methods, predictorMatrix = predictors, where = where,
  seed = 724, printFlag = FALSE)
model$loggedEvents
head(mice::convergence(model))
```

Five imputations and five iterations keep this executable example short; they
are not an adequacy recommendation. Inspect logged events, chain behavior,
observed versus imputed distributions, and substantive plausibility. A lack
of logged events does not establish convergence or a suitable model.
`mfrm_cluster_imputed()` checks consistency of the data, not these assumptions.

## Describe groups and compare choices

Reuse the **same fitted imputation model** for every setting. Changing the
imputations as well as the weights would change two things at once.

```{r grouping}
experience_weights <- setNames(rep(1, length(feature_names)), feature_names)
experience_weights["ExperienceYears"] <- 3
analyses <- list(
  TwoGroups = mfrm_cluster_imputed(features, model, eligible, k = 2),
  ThreeGroups = mfrm_cluster_imputed(features, model, eligible, k = 3),
  ExperienceWeighted = mfrm_cluster_imputed(features, model, eligible, k = 3,
                                          weights = experience_weights)
)
comparison <- mfrm_cluster_compare(analyses)
comparison$analysis_summary
summary(comparison)
weight_effect <- subset(comparison$comparison_summary,
  First == "ThreeGroups" & Second == "ExperienceWeighted")
```

`MeanChangedFraction` describes the fraction of rater pairs whose together/apart
status changes between settings, averaged over the supplied imputations.
The minimum and maximum show its range. The individual rows in
`comparison$comparisons` retain every imputation, including split/join counts
and adjusted Rand indices. An adjusted Rand index of one indicates identical
partitions, allowing arbitrary group numbering. Neither high agreement nor a
large silhouette proves that the groups are meaningful. Silhouettes under
different weights use different distances and should not automatically select
the weights.

In this fictional example, keeping three groups and raising the experience
weight from one to three changes the together/apart status of
`r if (available) sprintf("%.1f%%", 100 * weight_effect$MeanChangedFraction) else "the fraction shown in the table"`
of rater pairs on average across imputations. Thus, the selected weights
materially affect this particular description; presenting a single partition
would conceal that dependence. This percentage concerns pairs, not the
percentage of raters assigned to a different numbered group. It is not evidence
about an actual assessment program.

Also inspect `MinGroupSize`. A rare category can produce a very small group,
and repeated profiles can give high separation values. A high overall
silhouette can coexist with these conditions.

Inspect the actual profiles before attaching interpretations to the groups:

```{r interpretation}
one_completion <- analyses$ThreeGroups$analyses[[1]]
one_completion$profiles$numeric
subset(one_completion$profiles$categorical, Feature == "TrainingLevel")
head(one_completion$membership)
analyses$ThreeGroups$co_membership[1:5, 1:5]
```

These profiles describe **one completion**, not an average cluster or a final
assignment. Group numbers can change across completions. The `co_membership`
matrix instead records how often each pair is grouped together across all
imputations. For example, 0.8 means four of the five completions, conditional
on this model and these settings. It is not an 80% probability of a true shared
class. Numeric ranges are recalculated within each completion; their changes
also contribute to sensitivity.

The workflow answers whether proposed background descriptions depend strongly
on the supplied choices. It supplies no automatic preferred group count,
consensus assignment, resampling stability, Rubin-pooled inference, or
imputation of rating responses. Before using groups to make decisions about
raters, establish that the attributes and proposed use address the intended
question and assess the consequences with appropriate evidence.

## Read separation, feature profiles, and imputation sensitivity

The silhouette plot shows which raters fit their assigned group less clearly
under the selected distances. Negative widths warrant examining the rater's
profile and nearby groups; high widths do not establish useful or stable types.
The dashed line is the overall mean. Raters excluded from clustering have no
width; their IDs remain in the returned plot data.

```{r silhouette, fig.width=7, fig.height=5}
plot(one_completion)
```

Inspect one feature at a time so that its units and categories remain clear.
Numeric profiles show means and medians in original units, without confidence
intervals. Categorical profiles show proportions within each group, preserving
the order of training levels. These are descriptions of this completion only.

```{r profiles, fig.width=7, fig.height=5}
plot(one_completion, type = "profile", feature = "ExperienceYears")
plot(one_completion, type = "profile", feature = "TrainingLevel")
```

The next heatmap asks which pairs remain together across imputations. For
readability, this view explicitly selects the first 20 IDs in the input table;
it does not recluster them, change the denominator, or represent the remaining
pairs. Omit `ids` to display all raters. By default, ID labels are hidden above
50 displayed entities; the data are never sampled automatically.

```{r co-membership, fig.width=7, fig.height=6}
plot(analyses$ThreeGroups, ids = raters$Rater[1:20])
```

The colour scale always runs from zero to one. Grey denotes unavailable pairs
involving excluded raters, not pairs that never share a group. Rows retain
the requested ID order. The map does not infer a consensus partition or a
hierarchy, and it does not measure stability under new samples of raters.

For a custom figure, extract the same values without drawing or refitting:

```{r plot-values}
view <- plot(analyses$ThreeGroups, ids = raters$Rater[1:5], draw = FALSE)
plot_data(view)$matrix
```

## Compare PAM with a hierarchy

A hierarchy asks which groups merge as dissimilarity increases. Fit it
separately from PAM: `mfrm_cluster_hierarchical()` uses average linkage
(the mean of all cross-group Gower dissimilarities) or complete linkage
(their maximum). The same feature types, weights, and missingness rules apply.
Choose the group count explicitly; a tree does not automatically identify it.

For multiply imputed features, reuse the same model and eligible cells. This
holds the completed tables fixed while changing the grouping method.

```{r hierarchical-comparison}
average <- mfrm_cluster_imputed(features, model, eligible, k = 3,
  method = "hierarchical", linkage = "average")
complete <- mfrm_cluster_imputed(features, model, eligible, k = 3,
  method = "hierarchical", linkage = "complete")
method_comparison <- mfrm_cluster_compare(list(
  PAM = analyses$ThreeGroups, Average = average, Complete = complete))
method_comparison$analysis_summary
summary(method_comparison)
```

The comparison identifies methods and linkages, alongside group sizes and
separation. As before, pair-change summaries describe sensitivity to the
supplied choices; they do not select a preferred algorithm or establish true
rater types. Substantively different profiles or very small groups deserve
inspection even when the overall pair agreement is high.

```{r dendrogram, fig.width=7, fig.height=5}
plot(average$analyses[[1]])
```

This is the full tree for **one completion**, with boxes marking its three
groups. The 120 raters are all included; ID labels are hidden by default above
50 entities. For a smaller cohort, labels appear automatically. Use
`plot_data(plot(average$analyses[[1]], draw = FALSE))$table` to inspect IDs and
memberships in leaf order. Excluded raters retain unavailable memberships but
have no leaves. No entities are sampled to draw the tree.

Merge heights are linkage dissimilarities, not significance levels or branch
support. Tied distances can produce alternative trees; the group cut uses
merge order, so a unique horizontal cutting height may not exist. Inspect
each completion's profiles with `type = "profile"` or silhouettes with
`type = "silhouette"`. `plot(average)` displays pairwise co-membership across
imputations; it does not combine the trees into a consensus hierarchy.

## How this relates to MFRM rater-type research

Clustering has been used alongside MFRM, with both hierarchical and
nonhierarchical methods. In writing assessment,
[Eckes (2012), *Operational Rater Types in Writing Assessment*](https://doi.org/10.1080/15434303.2011.649381),
clustered MFRM rater-by-criterion bias estimates using hierarchical Ward
clustering, also checking complete and average linkage solutions. In music
performance assessment,
[Wesolowski (2019), *Predicting Operational Rater-Type Classifications Using
Rasch Measurement Theory and Random Forests*](https://doi.org/10.1111/jedm.12227),
reported nonhierarchical k-means clustering of 29 differential rater functioning
indices, followed by prediction of the resulting labels.

Those analyses ask about patterns of rating behaviour. This tutorial instead
groups external backgrounds such as experience and specialty. Its Gower
distances accommodate the selected mix of variable types, and PAM produces
a partition rather than a hierarchy. The separate hierarchical API uses
average or complete linkage. Ward's variance criterion requires appropriate
Euclidean geometry; this Gower interface does not define a Euclidean
conversion and does not offer Ward clustering (see
[Murtagh and Legendre, 2014](https://doi.org/10.1007/s00357-014-9161-z)). The cited
studies do not validate the present attributes, weights, imputation model, or
PAM groups, and do not establish that one method is preferable for all uses.
Grouping estimated MFRM effects would additionally require consideration of
their measurement uncertainty and comparability. The present external-feature
API does not propagate that uncertainty.

## Plan for the size of the analysis

Pairwise distances require memory that grows quadratically with the number
of raters. Ordinary clustering permits up to 5,000 included entities; the
imputation workflow permits up to 5,000 total entities because its comparison
matrix includes excluded IDs. These are input limits, not a guarantee of low
memory use or acceptable run time. Multiple imputations also retain every
completed analysis. Save the fitted imputation model and clustering results;
`mfrm_cluster_compare()` reuses them without repeating the fits.
