---
title: "Dependent ratings within a person: testlet models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Dependent ratings within a person: testlet models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
is_cran_check <- !isTRUE(as.logical(Sys.getenv("NOT_CRAN", "false")))
knitr::opts_chunk$set(eval = !is_cran_check, collapse = TRUE, comment = "#>", cache = FALSE,
                     dev.args = list(bg = "white"))
library(mfrmr)
```

A judge scores several criteria for one performance. The judge's impression
of that particular performance may affect all of those scores, even after
accounting for the performer's ability, criterion difficulty and the judge's
usual severity. Treating the scores as conditionally independent can count
that shared impression several times. A **testlet** identifies ratings that
share an additional local effect within a person.

The grouping could describe criteria scored by one language examiner, ratings
within a clinical station, or dimensions of a music performance. Choose the
group from the assessment process, not simply because scores correlate.
This model is appropriate only when one ability, the declared fixed facets
and the specified independent normal local effects are a defensible account
of the assessment.

For decisions about adding tasks versus criteria, unequal rubric lengths,
possible halo and task-specific dependence, see
[Tasks, rubric criteria and testlet dependence](mfrmr-testlet-applications.html).
That article includes runnable allocation and score-sensitivity examples.
A large local variance does not identify halo, and accounting for dependence
does not enforce equal task weights. The current API estimates one common
local variance, not a separate variance for each task.

## Declare what shares an effect

In this example, `Rater` has two distinct roles. Its **fixed effect** describes
that rater's severity across persons. Its **testlet membership** groups that
rater's criteria within each person. The local effect for Person P01 and Rater
R01 is independent of the local effect for Person P02 and Rater R01. It is
not one random rater effect shared across both persons. For the latter model,
see `vignette("mfrmr-random-raters", package = "mfrmr")`.

This example corresponds most directly to the distinction between fixed
rater severity and Person-by-rater local effects in
[Wang and Wilson's random-effects facet model (2005)](https://doi.org/10.1177/0146621605276281),
equations 14--15. Their [Rasch testlet model](https://doi.org/10.1177/0146621604271053)
provides the broader block-dependence formulation. The implementation here
uses one common local variance; it does not estimate a separate inconsistency
parameter for each rater.

```{r data}
ratings <- load_mfrmr_data("example_core")
head(ratings[c("Person", "Rater", "Criterion", "Score")], 8)
```

Every row belongs to exactly one testlet. A separate `Session` or `Station`
column can be used instead of `Rater`; it need not also be a fixed facet.
Labels may be reused across persons. If two sittings within the same person
have independent local effects, give them distinct testlet labels. The model
does not infer membership or support overlapping groups.

## Fit and check the calibration

```{r fit}
testlet_fit <- fit_mfrm_testlet(
  ratings, person = "Person", score = "Score", testlet = "Rater",
  facets = c("Rater", "Criterion"), score_levels = 1:4,
  quad_points = 121
)
testlet_fit$checks
summary(testlet_fit)$data_usage
testlet_fit$calibration$variance
testlet_fit$calibration$person_variance
stopifnot(testlet_fit$checks$NumericalReady,
          testlet_fit$checks$InformationPositive)
```

The model uses adjacent-category RSM probabilities, a mean-zero normal ability
population with estimated variance, and independent zero-mean normal testlet
effects with a common estimated variance. Fixed-facet effects sum to zero; free category steps set
the overall location. Estimation is frequentist marginal maximum likelihood,
with both ability and local effects integrated using Gaussian quadrature.
There are no calibration priors and no additional fitting dependency.

`person_sd = NULL`, the default, estimates ability variance as well as local
variance. Mean-zero ability and free step location avoid estimating two
interchangeable locations. Supply `person_sd = 1` to fit the earlier N(0,1)
model, or another positive SD only when treating it as known is justified.
With a unit Rasch slope, fixing ability variance is a population restriction,
not merely choosing units. Estimating it still assumes a common normal
population independent of assignment; it does not repair skewness or
different ability populations across rater panels.

`NumericalReady` checks optimization, search boundaries, and agreement with a
higher integration order. `InformationPositive` checks usable local curvature.
The default order is 31; this example uses 121. Inspect
the individual differences when a check fails; increasing `quad_points` can
resolve integration errors, but cannot repair an unidentified or unsuitable
model. No order is universally sufficient. Failed checks withhold intervals
and block Person scoring. Passing checks does not establish model fit or
frequentist coverage.

The variance can be exactly zero. Such a result does not prove that the
assessment has no local dependence. Ordinary calibration intervals are
withheld at an estimated zero variance; a regular variance interval is not
provided. `testlet_variance = 0` is a different decision: it specifies zero
as known and estimates under that restricted model. A known positive variance
can also be supplied; it should not be treated as known merely because another
analysis estimated it.

Ability variance can also be estimated at zero. Both the population estimate
and its boundary status remain visible, but `predict()` then returns all
requested Persons as `unavailable`, with an explanation. It does not report
zero-width ability intervals. `variance_max` and `person_variance_max` bound
the searches for local and ability variance, respectively; hitting either
upper bound requires numerical review.

For comparison with ordinary MFRM, preserve the observed events, category
coding, fixed Rater and Criterion effects, and the ability population and
location convention. With `person_sd = 1`, use the fixed N(0,1) ordinary
model; with estimated ability variance, the ordinary model must also
estimate its population and its location must be aligned. A known zero
local variance gives the corresponding ordinary
RSM; an estimated zero is a variance-boundary result. Standard chi-squared
likelihood-ratio calibration is not automatic. `compare_mfrm()` checks these
conditions and compares centered fixed-facet effects. Separately supplied
source-roster scores add Person comparisons; no preferred model is chosen. Changes in scores or interval widths alone
do not demonstrate better accuracy when the true abilities are unknown.

## Review fixed-rater feedback

```{r feedback, fig.width=7, fig.height=5, fig.alt="Fixed-rater severity estimates with explicitly requested 95 percent observed-information normal bounds. These approximations have unresolved finite-sample coverage and are not rater-quality classifications."}
subset(testlet_fit$calibration_table, Facet == "Rater")
# Bounds require an explicit choice of approximation and nominal level.
confint(testlet_fit, parm = "calibration", level = .95)
plot(testlet_fit, facet = "Rater", intervals = "normal", level = .95)
```

Positive severity means lower expected scores at the same ability and other
facet levels. Default tables retain estimates and approximate SEs with missing
bounds; default plots show points only. The explicitly requested whiskers above
are 95% observed-information normal approximations for the fixed effects. Their finite-sample coverage remains unresolved in the
estimated-population comparison below. They are not intervals for a newly sampled rater, tests of
rater quality, or simultaneous pairwise comparisons. Missing whiskers remain
missing. `plot(testlet_fit, facet = "Criterion")` reviews criterion effects.

## Score persons and interpret their intervals

```{r scoring, fig.width=7, fig.height=5, fig.alt="Four Person EAP estimates with conditional equal-tail intervals under the testlet model. Fitted calibration, including ability and local variances, is held fixed."}
first <- ratings[ratings$Person %in% unique(ratings$Person)[1:4], ]
scores <- predict(testlet_fit, first)
scores$table
plot(scores)
stopifnot(all(scores$table$Status == "available_conditional"))
```

`Estimate` is the posterior mean (EAP). `ConditionalSD` describes the Person
posterior spread. `Lower` and `Upper` delimit a continuous equal-tail posterior
interval. The plot therefore labels them **conditional intervals**: all
calibration values, including ability and local variances, are held fixed. These
intervals exclude calibration-estimation uncertainty. They are neither
calibration-adjusted confidence intervals nor tests of differences between
persons. A 95% conditional interval does not guarantee 95% coverage at every
fixed ability.

A simulation compared ordinary and testlet RSMs on the same observed ratings,
with both models estimating normal ability variance. There were 120 independent
datasets per condition, with 12 prespecified Persons scored from each complete
calibration sample. True ability SD was 1.3; positive local variance was .8.
The balanced design had two tasks with three criteria each.

The table reports **marginal coverage**, averaging over generated abilities
and ratings, after correcting a numerical start-selection issue. Bracketed
ranges are 95% Monte Carlo intervals for the simulation result, calculated
with datasets as the independent units. They are not Person intervals.

| Condition | Testlet coverage [95% MC interval] | Ordinary RSM coverage |
| --- | ---: | ---: |
| 120 Persons, no local dependence | 95.1% [93.9, 96.4] | 95.1% |
| 120 Persons, positive local dependence | 94.4% [93.1, 95.8] | 84.9% |
| 24 Persons, positive local dependence | 91.7% [89.7, 93.7] | 83.3% |
| 120 Persons, sparse unequal blocks | 93.3% [91.9, 94.7] | 85.7% |

All 120 panels per condition were available in the corrected replay. In the
original run, 115, 113, 119 and 119 panels were available: fourteen cases
selected an abnormally terminated optimization start even though a converged
start had the same likelihood within floating-point precision. All fourteen
passed after the selection repair, with unchanged numerical tolerances.
The selection rule leaves 466 of the 480 fits unchanged; four of these
successful controls were refitted and retained identical estimates and scores. This replay reuses the generated data;
it is not an independent confirmation study. The original outcomes remain
recorded. Fitting used 61 quadrature points, with a prespecified 121-point
refit for unresolved grid agreement; the results do not establish that the
default order of 31 suffices.

Accounting for local dependence improved interval coverage relative to the
ordinary model in these three positive-dependence conditions. **Point-score
accuracy did not consistently improve.** Mean squared error measures distance
from the generated abilities; smaller is better. On the matched datasets, testlet
minus ordinary EAP mean squared error was .00665 [95% MC interval .00190,
.01140] with 120 Persons and .05174 [.02582, .07766] with 24 Persons in the
balanced design. Thus point errors were larger in those two conditions even
though coverage improved. Differences in the null and sparse conditions
were inconclusive. Mean conditional interval widths increased by .78, .76
and .67 logits in the three positive-dependence conditions; a wider interval
alone does not establish better scoring.

The known-calibration reference had coverage of 94.2--95.7%. In the balanced
positive-dependence conditions, fitted-calibration testlet coverage was 1.25
percentage points lower [95% MC interval -2.02, -.48] with 120 Persons and
2.50 points lower [-4.35, -.65] with 24 Persons. This compares all calibration
parameters together; it does not isolate the effect of estimating ability
variance. The small-sample testlet coverage MCSE was 1.01 percentage points.
There was no increase in replication after seeing the outcomes.

The sparse condition retained a common three-criterion task for every Person,
plus either two criteria on a second task or one on a third. It changes workload
and block length together; it does not test weak/disconnected linking or
informative assignment. Unassigned ratings were absent, not imputed. These
results concern same-source conditional scores, not new-Person transport,
calibration-adjusted intervals, or 95% coverage at every fixed ability. The
small and sparse conditions did not establish nominal 95% marginal
coverage. Earlier fixed-N(0,1) results remain specific to that population
assumption. No sample-size threshold or automatic choice of model follows.

For each person, local-effect integration is checked at two orders, and the
interval endpoints invert a continuous ability CDF. A numerical failure
retains an `unavailable` row with a `Reason`; it does not silently remove the
person or replace the score. Use `summary(scores)$status_counts` to inspect
all statuses. Successful numerical integration alone does not validate the
ability prior or the local-dependence assumptions.

## Missing assigned scores and sparse designs

An unassigned rating is absent from the table. An assigned rating whose score
was not recorded has an explicit `NA`. The default rejects missing assigned
scores. Use `missing = "omit"` only after deciding that analyzing the observed
ratings is appropriate; it records omissions without imputing scores or
correcting informative nonresponse.

```{r missing, fig.width=7, fig.height=5, fig.alt="Person estimates after omitting missing assigned scores. An open circle identifies the entirely missing Person whose result comes only from the fitted population; other points use their remaining observed ratings."}
incomplete <- first
last_person <- tail(unique(incomplete$Person), 1)
incomplete$Score[incomplete$Person == last_person] <- NA_real_
incomplete$Score[1] <- NA_real_
review <- predict(testlet_fit, incomplete, missing = "omit")
review$table
review$data_usage
plot(review)
stopifnot(sum(review$table$Status == "prior_only") == 1)
```

The entirely missing person has status `prior_only`, mean zero and the SD
stored in `testlet_fit$calibration$person_sd`. These are population values,
not measured average performance.
The plot uses an open circle for this case. The partially missing person is
scored from the remaining ratings, with the omission count retained.

Unequal testlet sizes and absent testlets are allowed. For fitting, at least
two persons must have multiple observed testlets, and at least two must have
repeated ratings within a testlet. Fixed-facet levels must be observed and the
fixed design must have full rank. These are initial requirements, not a proof
of estimability. Sparse assignments can still provide little information
about the local variance. Ability, local effects and assignment are assumed
independent; choosing assignments based on unmodeled ability can violate that
assumption even without missing scores. Omission and sparse assignment are
therefore different issues.

## Reuse and save the result

`predict(testlet_fit)` reuses the source assignment and its omission policy.
For new data, new Person and testlet labels are allowed; fixed-facet levels
must already exist in the calibration. Supply **all ratings to use for each
Person in one call**. The function neither appends to cached responses nor
conditions on a saved local-effect estimate. To score additional ratings for
an existing person, supply that person's complete set with the correct shared
memberships. Do not duplicate the old rows when assembling it.

```{r saved}
path <- tempfile(fileext = ".rds")
saveRDS(testlet_fit, path)
restored <- readRDS(path)
stopifnot(identical(restored$parameters, testlet_fit$parameters))
saved_plot <- plot(scores, draw = FALSE)
plot_data(saved_plot)$table
```

## Connect to the common graphics and table workflow

The ordinary fixed-facet model, shared random raters and Person-specific
testlets answer different questions. Use
`mfrmr_output_guide("models")` to choose the route, then consult
`help("mfrmr_workflow_methods")` and `mfrmr_interval_guide()` for its supported
outputs. In particular, `predict()` for a shared-rater fit returns score
probabilities at supplied abilities, whereas `predict()` for a testlet fit
estimates Person abilities from ratings.

The shared plot accessors work directly on the testlet results. They return
the stored results without drawing or refitting:

```{r shared-accessors}
plot_data_components(scores)
score_table <- plot_data(scores, component = "table")
plot_data(testlet_fit, component = "table", facet = "Rater")
stopifnot(identical(score_table, scores$table))
```

Optional `ggplot2` conversion retains all labeled rows, prior-only symbols,
missing whiskers and the interpretation note. It does not change the model or
recalculate an interval. The incomplete-data result makes those distinctions
visible:

```{r shared-graphics, fig.width=7, fig.height=5, fig.alt="The same incomplete-roster Person estimates and conditional intervals rendered with ggplot2. The open point marks a prior-only result, not measured average performance."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  figure <- as_ggplot(review)
  print(figure + ggplot2::theme_minimal(base_size = 12))
  # For a file: ggplot2::ggsave("person-scores.png", figure,
  #                            width = 7, height = 5, dpi = 300)
}
```

`as_ggplot(testlet_fit, facet = "Rater")` uses the same route for fixed-rater
effects. For another question, `plot(scores, style = "precision")` shows
estimates against conditional interval width; `style = "distribution"` shows
the empirical cumulative distribution of response-based point estimates,
excluding prior-only and unavailable scores. This is not the latent population
distribution. Shrinkage and Person selection affect it.

The extended-model plots share `sort = "estimate"` or `"uncertainty"`,
`palette = "mono"`, `show_labels`, `reference`, `text_scale` and `point_size`.
Replace `title` or `caption`, use `""` to remove either, or set `show_title =
FALSE, show_notes = FALSE` when preparing a figure with a separate caption.
These controls also pass through `plot(res, type = "scores", ...)`. Explain
the conditional interval target in that caption or the surrounding text.
`plot_data(scores)$alt_text` and `$display_data` provide a text alternative
and row inclusion/reasons; supply them alongside images for accessible output.
The [shared-rater tutorial](mfrmr-random-raters.html) shows these views and
presentation controls with saved estimates.
`plot_data(scores)$settings` and `plot_data(scores)$notes` retain the basis for
custom displays. Plot data saved by earlier versions can be converted without
refitting; recreating a plot from the saved fit or scores also refreshes its
display metadata.

For table export, keep `Status` and `Reason` rather than selecting only finite
scores. An RDS object preserves settings that a CSV table alone cannot:

```{r table-export}
csv_path <- tempfile(fileext = ".csv")
utils::write.csv(review$table, csv_path, row.names = FALSE, na = "NA")
score_path <- tempfile(fileext = ".rds")
saveRDS(review, score_path)
stopifnot(identical(readRDS(score_path), review))
```

## Compare fixed effects with an ordinary RSM

For the same rating events, compare the usual rater and criterion effects
with an ordinary model that estimates a common normal ability SD. This asks
how modeling local dependence changes the fitted fixed effects; it does not
test whether the local variance is zero or select a preferred model.

```{r ordinary-reference}
ordinary_fit <- fit_mfrm(
  ratings, person = "Person", facets = c("Rater", "Criterion"), score = "Score",
  model = "RSM", method = "MML", rating_min = 1, rating_max = 4,
  keep_original = TRUE, population_formula = ~1,
  person_data = unique(ratings["Person"]), quad_points = 121
)
```

```{r model-comparison, fig.width=7, fig.height=5, fig.alt="Differences between centered rater effects from the testlet and ordinary RSM, plotted against their means. The horizontal zero line is descriptive, with no difference intervals or automatic decision threshold."}
model_comparison <- compare_mfrm(
  ordinary_fit, testlet_fit, labels = c("Ordinary RSM", "Testlet RSM")
)
model_comparison$checks
model_comparison$effects
plot(model_comparison, facet = "Rater", style = "difference")
```

The comparison retains both models' population and numerical-check metadata.
Within each facet it subtracts the unweighted mean over the same complete set
of levels; the raw values, centers and differences remain available. Missing
level estimates or failed numerical checks remain unavailable. No difference
SEs, limits of agreement, automatic AIC/BIC preference or chi-squared LRT are
computed. The equality or zero line is an orientation aid. Person scores,
step locations and predictive Infit/Outfit are not compared by this facet
output. Use `style = "paired"` for an equality-line display, or `as_ggplot()`
for further styling; title/caption and accessibility controls are shared.

Observed events, category coding and omitted identities must agree. New
ordinary fits retain `prep$omitted_data`; old fits that omitted scores without
this provenance need refitting from the complete assigned roster. Do not
match datasets by counts alone. Zero known testlet variance retains the fixed
rater effects, unlike zero shared-rater variance, which removes rater differences.

## Inspect posterior predictive residuals

Do some raters have more residual variability under the testlet model?
Compute the predicted category distribution by integrating both ability and
the Person-specific local effect. Its variance includes uncertainty in the
conditional mean, not just the conditional rating variance. The resulting
Infit and Outfit are **descriptive summaries of the same data used for
fitting**. An expected value of one and ordinary-model cutoffs are not
established; a lower value is not evidence that a rater should be removed.

```{r response-diagnostics, fig.width=8, fig.height=5, fig.alt="Paired descriptive posterior predictive Infit and Outfit for each rater; symbols distinguish the indices and no reference cutoffs are drawn."}
response_review <- mfrm_response_diagnostics(testlet_fit, group_by = "Rater")
response_review$measures
plot(response_review, style = "paired")
```

Use `style = "scatter"` to compare the two indices on separate axes.
`palette = "mono"`, `show_title = FALSE`, `show_notes = FALSE`,
`show_labels = FALSE` and custom `title`/`caption` control the presentation.
`as_ggplot(response_review, style = "scatter")` preserves the definition and
text alternative. `plot_data()` retains unavailable groups and their reasons.

`response_review$rows` contains original row numbers, predicted means,
predictive variances and residuals. `response_review$probabilities` contains
category probabilities on the declared score scale. Assigned missing scores
have no residual and are not imputed; absent assignments are never added.
Groups with any unresolved observed row have unavailable summaries.
`rows = ...` selects original rows for output, while the full observed roster
still informs the posterior. This is not held-out validation or a formal test;
ordinary plug-in Infit/Outfit cannot be compared directly with these values.

## Compare predictions on the same definition

Does accounting for Person-local dependence change expected scores and residual summaries?
Calculate the ordinary RSM's posterior predictive quantities, then compare
them with the saved extension results. Both models use the same observed
ratings and selected events. The ordinary plug-in indices returned by
`diagnose_mfrm()` have a different definition and cannot replace this step.

```{r predictive-comparison, fig.width=8, fig.height=5, fig.alt="Differences in descriptive posterior predictive Infit for matched rater summaries. Zero means agreement between models, not acceptable model fit."}
ordinary_response_review <- mfrm_response_diagnostics(ordinary_fit,
  group_by = "Rater")
model_comparison <- compare_mfrm(ordinary_fit, testlet_fit,
  labels = c("Ordinary RSM", "Testlet RSM"),
  response_diagnostics = list(ordinary_response_review, response_review))
model_comparison$responses$measures
plot(model_comparison, metric = "infit", style = "difference")
```

These differences are descriptive and hold each model's calibration fixed.
The zero line means the summaries agree. A smaller Infit does **not** show
that the model is better or the rater is more accurate: these predictions
reuse the observations being checked, and their mean squares have no
established reference value of one.

```{r predictive-probabilities, fig.width=9, fig.height=5.5, fig.alt="Paired category probabilities from the two models, one panel per score category. The dashed diagonal marks equal probabilities; this monochrome display has no quality flags."}
as_ggplot(model_comparison, metric = "probability", show_labels = FALSE,
  palette = "mono", show_title = FALSE, show_notes = FALSE)
```

Use `metric = "expected_score"`, `"variance"` or `"outfit"` to change the
quantity, and `category = 4` with `metric = "probability"` to focus on one
score. Expected scores remain on the original scale; variances are in
squared score units. `responses$rows` retains both original row numbers and
`responses$events` identifies each matched rating. Repeated events retain
their multiplicity. If one prediction is unavailable, its difference remains
missing with a reason. Group summaries never silently discard an unresolved
observed row. Row numbers can differ between fits; matching uses event contents.

The report below saves these comparisons and supports
`plot(res, type = "response_comparison", metric = "infit")`.
To collect the ordinary model's summaries separately without running its
plug-in diagnostics, use
`mfrm_results(ordinary_fit, response_diagnostics = ordinary_response_review,
compute = "never")`. Plotting, collecting and exporting saved comparisons
performs no new integration.

## Compare Person scores and inspect model-aware maps

How do individual scores change when the model accounts for local dependence?
Use conditional scores computed from the same complete source roster.
Selecting returned Persons does not remove the other observed responses.
The ordinary model's scoring retains its fitted normal mean and variance.

```{r person-comparison, fig.width=8, fig.height=5, fig.alt="Changes in conditional Person EAPs after centering each model at its fitted population mean. There are no model-difference confidence intervals or significance decisions."}
source_scores <- score_mfrm_persons(testlet_fit, persons = scores$table$Person)
ordinary_person_scores <- score_mfrm_persons(ordinary_fit,
  persons = source_scores$table$Person)
model_comparison <- compare_mfrm(ordinary_fit, testlet_fit,
  response_diagnostics = list(ordinary_response_review, response_review),
  person_scores = list(ordinary_person_scores, source_scores))
model_comparison$persons$table
plot(model_comparison, metric = "person", style = "difference", show_labels = TRUE)
```

The comparison subtracts each model's fitted population mean to remove its
arbitrary origin; it preserves the Rasch logit unit. It retains raw scores,
origins, posterior SDs and conditional endpoints. Shrinkage and population
variance can still differ. Separate score intervals do not constitute an
interval for their difference, a ranking of models or a test between Persons.
Prior-only and unavailable differences remain withheld.

```{r model-wright, fig.width=10, fig.height=7, fig.alt="Model-aware Wright map of selected Person EAPs, conditional facet reference locations and adjacent category boundaries. Only Person points have conditional interval whiskers."}
map_results <- mfrm_results(testlet_fit, scores = source_scores,
  response_diagnostics = response_review, comparison = model_comparison)
plot(map_results, type = "wright", show_labels = TRUE)
```

The Person column shows the selected source EAPs. Facet positions equal the
severity coefficient plus the **mean step location**: the average adjacent
category boundary when other effects are zero. The category column shows
each step at that zero reference. The Person-local testlet effect is set to zero for these reference locations; it has no global facet column.
These positions are conditional references, not posterior-averaged category
crossings. A higher reference location means more ability is needed to cross
the corresponding boundary. Adding several plotted facet locations together
would count the mean step more than once; use the model equation for a specific
rating profile.

Whiskers show only Person intervals conditional on calibration. Translating
an ordinary facet interval by the step mean would omit the covariance and
uncertainty of that mean, so no such composite interval is drawn. The separate
calibration/rater plots retain their original uncertainty targets.

```{r model-fit-pathway, fig.width=8, fig.height=5.5, fig.alt="Rater reference locations against descriptive posterior Infit for the selected ratings. Rater labels identify points; the monochrome plot has no acceptable-fit bands or reference value of one."}
as_ggplot(plot(map_results, type = "fit_pathway", facet = "Rater",
  palette = "mono", show_title = FALSE, draw = FALSE))
```

This pathway connects position with the previously defined **descriptive**
posterior residual index. Here the summaries include every observed source rating.
Use `fit_stat = "Outfit"` to change the horizontal index. A large or small value
is a prompt to inspect the selected data, not a rater-quality classification.
No ordinary fit cutoff, ZSTD or automatic exclusion is supplied.

Use `facet`, `persons`, `show_steps`, `show_intervals` and `show_labels` to
select the display. `show_title = FALSE` and `show_notes = FALSE` hide
annotations while retaining their meaning in `plot_data()`. Shapes accompany
colour; `palette = "mono"` supports grayscale output. Tables retain unavailable
rows and original/transformed coordinates. Older saved testlet scores lacking
a complete `scoring_data` roster must be regenerated from the saved fit before
these maps or score comparisons; no refit is needed. See `?mfrmr_model_maps`.

## Build a report from saved results

The common reporting route accepts the fit and **separately computed** scores.
It never runs `predict()`, refits the model or adds ordinary-model diagnostics.
The scoring table retains every requested Person, including prior-only and
unavailable rows with their reasons.

```{r report}
res <- mfrm_results(testlet_fit, scores = source_scores, comparison = model_comparison,
                    diagnostics = response_review)
res$tables$interval_basis
res$tables$scoring_status
report <- mfrm_report(res)
summary(report, view = "reader")
archive_dir <- tempfile("testlet-report-")
archive <- export_mfrm_results(res, archive_dir, preset = "starter",
  acknowledge_sensitive = TRUE) # Synthetic example; real exports retain IDs
stopifnot(nrow(archive$plot_errors) == 0)
restored_results <- readRDS(file.path(archive_dir, "mfrmr_results_results.rds"))
stopifnot(identical(restored_results$tables$person_scores, source_scores$table))
```

Open `index.html` in `archive_dir` for the model-specific figures and report.
CSV files keep all rows; RDS preserves the fit, scores and their settings.
Run the replay script from that folder to reload the report without fitting
or scoring. Scores made before prediction-source metadata was introduced
must be regenerated with `predict()` from their saved fit before attachment;
refitting is unnecessary. Matching checks prevent mixing different column
roles, facet levels, calibrations or model settings.

Saved fits can score in a new R session without a live optimizer. Scores and
plot data can also be saved separately. The result is a distinct
`mfrm_testlet` object: ordinary `mfrm_fit` diagnostics, response-MI pooling,
portable calibration objects and random-rater methods do not accept it.
Numerical checks are not model-fit diagnostics. The Shiny viewer, ordinary
Wright map and ordinary plug-in Infit/Outfit diagnostics are unavailable for
this class. The separate posterior predictive summaries above do not supply
formal fit tests or ordinary-model thresholds.
An earlier ordinary model needs a new fit with explicit testlet membership;
changing its saved class does not convert the model.

Earlier saved testlet fits retain their fixed N(0,1) population when rescored
or reported. A new fit estimates ability variance by default; specify
`person_sd = 1` to reproduce the earlier model. Keep a new fit and its newly
computed scores together when changing this assumption; do not attach scores
from a different population fit to a report.

The supported scope is one ability, RSM, unit weights, additive fixed facets
and one common variance for independent Person-specific testlets. PCM,
heterogeneous or correlated local effects, shared random raters and joint
multidimensional abilities need other models. The testlet modeling idea is
described by [Wang and Wilson (2005), *The Rasch Testlet Model*](https://doi.org/10.1177/0146621604271053).
