---
title: "Portable calibration and fresh-session scoring"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Portable calibration and fresh-session scoring}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
is_cran_check <- !isTRUE(as.logical(Sys.getenv("NOT_CRAN", "false")))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  eval = !is_cran_check
)
options(width = 78)
```

This workflow creates a portable calibration from an eligible fitted model,
validates and freezes it, saves it, and then scores new Persons without the
source fit or training responses. The scoring phase uses only the saved
artifact and the new response rows.

The portable workflow is deliberately narrower than fitted-object scoring.
It supports one observed score scale and one latent dimension for `RSM` or
`PCM` fitted by `MML` under the fixed standard-normal scoring basis. It does
not currently create portable artifacts for estimated-population or latent-
regression fits, `JML`, or bounded `GPCM`.

```{r capabilities}
library(mfrmr)

mfrm_calibration_capabilities()[, c(
  "Model", "Estimator", "ScoringBasis", "PortableCalibration"
)]
```

## Calibration session

The bundled `example_core` data are synthetic. This example uses 18 Persons
to keep vignette runtime modest; a substantive analysis should use its planned
sample and design rather than copying that count.

```{r create-calibration}
synthetic <- load_mfrmr_data("example_core")
person_ids <- unique(as.character(synthetic$Person))
training_ids <- person_ids[seq_len(18L)]
training <- synthetic[
  as.character(synthetic$Person) %in% training_ids,
  ,
  drop = FALSE
]

fit <- suppressWarnings(fit_mfrm(
  training,
  person = "Person",
  facets = c("Rater", "Criterion"),
  score = "Score",
  model = "RSM",
  method = "MML",
  quad_points = 5,
  maxit = 20
))

draft <- extract_mfrm_calibration(
  fit,
  calibration_id = "portable-rsm-example",
  source_fit_id = "synthetic-rsm-fit",
  scoring_quad_points = 31
)
review_mfrm_calibration(draft)

validated <- validate_mfrm_calibration(draft)
frozen <- freeze_mfrm_calibration(validated)
frozen

artifact_file <- tempfile(fileext = ".rds")
save_mfrm_calibration(frozen, artifact_file)
```

The fit-time quadrature and the scoring-time quadrature are distinct. Here the
small fit-time grid keeps the example quick, while the artifact explicitly
records the 31-point grid used for later posterior scoring.

## Fresh scoring session

In operational use, start a new R process and supply only the artifact path and
new response rows. The following chunk removes the fit, draft, validation
object, and training data before loading the artifact, so none can influence
the score.

```{r score-from-artifact, fig.alt = "Posterior EAP intervals for two newly scored Persons. A blue circle marks a conditionally ready score, an orange triangle marks a score requiring review, and a dashed vertical line marks the zero-logit scale origin."}
new_rows <- synthetic[
  as.character(synthetic$Person) %in% person_ids[19:20],
  c("Person", "Rater", "Criterion", "Score"),
  drop = FALSE
]
new_ids <- stats::setNames(
  c("NEW_PERSON_1", "NEW_PERSON_2"), person_ids[19:20]
)
new_rows$Person <- unname(new_ids[as.character(new_rows$Person)])
rownames(new_rows) <- NULL

# Create one explicit review case for the tutorial figure. In operational
# scoring, never alter observed responses to manufacture a disposition.
new_rows$Score[new_rows$Person == "NEW_PERSON_2"] <- max(training$Score)

rm(fit, draft, validated, frozen, training, synthetic)
invisible(gc())

calibration <- load_mfrm_calibration(artifact_file)
scores <- score_mfrm_calibration(calibration, new_rows)
summary(scores)

plot(
  scores,
  type = "interval",
  preset = "publication"
)

unlink(artifact_file)
```

The interval figure is the first score-batch review. Orange triangles identify
Persons whose `Disposition` is `scored_review`; inspect their `ReasonCodes`
before using the estimate. A Person with no valid response has no score
coordinate and therefore appears in `summary(scores)$review` and in the
draw-free plot payload's `unplotted_dispositions`, not as an artificial point.

Two focused follow-up views use the same score object without consulting the
source fit:

```{r score-review-payload, eval = FALSE}
# Valid response rows versus posterior SD.
plot(scores, type = "precision", preset = "publication")

# Posterior mass on the outer quadrature nodes versus the recorded threshold.
edge_review <- plot(scores, type = "edge_mass", draw = FALSE)
edge_review$data$selection_summary
edge_review$data$unplotted_dispositions
```

These displays review a returned score batch. They do not establish source-fit
quality, reliability, calibration transport, or validity. Every interval and
posterior SD remains conditional on the frozen point calibration and excludes
calibration-parameter uncertainty.

For a separate scoring script, the essential code is:

```r
library(mfrmr)

calibration <- load_mfrm_calibration("reviewed-calibration.rds")
new_rows <- read.csv("new-responses.csv", stringsAsFactors = FALSE)
scores <- score_mfrm_calibration(calibration, new_rows)
print(scores)
plot(scores, type = "interval", preset = "publication")
write.csv(scores$estimates, "person-scores.csv", row.names = FALSE)
```

Unknown facet levels, unknown score values, invalid weights, ambiguous
duplicates, and malformed artifacts fail closed. Missing responses require an
explicit `missing_response = "omit"` policy. Every returned estimate is a
posterior EAP conditional on the recorded fixed point calibration and prior;
its uncertainty interval excludes calibration-parameter uncertainty. Validate
the intended score use, population transport, and decision consequences
outside this numerical workflow.
