---
title: "GPCM scope and current limitations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GPCM scope and current limitations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

`mfrmr` includes a bounded implementation of the Generalized Partial Credit
Model (GPCM; Muraki 1992). The estimator and the main package-native
reporting routes are available, but several outputs still carry caveats
because score-side semantics under free discrimination differ from the
Rasch-family case. This vignette documents which helpers are available,
which are caveated, and what remains restricted.

## Checking the support boundary

`gpcm_capability_matrix()` is the canonical reference. It returns one row
per helper family with a `Status` column drawn from `supported`,
`supported_with_caveat`, `blocked`, and `deferred`, plus the rationale and
the evidence trail behind each classification.

```{r capability-supported}
library(mfrmr)
gpcm_capability_matrix("supported")[, c("Area", "Status")]
```

```{r capability-with-caveat}
gpcm_capability_matrix("supported_with_caveat")[, c("Area", "Status")]
```

```{r capability-blocked}
gpcm_capability_matrix("blocked")[, c("Area", "Status", "Boundary")]
```

```{r capability-deferred}
gpcm_capability_matrix("deferred")[, c("Area", "Status")]
```

The matrix is intentionally conservative. A row stays in `blocked` or
`deferred` even when some lower-level component already runs, because the
scope statement reflects the validation evidence rather than the raw code
path.

## What works today

The following routes are validated for bounded `GPCM`:

- **Fitting and core summaries** via `fit_mfrm(model = "GPCM", step_facet =
  ...)`. The validated default keeps `slope_facet == step_facet`, with the
  direct `MML` engine.
- **Posterior scoring and information** via `predict_mfrm_units()`,
  `sample_mfrm_plausible_values()`, `compute_information()`, and
  `plot_information()`.
- **Curve and category views** via `plot(fit, type = c("wright",
  "pathway", "ccc", "ccc_surface"))`, `category_structure_report()`, and
  `category_curves_report()`.
- **Slope-aware simulation specifications** via `build_mfrm_sim_spec()`
  and `simulate_mfrm_data()`.

## What works with caveats

The following are exposed for `GPCM` but should be read as exploratory
screens rather than as Rasch-style invariance evidence:

- `diagnose_mfrm()` and the residual and unexpected-response stack:
  `unexpected_response_table()`, `displacement_table()`,
  `measurable_summary_table()`, `rating_scale_table()`,
  `interrater_agreement_table()`, `facet_quality_dashboard()`,
  `plot_qc_dashboard()`, `plot_marginal_fit()`,
  `plot_marginal_pairwise()`.
- `reporting_checklist()` and `precision_audit_report()` route to the
  supported direct tables and plots only. `build_visual_summaries()` and
  `run_qc_pipeline()` bundle those screens and return a support caveat;
  they do not imply FACETS score-side compatibility.
- `build_apa_outputs()`, `build_mfrm_manifest()`,
  `build_mfrm_replay_script()`, and `export_mfrm_bundle()` run as
  package-native reproducibility/reporting routes and return an explicit
  GPCM support caveat.
- `evaluate_mfrm_design()`, `evaluate_mfrm_diagnostic_screening()`,
  `evaluate_mfrm_signal_detection()`, and `predict_mfrm_population()` can
  consume bounded GPCM simulation contracts. Treat the resulting planning
  summaries as model-based forecasts rather than validation of
  Rasch-family fair-score invariance.
- `build_misfit_casebook()` inherits the exploratory screening framing of
  its underlying sources.
- `fair_average_table()` and `estimate_bias()` use the slope-aware
  element-conditional GPCM kernel. Read their `caveat` fields before
  quoting SE, t, or probability columns.

`diagnose_mfrm()` now carries the same slope-aware fair-average table used by
`fair_average_table()`, so visual summaries and QC checks can route it without
recomputing a substitute table.

## What is intentionally restricted

The FACETS compatibility score-side exports remain restricted under `GPCM`
because the free discrimination parameter changes the relationship between
the latent measure and the expected category. Specifically:

- `facets_parity_report()` and `facets_output_file_bundle(include =
  "score")` still depend on a FACETS-style compatibility score contract
  that is not generalized to free discrimination.

`build_linking_review()` is deferred under `GPCM` for the same reason. The
underlying `audit_mfrm_anchors()` and `detect_anchor_drift()` helpers can
still be called directly as exploratory support.

## Recommended substitutes

When a restricted helper is needed for a `GPCM` report, the practical paths
are:

- Refit with `model = "PCM"` if the discrimination-free assumption is
  defensible for the data. The full APA / fair-average / parity stack
  becomes available, and `compare_mfrm()` quantifies the loss in fit.
- Keep the report on the `GPCM` fit itself but draft the manuscript section
  around the package-native GPCM outputs: `summary(fit)` for parameters,
  `diagnose_mfrm()` for residual fit, `facet_quality_dashboard()` for the
  per-facet quality summary, `build_apa_outputs()` for caveated manuscript
  text, `build_visual_summaries()` / `run_qc_pipeline()` for caveated
  screening bundles, and `compute_information()` for precision evidence.
- Generate the reproducibility manifest and replay script from the `GPCM`
  fit with `build_mfrm_manifest()` and `build_mfrm_replay_script()`, then
  keep the support caveat in the exported bundle.

## A worked example

The `example_core` dataset includes a small synthetic block that supports a
bounded `GPCM` fit:

```r
library(mfrmr)
toy <- load_mfrmr_data("example_core")

fit_gpcm <- fit_mfrm(
  data = toy,
  person = "Person",
  facets = c("Rater", "Criterion"),
  step_facet = "Criterion",
  score = "Score",
  model = "GPCM",
  method = "MML"
)

summary(fit_gpcm)

diag_gpcm <- diagnose_mfrm(fit_gpcm)
summary(diag_gpcm)

info <- compute_information(fit_gpcm)
plot_information(info)

fair <- fair_average_table(fit_gpcm, diagnostics = diag_gpcm)
fair$caveat
plot_fair_average(fair, draw = FALSE)

vis <- build_visual_summaries(fit_gpcm, diag_gpcm)
vis$support_status

qc <- run_qc_pipeline(fit_gpcm, diag_gpcm, include_bias = FALSE)
qc$support_status

apa <- build_apa_outputs(fit_gpcm, diag_gpcm)
apa$support_status

manifest <- build_mfrm_manifest(fit_gpcm, diagnostics = diag_gpcm)
manifest$support_status

replay <- build_mfrm_replay_script(fit_gpcm, diagnostics = diag_gpcm)
replay$support_status

tables <- build_summary_table_bundle(summary(fit_gpcm))
apa_table(tables, which = "overview")
```

The fit, summary, residual diagnostics, and information helpers all run
under `GPCM`. `fair_average_table(fit_gpcm)`, `plot_fair_average()`,
`build_visual_summaries()`, `run_qc_pipeline()`, and summary-table handoff
from supported summaries also run and return explicit caveats where needed.
`build_apa_outputs(fit_gpcm)`, `build_mfrm_manifest(fit_gpcm)`, and
`build_mfrm_replay_script(fit_gpcm)` run as package-native outputs and
record `supported_with_caveat` status.

## Roadmap

The boundary above is a release-scope statement, not a permanent design
choice. FACETS-compatibility score exports, linking reviews, posterior
predictive checks, and full joint-uncertainty fair-average intervals for
free-discrimination polytomous models remain future work. Until then, the
matrix returned by `gpcm_capability_matrix()` is the binding contract.
