---
title: "GPCM scope and current limitations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GPCM scope and current limitations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
is_cran_check <- !isTRUE(as.logical(Sys.getenv("NOT_CRAN", "false")))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  eval = !is_cran_check
)
```

`mfrmr` includes a Generalized Partial Credit Model (GPCM; Muraki 1992) in which
one selected facet supplies level-specific discriminations. MML permits a
different facet to supply category steps; JML requires a shared owner. A
provisional two-family MML--EM route is described separately below; its fitted
curves do not include intervals. For the one-family route, MML information
criteria can be compared when the likelihood and local-solution checks pass.
`confint(fit, parm = "slopes")` provides approximate relative-slope intervals
when its MML checks pass. Explicit options add standardized slopes, specified
comparisons, independent-cluster covariance and finite-family adjustment.
Separate functions provide fitted-model bootstrap inference and curve intervals. Several
downstream reporting helpers remain restricted because score-side semantics
under free discrimination differ from the Rasch-family case. This vignette
documents which helpers are available, which are not, and what to use as a
substitute when a helper is restricted.

We call this model **GPCM** and state its structural choices and output
availability separately. Unidimensionality is not a reason to give GPCM a
special restrictive name. The JML route uses an
unpenalized identified joint likelihood. A non-representable slope proposal can
be rejected during line search for numerical safety, but that is not a
statistical penalty. If the likelihood has a certified recession direction,
the finite optimizer iterate is retained only as a numerical trace and the
primary result carries an extended-real or typed boundary status.

## Follow the result through the existing API

The established GPCM route estimates **one slope family**. For example, criterion
slopes and rater-owned category steps are supported by MML; this does not also
estimate rater slopes. The following routes apply to that public model, subject
to each function's checks. A returned number is not by itself evidence that
its uncertainty or interpretation is adequate for the intended decision.

| Purpose | Existing route | What to check |
| --- | --- | --- |
| Fit and read the model | `fit_mfrm()` → `print(fit)` → `summary(fit)` | Estimator, actual optimization engine, slope/step owners, ability scale, numerical status and unavailable estimates. |
| Inspect locations, fit and category use | `plot(fit, type = "wright")`, `"fit_pathway"`, `"pathway"` or `"ccc"` | These describe locations, fit or response curves; they do not supply slope confidence intervals. GPCM fit screens remain exploratory. |
| Examine slope uncertainty | `confint(fit, parm = "slopes")` → `print()` / `plot()` | MML only, with separate numerical/inference checks. Profile intervals are an explicit experimental option for one relative slope. |
| Examine response uncertainty in a rating context | `mfrm_curve_intervals(fit, newdata)` → `print()` / `plot()` | Category probability or information at supplied abilities and facet levels; retain labels for combinations not observed together. |
| Report and reuse selected inference | `mfrm_results(fit, intervals = ...)` → `mfrm_report()` → `export_mfrm_results()` | Select the result needed for the question; preserve its scale, source, unavailable intervals and interpretation. |
| Score later persons with saved calibration | `extract_mfrm_calibration()` → `save_mfrm_calibration()` / `load_mfrm_calibration()` → `score_mfrm_calibration()` | Separate source restrictions apply to MML and JML. These are conditional EAP scores; their posterior intervals exclude calibration uncertainty and are not JML slope intervals. |

Use `gpcm_capability_matrix()` to check availability and
`mfrm_calibration_capabilities()` for portable source restrictions. The
**provisional two-slope-family route** below has a narrower output scope;
the one-family table above does not grant it the same diagnostics or inference.
**Corrected JML** also has a separate, experimental workflow below. It provides
adjusted point estimates and local root uncertainty, with no formal structural
intervals or ordinary fit tests. Conditional probabilities, descriptive
residuals and new-Person/portable EAP have separate saved-output routes below. The approximate location
SEs and conditional Person posterior intervals in the ordinary JML workflow
answer different questions.

## An explicit corrected-JML workflow

Suppose many learners each receive only a few ratings. Increasing the number
of learners does not necessarily remove the structural bias caused by estimating
an ability from each short rating record. `jml_correction_order` adjusts the
profile-score equation to address that bias. It changes the estimation method;
the shared-owner GPCM response model remains the same.

This is an experimental, explicitly chosen analysis. It does not automatically
replace ordinary JML or choose a correction order. A larger order may reduce
one source of bias while increasing variance or numerical difficulty; existing
studies do not establish a universally best order. Choose an order in the
analysis plan, examine sensitivity, and do not select by the smallest SE.

```r
# Example for your observed 0:3 ratings, with at least two levels per facet.
# Install the optional nleqslv package before using this route.
# An order of 2 illustrates the syntax; it is not a universal recommendation.
# fit <- fit_mfrm(
#   ratings, person = "Candidate", facets = c("Judge", "Criterion"),
#   score = "Score", model = "GPCM", method = "JML",
#   step_facet = "Criterion", rating_min = 0, rating_max = 3,
#   category_policy = "preserve", jml_correction_order = 2,
#   jml_correction_sampling = "fixed_rosters"
# )
# summary(fit)$tables
# plot(fit, type = "locations", facet = "Judge")
# as_ggplot(plot(fit, type = "slopes", draw = FALSE))
# res <- mfrm_results(fit)
# mfrm_report(res)
```

Here each Criterion owns both its positive discrimination and its step ladder;
Judge and Criterion locations and each ladder are centered, and discriminations
have geometric mean one. Names are arbitrary: a different assessment can supply
its own facet names. The current scope requires observed integer scores, fixed
additive facets and unit-weight rows. It excludes anchors, interactions,
separate slope/step owners, corrected RSM/PCM and correlated responses. Sparse
assignments are retained as observed; absent cells are not filled with scores.
Repeated rows are independent ratings under this model, not a treatment of
within-Person local dependence.

Read the output in this order:

1. **Was a local solution found?** The summary records agreement or disagreement
   across starting values. Numerical agreement does not prove a unique global
   solution or removal of statistical bias. Unresolved fits retain their
   attempts without displaying optimizer placeholders as estimates.
2. **What was estimated?** The tables contain adjusted locations, steps and
   relative slopes. Person abilities are reprofiled at those values; extreme
   scores have infinite limits. Person standard errors are unavailable.
3. **What does RootSE mean?** It describes local variation around the adjusted
   equation's solution. That solution may still be biased for the true
   structural value. RootSE is therefore not presented as a structural SE with
   established coverage, and the plots do not invent confidence intervals.
   When this covariance cannot be computed, the point estimates remain visible
   alongside the reason.
4. **How was allocation sampled?** `"fixed_rosters"` conditions on assignment
   counts and centers actual Person contributions within each pattern. It
   requires at least two Persons per pattern for covariance. `"random_rosters"`
   instead centers globally when assignment patterns are sampled. This choice
   must reflect the sampling design; it is not a workaround for sparse data.

The figures show slopes, one selected facet's locations, or steps as points
(`style = "points"`) or an empirical cumulative distribution
(`style = "distribution"`). Titles and notes can be suppressed with
`show_title = FALSE` and `show_notes = FALSE`. Reports and saved exports retain
the estimator and interpretation; they do not support ordinary fit tests,
Wright/Pathway maps, likelihood ranking or rater-quality classifications.
New-Person EAP uses the separate scoring workflow below.

### Score a new cohort using the corrected calibration

Suppose next year's learners use the same rubric and the same calibrated
raters. `predict_mfrm_units(fit, new_ratings)` uses only those supplied ratings
and the corrected facet, step and slope values. It adds a normal reference
prior to obtain EAP scores. This is a different question from estimating the
original learners' profile maxima: even all-minimum or all-maximum responses
can have finite EAPs. JML did not estimate this prior. Consider a substantive
alternative with `scoring_prior = list(mean = ..., sd = ...)`; sensitivity to
that choice matters particularly for short or extreme response patterns.

To score in another R session without retaining the training data:

```{r corrected-jml-portable}
# draft <- extract_mfrm_calibration(fit, scoring_quad_points = 101)
# calibration <- freeze_mfrm_calibration(validate_mfrm_calibration(draft))
# save_mfrm_calibration(calibration, "rubric-calibration.rds")
# calibration <- load_mfrm_calibration("rubric-calibration.rds")
# scores <- score_mfrm_calibration(calibration, new_ratings)
# summary(scores)
# plot(scores, type = "interval")
# plot(scores, type = "precision")
# as_ggplot(scores)
```

Here 101 is an example integration order, not a universal accuracy setting.
Each batch is checked against higher-order adaptive integration. Increase the
extraction order and recreate the artifact if those checks fail. When a
Person-by-facet cell legitimately contains independent repeated events,
supply a column of distinct event IDs through `event_id`; duplicate cells
without event IDs are refused. Missing scores cause an error by default;
`missing_response = "omit"` records omissions and leaves missing-only Persons
unscored. It does not fill unassigned design cells.

Extraction checks the corrected equation and its full Jacobian at the saved
solution; it does not refit the model or demand an ordinary JML likelihood
maximum. A covariance failure does not by itself prevent scoring. File format
5 preserves the correction order, source checks, scale and prior while
omitting original ratings and Person estimates. The posterior SD and intervals
hold calibration fixed: they do not propagate RootSE, resolve residual
calibration bias or provide corrected Person ML/WLE. Calibration uncertainty
and the suitability of the prior for the new cohort remain separate issues.
Saved score objects support their own summaries and plots; they are not
attachments to the corrected fit's `mfrm_results()` report.

### Describe the observed ratings at the corrected estimates

Use `mfrm_response_diagnostics()` to ask how the observed scores differ from
the scores expected at the corrected calibration and each learner's reprofiled
ability. For example, select Judge groups to inspect rating residuals without
automatically labelling anyone an inaccurate or overly predictable judge:

```r
# response <- mfrm_response_diagnostics(fit, group_by = "Judge")
# response$measures
# response$rows
# response$probabilities
# plot(response, style = "paired")
# as_ggplot(plot(response, style = "scatter", draw = FALSE))
# res <- mfrm_results(fit, response_diagnostics = response, compute = "never")
# mfrm_report(res)
```

These are **conditional fitted probabilities**. The calculation holds the
corrected facet/step/slope estimates and reprofiled abilities fixed. It does
not average over a posterior distribution or account for uncertainty in those
estimates. A comparison between observed and expected scores uses the same
data as fitting; it is not a test of performance on unseen ratings. Sparse
assignments and repeated observed events retain their original row identities.
Selecting `rows` changes the summarized events but does not re-estimate ability.

All-minimum or all-maximum learners have infinite profile limits. Their
conditional distribution concentrates on one category, so its variance is
zero. The probabilities, expected score and raw residual remain available,
but a residual divided by its SD is undefined at these saved values. No
automatic convention replaces `0/0` with zero. Consequently:

- **Infit** can still be returned for a group: its sum of squared raw residuals
  divided by the sum of conditional variances is defined when that denominator
  is positive and all required values are available. Zero-variance rows remain
  in those sums.
- **Outfit** is missing if any selected standardized residual is undefined.
  The result records `partially_available` and explains why. `Available` counts
  standardizable rows; it is not an Infit exclusion count.
- A group consisting entirely of zero-variance rows has neither index.
  Rows are never dropped silently. An explicit `rows` selection changes the
  target group and must be reported.

The paired plot retains an available Infit when Outfit is missing; the scatter
plot requires both. Both retain complete tables and reasons in `plot_data()`.
There are no reference cutoffs, automatic overfit warnings, p-values or
rater-quality decisions. Saved results and reports preserve the conditional
definition and check the source slopes, Person profiles and exact observed
roster before accepting an attachment. This does not enable ordinary
`diagnose_mfrm()` tests or corrected-JML model comparisons.

The adjustment uses the MLE plug-in profile-score construction discussed by
[Dhaene and Weidner (2023), section 8.1](https://doi.org/10.1007/s13209-023-00283-1).
Exact owner-total expectations and a full equation Jacobian are used within
the supported state limit (5,000 joint states per assignment pattern).
This numerical implementation does not establish general residual-bias or
sampling-coverage guarantees. See `?fit_mfrm` for the full input contract.

## A provisional two-family workflow

Suppose candidates complete speaking tasks scored by overlapping assessors.
Use this route to examine how fitted category probabilities change when both
the task and assessor have their own slopes. The first owner has centered
locations and geometric-mean-one slopes; the second has free locations and
slopes and owns centered category steps. This is an explicit fixed-N(0,1)
model choice, not an automatic change to the usual GPCM population model.

For an empirical writing-data example, including the observed assignment
pattern and the limits of the resulting fit, see
[Start from an actual rating table](#start-from-an-actual-rating-table).

The code below expects your observed ratings in `ratings`, with columns
`Candidate`, `Task`, `Assessor` and integer `Rating` on the declared 0--3 scale.
Keep only actually observed scored rows. Unassigned cells must not be filled
with zero. This route does not handle missing-score rows, weights, anchors,
interactions or population covariates. Column names may be changed to match
your application, including non-syntactic names.

```{r two-family-workflow, eval=FALSE}
fit_joint <- fit_mfrm(
  ratings, person = "Candidate", facets = c("Task", "Assessor"), score = "Rating",
  model = "GPCM", method = "MML",
  slope_facet = c("Task", "Assessor"), step_facet = "Assessor",
  noncenter_facet = "Assessor", gpcm_mml_identification = "fixed_standard_normal",
  mml_engine = "em", rating_min = 0, rating_max = 3,
  category_policy = "preserve", quad_points = 31, maxit = 500,
  em_score_tol = 1e-6
)
summary(fit_joint)
context <- expand.grid(
  Theta = seq(-3, 3, length.out = 61),
  Task = unique(ratings$Task)[1], Assessor = unique(ratings$Assessor)[1]
)
curves <- mfrm_curve_intervals(fit_joint, context)
plot(curves) # provisional fitted probabilities; no confidence intervals
ci_joint <- confint(fit_joint) # experimental component-slope intervals
plot(ci_joint)
apa_table(ci_joint, which = "checks")
# For a figure without headings or notes:
plot(curves, title = NULL, subtitle = NULL, caption = NULL)
res <- mfrm_results(fit_joint, include = c("fit", "plots"),
                    compute = "never", intervals = list(slopes = ci_joint, curves = curves))
mfrm_report(res)
# export_mfrm_results(res, output_dir = "joint_results")
```

`maxit` counts outer EM iterations. `em_score_tol` controls the largest absolute
marginal-score component divided by the number of candidates; `reltol` is not
a control for this route. EM uses ascent-checked numerical BFGS M steps.
Inspect the numerical stopping result and retain nonconverged outcomes.
Numerical convergence alone does not certify identification, an adequate
quadrature order, interval coverage or the one-ability assumption.

The curve function keeps its existing name and result class, but **does not
provide intervals for two families**: SEs and bounds are missing and
`CIEligible` is false. The default plot and saved reports say this explicitly.
Known facet levels not observed together are labelled in `curves$contexts`;
this label is not evidence that extrapolation is reliable. The first and
second family's numerical slopes are retained separately in `fit_joint$slopes`;
`PrimaryEstimate` remains unavailable pending inferential qualification.

For a separate uncertainty check, `confint(fit_joint)` gives **experimental
component-slope intervals** when its numerical checks pass. For example, the
task slope describes a task relative to the geometric-average task slope;
the assessor slope is on the fixed ability scale. Neither measures scoring
accuracy. A wide interval means that the data give little precision about
that component. The two sets have different scale references: do not rank
tasks and assessors together or treat one as a universal quality threshold.

The approximation uses the inverse full observed marginal information,
including uncertainty in locations and steps and covariance between the two
slope families. The default scale is `"standardized"`, with ability SD fixed at
one. The default supplies model-based log-Wald intervals, without contrasts or
cluster adjustments. `simultaneous = "bonferroni"` adjusts for all displayed
components; no individual unit-slope p-values are provided.

For one question chosen in advance, you can also examine a **profile likelihood
interval**. It fixes that component slope at successive values and reestimates
all remaining slopes, locations and steps. The ability population stays fixed
at N(0,1). This allows the likelihood shape to depart from the quadratic shape
used by Wald intervals; it does not automatically remove bias or improve coverage.

```r
# Choose an actual level of the fitted Assessor column, for example "A1":
# profile <- confint(fit_joint, method = "profile", slope = c(Assessor = "A1"))
# plot(profile) # likelihood loss, cutoff, and same-target Wald limits
# apa_table(profile, which = "profile_endpoints")
# res <- mfrm_results(fit_joint, include = c("fit", "plots"), compute = "never",
#                     intervals = list(profile = profile))
# mfrm_report(res)
```

The name `Assessor` identifies the owner; `"A1"` identifies its level. Use your
own column/level names. Specifying both avoids confusion when two facets share
the same level labels. A first-family component retains its geometric-mean-one
reference; a second-family component is free on the fixed ability scale. The
interval concerns one component, not the product of the two slopes. Neither
owner name nor a large slope establishes rater competence.

Profiling is an explicit, more expensive sensitivity analysis. Inspect the
endpoint and numerical tables, including failed searches. The integration grid
may need refinement even when it suffices for Wald intervals at the estimate:
the profile also visits other coefficient values. No automatic grid change or
fallback to Wald occurs. Missing endpoints mean unresolved calculations, not
infinite intervals. The plot retains failed points and separates them from
connected passing sections; headings and notes can be removed with `title =
NULL`, `subtitle = NULL`, and `caption = NULL`, without deleting saved checks.

In a synthetic 240-Person example, task t3 gave profile bounds 0.899--1.227
versus Wald 0.901--1.228; assessor r3 gave 1.070--1.585 versus 1.075--1.590.
These close results verify a calculation and do not demonstrate superior
coverage. Profiling does not make a numerically rejected near-zero-slope fit
eligible or certify that alternative boundary solutions are absent.

Sparse assignments and a narrow ability distribution can leave slopes poorly
determined. A fit may converge while a constrained search for an interval
endpoint does not. Obtaining endpoints by increasing numerical accuracy does
not establish that the intervals have their nominal coverage. In particular,
evaluating coverage only among successful searches can overstate accuracy.
There is no established coverage advantage over Wald intervals for this model.

Failed trials remain saved even when both endpoints are found. An inaccurate
trial beyond an endpoint need not prevent computing that endpoint: the search
can try nearer values. Failures within a bracket still prevent accepting the
interval. This is why a plot may contain crosses for unresolved trial points
and also show two successfully computed bounds.

If a profile is unresolved, first inspect the original fit's convergence, then
the profile-check and endpoint tables. The controls address different problems:
`maxit` in `fit_mfrm()` controls EM iterations, whereas
`profile_control = list(maxit = ...)` in `confint()` controls each constrained
BFGS stage. A finer `quad_points` setting requires refitting and addresses
integration error; it does not repair unsuccessful optimization or agreement
between starts. Keep the original failures when documenting a planned analysis;
a changed numerical procedure needs its own checks.

For a two-family fit that used 121 integration points, the following illustrates
how to compare with 181 points and explicitly repeat the profile calculation.
These counts illustrate the workflow; they are not universal accuracy settings.
Use the same rating data used for the original fit.

```r
# review <- mml_quadrature_sensitivity(fit_joint, ratings,
#                                      quad_points = c(121, 181))
# summary(review)
# profile_finer <- confint(review$fits$q181, method = "profile",
#                          slope = c(Assessor = "A1"))
# apa_table(profile_finer, which = "profile_endpoints")
```

`review$intervals` contains Wald intervals for each grid, not recalculated
profile intervals. Use the explicit `confint()` call above to assess the
selected profile. If the numerical tables instead show that constrained
optimization reached its iteration limit, increase the profile iteration
budget on the same fit:

```r
# profile_longer <- confint(fit_joint, method = "profile",
#                           slope = c(Assessor = "A1"),
#                           profile_control = list(maxit = 1600))
```

An increased budget allows more computation without relaxing the numerical
checks. It may still leave an endpoint unresolved. Report any changed settings
and the original outcome; do not select the result simply because its interval
is narrower or supports a preferred conclusion.

The checks re-evaluate the saved model, local score rank, convergence and
information. They also increase the quadrature grid from q to 2q-1 at the
saved parameters. **Passing is a numerical result, not a coverage guarantee
or proof of global identification.** A converged five-point fit of a retained
example failed this integration check; its 31-point fit passed. This example
shows why convergence alone is insufficient, not that 31 points always suffice.
See `?confint.mfrm_fit` for the tolerances. Use `apa_table(ci_joint, which =
"checks")` for pass/fail reasons and `which = "numerical_checks"` for measured
changes. Failed checks keep estimates with missing bounds. The experimental
warning, owner identities and checks follow plots, reports and saved exports;
the original fit's primary-estimate/readiness fields remain unchanged.

A **small optimization residual** warning means the retained solution has a
nonzero marginal score, but its local Newton displacement is at most 0.01 in
joint SE units. Values above 1e-4 now retain this warning. This is an
approximation at the saved estimate: `confint()` does not refit or move it.
The other mean-score, information, local-rank and quadrature checks still apply.
Do not repeatedly tighten `em_score_tol` without inspecting the result: very
strict settings can exhaust the EM iteration limit at nearly unchanged
estimates. A failed **quadrature sensitivity** check is a separate issue;
use `mml_quadrature_sensitivity()` to refit with more points while retaining
the same model and EM stopping controls:

```{r joint-quadrature, eval=FALSE}
# fit_joint above used 31 points. Include the original point count.
integration <- mml_quadrature_sensitivity(
  fit_joint, ratings, quad_points = c(31, 61)
)
summary(integration)
integration$slopes # estimates, raw diagnostic SEs, interval availability
ci_finer <- integration$intervals$q61 # already computed; no repeated fit
apa_table(ci_finer, which = "numerical_checks")
plot(ci_finer) # retains missing intervals and cautions
# integration$fits$q61 is the corresponding fitted model.
```

Here the 61-point interval checks also compare 61 with 121 points at the same
saved parameters; that check does not refit at 121 points. Each requested refit
uses the usual neutral start. The first family's geometric-mean reference and
the second family's fixed-population reference are retained in the slope table.
The comparison does not yet provide two-family Person-score changes or an
adaptive-grid review; missing EAP/SD changes mean unavailable, not agreement.
Neither warning handling nor additional quadrature points
establishes nominal coverage in a sparse assessment design.

An initial simulation checked two assignment designs at equal cost: 240
candidates, three tasks, six assessors and 1,440 ratings. A common-candidate
design had 48 candidates scored by all assessors and 192 scored by one;
a rotating design gave each candidate two adjacent assessors. Every assessor
scored 80 candidates. Candidate assignment was independent of ability, and
there were no fixed parameter anchors or missing assigned scores. The normal
ability SD was 1 or 0.5; truth was transformed to the fitted fixed-N(0,1) scale.

Using the current small-residual warning rule and a finer-grid follow-up for
the common-candidate SD=1 condition gave:

| Assignment | Generating ability SD | Fitting points | Intervals returned / 100 attempts | Task t1 covered / returned | Assessor r1 covered / returned |
| --- | ---: | ---: | ---: | ---: | ---: |
| Common candidates | 1 | 61 | 100/100 | 93/100 | 98/100 |
| Rotating pairs | 1 | 31 | 100/100 | 93/100 | 95/100 |
| Common candidates | 0.5 | 31 | 100/100 | 97/100 | 95/100 |
| Rotating pairs | 0.5 | 31 | 99/100 | 90/99 | 95/99 |

These are pointwise 95% intervals for two prespecified components, not a joint
coverage statement. All nine components were summarized separately. The
rotating-pair SD=0.5 task result, 90/99, has a Monte Carlo 95% interval of
83.4--95.8%; it does not establish nominal coverage. In that condition its
empirical log-slope SD was 0.178 versus root mean estimated variance 0.155.
Some log-slope bias also remained: common-candidate SD=0.5 assessor r3 had mean
error -0.111 (Monte Carlo SE 0.032), despite 99/100 coverage.

Originally, 89 common-candidate SD=1 intervals failed the 31-versus-61-point
integration check. Refitting **all 100 datasets**, including the 11 that had
passed, at 61 points resolved these failures; all also passed the 61-versus-121
check. The EM stopping tolerance, iteration limit, response data and model were
unchanged. This is a numerical correction using the same sample, not an
independent replication. Task t1 coverage of 93/100 has a Monte Carlo 95%
interval of 86.1--97.1%; assessor r1 coverage of 98/100 has an interval of
93.0--99.8%. These results do not establish nominal coverage or prove that
61 points will suffice in other datasets. Across all nine components in this
condition, coverage ranged from 90/100 to 98/100. Task t3 had 90/100 coverage
(Monte Carlo 95% interval 82.4--95.1%), with empirical log-slope SD 0.130
versus root mean estimated variance 0.110. Assessor r3 retained log bias
-0.058 (Monte Carlo SE 0.016). Thus, removing integration failures did not
remove the observed sampling discrepancies.

The original failures must not be described as an unidentified assignment
design. Equal integration-point counts did not give equivalent numerical
accuracy across these designs. The remaining
rotating SD=0.5 failure had a near-zero assessor slope and larger local
optimization displacement. The warning change was assessed on saved fits;
this is not independent confirmation of the revised rule. These results
support cautious exploration and numerical review, not a general guarantee
for sparse designs or a recommendation that one allocation is better.

### Inspect observed ratings with posterior residuals

Suppose you want to describe where a judge's observed scores differ from the
model's predictions. `mfrm_response_diagnostics()` integrates each candidate's
ability posterior, using all of that candidate's observed ratings and holding
the fitted calibration fixed. Both slope families enter the response equation.
It does not simply plug in the candidate's EAP score. Its predictive variance
includes both rating variation at a given ability and variation in the expected
rating across the ability posterior.

```r
# After the two-family fit has met its EM stopping criterion:
# response <- mfrm_response_diagnostics(fit_joint, group_by = "Assessor")
# response$rows     # review unavailable rows and integration differences first
# response$measures
# plot(response, style = "paired")
# as_ggplot(plot(response, style = "scatter", draw = FALSE))
# res <- mfrm_results(fit_joint, include = c("fit", "plots"), compute = "never",
#                     response_diagnostics = response)
# mfrm_report(res)
```

Use your fitted identifier name for `group_by`. If response integration is
unresolved, explicitly increase `quad_points` and inspect the new check; the
fitting grid need not suffice for every response prediction. This recomputes
probabilities at the saved calibration. In contrast,
`mml_quadrature_sensitivity()` refits the calibration across specified grids.
Neither numerical check establishes model adequacy.

With `rows`, select original input row numbers for display and grouping. All
observed ratings still inform each candidate's posterior. Incomplete assignments
stay incomplete; the function adds no unassigned ratings. A group with any
unresolved observed row has unavailable summaries rather than silently dropping
that row. Available output connects to paired/scatter plots, ggplot conversion,
saved reports and exports without refitting or reintegration.

These are **same-data descriptive summaries**: calibration and checking use
the same observations. Infit and Outfit have no established expectation-one
reference here. A value above one is not an automatic misfit flag, and a value
below one is not evidence that a judge is too predictable. No formal fit test,
confidence interval or rater-quality classification is supplied. These residuals
do not test unidimensionality or local independence and exclude calibration
uncertainty. No slope confidence interval is needed to compute the fixed-point
probabilities; its availability answers a different question.

`plot(fit_joint)`, ordinary fit diagnostics, Wright/Pathway maps, location/curve
intervals, model ranking, later-person scoring and portable two-family
calibration are not yet available. Unsupported fitting arguments produce an
error rather than being ignored. Saving and reopening `res` preserves this
analysis; it does not create a calibration that can score a new cohort.

## Model, estimation method and algorithm are different choices

A model describes how a rating depends on ability, difficulty, severity and
category steps. An estimation method defines what is fitted to the data. An
algorithm is the numerical procedure used to obtain that fit. Keep these
three choices separate when reading an analysis or selecting an option:

| Choice | Question it answers | Examples |
| --- | --- | --- |
| Response model | How can ability and rating context change category probabilities? | RSM shares category steps; PCM allows facet-level step sets; GPCM adds positive discriminations. A two-family formulation multiplies two facet-specific discriminations. |
| Estimation method | How is unknown person ability handled while estimating the model? | MML integrates over an ability distribution; JML estimates the training persons' abilities jointly with the other parameters. |
| Numerical algorithm | How is the likelihood optimized? | Direct optimization, EM, or an EM warm start followed by direct optimization. |

**RSM and PCM can also use MML and EM.** In EM for an MML model, the E-step
calculates a distribution over each person's ability given all their ratings
and the current parameters. The M-step updates the parameters using those
probabilistic contributions. It does not replace each unknown ability by an
EAP point score and treat that score as observed. Direct optimization instead
optimizes the marginal likelihood itself. With the same model, identification
and integration rule, both seek the same likelihood target, although different
starts, local optima and convergence tolerances can produce different results.

The current mfrmr routes are:

| Model route | Estimation and calculation in this version |
| --- | --- |
| Public RSM / PCM | MML is the default; JML is also available. MML uses direct optimization by default. EM and hybrid are available for supported additive, fixed-population, fixed-integration settings (`population = NULL`). |
| Public one-slope-family GPCM | MML uses direct optimization. Requests for EM or hybrid currently fall back to direct and record the choice in `fit$summary`. JML is available within its separate restrictions. |
| Provisional two-slope-family GMFRM | Explicit fixed-standard-normal MML with numerical generalized EM. The numerical M-step increases its objective rather than requiring an exact maximizer at every iteration. Summary, conditional fitted curves and experimental component-slope intervals are available; other parameter intervals, model ranking and new-person scoring are not. |

Muraki (1992) derives an EM estimation route for GPCM; this does not make EM
a defining property of the response model. The
[TAM documentation](https://alexanderrobitzsch.github.io/TAM/reference/tam.mml.html)
also provides MML for RSM/PCM, and
[`sirt::rm.facets()`](https://alexanderrobitzsch.github.io/sirt/reference/rm.facets.html)
uses MML--EM for facet models with optional item and rater slopes. Matching an
algorithm name does not establish matching response equations. In particular,
GMFRM is a family description, not a uniquely specified estimator.

## What happened to the name "bounded GPCM"?

**Use GPCM as the model name, and state the actual restrictions.** The older
package label described a limited implementation scope; it should not suggest
that unidimensional GPCM is a separate statistical model or that every estimated
slope is truncated at an arbitrary finite upper bound.

Three different issues should not be conflated:

- **Parameter range and identification.** Current slopes are positive and
  represented on a log scale. Public relative slopes have geometric mean one
  to identify the scale. There is no additional user-facing finite slope box;
  numerical overflow or underflow proposals are rejected, not clipped to an
  acceptable estimate. These constraints do not disappear when the old label
  is retired.
- **Available structure and outputs.** MML now allows different slope and step
  owners, approximate slope/curve inference, checked information-criterion and
  PCM/GPCM comparisons, and portable calibration within their documented
  one-family scopes. Two-family fitting adds provisional estimates and
  conditional curves on fixed N(0,1), with separately checked experimental
  component-slope intervals. Statistical qualification remains unfinished.
  JML does not acquire MML intervals automatically.
- **A boundary in a particular dataset.** Sparse or extreme data may not yield
  a finite, well-identified optimum. Boundary diagnostics and an unbounded
  interval describe that data/model fit; removing the word "bounded" does not
  remove these possibilities or guarantee coverage.

Prefer a concrete description such as "unidimensional many-facet GPCM,
criterion slopes and rater steps, fitted by MML with direct optimization".
Older NEWS entries retain historical terminology; they do not define the
current parameter range or available workflows.

## What changes if I use JML?

`fit_mfrm(..., model = "GPCM", method = "JML")` is an implemented estimation
route. It estimates each training Person's ability together with the structural
parameters, instead of integrating over a population distribution as MML does.
The recent MML inference and portable-calibration additions do not extend
automatically to JML.

| Task | GPCM MML | GPCM JML |
| --- | --- | --- |
| Estimate one family of relative slopes | Available, subject to fit diagnostics | Available, subject to fit diagnostics |
| Use different slope and step facets | Available | Not implemented; use the same facet |
| Slope intervals, bootstrap and curve uncertainty | Available within each method's scope; profile intervals are experimental | Not implemented |
| Inferential model ranking and the matched PCM/GPCM test | Available after their separate checks | Not implemented; numerical criteria alone are descriptive |
| Score new Persons from the fitted object | Conditional EAP under the retained or explicitly supplied prior | Post-hoc EAP under a reference prior; not direct ML/WLE scoring |
| Save a portable calibration without the fit | Available within the stated MML scope | Available for shared owners after JML-specific conditional source checks; incomplete global audits remain recorded |

For example, a JML calibration can describe how strictly existing raters score.
To score next year's learners with `predict_mfrm_units()`, the current route
holds those structural parameters fixed and adds a normal prior for EAP:
standard normal by default, or `scoring_prior = list(mean = ..., sd = ...)`.
That prior was **not estimated by JML**, and the posterior SD does not include
uncertainty in the estimated calibration. Neither these scores nor posterior
draws reproduce the training Person joint maximum-likelihood estimates.

JML also has an incidental-parameter issue: increasing the number of Persons
does not by itself increase the number of observations for each Person. Its
structural SEs currently use exploratory observation-table approximations,
not a joint-information calculation accounting for estimated Person
parameters. These default JML calls apply no bias correction; the explicit
corrected route described above has a different covariance target and excludes
portable scoring. Certified extreme Person
estimates can be infinite, while finite optimizer traces remain diagnostic.
See `?fit_mfrm` for these conventions.

When comparing JML software, align these choices before interpreting differences.
For example, [TAM's `tam.jml()` documentation](https://alexanderrobitzsch.github.io/TAM/reference/tam.jml.html)
specifies an extreme-score adjustment (`adj = 0.3`) and item-parameter bias
correction (`bias = TRUE`) by default. A common estimator label does not make
those defaults identical to mfrmr, nor does a PCM comparison validate free
GPCM slope inference. Portable reference-prior EAP is available for RSM/PCM and shared-owner GPCM
JML, with separate source checks. GPCM JML local curvature does not establish
a global finite maximum or qualify formal slope intervals; known boundary
certificates prevent automatic portable scoring. See the portable-calibration
vignette for the connected workflow.

## What can I report under MML?

The following distinctions apply to the current free-slope model under MML.
They describe mfrmr's available output, not a general impossibility of GPCM
inference.

| Output | Current use | What the number does not establish |
|---|---|---|
| Relative slopes and fitted curves | Inspect `OptimizerEstimate` and numerical diagnostics for descriptive sensitivity analysis. `Estimate` retains the same numerical slope for compatibility. | A finite optimizer result is not a qualified primary slope estimate. |
| Slope SEs and intervals | `confint(fit, parm = "slopes")` or `diagnose_mfrm()` returns pointwise relative log-Wald intervals by default. Explicit `confint()` options add standardized slopes, contrasts, sandwich covariance and Bonferroni adjustment. | Inspect `CIEligible` and `InferenceReview`. Each option changes the target or approximation; none supplies universal coverage. |
| Facet and step SEs | Some local SEs remain available for review; read fit readiness and the accompanying precision/interval eligibility. | A finite band does not override the source fit's inference restriction. |
| Person posterior summaries | Conditional scoring may be available; inspect source readiness and calibration stability. | Posterior SDs do not include uncertainty from estimating the calibration. |
| AIC/BIC/SABIC | `compare_mfrm()` returns ranking, deltas and weights when its MML comparison checks pass. | `ICSelectable` covers likelihood/integration, `ICFitEligible` covers the numerical solution, and `ICComparable` is the final comparison decision. This does not grant interval or LRT eligibility. |
| PCM/GPCM likelihood-ratio test | `compare_mfrm(..., nested = TRUE)` returns an asymptotic chi-square test for eligible matched MML fits. | A significant difference does not decide the scoring policy; small or sparse samples can invalidate the approximation. |

### Which uncertainty claims are supported?

All the inferential routes below are approximations. A numerical eligibility
check establishes whether a calculation can be returned under its stated
rules; it does not certify the nominal confidence level for your assessment.
The available studies support different claims for different targets:

| Question and API | Evidence and reporting limit |
| --- | --- |
| Relative or population-standardized discrimination: `confint(fit)` | Joint-information and independent-cluster sandwich calculations have numerical checks and simulation evidence for specified designs. Historical studies do not establish current-estimator coverage for every changed numerical branch. |
| A specified ratio or difference: `confint(fit, contrasts = ...)` | Use the declared scale and full covariance. Evidence for relative contrasts does not transfer automatically to standardized differences; the latter have a separate saved-fit evaluation below. |
| Category probability at fixed ability: `mfrm_curve_intervals()` | Numerical transformations are checked, but nominal coverage failed in some small incomplete designs, including Bonferroni families. Report these as approximate calibration-uncertainty intervals, not validated 95% bands. |
| Information per rating: `mfrm_curve_intervals(..., type = "information")` | Its own saved-fit evaluation is available below. This is neither total test information nor uncertainty in an estimated Person score. |
| Matched-model selection or equal slopes: `compare_mfrm()` | AIC/BIC ranking and the asymptotic PCM/GPCM test require their separate likelihood, nesting and numerical checks. The historical null study does not establish performance for arbitrary sparse designs. |
| Bootstrap slopes or a PCM-null test: `bootstrap_mfrm_gpcm()` | Quantile calculations and accounting for unresolved trials are checked. The one-dataset example does not establish repeated-dataset coverage or Type I error, or fix the probability-interval shortfall. |

For all routes, keep unavailable or unbounded outcomes and saved cautions in
your report. Bonferroni concerns the finite family requested in one call;
sandwich covariance requires the declared independent clusters. Neither option
removes model bias or guarantees small-sample calibration. The APIs remain
available with these limits; no data-dependent widening or arbitrary sample-size
cutoff is used to conceal the adverse findings.

### How uncertain is a relative slope?

Suppose three rubric criteria have relative slopes 0.8, 1.0 and 1.25.
Their geometric mean is one. The last criterion has the steepest response
curve on this model's common ability scale. An interval describes sampling
uncertainty in that relative discrimination; it does not say that the criterion
should receive more scoring weight or that its raters agree more closely.

```r
# After fitting a native GPCM with method = "MML":
# slope_ci <- confint(fit, parm = "slopes", level = 0.95)
# slope_ci
# attr(slope_ci, "diagnostics")[, c("SlopeFacet", "Estimate",
#   "CIEligible", "InferenceReview")]
```

The calculation inverts the full joint observed information, including
estimated population parameters. It then maps its free log-slope covariance
to all slope levels using the sum-zero constraint. The interval for slope
$a_g$ is $\exp[\log(a_g) \pm z SE(\log(a_g))]$. Working on the log scale keeps
bounds positive and accounts for the dependent slope coordinates. Inverting
only the slope block of the information matrix would ignore estimation of
the other parameters and generally understate uncertainty.

The current likelihood and gradient must match a numerically adequate fit,
and the information must be positive without regularization. Unit weights,
supported score categories and at least 31 quadrature points are required.
Missing bounds retain a reason; increasing iterations cannot repair every
failure. Examine different starts and a denser grid for consequential results.
`diagnose_mfrm(fit)$parameter_uncertainty$slopes` supplies the same 95% intervals;
`attach_diagnostics = TRUE` also makes them available in the fitted slope table
and subsequent weighting reviews. Ordinary Wright/Pathway maps display ability,
facet location or fit, not these discriminations. Existing information plots
remain plug-in curves; `mfrm_curve_intervals()` adds calibration intervals.

These defaults are **pointwise model-based approximations**. The optional
methods below explicitly change the target, covariance or multiplicity rule.
An interval for `PopulationSD * Estimate` needs the variance and covariance
of the estimated population SD. A slope interval
containing one is not a test that all slopes are equal; use the matched
PCM/GPCM test for that question.

The following targeted check used saved fits from an earlier implementation.
Later numerical repairs can change fitted solutions or interval availability;
these percentages are not an evaluation of refitting every dataset with the
current estimator. The check used 100 independent datasets per condition, with three
criteria, three raters and three score categories. It crossed the slope owner
(Criterion or Rater), equal versus unequal slopes (approximately 0.67, 1 and
1.49), and two designs. With 100 Persons rated by all three raters, intervals
were available in every dataset; pointwise coverage across levels was 92--98%.
With 40 Persons each rated by two of the three raters, equal-slope availability
was 100% and coverage was 95--97%. Unequal-slope availability fell to 96--97%,
with coverage among available intervals from 90.6% to 97.9%; the lowest
covered-and-returned fraction over all planned datasets was 87%. Some intervals
were extremely wide. Four datasets were withheld for regularized information
and three for category-support review. These results support an approximate
procedure with explicit limits, not a general 95% coverage claim. Each coverage
estimate has only 96--100 independent repetitions; the three intervals within
a dataset are dependent. Sample size and incomplete assignment changed
together, so their effects cannot be isolated from this check.

### Choosing the interval for the question

For rubric criteria, a comparison of two slopes asks whether their response
curves differ in steepness, conditional on this model. It is not a comparison
of rater agreement or a reason to change scoring weights. Prespecify the
comparison before looking at which estimates differ most.

The following fictional rubric data give a reproducible example. The simulated
model is correctly specified; this illustrates the API, not field performance.

```{r gpcm-interval-example}
library(mfrmr)
ratings <- simulate_mfrm_data(
  n_person = 80, n_rater = 3, n_criterion = 3, score_levels = 3,
  model = "GPCM", step_facet = "Criterion", slope_facet = "Criterion",
  slopes = c(C01 = 0.8, C02 = 1, C03 = 1.25), seed = 924090
)
fit <- fit_mfrm(
  ratings, "Person", c("Rater", "Criterion"), "Score",
  model = "GPCM", method = "MML", step_facet = "Criterion",
  slope_facet = "Criterion", quad_points = 31, maxit = 400, reltol = 1e-10
)
```

```{r gpcm-slope-comparisons}
confint(fit, scale = "standardized")
slope_levels <- fit$slopes$SlopeFacet
comparison <- matrix(0, 1, length(slope_levels),
                     dimnames = list("first versus second", slope_levels))
comparison[1, 1:2] <- c(1, -1)
ratio <- confint(fit, contrasts = comparison)
difference <- confint(fit, contrasts = comparison,
                      contrast_scale = "difference")
attr(ratio, "diagnostics")[, c("Estimate", "CI_Lower", "CI_Upper", "PValue")]
```

The ratio null is one; the difference null is zero. More rows specify a finite
family. `simultaneous = "bonferroni"` adjusts both intervals and the corresponding
Wald p-values for all requested rows, not for other analyses or data-selected
comparisons. Standardized slopes multiply the relative slope by estimated
population SD, propagating its uncertainty and cross-covariances. With latent
regression this is the **residual** population SD, not the marginal spread
across different covariate groups. Ratios cancel a common scale; differences
do not. JML inference is outside these MML methods.

```{r gpcm-sandwich-example}
confint(fit, method = "sandwich", simultaneous = "bonferroni")
# If Persons belong to independent schools, supply a complete Person/Cluster
# mapping and use clusters = schools. Do not cluster separate rows of one Person.
```

Sandwich covariance uses the marginal-likelihood score vector for each Person,
or sums those vectors within a supplied cluster. It needs many independent
clusters with enough score-vector rank. Its target is the fitted working-model
parameter. It does not repair bias from a wrong population model, informative
assignment or dependence between clusters. The optional `adjust = TRUE` only
multiplies the covariance by G/(G-1); it is not a small-sample correction with
guaranteed coverage. The earlier 800-dataset relative-interval study does not
qualify these additional procedures by itself. An earlier reanalysis recalculated
intervals on all 800 saved fits, without new simulations or optimizer runs, and found
90.7--99% pointwise coverage for standardized model intervals and 90.7--100%
for standardized sandwich intervals. The minimum was 88/97 available intervals
in a small incomplete, unequal-rater-slope condition (Monte Carlo 95% interval
83.1--95.7%; 88/100 covered-and-returned). For three standardized slopes,
Bonferroni **family** coverage ranged from 91.7--98% with model covariance and
90--97% with sandwich covariance. Each cell has only 100 independent datasets.
These correctly specified normal-population results show that neither
sandwich covariance nor multiplicity adjustment guarantees nominal coverage
in small samples; they do not test robustness under misspecification.

### Examine one relative discrimination with a profile likelihood

The default slope interval uses the local curvature near the fitted solution.
If the likelihood is asymmetric, that approximation can differ substantially
from an interval obtained by fixing a candidate discrimination and refitting
all remaining parameters. Development mfrmr offers the latter as an explicit,
experimental option for **one prespecified relative slope**. It does not
establish better small-sample coverage.

```{r slope-profile, eval=FALSE}
# Choose a level from fit$slopes$SlopeFacet before looking for a favorable result.
ci <- confint(fit, method = "profile", slope = "R01")
print(ci)
plot(ci, type = "profile")
# Hide text when composing a larger figure:
as_ggplot(ci, type = "profile", title = NULL, subtitle = NULL, caption = NULL)
apa_table(ci, which = "profile_endpoints")
apa_table(ci, which = "profile_checks", digits = 8)

# Keep the calculated profile; reporting does not fit the model again.
res <- mfrm_results(fit, intervals = list(profile = ci), compute = "never")
report <- mfrm_report(res)
saveRDS(ci, "one-slope-profile.rds")
```

The curve shows twice the log-likelihood loss. The horizontal dashed line is
the requested chi-square(1) cutoff; its crossings define the connected local
interval. Dotted vertical lines show the same-target Wald limits. The slope
axis is logarithmic. Crosses identify unresolved calculations; lines stop at
failed points, and evaluations without finite coordinates remain in the tables.
A steeper discrimination is not a measure of rater agreement or a scoring
weight recommendation.

The relative slopes still have geometric mean one. Fixing even the dependent
last slope preserves that constraint, while the population mean and variance
and other nuisance parameters remain free. Two starting values and a finer
quadrature are checked at each candidate. A better source likelihood, failed
numerical checks, observed nonmonotonicity or an exhausted search produce
explicitly unresolved bounds. Increase `profile_control$max_steps` only to
extend a stated search; failing to find a crossing does not prove an infinite
interval. Successful finite searches cannot exclude other, disconnected regions.

The initial scope is an eligible GPCM MML fit with an estimated intercept-only
normal population, unit weights, and no anchors or interactions, with shared or
separate slope/step owners. Standardized targets, ratios/differences, JML,
latent regression and simultaneous profile intervals are not included.
The source fit and its global audit states remain unchanged. This method
requires additional nuisance fits and is not assumed cheaper than bootstrap.
Regular likelihood-ratio asymptotics still matter near boundaries; a profile
is not a correction for a wrong model or a guarantee of interval coverage.

A comparison using **all 100 saved calibrations** from one small-sample
condition illustrates why successful profiles alone are insufficient. Each
dataset had 40 people, three criteria and a linked rotating assignment of two
of three raters. Rater slopes and steps shared their owner. The target was the
first rater's relative slope, with true value `exp(-0.4)`; both methods used
nominal 95% intervals on the same stored estimate.

| Outcome | Log-Wald | Profile |
| --- | ---: | ---: |
| Complete finite intervals / all datasets | 98/100 | 53/100 |
| True slope inside / available intervals | 89/98 (90.8%) | 53/53 (100%) |
| Interval both available and containing truth / all datasets | 89/100 | 53/100 |
| Median calculation time, including unsuccessful calls | 0.38 seconds | 8.80 seconds |

The exact 95% Monte Carlo interval for profile availability is 42.8%–63.1%;
for coverage among available profiles it is 93.3%–100%. The latter describes
the selected 53 datasets, not a reliable procedure for all datasets. Among the
same 53 datasets, Wald covered 49; their median interval widths were 1.010 for
Wald and 1.035 for profile. This comparison does not establish general profile
superiority. Wald also remains approximate: its available-interval coverage has
a Monte Carlo interval of 83.3%–95.7%, and some intervals were extremely wide.

Of the 47 unresolved profiles, 11 failed before profiling (source integration,
information or category checks) and 36 failed during constrained searches.
These are limitations of this numerical procedure on the stored calibrations;
they do not show that a statistical interval cannot exist. No failed case was
replaced, and no tolerance was relaxed. Times include the interval method's
checks, exclude original fitting, and were measured with two local workers.

These previously analysed data provide a diagnostic comparison, not independent
confirmation of the current estimator. They do not assess separate owners,
probability intervals or all sparse designs. Retain profile intervals as an
explicit experimental sensitivity analysis; do not use them as an automatic
Wald replacement or report their successful subset as a coverage guarantee.

### A fitted-model bootstrap alternative

The parametric bootstrap asks what variation occurs if independent Persons
and scores follow the fitted model on the same observed rating assignment.
It preserves covariates and facet labels, and reestimates the free population
parameters. Unassigned ratings stay unassigned. It does not test whether the
assignment mechanism or the normal population assumption is correct.

```r
boot <- bootstrap_mfrm_gpcm(fit, nsim = 499, seed = 92401)
confint(boot, scale = "standardized")
confint(boot, contrasts = comparison, simultaneous = "bonferroni")
boot$trials  # Includes every planned trial, its last stage and any reason/warning.
apa_table(boot, which = "checks")  # Returned fits: categories, convergence and information.
# boot$refit_draws retains returned parameters for diagnosis, not interval calculation.
# For an eligible matched PCM fit with the same population model:
lr_boot <- bootstrap_mfrm_gpcm(fit, null_fit = pcm_fit,
                              nsim = 499, seed = 92402)
lr_boot$test
```

The interval uses basic bootstrap error quantiles, on a log scale for positive
slopes/ratios and on the original scale for differences. Failed refits and
returned estimates rejected by eligibility checks are retained as unknown outcomes. The reported limits enclose all completions,
possibly reaching zero or infinity; deleting failures could instead conceal
the least stable cases. The LRT reports Monte Carlo precision and p-value
bounds when any refit is unresolved. Its null-model draws cannot supply slope
intervals around the alternative fit. Saved results support new confidence
levels without refitting. More repetitions improve tail resolution, not
model validity; inspect `attr(confint(boot), "expected_tail_draws")`, especially
with many simultaneous comparisons. This is an approximate bootstrap method,
not a profile-likelihood implementation or an exact small-sample test.

For basic slope bootstrapping, a category observed once is now a warning
rather than an automatic exclusion. The source fit and each refit must still
pass model-identity and numerical checks, including unregularized joint
information. The exception requires a fresh audit confirming that every
category is observed in each step scope and no step coordinate or category
contrast is unsupported. It does not change the fit's saved readiness or
permission for Wald intervals, IC comparisons or LRTs. `checks$BootstrapEligible`
and `checks$WaldEligible` distinguish these decisions. Source decisions are in
`source_checks`. Accepted singleton warnings remain in the interval diagnostics,
printed output, default plot subtitle and reports, even when limits are finite.
A supplied subtitle, including `NULL`, overrides the default plot text.

The basic interval formula itself does not require a variance estimate in
every replicate; variance estimates are required for studentized intervals,
as explained in the [R boot documentation](https://stat.ethz.ch/R-manual/R-devel/library/boot/html/boot.ci.html).
Separating those formulas does not establish that an unstable estimate is
suitable for basic-bootstrap inference. Use `trials$Stage`,
`trials$AlternativeReturned` and `checks` to distinguish the reasons. Missing
check fields mean that the quantity was not recorded. Older saved results may
have only a combined reason and no rejected parameter vectors.

A small-sample example illustrates why unresolved replicates matter. Its
original 499-replicate run accepted 408 estimates. Rechecking 62
category-support cases under a revised acceptance rule produced an assembled
result with 470 accepted and 29 unresolved estimates. **That assembled result
is a comparison of procedures, not a complete bootstrap run under the current
estimator.** All three standardized 95% intervals still extended from zero to
infinity. This is one dataset, not evidence of repeated-dataset coverage or a
failure rate for other designs.

Preserve an original bootstrap when changing estimation or acceptance rules.
A complete new run is needed to evaluate the changed procedure; replacing
selected failed draws cannot establish its performance. Saved reanalysis
history, when present, follows printing, slope intervals and reporting tables.
Missing history does not establish that an older result used the current
procedure. Neither more replicates nor selective replacement automatically
resolves poor identification or validates interval coverage.

### Why an information warning needs a parameter-specific review

Inspect `boot$checks` (or `apa_table(boot, which = "checks")`) for
`PopulationSD`, `MinimumStandardizedSlope` and `MaximumStandardizedSlope`.
These describe the returned optimizer estimates, including rejected refits.
For an estimated population, standardized slopes multiply relative slopes
by the residual population SD; for a fixed standard-normal population that
SD is one. Missing entries indicate missing retained estimates. These fields
are diagnostics, not thresholds for judging raters or permission to use a CI.
APA check tables use significant digits for these values and the gradient/
information diagnostics, preserving small positive values in table conversion.
The original `boot$checks` columns remain unrounded numeric values.

The 29 unresolved trials illustrate different reasons for withholding
inference. Independent likelihood and derivative calculations helped separate
finite numerical stopping points from evidence for boundary approaches.
These checks diagnose the selected examples; they do not certify global
maxima or repeated-sample interval accuracy.

| Main finding in this one-dataset audit | Cases | Consequence |
| --- | ---: | --- |
| An unused top category in an unanchored step scope gives a monotone likelihood direction as its intercept cost increases | 15 | A threshold has no finite maximizer. In one checked example the slope estimates remained stable across starts; that does not validate all slope intervals. |
| Restarts and fixed-slope profiles support approach to zero standardized discrimination; the direction persists at 61 and 101 quadrature nodes | 11 | Evidence for a boundary limit, not just a large standard error. These numerical checks do not certify a global maximum. |
| Reparameterized optimization finds a better solution with a positive standardized slope and positive local curvature | 2 | The original numerical stopping point needs improvement. A warning attached to that original point is insufficient. The current optimizer and information refinement recover eligible finite solutions in these two selected examples; this does not replace a complete bootstrap evaluation. |
| A very steep standardized slope is highly sensitive to quadrature | 1 | At the same saved parameters, changing 61 to 101 nodes changes negative log-likelihood from about 185 to 858. Neither grid is thereby validated. |

These are primary findings; numerical and boundary problems can overlap.
An unused category in a replicate is a sampling outcome, not evidence that
the scoring category should be removed. A near-zero standardized slope can
coexist with other finite standardized slopes. Keeping the geometric mean
of *relative* slopes at one can then make relative slopes and location
coordinates extreme. Conversely, a stable discrimination estimate does not
establish a finite threshold. The assembled example retains all 29 unresolved
trials. The two improved finite solutions are separate numerical evidence;
they have not been inserted into its draws or counted as a current-procedure
bootstrap result.

For small fixed-grid GPCM MML fits (up to 64 free parameters, with
`reltol <= 1e-9`), the direct optimizer now reviews numerical curvature even
after a small raw gradient. Negative curvature triggers up to three BFGS
restarts in rescaled search coordinates. Each uses the requested iteration
ceiling. Only a non-worsening, numerically converged candidate without detected
negative curvature can replace the selected point. Otherwise the estimate is
retained with a numerical warning. Inspect `fit$opt$optimizer_polish$Stages`:
`StageLabel == "curvature_rescale"` identifies these attempts, and
`SmallestCurvature`, `CurvatureScale` and `CurvatureReviewError` explain the
curvature review. Missing curvature entries mean it was not evaluated.
The search rescaling does not provide a regularized covariance estimate.
An improved optimizer result still needs separate information and integration
checks; positive numerical curvature alone is insufficient.

Poor conditioning does not by itself mean that an information matrix must be
regularized or discarded. It can reflect parameter units, weak information,
or numerical differentiation error. The shared MML calculation now compares
two refined Hessians, checks their unregularized inverse and reviews the
gradient in the curvature metric. If these checks pass, intervals can be
returned with a saved caution. Otherwise they remain unavailable. The
information workspace budget covers the additional calculation too.

In the remaining finite-solution example, the standardized log-slope SE
changed from about 30 with the default numerical differentiation to about 27
after refinement, agreeing with the reference parameterization. Nevertheless,
the weakest discrimination's 95% log-Wald interval spans roughly
\(10^{-25}\) to \(10^{21}\). Returning that interval makes the lack of precision
visible; it does not make the approximation trustworthy near zero. Examine
`attr(intervals, "cautions")` and `InferenceReview`, which also appear in
printing, default plot subtitles and reports. This numerical review does not
establish sampling coverage or turn a boundary limit into a finite estimate.

Further requirements include integration and parameter-specific boundary
estimates and inference. The examples
support different treatment of different targets; they do not validate ordinary
Wald intervals or demonstrate repeated-dataset coverage.

### Uncertainty in probability and information curves

For example, examine a single rater/criterion context across known ability
values. The existing probability and information plots still describe fitted
points. This API propagates the full calibration covariance into each curve
point, using logit intervals for probabilities and log intervals for information.

```{r gpcm-probability-intervals, fig.alt="Three fitted category probability curves for one rater and criterion, with pointwise calibration-uncertainty ribbons at fixed ability values. Colors and line types distinguish categories; these are not Person-score intervals."}
grid <- expand.grid(Theta = seq(-3, 3, length.out = 41),
                    Rater = "R01", Criterion = "C01")
curves <- mfrm_curve_intervals(fit, grid)
plot(curves, title = NULL, subtitle = NULL)
```

```{r gpcm-information-intervals, fig.alt="Fitted information per rating for the same rater and criterion across ability values, with a pointwise calibration-uncertainty ribbon. The curve is not total information across all ratings."}
information <- mfrm_curve_intervals(fit, grid, type = "information")
plot(information)
```

Crosses identify fitted points whose intervals are unavailable. The caption
counts these points; a missing ribbon does not mean negligible uncertainty.
Ribbons stop at unavailable grid points. Inspect `plot_data(curves, "table")`
for the saved `InferenceReview` reasons. Use `caption = NULL` to hide the
caption; the crosses and saved reasons remain. Long default warnings wrap
within the figure.

For slope or ratio intervals spanning many orders of magnitude, a logarithmic
axis can make the estimates easier to compare:

```r
# When estimates and finite interval bounds are strictly positive:
# as_ggplot(ci) + ggplot2::scale_x_log10()
```

Keep a linear axis for signed differences. A logarithmic display does not
reduce uncertainty or establish interval accuracy.

Replace these labels with levels in the fitted model and include every
non-Person facet. Information is **per rating**: slope squared times the
conditional category variance. It is not total information summed over the
observed exposures. `Theta` is fixed on the fitted native ability scale;
uncertainty in a Person score is not included. Lines and ribbons retain their
numerical table, use color plus line type, and accept custom palettes and
omitted titles. Sandwich covariance is optional. Bonferroni adjustment is applied
over the finite output grid, including all categories; connecting points does not
produce a simultaneous confidence band for every value between them.

### What the additional interval checks found

Do the slope-interval results also describe uncertainty in differences and
curves? To examine that question, a further reanalysis used **all 800 saved
fits**, with the current interval calculations and their availability checks.
It did not generate new datasets or refit the historical estimates. Numerical
repairs to estimation can change those estimates, so this is evidence about
the saved-fit analysis, not an independent evaluation of the current estimator.

The generating facet effects were saved for every dataset. They were centered
to the fitted reference before comparing curves at native ability values
`Theta = -2, 0, 2`. Each slope-owner level was evaluated with the other facet
at its second level. This gives nine rating contexts, 27 category probabilities
and nine information values. Standardized differences compare the first with
the second slope and the second with the third; they include population-SD
uncertainty. Model and independent-Person sandwich covariance used the same
fits. The confidence level was 95%, without a small-cluster adjustment.

The table gives ranges across the eight original conditions and the specified
targets. Pointwise coverage concerns one target at a time. Family coverage
requires **every** interval in the specified call to cover its truth; a family
contains two differences, 27 probabilities or nine information values.

| Target | Pointwise: model | Pointwise: sandwich | Bonferroni family: model | Bonferroni family: sandwich |
| --- | ---: | ---: | ---: | ---: |
| Standardized slope differences | 91.0–98.0% | 91.0–99.0% | 92.9–98.0% | 93.9–98.0% |
| Category probabilities | 88.0–100.0% | 85.6–99.0% | 84.7–96.0% | 83.7–97.0% |
| Information per rating | 90.7–100.0% | 89.8–99.0% | 95.0–100.0% | 92.0–98.0% |

These percentages condition on available intervals or complete available
families. Each condition has 100 independent datasets, not one independent
replication per curve point. Depending on the target and condition, 96–100
pointwise intervals were available. In the small incomplete unequal-rater-slope
condition, the sandwich probability family covered in 82 of 98 available
datasets: 83.7%, with a Monte Carlo 95% interval of 74.8–90.4%. It was covered
and returned in 82 of all 100 planned datasets. That Monte Carlo interval
describes this cell, not simultaneous uncertainty for the minimum across cells.

**Bonferroni does not fix an inaccurate marginal approximation.** The curve
results show a limitation even under the generating model's assumptions;
sandwich covariance is not an automatic solution. Read a returned interval
as an asymptotic approximation, especially in small incomplete designs.
Do not interpret numerical eligibility as demonstrated 95% coverage, turn
curve differences into rater-quality verdicts, or assume that the bootstrap
has resolved this limitation without its own repeated-dataset evidence.
The two source designs also change sample size and assignment together, so
the results cannot isolate an effect of sparsity. They do not establish
performance for other grids, misspecified models or the latest refitted solutions.

### Checking the updated estimator on the affected design

The entire small incomplete, unequal-rater-slope condition was then refitted:
all 100 datasets, including cases whose intervals missed the truth or were
unavailable. The updated numerical implementation used ordinary initialization
and the same model, 61-point quadrature, iteration ceiling and convergence
tolerance. All fitted parameters, objective values, probability estimates and
interval endpoints were identical to the earlier results. Model and sandwich
Bonferroni families still covered 83/98 and 82/98 available datasets,
respectively. Updating the optimizer therefore did not resolve this shortfall
in the selected condition.

These are the same 100 datasets already included in the 800-fit study, not
another 100 independent replications or an independent confirmation after
selecting a condition. Analytic probability derivatives and full-covariance
delta intervals were also checked on two prespecified refits. The largest
endpoint discrepancy from the API was below `2e-9`. This checks the probability
transformation; it does not establish global optimization, quadrature accuracy
or finite-sample coverage, nor identify approximation error as the only cause.

For a small incomplete assessment, use these bands as approximate descriptions
of calibration uncertainty. Do not rely on their nominal 95% label alone to
claim that an entire curve is covered or that two curves clearly differ.
Such a conclusion needs an uncertainty method justified for the study's design
and model. A larger sample, a simpler model or bootstrap may be worth assessing,
but none is an automatic correction established by this check. Printing and
default figure subtitles now identify the intervals as approximate;
`CIEligible` continues to describe numerical availability.

### Carry the selected inference into reports and figures

A rater's severity, discrimination and residual fit answer different questions.
Wright maps locate fitted abilities and facet effects; Pathway maps relate
location to fit statistics. A slope interval instead describes discrimination
under its stated scale and covariance method. Adding a slope interval must not
replace severity intervals, Infit/Outfit, or a feedback decision with a different
target. In particular, a larger slope does not automatically mean a better rater
or a larger recommended scoring weight.

```{r gpcm-saved-inference, fig.alt="Population-SD-standardized discrimination estimates with explicitly selected model-based pointwise 95 percent intervals. The plot concerns discrimination, not rater severity or quality."}
ci <- confint(fit, scale = "standardized")
results <- mfrm_results(fit, intervals = list(slopes = ci, curves = curves),
                        include = c("fit", "plots"), compute = "never")
plot(results, type = "gpcm_slopes", title = NULL)
apa_table(ci)
head(plot_data(results, type = "gpcm_curves", component = "table"))
```

The reproduction code in `summary(results)$reproducible_code` saves and reloads
the complete result, preserving attached inference. In that code, `res` denotes
your result object (`res <- results` here); choose the file path before saving.
The starter export index also links the saved GPCM inference figures alongside
the ordinary maps, with separate interpretation guidance.

The same stored values are available through `mfrm_report(results)` and
`export_mfrm_results(results, ...)`. Named results define lowercase plot routes.
Use `as_ggplot(results, type = "gpcm_curves")` for further ggplot customization;
`title = NULL` or `subtitle = NULL` removes the corresponding annotation.
Plots distinguish unavailable intervals by crosses and unbounded intervals by
open circles, retaining their full limits and reasons in the tables. Color and
line type distinguish probability curves. No reference cutoff is chosen for a
slope plot; `reference` is an explicit optional plotting choice.

Source matching checks fitted data, parameters, population and integration
settings before attachment. Reusing a saved bundle does not change the selected
method, recompute intervals or rerun a bootstrap. For bootstrap slopes, pass
`confint(boot, scale = "standardized", level = .90)` to preserve that choice;
passing the raw `boot` object selects its default relative 95% pointwise
intervals. The export replay reloads the saved RDS. The full archive contains
the fitted data and identifiers, so review it before sharing.

### Check that the integration grid does not change the decision

`mml_quadrature_sensitivity(fit, data, quad_points = c(41, 61))` preserves
explicit population formulas, covariates and factor coding. Its slope table
contains both interval endpoints and eligibility; the summary reports their
changes. For a PCM/GPCM comparison, apply the helper to each model and rerun
`compare_mfrm()` on their fits at each **common** grid. Compare the IC/LRT
decisions as well as estimates; increasing the GPCM grid alone changes the
comparison basis. A small observed change is evidence for that fit and grid,
not a universal integration certificate.

Information criteria and parameter intervals need different evidence. The
[R documentation for AIC](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/AIC.html)
requires maximized likelihoods on comparable data. Computing an AIC number
does not require a slope confidence interval. In mfrmr, the comparison uses
the declared parameter count, numerical solution and integration settings.
Users should assess competing starts and integration sensitivity, particularly
when a model-choice decision is close.
An interval procedure also needs a justified covariance or other uncertainty
method, the correct constrained-parameter transformation, and evidence about
its repeated-sampling behavior. Neither decision requires a theorem covering
every conceivable GPCM likelihood path.

An unavailable slope interval does not automatically invalidate a separately
eligible IC comparison or matched PCM/GPCM test. Older saved fits must not gain
intervals from stored flags alone; rerun `confint()` or `diagnose_mfrm()` to
check the retained solution and refresh its uncertainty tables.

### Why intervals, information criteria and tests need separate checks

**Slope confidence intervals.** The package uses the joint covariance in
free coordinates, maps it to sum-zero log slopes and exponentiates the log-Wald
limits. Its separate qualification checks preserve missing results for singular,
regularized or unstable information. The result is an asymptotic approximation,
not exact coverage for every possible dataset. The
[mirt documentation](https://philchalmers.github.io/mirt/reference/mirt.html)
also provides GPCM fitting and information-matrix SE methods; that establishes
a methodological precedent, not numerical validation of mfrmr.

**Information-criterion comparison.** The package already calculates
\(\mathrm{AIC}=-2\widehat\ell+2p\) and
\(\mathrm{BIC}=-2\widehat\ell+p\log N\), with \(N\) equal to the number of
Persons for the ordinary MML route and \(p\) the free-parameter count.
For GPCM, the comparison reevaluates the retained likelihood, gradient and
observed information without rerunning the optimizer. It checks agreement
with the stored objective, the existing gradient tolerance (at most
\(10^{-4}\)), and positive unregularized information using the same relative
eigenvalue tolerance as covariance inversion. IC/LRT and slope intervals share
the full joint information calculation. The old IC/LRT-specific 80-coordinate
cutoff is replaced by a dense-matrix workspace budget, configurable with
`options(mfrmr.max_information_bytes = 256 * 1024^2)`. This estimates eight
double-precision square matrices; it excludes data, integration arrays and
other process memory. An unavailable check returns its reason. This local check is separate from `InferenceReady` and
does not qualify a slope interval. Same-data/likelihood, free-parameter-count,
unit-weight and integration checks still apply. There is no requirement that
the candidates be nested. Local curvature does not prove global optimality,
and sorting candidate IC values is not a test that the winning model is true.

**PCM/GPCM likelihood-ratio test.** For the current relative-slope
parameterization, let \(\lambda_g=\log\alpha_g\), with
\(\sum_g\lambda_g=0\). The PCM null is
\(\lambda_1=\cdots=\lambda_G=0\). It imposes \(G-1\) independent restrictions
when the population model, steps, other facets and constraints are identical.
The null value \(\alpha_g=1\) is inside the positive slope space; it is not a
zero-variance boundary. Under the usual identified, interior and regular MML
conditions, the candidate reference is
\(2(\widehat\ell_{GPCM}-\widehat\ell_{PCM})\sim\chi^2_{G-1}\) asymptotically.
The implemented test verifies the matched null/alternative population model,
step and facet constraints, fixed interactions and free dimensions. It reuses
the likelihood, gradient and unregularized-information checks described above.
A negative likelihood gain is a numerical failure and does not become p = 1.
The reference assumes independent Persons and correctly specified response
and population models, with regular interior solutions. It is an asymptotic
approximation, especially uncertain in small or weakly linked designs.
Default calls can differ in population assumptions, so changing only
`model = "PCM"` to `model = "GPCM"` does not necessarily test slopes alone.
The [mirt model-comparison documentation](https://philchalmers.github.io/mirt/reference/anova-method.html)
distinguishes ordinary likelihood-ratio comparisons from boundary-null tests.
That distinction does not require a mixture chi-square test merely because
GPCM slopes are positive.

A simulation illustration examined the approximation under a true PCM,
using three raters, three criteria, three score categories and normally
distributed Person abilities. In the crossed design, each of 100 Persons was
rated by all three raters on all three criteria. In the rotating design,
each of 40 Persons was rated by two of the three raters on all three criteria.
The latter changes both sample size and assignment density, so it does not
isolate the effect of sparsity. Rater and criterion locations varied across
replications; the slope null was always true. Both fitted models estimated
the population mean and variance on the same 61-point integration grid.

| Slope facet | Rating design | Rejections at 5%, out of 100 | 95% Monte Carlo interval |
|---|---|---:|---:|
| Criterion | Crossed, 100 Persons | 5 | 1.6%--11.3% |
| Rater | Crossed, 100 Persons | 7 | 2.9%--13.9% |
| Criterion | Rotating, 40 Persons | 6 | 2.2%--12.6% |
| Rater | Rotating, 40 Persons | 7 | 2.9%--13.9% |

All 400 pairs supplied a test in these conditions. The intervals above are
exact binomial intervals describing uncertainty in the simulated rejection
rates, not parameter confidence intervals. The point estimates do not show a
large departure from 5%, but 100 replications per condition give limited
precision. They do not guarantee calibration under other sample sizes,
category support, dependence, assignment or population distributions. In the
first replication of each condition, increasing both fits to 121 integration
points changed the likelihood-ratio statistic by less than \(10^{-6}\); this
illustrates a sensitivity check, not a universal integration-error bound.

These explanations separate missing mfrmr functionality from failures of a
particular fit. They do not imply that every model needs every downstream
workflow, or that JML inherits the MML reference distribution.

## Relationship to generalized many-facet models

Ordinary GPCM adds a positive discrimination parameter to the partial-credit
kernel; it does not replace the item position or step parameters. In the
current many-facet notation,

\[
\log\frac{P(Y_o=k)}{P(Y_o=k-1)}
  = \alpha_{s(o)}\{\eta_o-\tau_{t(o),k}\},
\qquad \prod_g\alpha_g=1,
\]

where \(s(o)\) selects the slope owner and \(t(o)\) selects the step owner.
They must coincide for JML; MML also permits separate owners. The additive
predictor \(\eta_o\) contains person ability and the signed
locations of the observed facets. Thus the selected slope multiplies the
complete adjacent-category predictor: person ability, every facet location or
fitted interaction inside \(\eta_o\), and the owned step. It is not a
loading-only formulation in which discrimination multiplies ability but leaves
rater severity and other intercept effects unscaled. Unit slopes reduce this
expression to the equal-discrimination PCM kernel.

The phrase *generalized many-facet Rasch model* is broader and is not a unique
synonym for this implementation. For example, Uto and Ueno (2020, equation 9)
use multiplicative task and rater slope blocks,
\(\alpha_i\alpha_r\), together with rater severity and rater-specific steps.
The one-family mfrmr GPCM can separate the slope owner from the step owner
under MML; JML requires `slope_facet == step_facet`. The provisional two-family
MML--EM route jointly estimates the two slope blocks in this equation on a
fixed standard-normal ability scale. Neither route introduces a
multidimensional ability vector. Selecting criterion slopes and rater steps
therefore remains a restricted many-facet GPCM, with the entire adjacent-category
predictor multiplied by the selected criterion slope.

A rater-owned fit is close to a restricted Uto--Ueno form after suppressing the
task-slope block, provided the remaining step and identification contracts are
also aligned. A fit with both slopes and steps owned by Criterion should be
described as an analogous restricted many-facet GPCM, not
as an exact fit of Uto and Ueno's equation. Because free discrimination relaxes
the defining equal-discrimination Rasch restriction, "generalized MFRM" means
a model generalized from the MFRM baseline rather than a strict Rasch model.

### When would two slope families help in practice?

Consider a speaking assessment: each candidate completes several tasks, and
several assessors score overlapping sets of performances. Difficulty concerns
which tasks tend to receive lower scores. Severity concerns which assessors
tend to give lower scores. Discrimination asks a different question: how
strongly do the probabilities of higher categories change with ability?
Two slope families let that change depend on both the task and the assessor.
The provisional `fit_mfrm()` route supports exploring this formulation.
Experimental component-slope intervals are available when their numerical
checks pass. They do not establish a rule for feedback decisions; curve
intervals and individual two-family feedback sheets remain unavailable.

| Practical question | What the two-family model can describe | What it does not establish by itself |
| --- | --- | --- |
| Should a speaking task be reviewed? | A task slope describes its contribution to response sensitivity across assessors, alongside difficulty and predicted category curves. | A low slope does not establish poor content validity or require deleting the task. |
| Which assessors may benefit from feedback? | Severity, category steps and an assessor slope distinguish a shift in scores from a change in response sensitivity. | A large slope is not proof of scoring accuracy or rater competence; training decisions also need substantive review and, where appropriate, reference ratings. |
| Does the pattern recur in music assessment or clinical skills examinations? | Task and assessor labels can be replaced by piece/station and judge/examiner while keeping their declared model roles. | Renaming columns does not justify a single ability dimension across unrelated musical or clinical skills. |

For a simple numerical illustration, let a task slope be 0.8 and an assessor
slope be 1.2. Their effective slope is 0.96. Combining the same task with an
assessor slope of 0.7 gives 0.56. These are multipliers in the adjacent-category
log-odds equation, not score weights, accuracy percentages or observed score
increases. The actual expected-score change also depends on ability, locations
and category steps. In particular, a large slope can coexist with little
information where almost everyone receives an extreme category. Read the
probability and information curves over the ability range relevant to the
assessment, not just a ranked slope table.

The multiplication is also a substantive restriction. The task/assessor
sensitivity factors combine as a product; they are not a freely estimated
slope for every task-by-assessor pair. A particular assessor's unusual reaction
to only one task may require a different model. Nor do two slopes create two
latent abilities: the formulation discussed here has one ability dimension.

A useful analysis starts with the decision and rating design:

1. Define the ability being assessed, the meaning of each facet and the proposed
   feedback or task-review decision. Keep the column labels distinct from the
   chosen slope reference and step owner.
2. Inspect who rated which performances. Overlapping performances and multiple
   ratings per person help separate effects. A connected assignment graph alone
   does not establish that both slope families are identified; inspect weak
   links and systematic differences in the assigned candidate groups too.
3. Compare equal-slope and simpler one-family explanations on matched data,
   response equations and scales before attributing a pattern to two families.
   Supported one-family PCM/GPCM comparisons are available now; the current
   comparison API does not qualify a two-family test or ranking.
4. For a qualified two-family analysis, inspect uncertainty, starting-value and
   integration sensitivity, curves and substantive evidence together. Mark
   predictions for combinations not observed together. Use the findings to
   investigate tasks or feedback needs, not as an automatic removal or
   competence rule.

### Start from an actual rating table

Suppose an assessment coordinator wants to review how raters use a writing
rubric. The question is whether differences concern consistently higher/lower
scores, category use, or different sensitivity to writing ability. A two-slope
model is a candidate explanation when both criteria and raters may differ in
sensitivity. It is not automatically preferable to an equal-slope or
one-slope-family model, and a high rater slope is not evidence of accurate
scoring. Reference performances and a substantive review of the rubric are
needed before deciding what feedback to give.

The separately installed `sirt` package supplies the empirical
`data.ratings1` table from a 2009 Austrian grade-8 German writing survey
([sirt data documentation](https://alexanderrobitzsch.r-universe.dev/sirt/doc/manual.html#data.ratings)).
In sirt 4.2.133 it contains 135 students, seven observed raters, five criteria
(`k1`–`k5`) and scores 0–3. The stored rater factor has additional unused
levels; these are not additional observed raters. The documentation does not
provide descriptive names for the five criteria, so retain their labels
instead of inventing rubric meanings. Using this dataset does not make
mfrmr's response equation equivalent to `sirt::rm.facets()`.

The following conversion preserves every supplied score and identifier. It
requires `sirt` to be installed; it does not download data during analysis.

```{r empirical-writing-review, eval=FALSE}
library(mfrmr)
data("data.ratings1", package = "sirt")
writing_ratings <- do.call(rbind, lapply(paste0("k", 1:5), function(k) {
  data.frame(
    Person = as.character(data.ratings1$idstud),
    Rater = as.character(data.ratings1$rater),
    Criterion = k, Score = data.ratings1[[k]]
  )
}))
writing_review <- describe_mfrm_data(
  writing_ratings, person = "Person", facets = c("Criterion", "Rater"),
  score = "Score", rating_min = 0, rating_max = 3,
  category_policy = "preserve", rater_facet = "Rater"
)
writing_review$design_connectivity
writing_review$structural_missingness$summary
# Count distinct raters for each student, independently of criterion rows.
writing_pairs <- unique(writing_ratings[c("Person", "Rater")])
table(table(writing_pairs$Person))
xtabs(~ Rater + Score, writing_ratings)
```

There are 1,370 observed ratings: 89 students have one rater, 27 have two,
two have six, and 17 have all seven. The rater network is connected, but its
information is concentrated unevenly across students. No score is missing
within the supplied rows. That does **not** establish that every intended
rating was collected: a separate planned assignment roster is unavailable.
The unobserved combinations must not become zeros or automatically be labelled
missing at random. Each criterion appears with each rater, but five
rater-by-criterion-by-category counts are zero. These observations matter
more than a single overall sample-size recommendation.

To examine criterion and rater slopes, use the preceding two-family call with
`data = writing_ratings`, `person = "Person"`,
`facets = c("Criterion", "Rater")`, `score = "Score"`,
`slope_facet = c("Criterion", "Rater")`, `step_facet = "Rater"` and
`noncenter_facet = "Rater"`. Keep the declared 0–3 categories. This assumes
one writing ability, a common standard-normal population, conditional
independence of ratings, and rater-owned steps shared across criteria.
Five analytic scores on the same performance may challenge independence;
the fitted slope table does not test that assumption.

If your records distinguish Task, Criterion and Rater, preserve all three
roles during design review. The current two-family route admits exactly two
non-Person facets. Pooling task and criterion labels or dropping a facet
changes the model and can conceal dependence; such a dataset needs an
explicit model-scope decision before fitting.

An analysis of the complete table at 61 and 121 quadrature points converged
under the same EM stopping rule, yet the maximum category-probability
difference on a common ability grid from -4 to 4 was about 0.17. This compares
fitted curves, not empirical predictive accuracy; the discrepancy also occurs
within -2 to 2. Component-slope intervals were unavailable for both fits.
At the 61-point calibration,
the response-diagnostic check at 121 versus 243 points retained only 318 of
1,370 rows under its probability tolerance; all grouped Infit/Outfit values
remained unavailable. Thus this example does **not** support issuing rater
feedback from the returned estimates. Integration and solution sensitivity
need resolution first. More convergence messages, a low in-sample residual,
or a ranked slope plot cannot replace that review.

For a different operational setting, Uto et al. (2024) publish medical
interview OSCE scores and their data description on
[Dryad](https://doi.org/10.5061/dryad.tmpg4f56q): two raters cover every examinee
and three additional raters cover different subsets. This asks whether rubric
use and rater effects differ in a small assessment cohort. Their
[published model](https://doi.org/10.1371/journal.pone.0309887) includes
rater-by-item location interactions and item-specific steps, which the
current two-family route does not implement. A native fit would be a
restricted analysis, not a reproduction of that study. The writing example
does not qualify OSCE, music or sport applications by renaming facets.

### Can EM estimate two slope families?

Yes. With fixed task and rater effects and one latent ability, adding a second
slope family does not add another latent integration dimension. MML--EM can
combine all ratings of each person in the E-step and jointly update task
slopes, rater slopes, locations and steps in a constrained numerical M-step.
This extends the estimation approach used for GPCM by Muraki (1992); it does
not turn the model into JML. Numerical EM can converge slowly or to a local
solution, and small likelihood changes alone do not establish convergence.
Uncertainty must account for unobserved ability: the curvature of a single
M-step is not the observed marginal information matrix.

In the two-family formulation discussed here, fixing the ability distribution
to N(0,1) and the geometric mean of task slopes to one leaves the rater slopes
free. A task slope of one is the task-family geometric reference. A rater
slope of one is a unit multiplier on this model's identified scale, not the
average rater's slope. Neither is a cutoff for rater competence. Tables and
plots must identify both the owning facet and its level, even when task and
rater levels use the same labels. The effective slope for a rating is their
product, not either component alone.

"Task" and "Rater" describe the example's roles, not required column names.
The same roles could be labelled "Criterion" and "Assessor". Keep the column
names separate from the choice of slope reference and step owner: changing a
label should not change the model, whereas assigning steps to a different
facet generally does. Neither the order of columns in a data frame nor words
such as "rater" should silently select the model's constraints. The public
`fit_mfrm()` accepts chosen person, facet and score column names. For two
families, the order of `slope_facet = c(first_owner, second_owner)` declares
their roles explicitly; it is independent of data-column or `facets` order.

[`sirt::rm.facets()`](https://alexanderrobitzsch.github.io/sirt/reference/rm.facets.html)
provides an existing MML--EM route with item and rater slopes. Its documented
slopes multiply ability, with item category intercepts and unscaled rater
severity. The Uto--Ueno equation instead multiplies the whole adjacent-category
predictor and uses rater steps. They are generally different models, so their
estimates should not be compared as alternative optimizers without first
matching the response equation and identification. Availability of both
fitting routes does not establish software equivalence or qualify uncertainty
for mfrmr's provisional two-family route.

Assignment also matters. Having every task rated by every rater is not enough
to distinguish two slope families. For example, with two tasks, two raters,
binary scores and a fixed standard-normal ability distribution, the two-family
model above has six free location/slope parameters. If each person supplies
only one rating, the data provide just four marginal success probabilities,
one per task-rater pair. More people with that same assignment cannot resolve
all six parameters. Several ratings of the same person provide additional
information through their joint response patterns. This does not guarantee
precise estimates: category use, sample size, the population assumption and
numerical integration still matter. Optimizer convergence alone cannot resolve
an insufficient assignment design.

### What does changing the ability SD test?

A narrow ability distribution may give different category exposure and less
information about a response curve than a broad distribution. But parameter
recovery must first be compared on the same identified scale. For the two-owner
equation above, write ability as `theta = mu + sigma*z`, with standard-normal
`z`. The same response probabilities result when task slopes are unchanged,
rater slopes are multiplied by `sigma`, task locations and rater steps are
divided by `sigma`, and rater locations become `(location - mu)/sigma`.
This preserves the task-slope geometric mean of one. A change in raw slope
values across these representations is not by itself estimator bias.

A single correctly transformed normal population is different from rater
panels assigned different ability means or SDs, or a skewed/mixture population
fitted with a normal model. Common rated performances help connect panels, but
are not fixed ability anchors unless their values are explicitly supplied.
Neither overlap counts nor fixed severity anchors alone establish the ability
scale and both slope families. The one-family GPCM and provisional
two-family formulation have distinct identification constraints; do
not transfer parameter values or anchor settings without matching them.

### Do two slopes handle testlets or local dependence?

No. Two fixed slope families allow the response curve to differ across two
facets. They still assume independent ratings after conditioning on ability
and the specified fixed parameters. Two scores from the same performance may
remain related beyond ability: for example, fluency and accuracy ratings may
both reflect an unusually successful response to one speaking task. Merely
adding task and assessor slopes does not model this shared fluctuation.

**Local independence is a conditional assumption.** A testlet model adds a
latent effect for a block of related ratings and assumes independence after
conditioning on ability and that effect. Integrating out the effect leaves
dependence among the block's ratings. This is compatible with one substantive
ability, but it introduces additional latent variables for dependence. It
should not be described as having only one latent variable or as measuring
several substantive abilities automatically.

| Feature | Speaking-assessment example | Current mfrmr route |
| --- | --- | --- |
| Two fixed slope families | Task and assessor jointly determine the effective response slope. | Provisional fixed-population MML--EM with summary and conditional curves; no local-dependence correction or interval inference. |
| Person-local testlet | One candidate's performance affects multiple scores within a task. | `fit_mfrm_testlet()` for RSM, one common local variance and one nonoverlapping block membership per row. |
| Shared random rater severity | An assessor's usual severity affects everyone they assess. | `fit_mfrm_random_rater()` for approximate RSM MML, with its documented uncertainty limitations. |
| Random discrimination | Assessors' slopes arise from a specified population distribution. | Not supplied by either two fixed slope families or random rater severity; no current public route. |

A testlet effect is itself a random effect; the distinction from random rater
severity is which ratings share it, not whether the model uses random effects.
Block identity must follow the assessment process. A Person-by-Task effect
shared across assessors represents something different from a
Person-by-Task-by-Assessor impression shared only across that assessor's rubric
scores. With `testlet = "Rater"`, the current testlet API assigns a separate
effect to each Person/rater pair; it does not create one rater effect shared
across persons. Repeated ratings and assignment overlap are needed to separate
such effects. Nonzero testlet variance is not proof of halo, and zero estimated
variance in a weak design is not proof of independence.

The current testlet and random-rater routes are separate RSM models. Neither
adds dependence effects to a GPCM fit, and they cannot yet be combined.
A future combined model must specify how effects are shared, whether a local
effect is multiplied by discrimination, and how all shared effects are
integrated. Independently redrawing a common rater for each person would change
the model. EM cannot resolve that problem by itself.

For feedback and planning, specify whether a prediction concerns an observed
performance/rater or a new one. An observed block's fitted effect is not the
predictive distribution of a new block. Read model-specific residuals and
uncertainty alongside the response curves; ordinary infit thresholds do not
automatically become calibrated tests in the extended models. Do not sum
per-rating marginal information as if dependent ratings were independent, or
interpret testlet variance as a G-theory coefficient. A testlet model also does
not automatically make differently sized tasks equally weighted.

The [Rasch testlet model of Wang and Wilson (2005)](https://doi.org/10.1177/0146621604271053)
and their [random-effects facet model](https://doi.org/10.1177/0146621605276281)
provide the local-effect background. [Rijmen (2009)](https://www.ets.org/research/policy_research_reports/publications/report/2009/hycg.html)
explains how testlet and bifactor formulations can be related under particular
loading constraints. These relationships do not mean every latent effect or
multilevel model is interchangeable. For available operations, consult
`?fit_mfrm_testlet`, `?fit_mfrm_random_rater` and their prediction methods.

The one-ability assumption can be examined before a multidimensional GMFRM
is available. The [visual diagnostics guide](mfrmr-visual-diagnostics.html#checking-one-substantive-dimension-without-fitting-a-multidimensional-model)
compares residual PCA, Q3-style and marginal screens, model-generated reference
checks and explicit alternative-model comparisons. It also distinguishes
residual networks/EGA from assignment networks. Current residual screens are
exploratory; neither a residual component count nor a network community count
is a certified number of abilities.

### How does this relate to multivariate G theory?

The two approaches answer different questions about the same assessment.
A many-facet GPCM describes category responses in terms of ability, task
location, rater severity, steps and discrimination. Multivariate G theory
instead asks how observed scores on several criteria, such as fluency and
accuracy, vary together across persons, raters and tasks. It can then compare
assessment plans, for example adding raters or tasks for a prespecified
weighted score. Two slope families still describe one latent ability; they
do not create two score dimensions or estimate their G-study covariances.

| Question | Existing route | Interpretation |
| --- | --- | --- |
| How do category responses and rater effects vary with ability? | `fit_mfrm()` and its response/diagnostic outputs | A latent response model; the provisional two-family route has a narrower output scope than one-family GPCM. |
| How much observed-score variation comes from persons, raters, tasks and their interactions across criteria? | `mfrm_multivariate_gstudy()` | Score-specific variance and between-score covariance components. |
| What changes when the assessment uses more raters/tasks or different prespecified score weights? | `mfrm_multivariate_d_study()` and `plot()` | G, Phi and SEM for the declared future design. |
| How uncertain is the difference between two prespecified plans? | `mfrm_multivariate_d_compare()` | Approximate paired intervals under explicit normal random-effect assumptions for two common crossed facets. |

Start the G-study from numeric score columns and the declared design, not a
matrix of GPCM parameter estimates. The current G-study supports one or two
common random measurement facets, including its documented nested structure.
Balanced ANOVA and identifiable incomplete-design MINQUE(0) analyses are
available. The latter does not correct informative missingness. D-study plans
use the documented balanced future designs; their help describes the distinct
restrictions on point estimates and plan-difference intervals. G/D-study objects
have their own `summary()`, `plot()` and `plot_data()` routes and are not
`mfrm_results()` inputs.

For design decisions, begin with the existing two-dimensional D-study plots:
condition count on the horizontal axis, G/Phi or SEM on the vertical axis,
with separate lines for the other count. Inspect one criterion or declared
composite at a time. Multiple score criteria do not require a 3D surface.
An increase in G/Phi across populations need not mean smaller measurement
error: coefficients depend on person variation as well as error variance.
Read SEM in score units alongside them. The plot help explains the distinction
between point projections and intervals for prespecified plan differences.

Individual rater feedback sheets currently support native additive RSM/PCM,
not GPCM. Use the supported GPCM-specific outputs for discrimination and
response curves; do not relabel an ordinary-model sheet as a GMFRM report.

The separate `mfrm_generalizability(fit)` convenience route does accept an
MFRM fit, but uses its stored ratings (or supplied data) for an observed-score
main-effects analysis. It does not convert latent model parameters to
G-theory components or fit the multivariate model.

There is direct precedent for complementary use. [Hirai and Koizumi
(2013)](https://doi.org/10.1080/15434303.2013.824973) used multivariate G theory
with mGENOVA to examine criterion-specific dependability across stories and
MFRM with Facets to study ratings and scales. Their G-study excluded raters;
that design does not by itself establish generalization to new raters. The
paper also set negative variance estimates to zero, whereas mfrmr retains
raw component estimates and reports when a requested coefficient is
unavailable. Using the same broad methods is not a numerical reproduction of
that analysis. [Jiang et al. (2020)](https://doi.org/10.3758/s13428-020-01399-z)
explain estimation of multivariate G-theory parameters through a linear mixed
model framework; this is not an automatic conversion of GPCM slopes.

There are also joint IRT/G-theory approaches, including [Briggs and Wilson
(2007)](https://doi.org/10.1111/j.1745-3984.2007.00031.x). Such integration
requires specifying the population of tasks/raters, the score or latent scale
and the quantity to generalize. It is not implemented by passing the sampling
covariance of estimated GPCM slopes to a G-study: that covariance describes
estimation error, not sources of variation in observed scores. G-study score
weights express the intended assessment composite, not estimated GPCM
identification or discrimination weights.

For complementary analyses, keep the person, rater and task identifiers and
state which ratings and score criteria each analysis uses, including any
aggregation or exclusions. Do not silently replace unassigned ratings with
predictions or analyze EAP scores as error-free observations. Those choices
change the target and its uncertainty. A joint GMFRM/G-theory model with
propagated calibration uncertainty and a declared generalization population
remains a separate extension; the current APIs do not claim that integration.

### What can the related facets literature add?

Wang and Liu (2007) combine item discrimination and item steps with latent
regression: person background variables explain the latent ability distribution
while response uncertainty remains in the model. The existing MML arguments
`population_formula` and `person_data` provide a conditional-normal latent
regression route in mfrmr. Regressing fitted EAP scores as if they were observed
without error is a different procedure. Their model does not itself add a
second, multiplicative rater-slope family, and its item steps differ from the
rater steps in the Uto--Ueno example.

Wang, Wu and Qiu (2025, preprint) propose a derivative-based rater capability
index for binary ratings. The useful practical question is how a rater's
response changes across the ability range. In a GPCM context with effective
slope A, expected-score sensitivity is `A * Var(Score | Theta, context)`, whereas
information is `A^2 * Var(Score | Theta, context)`. A steep curve and a large
amount of information are related but not identical. Existing probability and
information curves can help examine this distinction; mfrmr does not currently
provide the paper's normalized capability index.

Population-averaged feedback also depends on the ability distribution and the
tasks used for averaging. Comparing raters on different assigned populations
can change the comparison; a common reference population answers a different
question from each rater's actual assignment. Such summaries describe a fitted
response model, not agreement with an independently established correct score.
The preprint's binary normalization and Laplace approximation should not be
treated as validated defaults for polytomous GPCM or sparse rating panels.
EM and Laplace address different numerical choices: EM is an optimization
strategy, whereas Laplace approximates integration over the latent ability.
A Laplace-based marginal likelihood is not ordinary JML, and its accuracy
needs checking separately from optimizer convergence.

### Which facet receives different slopes?

The slope is not one common number for the entire fitted GPCM. The selected
`slope_facet` contributes one positive slope for each of its levels. If it has
\(G\) levels, the returned table contains \(G\) slopes and the model estimates
\(G-1\) free relative log-slope contrasts because their geometric mean is fixed
to one.

| Configuration | Meaning | Current status |
|---|---|---|
| `step_facet = "Criterion"`, `slope_facet = "Criterion"` | Each criterion has its own relative discrimination and criterion-specific steps. | Supported; inference availability is listed above. |
| `step_facet = "Rater"`, `slope_facet = "Rater"` | Each rater has its own model-conditional relative discrimination and rater-specific steps. | Code-supported with rater-interpretation caveats; a slope is not automatically evidence of rater consistency. |
| Criterion steps with rater slopes, or conversely | Slope owner and step owner differ. | Supported with MML; one slope family and existing output-specific inference checks. JML, weighting reviews and simulation/design workflows are unavailable for this structure. |
| Criterion slopes and rater slopes together | Effective discrimination is the product of two slope blocks. | Provisional MML--EM on fixed N(0,1), exactly two facets and second-owner steps; summary, fitted curves without intervals and separately checked experimental component-slope intervals. |

Every facet that is not selected remains an additive location term inside
\(\eta\). For example, a criterion-owned GPCM can still estimate rater severity,
but it does not give each rater a separate discrimination parameter. The
criterion slope nevertheless scales that rater-severity contribution because
the contribution is inside \(\eta\). “No separate rater slope” and “rater
severity is unscaled” are therefore not the same statement.

### Criterion discrimination with rater-specific category use

Suppose several raters score each performance on several rubric criteria.
Use `step_facet` for **whose category transitions may differ** and `slope_facet`
for **whose responses may distinguish ability more sharply**. These are distinct
questions from overall rater severity, which remains a location effect.

```{r separate-owners, eval=FALSE}
fit <- fit_mfrm(ratings, "Person", c("Rater", "Criterion"), "Score",
  model = "GPCM", method = "MML",
  step_facet = "Rater", slope_facet = "Criterion")
slopes <- confint(fit)
grid <- expand.grid(Theta = seq(-2, 2, length.out = 21),
                    Rater = "R01", Criterion = "C01") # use your fitted labels
curves <- mfrm_curve_intervals(fit, grid)
plot(curves, title = NULL, subtitle = NULL)
results <- mfrm_results(fit, intervals = list(slopes = slopes, curves = curves),
                        include = c("fit", "plots"), compute = "never")
saveRDS(results, "criterion-slopes-rater-steps.rds")
mfrm_report(readRDS("criterion-slopes-rater-steps.rds"))
```

Relative slopes have geometric mean one. Step contrasts are centered within
each rater, separately from severity. Adequate crossing and category use are
needed to distinguish these effects; a sparse or confounded design can leave
intervals unavailable. A large slope or unusual step is not, by itself, a
rater-quality or training diagnosis. The numerical workflow checks do not
establish finite-sample coverage for this new structure.

For a PCM comparison, keep the same data, rater steps, constraints and population
model; the GPCM adds one fewer slope contrast than there are criteria. The default
GPCM estimates a population mean and variance, whereas default PCM fixes its
population, so defaults alone do not form that matched comparison. Follow the
matched population example earlier in this vignette. Existing likelihood and
local-information checks still apply. `bootstrap_mfrm_gpcm()` preserves both
owners and the observed assignment for parametric resampling; this is not a
new-design simulation. `build_weighting_review()`, `extract_mfrm_sim_spec()`
and the simulation/design workflows still require a shared owner. Portable
GPCM calibration and simultaneous rater/criterion slope families remain unavailable.

### An explanation for a high-school reader

Imagine that several teachers grade presentations on several criteria.

- **PCM** gives everyone rulers with the same magnification. One criterion may
  be harder and one teacher may be stricter, but a one-unit ability difference
  changes the rating probabilities at the same basic sharpness everywhere.
- **The current mfrmr GPCM** lets you choose one box of rulers. If the box is
  `Criterion`, each criterion can have a different magnification. If the box is
  `Rater`, each rater can have a different magnification. You may choose only
  one box for magnification. With MML, another box may supply the category
  steps: for example, criteria can differ in magnification while teachers
  differ in how they use the score categories. With JML, the same box must
  supply both magnification and category steps.
- **A broader generalized MFRM** can give both the task/criterion and the rater
  their own magnification knobs. In the Uto--Ueno example, the effective
  sharpness is the product of the task and rater settings. It can also model
  rater severity and rater-specific use of rating categories.

Thus the current GPCM is a smaller branch inside the broader family: it adds a
carefully controlled slope vector to a many-facet PCM, but it is not the full
two-knob generalized MFRM. Also, once discrimination is allowed to vary, the
model is generalized *from* the Rasch baseline; it no longer retains the strict
equal-discrimination Rasch restriction.

## Before fitting: model-choice triage

Do not choose `GPCM` only because it is the most flexible model in the menu.
Start with the score interpretation.

| Model | Use when | Main risk if over-used |
|---|---|---|
| `RSM` | The rating scale is intended to share one category-threshold structure across the step facet. | Real threshold differences can be hidden in residual diagnostics. |
| `PCM` | Thresholds may differ by item, criterion, task, or another designated step facet, but rating events should still contribute equally after conditioning on the modeled facets. | It can absorb threshold heterogeneity without asking whether some levels are more discriminating. |
| `GPCM` | The analysis explicitly allows discrimination-based reweighting and treats slopes as part of the substantive sensitivity question. | Better statistical fit can be mistaken for a better operational scoring rule. |

This ordering matters for reporting. `RSM` and `PCM` are the package's
equal-weighting reference route; `GPCM` is a slope-aware extension.
If equal contribution of items, criteria, or raters is part of the validity
argument, a better-fitting `GPCM` should be reported as sensitivity
evidence rather than as an automatic replacement.

Category avoidance is a separate decision. For example, a rater who uses only
3--8 on a declared 1--10 scale shows local restriction of range if the omitted
categories are used elsewhere. Start with `data_quality_report()` and inspect
`category_usage_by_facet` and `category_usage_summary`; also examine assignment
and case mix, fit, and unexpected responses. Do not move to `GPCM` merely
because a rater avoids extreme categories. A rater-owned PCM/GPCM may expose
owner-specific threshold or slope support problems, but it changes the model
being estimated and does not by itself explain or repair the scoring practice.

### Comparing PCM and GPCM

The package provides three complementary comparison layers:

1. `compare_mfrm()` returns AIC, Person-based BIC and SABIC comparisons when
   `comparison$table$ICComparable` is true. Its local-solution check is
   independent of slope-interval availability. Add `nested = TRUE` to test
   equal relative slopes under the matched PCM/GPCM specifications below;
   inspect `$lrt` and `$comparison_basis$lrt_reason`.
2. `build_weighting_review()` shows what the GPCM slopes changed: the slope
   profile, facet-measure shifts, and redistribution of information across
   levels.
3. `build_model_choice_review(..., run_weighting_review = TRUE)` bundles the
   statistical comparison, operational model roles, support boundaries, and
   the weighting review.

For example, `dat` is a long-format rating table with `Person`, `Rater`,
`Criterion` and `Score` columns. Fit the same estimated-normal population
in both models; `~1` estimates its mean and variance without covariates.

```r
persons <- data.frame(Person = unique(dat$Person))
fit_pcm <- fit_mfrm(
  dat, "Person", c("Rater", "Criterion"), "Score",
  method = "MML", model = "PCM",
  step_facet = "Criterion", quad_points = 61,
  population_formula = ~1, person_data = persons, person_id = "Person"
)

fit_gpcm <- fit_mfrm(
  dat, "Person", c("Rater", "Criterion"), "Score",
  method = "MML", model = "GPCM",
  step_facet = "Criterion", slope_facet = "Criterion",
  quad_points = 61,
  population_formula = ~1, person_data = persons, person_id = "Person"
)

comparison <- compare_mfrm(fit_pcm, fit_gpcm,
                           labels = c("PCM", "GPCM"), nested = TRUE)
comparison$lrt
comparison$comparison_basis$lrt_reason
comparison$table[, c(
  "Label", "LogLik", "AIC", "BIC", "SABIC", "ICComparable"
)]

weighting <- build_weighting_review(fit_pcm, fit_gpcm, nested = TRUE)
summary(weighting)
weighting$comparison_contract

review <- build_model_choice_review(
  PCM = fit_pcm,
  GPCM = fit_gpcm,
  run_weighting_review = TRUE
)
summary(review)
review$model_roles[, c(
  "Model", "StepStructure", "StepCoordinates", "FreeStepParameters",
  "SlopeStructure", "SlopeCoordinates", "FreeSlopeParameters",
  "FitReadiness", "FormalInference", "Interpretation"
)]
```

`StepCoordinates` and `SlopeCoordinates` count the values shown in fitted
tables. `FreeStepParameters` and `FreeSlopeParameters` count the independent
coordinates optimized after identification constraints. For example, with
four score categories and four levels of the step facet, PCM and GPCM display
12 step coordinates but optimize 8 free step parameters because each
three-step ladder is sum-zero. GPCM additionally displays four relative slopes
but optimizes three free log slopes because their geometric mean is one.

The `comparison_contract` describes the evidence available from the two fits.
Its `FormalModelSelectionAvailable` field follows the verified IC comparison,
not numerical convergence alone:

| Evidence tier | What may be concluded? |
|---|---|
| `mml_information_criterion_comparison` | The IC and solution checks pass. Compare candidate AIC/BIC and their differences; choose an operational weighting policy on separate substantive grounds. Use `confint(fit, parm = "slopes")` for separately checked relative-slope intervals and `nested = TRUE` for the equal-slope test. |
| `mml_descriptive_not_inference_ready` | Both fits passed numerical checks, but inferential qualification is incomplete. Available raw fit values and weighting changes remain descriptive. More quadrature points alone do not qualify GPCM selection. |
| `mml_screening_or_review_grid_only` or `mml_noncomparable_descriptive_only` | Other MML comparison restrictions remain; neither tier enables automatic model selection. |
| `jml_descriptive_reweighting_only` | The fitted slopes, measure shifts, and information redistribution may be reviewed, but the unpenalized JML likelihood gain does not select GPCM. |
| `mml_numerical_review_only` or `jml_numerical_review_only` | At least one fit still needs numerical review. A difference between retained optimizer traces is not inferential model evidence. |
| `cross_method_or_data_not_comparable` | The methods or prepared data do not supply the matched comparison required here. |

For a JML question, follow that table with recovery conditions rather than an
IC or LRT claim. `evaluate_mfrm_recovery()` already distinguishes
`unit_slopes`, `near_flat`, `moderate`, and `high_dispersion` generator
conditions. The unit-slope condition asks how often harmless sampling and
optimizer variation can create an apparent slope spread; the non-unit
conditions ask whether the declared heterogeneity can be recovered. FACETS
can support the aligned PCM/JML side of this review, but its Table 7
discrimination remains a post-fit diagnostic and not a free-slope GPCM fit.

PCM is obtained from the aligned GPCM response kernel by setting all slopes to
one. Calling `compare_mfrm(..., nested = TRUE)` records `PCM_in_GPCM` when the
structural restrictions are verified. The result in `$lrt` is returned only
when the numerical and likelihood checks also pass; otherwise read
`$comparison_basis$lrt_reason`. Older saved comparisons must be rebuilt to
obtain the new decision. Weighting consequences still need substantive review.

## Response-family boundary

Every current `fit_mfrm()` likelihood is ordered categorical. The distinction
between response values and observation frequencies is essential:

| Input or estimand | Current mfrmr treatment |
|---|---|
| Ordered binary response | Supported as a two-category ordered response. |
| Ordered polytomous response | Supported through RSM, PCM, or GPCM. |
| Unordered nominal/multinomial response | Unsupported; category order enters the current likelihood. |
| Poisson, negative-binomial, or grouped binomial-trial count | Unsupported as a response family. An integer `Score` is only an ordered category code. |
| Row-frequency/replication weight for an ordered rating | A positive numeric `weight` can weight that row's conditional ordered-category likelihood contribution. |

Thus, the fact that the fitted category probabilities sum to one does not make
the model a nominal-response or multinomial-logit model. Likewise, a frequency
weight does not turn the score into a Poisson count or account for dependence
among replicated ratings. Nor is it a general collapsed-person frequency
table: under MML, powering responses inside one Person's conditional pattern
is not equivalent to replicating a complete Person pattern after
marginalization. FACETS provides separate `Bn` and `P` response models; those
are outside the current mfrmr estimator even though FACETS-style coverage
documentation describes the handoff boundary.

Within that supported scope, an explicit identification contract is still
required. With default MML,
`gpcm_mml_identification = "free_population"` estimates an intercept-only
person distribution \(N(\beta_0,\sigma^2)\), while the level-specific log
slopes are constrained to sum to zero (geometric mean 1). The fitted slopes are
therefore relative discriminations with \(G-1\) free contrasts, and
\(\sigma\alpha_g\) gives the equivalent slope on a fixed-latent-standard-
deviation scale. This retains the conventional common-discrimination degree
of freedom. The explicit legacy mode
`gpcm_mml_identification = "fixed_standard_normal"` instead fixes both
\(\sigma=1\) and the slope geometric mean to one and is a narrower model.

## GPCM is not one estimator

Muraki (1992) supplies the GPCM response model and an MML-EM estimation route;
it is not a source for the present JML implementation. The present mfrmr JML
route is an identified fixed-effects, unpenalized joint likelihood with
post-fit boundary/recession auditing. This differs from penalized JML for GPCM
(Wijayanto et al. 2021), whose quadratic regularization changes the objective,
and from software routes that impose finite parameter boxes. A penalized,
adjusted, or box-constrained point estimate can be a useful comparator, but it
is not the maximizer of mfrmr's original unpenalized JML objective.

This distinction matters even when all methods return finite numbers. With a
fixed number of observations per person, jointly estimated person coordinates
are incidental parameters; structural JML estimates can retain finite-test-
length bias as the number of persons grows (Hessen 2025). An extreme response
pattern can additionally make the ordinary finite JML maximum unattained.
mfrmr therefore keeps recovery/bias evidence, boundary handling, numerical
convergence, and cross-software agreement as separate checks.
Under JML, the geometric-mean constraint resolves the scale of the jointly
estimated person coordinates and slopes.

## Report wording templates

Use wording that matches the model actually fitted:

- `RSM`: "We fit a many-facet rating-scale Rasch model, treating category
  thresholds as common across the step facet."
- `PCM`: "We fit a many-facet partial-credit Rasch model, allowing thresholds
  to vary by the designated step facet while retaining equal discrimination."
- `GPCM`: "We fit a generalized partial-credit many-facet model
  as a slope-aware sensitivity analysis; interpretation focused on whether
  discrimination-based reweighting changed the substantive conclusions."

Avoid wording that says `GPCM` "improves the score" solely because it
improves log-likelihood, `AIC`, or `BIC`. The model can fit better while changing
the scoring contract.

## Checking the support boundary

`gpcm_capability_matrix()` is the canonical reference. It returns one row
per helper family with a `Status` column drawn from `supported`,
`supported_with_caveat`, `blocked`, and `deferred`. Read `Boundary` for the
interpretive limit and `RecommendedRoute` for the route to use next. The
default print is deliberately compact; subset by status to inspect a focused
set of rows.

```{r capability-supported}
library(mfrmr)
gpcm_capability_matrix("supported")[, c("Area", "Status")]
```

```{r capability-with-caveat}
gpcm_capability_matrix("supported_with_caveat")[, c("Area", "Status")]
```

```{r capability-blocked}
gpcm_capability_matrix("blocked")[, c("Area", "Status", "RecommendedRoute")]
```

```{r capability-deferred}
gpcm_capability_matrix("deferred")[, c("Area", "Status", "Boundary", "RecommendedRoute")]
```

The matrix is intentionally conservative. A row stays in `blocked` or
`deferred` even when some individual computation is already available, because the
scope statement includes the interpretation needed for a complete public
workflow rather than only checking whether code executes.

## Slope values, boundaries, and uncertainty

The finite `LogEstimate` and `Estimate` columns in `fit$slopes` are optimizer
outputs, not automatic evidence of a finite estimand. Use `ParameterStatus` and
the `Primary*` columns first. Under JML, mfrmr can certify a monotone
sum-to-zero log-slope path while holding the retained additive coordinates
fixed. Such a result yields `unbounded_low`, `unbounded_high`, or
`unbounded_both`. If no such path is certified, the joint Person--facet--step--
slope problem remains unevaluated; the finite optimizer output is retained only
as a numerical trace. A second bounded check lets additive coordinates move
with an ordered positive/negative log-slope pair. A competitive path from this
family is recorded as a candidate, but `ParameterStatus` remains
`not_evaluated` and the primary value remains unavailable: this sufficient
path result does not determine the global maximum of the non-concave GPCM
likelihood. Failure to find such a path is equally limited to that family.
Under MML, Persons are integrated out, so neither conditional JML result is
reused as marginal-likelihood evidence. The MML branch records a separate
sufficient-path check for the fitted finite-quadrature objective, but that
check is not a complete analysis of the continuous marginal likelihood.
That global boundary question is separate from the local MML approximation.
`confint()` or `diagnose_mfrm()` qualifies the relative slopes using the current
likelihood, gradient, unregularized information and the other checks described
above. Qualified diagnostic tables use `ParameterStatus = "finite_local_solution"`
and `PrimaryEstimateBasis = "verified_local_mml_solution"`; this does not claim
a global maximum. The unmodified fit may still carry its original unevaluated
parameter audit. Attaching diagnostics synchronizes its slope parameter rows,
while the global fit-readiness record remains unchanged.

Local `Optimizer*SE` and `Optimizer*CI` calculations remain available for
numerical review where possible, even if ordinary intervals are withheld.
`SEEligible`, `CIEligible`, `CIUse` and `InferenceReview` identify the separate
uncertainty decision. A positive information matrix alone is insufficient,
and a regularized inverse cannot qualify ordinary intervals.

## What can and cannot be compared across programs

A column labelled "discrimination" is not sufficient evidence that two
programs fitted the same GPCM. FACETS reports an element-discrimination
diagnostic after estimating its Rasch measures and explicitly does not allow
that diagnostic to alter the other estimates. It is therefore not a matched
counterpart of a jointly estimated free slope in `mfrmr`. FACETS remains useful
for unit- or explicitly fixed-discrimination reductions, category/threshold
and facet-measure comparisons within a common Rasch specification, and as a
deliberately different-model diagnostic benchmark.

### Model structure: which effect does discrimination change?

TAM and ConQuest both support more flexible slope designs than the current
mfrmr interface. That does not make every many-facet GPCM in these programs
the same model.

| Route | Where slopes are assigned | Relationship to current mfrmr |
|---|---|---|
| `mfrmr::fit_mfrm(model = "GPCM")` | One selected facet supplies positive relative slopes. MML permits a different step owner; JML requires the same owner. | Each slope multiplies the complete adjacent-category predictor. |
| TAM `tam.mml.2pl()` | Item slopes, shared slope groups or a design `E`; Example 14c combines this with a facet intercept design `A`. | The example has slopes on ability and separately additive facet intercepts. `tam.mml.mfr()` alone does not estimate slopes. |
| ConQuest `scoresfree` | By default, one slope for each generalized item, meaning a combination of facets; a `C` design can constrain sharing. | Sharing slopes by criterion still does not impose their products with estimated rater effects in the intercept design. |

The TAM construction is documented in
[Example 14, Model 14c](https://alexanderrobitzsch.github.io/TAM/reference/tam.mml.html).
ConQuest's [Note 8, pages 1--5](https://www.acer.org/files/Note_8--The_ConQuest_4_Model.pdf)
describes its scoring design, population model and identification. It updates
the earlier ConQuest 3 note; limitations described only in that older note
should not be assigned to current ConQuest.

For a rubric criterion \(c\), rater \(r\) and category transition \(k\),
compare these adjacent log odds, using minus signs for difficulty and severity:

\[
\begin{aligned}
\text{Current mfrmr:}\quad &\alpha_c(\theta_p-d_c-\rho_r-\tau_{ck}),\\
\text{Additive-intercept design:}\quad &a_c z_p-b_c-u_r-v_{ck}.
\end{aligned}
\]

The second expression represents the TAM example and a ConQuest design with
criterion-shared scores and additive rater intercepts. It is not every model
those programs can fit. The comparison is an algebraic consequence of their
documented designs: the first expression contains the product
\(\alpha_c\rho_r\); independent linear intercept and slope designs do not
impose that product of two estimated parameters.

For example, on the mfrmr scale, two raters differing by 0.5 severity units
produce an absolute log-odds difference of 0.4 for a criterion with slope 0.8 and 0.625
for one with slope 1.25. In the additive-intercept model, the shared rater
contrast has the same log-odds effect at every criterion. These are different
assumptions about how rater severity acts. Changing which facet supplies
slopes and steps does not by itself change this assumption.

`sirt::rm.facets()` provides a useful second MML
reference: it estimates item and/or rater slopes under product-one centering and
an estimated normal trait spread. Its item-only GPCM and equal-discrimination
many-facet reductions are candidate matched comparisons after category, threshold,
quadrature, and centering conventions are aligned. Its general free-slope rater
kernel is not the current mfrmr kernel: sirt uses the product of item and rater
slopes on the trait term while rater severity remains a separate location term,
whereas mfrmr assigns one slope owner and multiplies the complete adjacent-
category predictor. Such results are different-model sensitivity evidence, not
automatic numerical validation. `immer` provides PCM-design JML/CML/CCML
routes and a hierarchical rater model, not a direct free-GPCM many-facet match.
Its role is consequently a unit-slope reduction or alternative-model
sensitivity analysis rather than free-slope numerical validation.

### Exact overlap and what has been checked

With items only, free item-specific steps can absorb the slope multiplication.
For an estimated normal ability population, write
\(\theta=\mu+X\beta+\sigma\epsilon\), with standard-normal \(\epsilon\), and
\(\prod_i\alpha_i=1\). Set
\(z=(\theta-\mu)/\sigma\), \(a_i=\sigma\alpha_i\),
\(b_i=\alpha_i(d_i-\mu)\), \(v_{ik}=\alpha_i\tau_{ik}\) and
\(\beta_z=\beta/\sigma\). Then
\(\alpha_i(\theta-d_i-\tau_{ik})=a_i z-b_i-v_{ik}\).
Here \(\sigma\) is the residual population SD given any covariates \(X\);
without covariates it is the population SD. This derives an identical response
model on a unit residual-variance scale;
the intercept/step signs and centering must still match each program.
Fixing both population variance and the geometric mean of slopes to one
instead imposes an additional restriction, so it is not the same conversion.
The overlap is for positive slopes; ConQuest also exposes a `positivescores`
setting, so the permitted parameter space must be checked.

A retained validation comparison exercises the exact item-only TAM
overlap on TAM's `data.gpcm` fixture at q=31 and q=41. TAM fixes the latent
variance and estimates absolute slopes; mfrmr fixes the geometric mean of its
relative slopes and estimates the latent variance. After applying the analytic
scale and location transformation, maximum q-specific differences were about
`1.8e-5` for relative slopes, `1.3e-5` for transition thresholds, `6e-6` for
fitted category probabilities, and `1.4e-6` for deviance. The transformed TAM
probability formula itself agreed to binary64 precision. This is external
evidence for the item-only GPCM kernel, coordinate map, and continuous MML
target; it is not a comparison of the rater-plus-slope model, a cross-engine
SE validation, or permission to override mfrmr readiness. These results use
TAM 4.3-25. A separate q=31 many-facet re-expression was retained as a
different-model comparison after the algebraic distinction above was found.

An earlier local ConQuest 5.47.5 MML run checked the item-only overlap with
one population covariate, 120 Persons, five items and four categories. The
deviances differed by about `1.3e-5` after matched identification. Its exact
coordinate transformation reproduces probabilities within `1e-14`. This is
historical, version-specific evidence for that model and mapping, not a fresh
check of the current source, interval coverage or a multifacet slope model.
The public ConQuest overlap bundle does not automate this GPCM comparison.

### Inference: same estimates do not imply the same intervals

The saved TAM 4.3-25 fits retain marginal slope and intercept SEs, but not the
joint slope covariance or
slope--intercept cross-covariance required by the nonlinear transformation to
mfrmr's relative slopes, thresholds, and free population scale. Two positive-
definite covariance matrices can preserve exactly those reported marginal SEs
while producing different transformed SEs. This non-identification means the
comparison must be withheld rather than completed by assuming zero covariance.
TAM's documented [`tam.se()`](https://alexanderrobitzsch.github.io/TAM/reference/tam.se.html)
also omits parameter covariances and labels loading SEs for `GPCM.design`
experimental. This describes that route and the saved objects; it does not
rule out obtaining a joint matrix through additional methods.

ConQuest's [command reference](https://conquestmanual.acer.org/s4-00.html)
provides `estimatecovariances` with `stderr=empirical` and `matrixout`, and
`show` table 10 reports parameter-estimate error covariances. In contrast,
`export covariance` exports the **latent-population** covariance. It is not
the sampling covariance needed for transformed calibration SEs. Parameter
ordering, constraints and the inclusion of free score parameters must be
checked in the actual exported matrix before using it. No such current-source
joint-covariance comparison is claimed here. ConQuest's estimated-score route
is an MML comparator; its JML route does not estimate free scores.

For example, let \(s=\log\sigma\) and \(\ell_i=\log\alpha_i\). A
population-SD-standardized slope has \(\log a_i=s+\ell_i\), so its delta-method
variance requires

\[
\operatorname{Var}(\log a_i)=\operatorname{Var}(s)+
\operatorname{Var}(\ell_i)+2\operatorname{Cov}(s,\ell_i).
\]

Multiplying the current relative-slope interval endpoints by the estimated
\(\sigma\) omits two of these terms. Use
`confint(fit, parm = "slopes", scale = "standardized")` to include all three;
the default still reports relative slopes. Conversely, converting absolute
slopes to geometric-mean-one slopes requires their joint covariance.

TAM's [`anova()` method](https://alexanderrobitzsch.github.io/TAM/reference/anova.tam.html)
reports information criteria and likelihood-ratio comparisons. These methods
and mfrmr's current IC/LRT operations still require comparable likelihoods.
Replicating another program's fit requires the same model. IC selection can
instead compare different, nonnested models when their likelihoods refer to
the same observed data with compatible likelihood definitions and constants.
Verify category coding, population specification, each model's free-parameter
count and integration accuracy. An LRT additionally requires a nested null;
different many-facet slope actions are not automatically nested.
The simple PCM/GPCM null with \(G-1\) relative-slope
restrictions does not establish a reference distribution for arbitrary
cross-program design changes.

These distinctions also govern diagnostics. FACETS' discrimination, `mfrmr`
residual PCA or bias screens, and hierarchical-rater variance components answer
different questions. Agreement among them can be supporting evidence; a
difference is not a software error until the fitted likelihood, estimand,
identification, data rows, and output transformation have first been shown to
match.

A population-level validation also asked whether the current
complete-predictor slope and a loading-only slope are merely two coordinate
systems. They reduce exactly when slopes are all one or when the non-owner
Rater contrast is zero. In a fully crossed four-Criterion/four-Rater design,
however, their identifying adjacent-logit difference in differences is zero
for loading-only and
\(-\{\alpha_c-\alpha_d\}\{\rho_r-\rho_s\}\) for the current kernel. After all
available slopes, severities, and transition boundaries were re-estimated,
moderate and strong fixed conditions retained population KL losses of about
0.0016 and 0.0085--0.0088 per response, with maximum common-grid probability
differences of about 0.047--0.184. q=31 and q=41 conclusions agreed closely.
These are fixed model-form demonstrations, not practical cutoffs or
finite-sample operating characteristics. They support treating a loading-only
kernel as a separate possible future family rather than as an interpretation
switch for an existing `GPCM` fit.

Sparsity changes this distinction structurally, not merely by reducing sample
size. On an observed Rater-by-Criterion graph with no cycles, the products
\(\alpha_c\rho_r\) can be represented exactly by new additive Rater and
Criterion coordinates. Thus even a connected tree cannot distinguish the two
slope actions. A known-ability validation compared four connected
graphs. At 250 Persons, truth-selection rates were about 0.96 under complete
crossing, 0.76--0.78 for a balanced eight-edge cycle, and only 0.58 for an
eight-edge cycle localized to two Raters and two Criteria; the connected tree
remained observationally equivalent. Each Person was observed on every edge,
and abilities and both parameter vectors were supplied to the comparison, so
these are optimistic fixed-parameter results rather than JML or MML model-
selection performance. The design lesson is to inspect which slope and
severity contrasts are traversed by cycles, not to infer identifiability from
connectedness, density, or an anchor percentage alone.

A subsequent finite-sample validation removed the known-ability advantage and
refitted both slope actions by direct MML. The common identification estimated
geometric-mean-one relative slopes, sum-zero Rater severities, mean-zero
transition boundaries, and a normal population mean and standard deviation.
Across 12 replications per truth and design at 250 training Persons, all 96
candidate fits converged and all full 19-coordinate Hessians were positive
definite. Two starts differed by at most about `7.9e-6` log-likelihood units,
and q=31 versus q=41 changed none of the 48 paired family decisions. Complete
crossing selected the true family in every training and independent holdout
dataset. Under the balanced cycle, training selected the truth in 7--9 of 12
datasets and median relative-slope RMSE increased from about 0.06 to as much as
0.12. A stricter retained-solution polish reduced the largest gradient by
roughly an order of magnitude without materially changing likelihoods. These
small fixed counts clarify optimizer, curvature, and design contributions;
they do not calibrate a universal selection, standard-error, or readiness
rule.

## Source-grounded recovery interpretation

The `GPCM` route follows Muraki's generalized partial credit model
and its information-function extension. The package-specific `slope_regime`
labels are narrower than that model theory: they summarize the centered
log-slope spread of the simulation generator so recovery evidence can be
read against a declared stress condition. They are not model-fit tests and
they are not literature-derived adequacy cut points.

For simulation reporting, read direct recovery checks in an ADEMP-style order:
the data-generating mechanism first, then the estimands and performance
measures, and only then the row-level recovery diagnostics. In practice, this
means:

1. Build or extract an explicit `mfrm_sim_spec`.
2. Run `evaluate_mfrm_recovery()` for the direct parameter-recovery question.
3. Run `assess_mfrm_recovery()` with practical RMSE/bias limits.
4. Read `summary(recovery_review)`, then
   `recovery_review$condition_reporting_notes` and
   `recovery_review$condition_review`, then
   `recovery_review$diagnostic_reporting_notes` and
   `recovery_review$diagnostic_review` when optional diagnostics were retained,
   then
   `plot(recovery_review, type = "status")`, then
   `plot(recovery_review, type = "metrics", metric = "rmse")`.

## Supported routes and verification evidence

The following `GPCM` routes are documented and verified within the
stated constraints:

- **Fitting and core summaries** via `fit_mfrm(model = "GPCM", step_facet =
  ...)`, with the direct `MML` engine. Omitting `slope_facet` keeps the step
  owner; naming another facet separates the two owners in MML.
- **Posterior scoring and information** via `predict_mfrm_units()`,
  `sample_mfrm_plausible_values()`, `compute_information()`, and
  `plot_information()`.
- **Curve and category views** via `plot(fit, type = c("wright",
  "pathway", "ccc", "ccc_surface"))`, `category_structure_report()`, and
  `category_curves_report()`.
- **Slope-aware simulation specifications** via `build_mfrm_sim_spec()`
  and `simulate_mfrm_data()`.
- **Direct recovery checks** via `evaluate_mfrm_recovery()` and
  `assess_mfrm_recovery()`, including fitted GPCM slope recovery on the
  log-slope scale.

## What works with caveats

The following are exposed for `GPCM` but should be read as exploratory
screens rather than as Rasch-style invariance evidence:

- `diagnose_mfrm()` and the residual and unexpected-response stack:
  `unexpected_response_table()`, `displacement_table()`,
  `measurable_summary_table()`, `rating_scale_table()`,
  `interrater_agreement_table()`, `facet_quality_dashboard()`,
  `plot_qc_dashboard()`, `plot_marginal_fit()`,
  `plot_marginal_pairwise()`.
- `reporting_checklist()` and `precision_review_report()` route to the
  supported direct tables and plots. The broader APA/QC/export family is
  available as caveated sensitivity-reporting output with explicit
  `gpcm_boundary` rows.
- `build_misfit_casebook()` inherits the exploratory screening framing of
  its underlying sources.
- `estimate_bias()` now provides GPCM conditional screening rows with
  slope-aware information and profile-likelihood columns. Treat these rows as
  screening evidence for follow-up, not as standalone confirmatory fairness
  tests. `unexpected_after_bias_table()` provides a descriptive in-sample
  before/after flag comparison; a lower flag count does not show that bias has
  been removed.
- `estimation_iteration_report()` provides a slope-aware reconstructed
  optimization trajectory. It is a diagnostic replay, not the exact optimizer
  history or an additional convergence test.
- `analyze_dff()`, `analyze_dif()`, `dif_interaction_table()`,
  `dif_report()`, `plot_dif_heatmap()`, and `plot_dif_summary()` provide
  GPCM DFF/DIF screening and reporting surfaces with explicit
  `gpcm_boundary` rows.
- `build_apa_outputs()`, `build_visual_summaries()`, `run_qc_pipeline()`,
  `build_mfrm_manifest()`, `build_mfrm_replay_script()`,
  `export_mfrm_bundle()`, package-native scorefile export, and
  `build_linking_review()` return caveated `GPCM` reporting or
  exploratory-review objects with explicit `gpcm_boundary` rows. The
  package-native scorefile can include native structural delta-method
  expected-score SEs and score-side delta SEs selected by `score_se_method`
  when the required MML diagnostics are available, but those SEs are not
  FACETS-equivalent score-side uncertainty.
- `evaluate_mfrm_design()`, `predict_mfrm_population()`,
  `evaluate_mfrm_diagnostic_screening()`, and
  `evaluate_mfrm_signal_detection()` are available as caveated role-based
  repeated simulation/refit routes. Treat their outputs as design-level or
  screening sensitivity evidence, not as operational scoring, calibrated
  inferential testing, or arbitrary-facet planning validation.

The dashboard marks the fair-average panel unavailable under `GPCM`; use
`fair_average_table()` directly for the slope-aware element-conditional table
and `fair_average_table(fit_gpcm, fair_se = TRUE)` to inspect structural
fair-average SEs for non-person rows. These condition on Person EAP/reference
means and remain diagnostic-only: `FairCIEligible = FALSE`, including when
the covariance is computable or regularized. Full-refit coverage remains
unverified. Historical measure-level SE columns are not fair-score SEs.

## What is intentionally restricted

The slope-aware `fair_average_table()` route and package-native scorefile route
are available under `GPCM`, including native expected-score uncertainty and
score-side delta SEs where the required MML diagnostics support them. Full
FACETS-style score-side compatibility remains restricted because free
discrimination changes the relationship between the latent measure and
operational score-side summaries. Specifically:

- `facets_output_contract_review()` still depends on FACETS-style
  compatibility semantics that are not generalized to free discrimination.
- posterior-predictive checks and MCMC estimation are not available in mfrmr;
  use external Bayesian software when those analyses are required.
- Caveated reporting, export, linking, design-forecast, and screening helpers
  must keep their `gpcm_boundary` wording visible and must not imply
  FACETS-equivalent score-side uncertainty, operational scoring, calibrated
  screening criteria, or arbitrary-facet planning validation.

## Recommended substitutes

When a restricted helper is needed for a `GPCM` report, the practical paths
are:

- Refit with `model = "PCM"` if the discrimination-free assumption is
  defensible for the data and a full FACETS score-side review is required;
  `compare_mfrm()` can describe available likelihood differences for matched
  data and model specifications. A shared quadrature grid and a denser-grid
  sensitivity check address numerical comparability. The PCM/GPCM test also
  requires matching population and constraint settings. Relative-slope intervals
  and IC comparison have their own solution and likelihood checks.
- Keep the report on the `GPCM` fit itself but draft the manuscript section
  manually around the supported tables: `summary(fit)` for parameters,
  `diagnose_mfrm()` for residual fit, `facet_quality_dashboard()` for the
  per-facet quality summary, and `compute_information()` for precision
  evidence.
- When a Rasch-family baseline comparison is substantively useful, optionally
  generate a second manifest from a parallel `RSM` or `PCM` fit. The two fits
  can be reported side by side, with the `GPCM` fit identified as the
  discrimination-aware counterpart. This is not required to use the caveated
  `GPCM` manifest/replay/export route.

Restricted helpers use the capability matrix at runtime. An unsupported
`GPCM` call stops with the relevant limitation and a supported
alternative instead of producing a partial score-side or backend result. Use
`gpcm_capability_matrix()` or `mfrmr_output_guide("gpcm")` before choosing a
downstream route.

## A worked example

The `example_core` dataset includes a small synthetic block that supports a
`GPCM` fit. This example uses compact quadrature and iteration settings
for a shorter runtime; for substantive analysis, rerun with the package default
or a higher quadrature setting and a larger recovery design.

```r
library(mfrmr)
toy <- load_mfrmr_data("example_core")

fit_gpcm <- fit_mfrm(
  data = toy,
  person = "Person",
  facets = c("Rater", "Criterion"),
  step_facet = "Criterion",
  slope_facet = "Criterion",
  score = "Score",
  model = "GPCM",
  method = "MML",
  quad_points = 7,
  maxit = 20
)

fit_gpcm
gpcm_summary <- summary(fit_gpcm, profile = "fit", detail = "brief")
gpcm_summary$decision
gpcm_summary$inference_evidence

diag_gpcm <- diagnose_mfrm(fit_gpcm)
summary(diag_gpcm)

info <- compute_information(fit_gpcm)
plot_information(info)

rec_gpcm <- evaluate_mfrm_recovery(
  sim_spec = build_mfrm_sim_spec(
    n_person = 30,
    n_rater = 3,
    n_criterion = 4,
    raters_per_person = 2,
    model = "GPCM",
    step_facet = "Criterion",
    slope_facet = "Criterion",
    slopes = c(0.8, 1.0, 1.15, 1.05)
  ),
  reps = 10,
  model = "GPCM",
  fit_method = "MML",
  quad_points = 7,
  maxit = 20,
  include_diagnostics = TRUE,
  diagnostic_fit_df_method = "both",
  seed = 1
)
review_gpcm <- assess_mfrm_recovery(
  rec_gpcm,
  max_rmse = c(facet = 0.5, step = 0.5, slope = 0.25),
  max_abs_bias = c(default = 0.25)
)

summary(review_gpcm)$overview
summary(review_gpcm)$reading_order
review_gpcm$condition_reporting_notes[, c(
  "ConditionArea", "ReportingAttention", "ConditionFinding"
)]
review_gpcm$condition_review[, c(
  "Model", "GPCMSlopeRegime", "StressLevel", "ScoreSupportStatus"
)]
review_gpcm$diagnostic_reporting_notes[, c(
  "Facet", "ReportingAttention", "DiagnosticFinding"
)]
summary(review_gpcm)$diagnostic_review
plot(review_gpcm, type = "status")
plot(review_gpcm, type = "metrics", metric = "rmse")

```

Use the same `Decision` fields as for RSM and PCM. In particular, optimizer
convergence does not override incomplete GPCM identifiability, curvature, or
slope-boundary evidence. If `FormalInference` is `"No"`, the diagnostic and
information calls above are investigative routes for resolving the recorded
reason; they do not promote the fit to an inferential result.

The fit, summary, residual diagnostics, information, recovery, fair-average, and
conditional bias-screening helpers all run under `GPCM` with the caveats listed
above. `build_apa_outputs(fit_gpcm)` returns a caveated sensitivity-reporting
object with a `gpcm_boundary`; full FACETS score-side review remains on the
`RSM` / `PCM` route.

## Estimator references

- Wang, W.-C., & Liu, C.-Y. (2007). *Formulation and Application of the
  Generalized Multilevel Facets Model*. Educational and Psychological
  Measurement, 67(4), 583--605.
  <https://doi.org/10.1177/0013164406296974>
- Wang, Y.-G., Wu, J., & Qiu, X. (2025). *A Differential Index Measuring
  Rater's Capability in Educational Assessment*. Preprint, arXiv:2502.09099v1.
  <https://arxiv.org/abs/2502.09099v1>
- Muraki, E. (1992). *A generalized partial credit model: Application of an EM
  algorithm*. Applied Psychological Measurement, 16, 159--176.
- Uto, M., & Ueno, M. (2020). *A generalized many-facet Rasch model and its
  Bayesian estimation using Hamiltonian Monte Carlo*. Behaviormetrika, 47,
  469--496.
- Wijayanto, F., Mul, K., Groot, P., van Engelen, B. G. M., & Heskes, T.
  (2021). *Semi-automated Rasch analysis using in-plus-out-of-questionnaire log
  likelihood*. British Journal of Mathematical and Statistical Psychology, 74,
  313--339.
- Hessen, D. J. (2025). *A convexity-constrained parameterization of the random
  effects generalized partial credit model*. British Journal of Mathematical
  and Statistical Psychology, 78, 401--419.

## Current boundary

Score-side semantics for free-discrimination polytomous models differ from the
Rasch-family route. Use the current matrix returned by
`gpcm_capability_matrix()` as the workflow contract and
`gpcm_score_side_contract()` for score-side alternatives.
