---
title: "mfrmr Reporting and APA"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{mfrmr Reporting and APA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
is_cran_check <- !isTRUE(as.logical(Sys.getenv("NOT_CRAN", "false")))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  eval = !is_cran_check
)
```

This vignette shows the package-native route from a fitted many-facet Rasch
model to manuscript-oriented prose, tables, figure notes, and revision checks.

The reporting stack in `mfrmr` is organized around four objects:

- `fit`: the fitted model from `fit_mfrm()`
- `diag`: diagnostics from `diagnose_mfrm()`
- `chk`: the revision guide from `reporting_checklist()`
- `apa`: structured manuscript outputs from `build_apa_outputs()`

Before those fit-specific objects exist, use
`mfrmr_minimum_report_checklist()` to outline the method, design, estimation,
fit, precision, category, rater/facet, fairness, linking, visual,
FACETS/external-software, validity-boundary, and limitation topics that a
critical psychometric reader may expect.

For a broader workflow view, see `vignette("mfrmr-workflow", package = "mfrmr")`.
For a plot-first route, see
`vignette("mfrmr-visual-diagnostics", package = "mfrmr")`.

## Minimal setup

```{r setup}
library(mfrmr)

toy <- load_mfrmr_data("example_core")

# The vignette uses compact quadrature so optional local execution stays fast.
# For final manuscript reporting, refit with the package default or a higher
# quadrature setting and record that setting in the analysis log.
fit <- fit_mfrm(
  toy,
  person = "Person",
  facets = c("Rater", "Criterion"),
  score = "Score",
  method = "MML",
  model = "RSM",
  quad_points = 7
)

diag <- diagnose_mfrm(fit, residual_pca = "none")
```

## 1. Start with the minimum report checklist

Use `mfrmr_minimum_report_checklist()` first when the question is "what must
the report cover?" This checklist is fit-independent and should be read as a
planning or reviewer-facing map.

```{r minimum-checklist}
minimum <- mfrmr_minimum_report_checklist()
minimum[, c("Section", "ReportItem", "Required")]
```

## 2. Move to the fit-specific revision guide

Use `reporting_checklist()` first when the question is "what is still missing?"
rather than "how do I phrase the results?"

```{r checklist}
chk <- reporting_checklist(fit, diagnostics = diag)

head(
  chk$checklist[, c("Section", "Item", "DraftReady", "Priority", "NextAction")],
  10
)
```

Interpretation:

- `DraftReady` flags whether the current objects already support a section for
  drafting with the package's documented caveats.
- `Priority` shows what to resolve first.
- `NextAction` is the shortest package-native instruction for closing the gap.

## 3. Check the precision layer before strong claims

`mfrmr` intentionally distinguishes `model_based`, `hybrid`, and
`exploratory` precision tiers.

```{r precision}
prec <- precision_review_report(fit, diagnostics = diag)

prec$profile
prec$checks
prec$fit_separation_basis
```

Interpretation:

- Use stronger inferential phrasing only when the reported tier is
  `model_based`.
- Treat `hybrid` and `exploratory` outputs more conservatively, especially for
  SE-, CI-, and reliability-heavy prose.
- Read `fit_separation_basis` as a boundary table: MnSq fit, ZSTD
  standardization, Rasch/FACETS-style separation, and package QC thresholds
  have different source bases and should not be collapsed into a single
  validation pass/fail claim.

## 4. Build structured manuscript outputs

`build_apa_outputs()` is the writing engine. It returns report text plus a
section map, note map, and caption map that all share the same output
contract.

```{r apa}
apa <- build_apa_outputs(
  fit,
  diagnostics = diag,
  context = list(
    assessment = "Writing assessment",
    setting = "Local scoring study",
    scale_desc = "0-4 rubric scale",
    rater_facet = "Rater"
  )
)

cat(apa$report_text)
```

```{r section-map}
apa$section_map[, c("SectionId", "Heading", "Available")]
```

Interpretation:

- `report_text` is the compact narrative output.
- `section_map` is the machine-readable map of what text blocks are available.
- The same contract also feeds captions and notes, which reduces wording drift.

## 5. Make the MML population metric explicit

For ordinary MML fits without latent regression, the default person metric fixes
the normal marginal population SD at the configured `population_prior_sd`.
When `estimate_population_sd = TRUE`, that SD is estimated by the EM variance
M-step. This is a metric-calibration choice, not a fit-status decision or a
bounded-`GPCM` slope effect, so it should be visible in Method wording, result
notes, and any first-screen report.

```{r population-sd-apa}
fit_free_sd <- fit_mfrm(
  toy,
  person = "Person",
  facets = c("Rater", "Criterion"),
  score = "Score",
  method = "MML",
  model = "RSM",
  quad_points = 7,
  maxit = 40,
  estimate_population_sd = TRUE
)

diag_free_sd <- diagnose_mfrm(
  fit_free_sd,
  residual_pca = "none",
  diagnostic_mode = "legacy"
)

summary(fit_free_sd)$population_overview[
  , c("PopulationSDMode", "EstimatedPopulationSD",
      "PopulationSDSE", "PopulationSDSEStatus")
]

apa_free_sd <- build_apa_outputs(fit_free_sd, diagnostics = diag_free_sd)
summary(apa_free_sd)$content_checks[
  summary(apa_free_sd)$content_checks$Check == "MML population SD wording alignment",
  c("Check", "Passed", "Detail")
]
```

The same information is also carried by the result/report wrapper.

```{r population-sd-results}
res_free_sd <- mfrm_results(
  fit_free_sd,
  include = c("fit", "diagnostics", "reporting", "apa")
)

summary(res_free_sd)$overview[
  , c("PopulationSDMode", "EstimatedPopulationSD", "PopulationMetric")
]

report_free_sd <- mfrm_report(res_free_sd, style = "apa")
report_free_sd$sections[
  report_free_sd$sections$Section == "Model and data setup",
  c("Section", "Evidence", "Boundary")
]
```

For APA-style reporting, use the generated wording as a draft. Estimated-SD
runs should report the estimated SD and profile SE/CI when available, and keep
the profile-SE caveat: the profile uncertainty currently holds other structural
parameters fixed.

## 6. Build tables from the same contract

Use `apa_table()` when you want reproducible handoff tables without rebuilding
captions or notes by hand.

```{r apa-tables}
tbl_summary <- apa_table(fit, which = "summary")
tbl_reliability <- apa_table(fit, which = "reliability", diagnostics = diag)

tbl_summary$caption
tbl_reliability$note
```

The actual table data are stored in `tbl_summary$table` and
`tbl_reliability$table`.

## 7. Add figure-ready visual data

For reporting workflows, `build_visual_summaries()` is the bridge between
statistical results and figure-ready plot data.

```{r visuals}
vis <- build_visual_summaries(
  fit,
  diagnostics = diag,
  threshold_profile = "standard"
)

names(vis)
names(vis$warning_map)
```

## 8. Reporting route when interaction screening matters

When bias or local interaction screens matter, keep the wording conservative.
The package treats these outputs as screening-oriented unless the current
precision and design evidence justify stronger claims.

```{r bias-screen}
bias_df <- load_mfrmr_data("example_bias")

fit_bias <- fit_mfrm(
  bias_df,
  person = "Person",
  facets = c("Rater", "Criterion"),
  score = "Score",
  method = "MML",
  model = "RSM",
  quad_points = 7
)

diag_bias <- diagnose_mfrm(fit_bias, residual_pca = "none")
bias <- estimate_bias(fit_bias, diag_bias, facet_a = "Rater", facet_b = "Criterion")
apa_bias <- build_apa_outputs(fit_bias, diagnostics = diag_bias, bias_results = bias)

apa_bias$section_map[, c("SectionId", "Available", "Heading")]
```

## Recommended sequence

For a compact manuscript-oriented route:

1. `fit_mfrm()`
2. `diagnose_mfrm()`
3. `precision_review_report()`
4. `reporting_checklist()`
5. `build_apa_outputs()`
6. Check `summary(fit)$population_overview` and `summary(apa)$content_checks`
   when MML population-SD mode affects the report.
7. `apa_table()`
8. `build_visual_summaries()`

## Related help

- `help("mfrmr_reporting_and_apa", package = "mfrmr")`
- `help("mfrmr_reports_and_tables", package = "mfrmr")`
- `help("reporting_checklist", package = "mfrmr")`
- `help("build_apa_outputs", package = "mfrmr")`
