---
title: "GPCM scope and current limitations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{GPCM scope and current limitations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
is_cran_check <- !isTRUE(as.logical(Sys.getenv("NOT_CRAN", "false")))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  eval = !is_cran_check
)
```

`mfrmr` includes a Generalized Partial Credit Model (GPCM; Muraki 1992) in which
one selected facet supplies level-specific discriminations. MML permits a
different facet to supply category steps; JML requires a shared owner. MML information
criteria can be compared when the likelihood and local-solution checks pass.
`confint(fit, parm = "slopes")` provides approximate relative-slope intervals
when its MML checks pass. Explicit options add standardized slopes, specified
comparisons, independent-cluster covariance and finite-family adjustment.
Separate functions provide fitted-model bootstrap inference and curve intervals. Several
downstream reporting helpers remain restricted because score-side semantics
under free discrimination differ from the Rasch-family case. This vignette
documents which helpers are available, which are not, and what to use as a
substitute when a helper is restricted.

We call this model **GPCM** and state its structural choices and output
availability separately. Unidimensionality is not a reason to give GPCM a
special restrictive name. The JML route uses an
unpenalized identified joint likelihood. A non-representable slope proposal can
be rejected during line search for numerical safety, but that is not a
statistical penalty. If the likelihood has a certified recession direction,
the finite optimizer iterate is retained only as a numerical trace and the
primary result carries an extended-real or typed boundary status.

## What can I report?

The following distinctions apply to the current free-slope model under MML.
They describe mfrmr's available output, not a general impossibility of GPCM
inference.

| Output | Current use | What the number does not establish |
|---|---|---|
| Relative slopes and fitted curves | Inspect `OptimizerEstimate` and numerical diagnostics for descriptive sensitivity analysis. `Estimate` retains the same numerical slope for compatibility. | A finite optimizer result is not a qualified primary slope estimate. |
| Slope SEs and intervals | `confint(fit, parm = "slopes")` or `diagnose_mfrm()` returns pointwise relative log-Wald intervals by default. Explicit `confint()` options add standardized slopes, contrasts, sandwich covariance and Bonferroni adjustment. | Inspect `CIEligible` and `InferenceReview`. Each option changes the target or approximation; none supplies universal coverage. |
| Facet and step SEs | Some local SEs remain available for review; read fit readiness and the accompanying precision/interval eligibility. | A finite band does not override the source fit's inference restriction. |
| Person posterior summaries | Conditional scoring may be available; inspect source readiness and calibration stability. | Posterior SDs do not include uncertainty from estimating the calibration. |
| AIC/BIC/SABIC | `compare_mfrm()` returns ranking, deltas and weights when its MML comparison checks pass. | `ICSelectable` covers likelihood/integration, `ICFitEligible` covers the numerical solution, and `ICComparable` is the final comparison decision. This does not grant interval or LRT eligibility. |
| PCM/GPCM likelihood-ratio test | `compare_mfrm(..., nested = TRUE)` returns an asymptotic chi-square test for eligible matched MML fits. | A significant difference does not decide the scoring policy; small or sparse samples can invalidate the approximation. |

### Which uncertainty claims are supported?

All the inferential routes below are approximations. A numerical eligibility
check establishes whether a calculation can be returned under its stated
rules; it does not certify the nominal confidence level for your assessment.
The available studies support different claims for different targets:

| Question and API | Evidence and reporting limit |
| --- | --- |
| Relative or population-standardized discrimination: `confint(fit)` | Joint-information and independent-cluster sandwich calculations have numerical checks and simulation evidence for specified designs. Historical studies do not establish current-estimator coverage for every changed numerical branch. |
| A specified ratio or difference: `confint(fit, contrasts = ...)` | Use the declared scale and full covariance. Evidence for relative contrasts does not transfer automatically to standardized differences; the latter have a separate saved-fit evaluation below. |
| Category probability at fixed ability: `mfrm_curve_intervals()` | Numerical transformations are checked, but nominal coverage failed in some small incomplete designs, including Bonferroni families. Report these as approximate calibration-uncertainty intervals, not validated 95% bands. |
| Information per rating: `mfrm_curve_intervals(..., type = "information")` | Its own saved-fit evaluation is available below. This is neither total test information nor uncertainty in an estimated Person score. |
| Matched-model selection or equal slopes: `compare_mfrm()` | AIC/BIC ranking and the asymptotic PCM/GPCM test require their separate likelihood, nesting and numerical checks. The historical null study does not establish performance for arbitrary sparse designs. |
| Bootstrap slopes or a PCM-null test: `bootstrap_mfrm_gpcm()` | Quantile calculations and accounting for unresolved trials are checked. The one-dataset example does not establish repeated-dataset coverage or Type I error, or fix the probability-interval shortfall. |

For all routes, keep unavailable or unbounded outcomes and saved cautions in
your report. Bonferroni concerns the finite family requested in one call;
sandwich covariance requires the declared independent clusters. Neither option
removes model bias or guarantees small-sample calibration. The APIs remain
available with these limits; no data-dependent widening or arbitrary sample-size
cutoff is used to conceal the adverse findings.

### How uncertain is a relative slope?

Suppose three rubric criteria have relative slopes 0.8, 1.0 and 1.25.
Their geometric mean is one. The last criterion has the steepest response
curve on this model's common ability scale. An interval describes sampling
uncertainty in that relative discrimination; it does not say that the criterion
should receive more scoring weight or that its raters agree more closely.

```r
# After fitting a native GPCM with method = "MML":
# slope_ci <- confint(fit, parm = "slopes", level = 0.95)
# slope_ci
# attr(slope_ci, "diagnostics")[, c("SlopeFacet", "Estimate",
#   "CIEligible", "InferenceReview")]
```

The calculation inverts the full joint observed information, including
estimated population parameters. It then maps its free log-slope covariance
to all slope levels using the sum-zero constraint. The interval for slope
$a_g$ is $\exp[\log(a_g) \pm z SE(\log(a_g))]$. Working on the log scale keeps
bounds positive and accounts for the dependent slope coordinates. Inverting
only the slope block of the information matrix would ignore estimation of
the other parameters and generally understate uncertainty.

The current likelihood and gradient must match a numerically adequate fit,
and the information must be positive without regularization. Unit weights,
supported score categories and at least 31 quadrature points are required.
Missing bounds retain a reason; increasing iterations cannot repair every
failure. Examine different starts and a denser grid for consequential results.
`diagnose_mfrm(fit)$parameter_uncertainty$slopes` supplies the same 95% intervals;
`attach_diagnostics = TRUE` also makes them available in the fitted slope table
and subsequent weighting reviews. Ordinary Wright/Pathway maps display ability,
facet location or fit, not these discriminations. Existing information plots
remain plug-in curves; `mfrm_curve_intervals()` adds calibration intervals.

These defaults are **pointwise model-based approximations**. The optional
methods below explicitly change the target, covariance or multiplicity rule.
An interval for `PopulationSD * Estimate` needs the variance and covariance
of the estimated population SD. A slope interval
containing one is not a test that all slopes are equal; use the matched
PCM/GPCM test for that question.

The following targeted check used saved fits from an earlier implementation.
Later numerical repairs can change fitted solutions or interval availability;
these percentages are not an evaluation of refitting every dataset with the
current estimator. The check used 100 independent datasets per condition, with three
criteria, three raters and three score categories. It crossed the slope owner
(Criterion or Rater), equal versus unequal slopes (approximately 0.67, 1 and
1.49), and two designs. With 100 Persons rated by all three raters, intervals
were available in every dataset; pointwise coverage across levels was 92--98%.
With 40 Persons each rated by two of the three raters, equal-slope availability
was 100% and coverage was 95--97%. Unequal-slope availability fell to 96--97%,
with coverage among available intervals from 90.6% to 97.9%; the lowest
covered-and-returned fraction over all planned datasets was 87%. Some intervals
were extremely wide. Four datasets were withheld for regularized information
and three for category-support review. These results support an approximate
procedure with explicit limits, not a general 95% coverage claim. Each coverage
estimate has only 96--100 independent repetitions; the three intervals within
a dataset are dependent. Sample size and incomplete assignment changed
together, so their effects cannot be isolated from this check.

### Choosing the interval for the question

For rubric criteria, a comparison of two slopes asks whether their response
curves differ in steepness, conditional on this model. It is not a comparison
of rater agreement or a reason to change scoring weights. Prespecify the
comparison before looking at which estimates differ most.

The following fictional rubric data give a reproducible example. The simulated
model is correctly specified; this illustrates the API, not field performance.

```{r gpcm-interval-example}
library(mfrmr)
ratings <- simulate_mfrm_data(
  n_person = 80, n_rater = 3, n_criterion = 3, score_levels = 3,
  model = "GPCM", step_facet = "Criterion", slope_facet = "Criterion",
  slopes = c(C01 = 0.8, C02 = 1, C03 = 1.25), seed = 924090
)
fit <- fit_mfrm(
  ratings, "Person", c("Rater", "Criterion"), "Score",
  model = "GPCM", method = "MML", step_facet = "Criterion",
  slope_facet = "Criterion", quad_points = 31, maxit = 400, reltol = 1e-10
)
```

```{r gpcm-slope-comparisons}
confint(fit, scale = "standardized")
slope_levels <- fit$slopes$SlopeFacet
comparison <- matrix(0, 1, length(slope_levels),
                     dimnames = list("first versus second", slope_levels))
comparison[1, 1:2] <- c(1, -1)
ratio <- confint(fit, contrasts = comparison)
difference <- confint(fit, contrasts = comparison,
                      contrast_scale = "difference")
attr(ratio, "diagnostics")[, c("Estimate", "CI_Lower", "CI_Upper", "PValue")]
```

The ratio null is one; the difference null is zero. More rows specify a finite
family. `simultaneous = "bonferroni"` adjusts both intervals and the corresponding
Wald p-values for all requested rows, not for other analyses or data-selected
comparisons. Standardized slopes multiply the relative slope by estimated
population SD, propagating its uncertainty and cross-covariances. With latent
regression this is the **residual** population SD, not the marginal spread
across different covariate groups. Ratios cancel a common scale; differences
do not. JML inference is outside these MML methods.

```{r gpcm-sandwich-example}
confint(fit, method = "sandwich", simultaneous = "bonferroni")
# If Persons belong to independent schools, supply a complete Person/Cluster
# mapping and use clusters = schools. Do not cluster separate rows of one Person.
```

Sandwich covariance uses the marginal-likelihood score vector for each Person,
or sums those vectors within a supplied cluster. It needs many independent
clusters with enough score-vector rank. Its target is the fitted working-model
parameter. It does not repair bias from a wrong population model, informative
assignment or dependence between clusters. The optional `adjust = TRUE` only
multiplies the covariance by G/(G-1); it is not a small-sample correction with
guaranteed coverage. The earlier 800-dataset relative-interval study does not
qualify these additional procedures by itself. An earlier reanalysis recalculated
intervals on all 800 saved fits, without new simulations or optimizer runs, and found
90.7--99% pointwise coverage for standardized model intervals and 90.7--100%
for standardized sandwich intervals. The minimum was 88/97 available intervals
in a small incomplete, unequal-rater-slope condition (Monte Carlo 95% interval
83.1--95.7%; 88/100 covered-and-returned). For three standardized slopes,
Bonferroni **family** coverage ranged from 91.7--98% with model covariance and
90--97% with sandwich covariance. Each cell has only 100 independent datasets.
These correctly specified normal-population results show that neither
sandwich covariance nor multiplicity adjustment guarantees nominal coverage
in small samples; they do not test robustness under misspecification.

### A fitted-model bootstrap alternative

The parametric bootstrap asks what variation occurs if independent Persons
and scores follow the fitted model on the same observed rating assignment.
It preserves covariates and facet labels, and reestimates the free population
parameters. Unassigned ratings stay unassigned. It does not test whether the
assignment mechanism or the normal population assumption is correct.

```r
boot <- bootstrap_mfrm_gpcm(fit, nsim = 499, seed = 92401)
confint(boot, scale = "standardized")
confint(boot, contrasts = comparison, simultaneous = "bonferroni")
boot$trials  # Includes every planned trial, its last stage and any reason/warning.
apa_table(boot, which = "checks")  # Returned fits: categories, convergence and information.
# boot$refit_draws retains returned parameters for diagnosis, not interval calculation.
# For an eligible matched PCM fit with the same population model:
lr_boot <- bootstrap_mfrm_gpcm(fit, null_fit = pcm_fit,
                              nsim = 499, seed = 92402)
lr_boot$test
```

The interval uses basic bootstrap error quantiles, on a log scale for positive
slopes/ratios and on the original scale for differences. Failed refits and
returned estimates rejected by eligibility checks are retained as unknown outcomes. The reported limits enclose all completions,
possibly reaching zero or infinity; deleting failures could instead conceal
the least stable cases. The LRT reports Monte Carlo precision and p-value
bounds when any refit is unresolved. Its null-model draws cannot supply slope
intervals around the alternative fit. Saved results support new confidence
levels without refitting. More repetitions improve tail resolution, not
model validity; inspect `attr(confint(boot), "expected_tail_draws")`, especially
with many simultaneous comparisons. This is an approximate bootstrap method,
not a profile-likelihood implementation or an exact small-sample test.

For basic slope bootstrapping, a category observed once is now a warning
rather than an automatic exclusion. The source fit and each refit must still
pass model-identity and numerical checks, including unregularized joint
information. The exception requires a fresh audit confirming that every
category is observed in each step scope and no step coordinate or category
contrast is unsupported. It does not change the fit's saved readiness or
permission for Wald intervals, IC comparisons or LRTs. `checks$BootstrapEligible`
and `checks$WaldEligible` distinguish these decisions. Source decisions are in
`source_checks`. Accepted singleton warnings remain in the interval diagnostics,
printed output, default plot subtitle and reports, even when limits are finite.
A supplied subtitle, including `NULL`, overrides the default plot text.

The basic interval formula itself does not require a variance estimate in
every replicate; variance estimates are required for studentized intervals,
as explained in the [R boot documentation](https://stat.ethz.ch/R-manual/R-devel/library/boot/html/boot.ci.html).
Separating those formulas does not establish that an unstable estimate is
suitable for basic-bootstrap inference. Use `trials$Stage`,
`trials$AlternativeReturned` and `checks` to distinguish the reasons. Missing
check fields mean that the quantity was not recorded. Older saved results may
have only a combined reason and no rejected parameter vectors.

A small-sample example illustrates why unresolved replicates matter. Its
original 499-replicate run accepted 408 estimates. Rechecking 62
category-support cases under a revised acceptance rule produced an assembled
result with 470 accepted and 29 unresolved estimates. **That assembled result
is a comparison of procedures, not a complete bootstrap run under the current
estimator.** All three standardized 95% intervals still extended from zero to
infinity. This is one dataset, not evidence of repeated-dataset coverage or a
failure rate for other designs.

Preserve an original bootstrap when changing estimation or acceptance rules.
A complete new run is needed to evaluate the changed procedure; replacing
selected failed draws cannot establish its performance. Saved reanalysis
history, when present, follows printing, slope intervals and reporting tables.
Missing history does not establish that an older result used the current
procedure. Neither more replicates nor selective replacement automatically
resolves poor identification or validates interval coverage.

### Why an information warning needs a parameter-specific review

Inspect `boot$checks` (or `apa_table(boot, which = "checks")`) for
`PopulationSD`, `MinimumStandardizedSlope` and `MaximumStandardizedSlope`.
These describe the returned optimizer estimates, including rejected refits.
For an estimated population, standardized slopes multiply relative slopes
by the residual population SD; for a fixed standard-normal population that
SD is one. Missing entries indicate missing retained estimates. These fields
are diagnostics, not thresholds for judging raters or permission to use a CI.
APA check tables use significant digits for these values and the gradient/
information diagnostics, preserving small positive values in table conversion.
The original `boot$checks` columns remain unrounded numeric values.

The 29 unresolved trials illustrate different reasons for withholding
inference. Independent likelihood and derivative calculations helped separate
finite numerical stopping points from evidence for boundary approaches.
These checks diagnose the selected examples; they do not certify global
maxima or repeated-sample interval accuracy.

| Main finding in this one-dataset audit | Cases | Consequence |
| --- | ---: | --- |
| An unused top category in an unanchored step scope gives a monotone likelihood direction as its intercept cost increases | 15 | A threshold has no finite maximizer. In one checked example the slope estimates remained stable across starts; that does not validate all slope intervals. |
| Restarts and fixed-slope profiles support approach to zero standardized discrimination; the direction persists at 61 and 101 quadrature nodes | 11 | Evidence for a boundary limit, not just a large standard error. These numerical checks do not certify a global maximum. |
| Reparameterized optimization finds a better solution with a positive standardized slope and positive local curvature | 2 | The original numerical stopping point needs improvement. A warning attached to that original point is insufficient. The current optimizer and information refinement recover eligible finite solutions in these two selected examples; this does not replace a complete bootstrap evaluation. |
| A very steep standardized slope is highly sensitive to quadrature | 1 | At the same saved parameters, changing 61 to 101 nodes changes negative log-likelihood from about 185 to 858. Neither grid is thereby validated. |

These are primary findings; numerical and boundary problems can overlap.
An unused category in a replicate is a sampling outcome, not evidence that
the scoring category should be removed. A near-zero standardized slope can
coexist with other finite standardized slopes. Keeping the geometric mean
of *relative* slopes at one can then make relative slopes and location
coordinates extreme. Conversely, a stable discrimination estimate does not
establish a finite threshold. The assembled example retains all 29 unresolved
trials. The two improved finite solutions are separate numerical evidence;
they have not been inserted into its draws or counted as a current-procedure
bootstrap result.

For small fixed-grid GPCM MML fits (up to 64 free parameters, with
`reltol <= 1e-9`), the direct optimizer now reviews numerical curvature even
after a small raw gradient. Negative curvature triggers up to three BFGS
restarts in rescaled search coordinates. Each uses the requested iteration
ceiling. Only a non-worsening, numerically converged candidate without detected
negative curvature can replace the selected point. Otherwise the estimate is
retained with a numerical warning. Inspect `fit$opt$optimizer_polish$Stages`:
`StageLabel == "curvature_rescale"` identifies these attempts, and
`SmallestCurvature`, `CurvatureScale` and `CurvatureReviewError` explain the
curvature review. Missing curvature entries mean it was not evaluated.
The search rescaling does not provide a regularized covariance estimate.
An improved optimizer result still needs separate information and integration
checks; positive numerical curvature alone is insufficient.

Poor conditioning does not by itself mean that an information matrix must be
regularized or discarded. It can reflect parameter units, weak information,
or numerical differentiation error. The shared MML calculation now compares
two refined Hessians, checks their unregularized inverse and reviews the
gradient in the curvature metric. If these checks pass, intervals can be
returned with a saved caution. Otherwise they remain unavailable. The
information workspace budget covers the additional calculation too.

In the remaining finite-solution example, the standardized log-slope SE
changed from about 30 with the default numerical differentiation to about 27
after refinement, agreeing with the reference parameterization. Nevertheless,
the weakest discrimination's 95% log-Wald interval spans roughly
\(10^{-25}\) to \(10^{21}\). Returning that interval makes the lack of precision
visible; it does not make the approximation trustworthy near zero. Examine
`attr(intervals, "cautions")` and `InferenceReview`, which also appear in
printing, default plot subtitles and reports. This numerical review does not
establish sampling coverage or turn a boundary limit into a finite estimate.

Further requirements include integration and parameter-specific boundary
estimates and inference. The examples
support different treatment of different targets; they do not validate ordinary
Wald intervals or demonstrate repeated-dataset coverage.

### Uncertainty in probability and information curves

For example, examine a single rater/criterion context across known ability
values. The existing probability and information plots still describe fitted
points. This API propagates the full calibration covariance into each curve
point, using logit intervals for probabilities and log intervals for information.

```{r gpcm-probability-intervals, fig.alt="Three fitted category probability curves for one rater and criterion, with pointwise calibration-uncertainty ribbons at fixed ability values. Colors and line types distinguish categories; these are not Person-score intervals."}
grid <- expand.grid(Theta = seq(-3, 3, length.out = 41),
                    Rater = "R01", Criterion = "C01")
curves <- mfrm_curve_intervals(fit, grid)
plot(curves, title = NULL, subtitle = NULL)
```

```{r gpcm-information-intervals, fig.alt="Fitted information per rating for the same rater and criterion across ability values, with a pointwise calibration-uncertainty ribbon. The curve is not total information across all ratings."}
information <- mfrm_curve_intervals(fit, grid, type = "information")
plot(information)
```

Crosses identify fitted points whose intervals are unavailable. The caption
counts these points; a missing ribbon does not mean negligible uncertainty.
Ribbons stop at unavailable grid points. Inspect `plot_data(curves, "table")`
for the saved `InferenceReview` reasons. Use `caption = NULL` to hide the
caption; the crosses and saved reasons remain. Long default warnings wrap
within the figure.

For slope or ratio intervals spanning many orders of magnitude, a logarithmic
axis can make the estimates easier to compare:

```r
# When estimates and finite interval bounds are strictly positive:
# as_ggplot(ci) + ggplot2::scale_x_log10()
```

Keep a linear axis for signed differences. A logarithmic display does not
reduce uncertainty or establish interval accuracy.

Replace these labels with levels in the fitted model and include every
non-Person facet. Information is **per rating**: slope squared times the
conditional category variance. It is not total information summed over the
observed exposures. `Theta` is fixed on the fitted native ability scale;
uncertainty in a Person score is not included. Lines and ribbons retain their
numerical table, use color plus line type, and accept custom palettes and
omitted titles. Sandwich covariance is optional. Bonferroni adjustment is applied
over the finite output grid, including all categories; connecting points does not
produce a simultaneous confidence band for every value between them.

### What the additional interval checks found

Do the slope-interval results also describe uncertainty in differences and
curves? To examine that question, a further reanalysis used **all 800 saved
fits**, with the current interval calculations and their availability checks.
It did not generate new datasets or refit the historical estimates. Numerical
repairs to estimation can change those estimates, so this is evidence about
the saved-fit analysis, not an independent evaluation of the current estimator.

The generating facet effects were saved for every dataset. They were centered
to the fitted reference before comparing curves at native ability values
`Theta = -2, 0, 2`. Each slope-owner level was evaluated with the other facet
at its second level. This gives nine rating contexts, 27 category probabilities
and nine information values. Standardized differences compare the first with
the second slope and the second with the third; they include population-SD
uncertainty. Model and independent-Person sandwich covariance used the same
fits. The confidence level was 95%, without a small-cluster adjustment.

The table gives ranges across the eight original conditions and the specified
targets. Pointwise coverage concerns one target at a time. Family coverage
requires **every** interval in the specified call to cover its truth; a family
contains two differences, 27 probabilities or nine information values.

| Target | Pointwise: model | Pointwise: sandwich | Bonferroni family: model | Bonferroni family: sandwich |
| --- | ---: | ---: | ---: | ---: |
| Standardized slope differences | 91.0–98.0% | 91.0–99.0% | 92.9–98.0% | 93.9–98.0% |
| Category probabilities | 88.0–100.0% | 85.6–99.0% | 84.7–96.0% | 83.7–97.0% |
| Information per rating | 90.7–100.0% | 89.8–99.0% | 95.0–100.0% | 92.0–98.0% |

These percentages condition on available intervals or complete available
families. Each condition has 100 independent datasets, not one independent
replication per curve point. Depending on the target and condition, 96–100
pointwise intervals were available. In the small incomplete unequal-rater-slope
condition, the sandwich probability family covered in 82 of 98 available
datasets: 83.7%, with a Monte Carlo 95% interval of 74.8–90.4%. It was covered
and returned in 82 of all 100 planned datasets. That Monte Carlo interval
describes this cell, not simultaneous uncertainty for the minimum across cells.

**Bonferroni does not fix an inaccurate marginal approximation.** The curve
results show a limitation even under the generating model's assumptions;
sandwich covariance is not an automatic solution. Read a returned interval
as an asymptotic approximation, especially in small incomplete designs.
Do not interpret numerical eligibility as demonstrated 95% coverage, turn
curve differences into rater-quality verdicts, or assume that the bootstrap
has resolved this limitation without its own repeated-dataset evidence.
The two source designs also change sample size and assignment together, so
the results cannot isolate an effect of sparsity. They do not establish
performance for other grids, misspecified models or the latest refitted solutions.

### Checking the updated estimator on the affected design

The entire small incomplete, unequal-rater-slope condition was then refitted:
all 100 datasets, including cases whose intervals missed the truth or were
unavailable. The updated numerical implementation used ordinary initialization
and the same model, 61-point quadrature, iteration ceiling and convergence
tolerance. All fitted parameters, objective values, probability estimates and
interval endpoints were identical to the earlier results. Model and sandwich
Bonferroni families still covered 83/98 and 82/98 available datasets,
respectively. Updating the optimizer therefore did not resolve this shortfall
in the selected condition.

These are the same 100 datasets already included in the 800-fit study, not
another 100 independent replications or an independent confirmation after
selecting a condition. Analytic probability derivatives and full-covariance
delta intervals were also checked on two prespecified refits. The largest
endpoint discrepancy from the API was below `2e-9`. This checks the probability
transformation; it does not establish global optimization, quadrature accuracy
or finite-sample coverage, nor identify approximation error as the only cause.

For a small incomplete assessment, use these bands as approximate descriptions
of calibration uncertainty. Do not rely on their nominal 95% label alone to
claim that an entire curve is covered or that two curves clearly differ.
Such a conclusion needs an uncertainty method justified for the study's design
and model. A larger sample, a simpler model or bootstrap may be worth assessing,
but none is an automatic correction established by this check. Printing and
default figure subtitles now identify the intervals as approximate;
`CIEligible` continues to describe numerical availability.

### Carry the selected inference into reports and figures

A rater's severity, discrimination and residual fit answer different questions.
Wright maps locate fitted abilities and facet effects; Pathway maps relate
location to fit statistics. A slope interval instead describes discrimination
under its stated scale and covariance method. Adding a slope interval must not
replace severity intervals, Infit/Outfit, or a feedback decision with a different
target. In particular, a larger slope does not automatically mean a better rater
or a larger recommended scoring weight.

```{r gpcm-saved-inference, fig.alt="Population-SD-standardized discrimination estimates with explicitly selected model-based pointwise 95 percent intervals. The plot concerns discrimination, not rater severity or quality."}
ci <- confint(fit, scale = "standardized")
results <- mfrm_results(fit, intervals = list(slopes = ci, curves = curves),
                        include = c("fit", "plots"), compute = "never")
plot(results, type = "gpcm_slopes", title = NULL)
apa_table(ci)
head(plot_data(results, type = "gpcm_curves", component = "table"))
```

The reproduction code in `summary(results)$reproducible_code` saves and reloads
the complete result, preserving attached inference. In that code, `res` denotes
your result object (`res <- results` here); choose the file path before saving.
The starter export index also links the saved GPCM inference figures alongside
the ordinary maps, with separate interpretation guidance.

The same stored values are available through `mfrm_report(results)` and
`export_mfrm_results(results, ...)`. Named results define lowercase plot routes.
Use `as_ggplot(results, type = "gpcm_curves")` for further ggplot customization;
`title = NULL` or `subtitle = NULL` removes the corresponding annotation.
Plots distinguish unavailable intervals by crosses and unbounded intervals by
open circles, retaining their full limits and reasons in the tables. Color and
line type distinguish probability curves. No reference cutoff is chosen for a
slope plot; `reference` is an explicit optional plotting choice.

Source matching checks fitted data, parameters, population and integration
settings before attachment. Reusing a saved bundle does not change the selected
method, recompute intervals or rerun a bootstrap. For bootstrap slopes, pass
`confint(boot, scale = "standardized", level = .90)` to preserve that choice;
passing the raw `boot` object selects its default relative 95% pointwise
intervals. The export replay reloads the saved RDS. The full archive contains
the fitted data and identifiers, so review it before sharing.

### Check that the integration grid does not change the decision

`mml_quadrature_sensitivity(fit, data, quad_points = c(41, 61))` preserves
explicit population formulas, covariates and factor coding. Its slope table
contains both interval endpoints and eligibility; the summary reports their
changes. For a PCM/GPCM comparison, apply the helper to each model and rerun
`compare_mfrm()` on their fits at each **common** grid. Compare the IC/LRT
decisions as well as estimates; increasing the GPCM grid alone changes the
comparison basis. A small observed change is evidence for that fit and grid,
not a universal integration certificate.

Information criteria and parameter intervals need different evidence. The
[R documentation for AIC](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/AIC.html)
requires maximized likelihoods on comparable data. Computing an AIC number
does not require a slope confidence interval. In mfrmr, the comparison uses
the declared parameter count, numerical solution and integration settings.
Users should assess competing starts and integration sensitivity, particularly
when a model-choice decision is close.
An interval procedure also needs a justified covariance or other uncertainty
method, the correct constrained-parameter transformation, and evidence about
its repeated-sampling behavior. Neither decision requires a theorem covering
every conceivable GPCM likelihood path.

An unavailable slope interval does not automatically invalidate a separately
eligible IC comparison or matched PCM/GPCM test. Older saved fits must not gain
intervals from stored flags alone; rerun `confint()` or `diagnose_mfrm()` to
check the retained solution and refresh its uncertainty tables.

### Why intervals, information criteria and tests need separate checks

**Slope confidence intervals.** The package uses the joint covariance in
free coordinates, maps it to sum-zero log slopes and exponentiates the log-Wald
limits. Its separate qualification checks preserve missing results for singular,
regularized or unstable information. The result is an asymptotic approximation,
not exact coverage for every possible dataset. The
[mirt documentation](https://philchalmers.github.io/mirt/reference/mirt.html)
also provides GPCM fitting and information-matrix SE methods; that establishes
a methodological precedent, not numerical validation of mfrmr.

**Information-criterion comparison.** The package already calculates
\(\mathrm{AIC}=-2\widehat\ell+2p\) and
\(\mathrm{BIC}=-2\widehat\ell+p\log N\), with \(N\) equal to the number of
Persons for the ordinary MML route and \(p\) the free-parameter count.
For GPCM, the comparison reevaluates the retained likelihood, gradient and
observed information without rerunning the optimizer. It checks agreement
with the stored objective, the existing gradient tolerance (at most
\(10^{-4}\)), and positive unregularized information using the same relative
eigenvalue tolerance as covariance inversion. IC/LRT and slope intervals share
the full joint information calculation. The old IC/LRT-specific 80-coordinate
cutoff is replaced by a dense-matrix workspace budget, configurable with
`options(mfrmr.max_information_bytes = 256 * 1024^2)`. This estimates eight
double-precision square matrices; it excludes data, integration arrays and
other process memory. An unavailable check returns its reason. This local check is separate from `InferenceReady` and
does not qualify a slope interval. Same-data/likelihood, free-parameter-count,
unit-weight and integration checks still apply. There is no requirement that
the candidates be nested. Local curvature does not prove global optimality,
and sorting candidate IC values is not a test that the winning model is true.

**PCM/GPCM likelihood-ratio test.** For the current relative-slope
parameterization, let \(\lambda_g=\log\alpha_g\), with
\(\sum_g\lambda_g=0\). The PCM null is
\(\lambda_1=\cdots=\lambda_G=0\). It imposes \(G-1\) independent restrictions
when the population model, steps, other facets and constraints are identical.
The null value \(\alpha_g=1\) is inside the positive slope space; it is not a
zero-variance boundary. Under the usual identified, interior and regular MML
conditions, the candidate reference is
\(2(\widehat\ell_{GPCM}-\widehat\ell_{PCM})\sim\chi^2_{G-1}\) asymptotically.
The implemented test verifies the matched null/alternative population model,
step and facet constraints, fixed interactions and free dimensions. It reuses
the likelihood, gradient and unregularized-information checks described above.
A negative likelihood gain is a numerical failure and does not become p = 1.
The reference assumes independent Persons and correctly specified response
and population models, with regular interior solutions. It is an asymptotic
approximation, especially uncertain in small or weakly linked designs.
Default calls can differ in population assumptions, so changing only
`model = "PCM"` to `model = "GPCM"` does not necessarily test slopes alone.
The [mirt model-comparison documentation](https://philchalmers.github.io/mirt/reference/anova-method.html)
distinguishes ordinary likelihood-ratio comparisons from boundary-null tests.
That distinction does not require a mixture chi-square test merely because
GPCM slopes are positive.

A simulation illustration examined the approximation under a true PCM,
using three raters, three criteria, three score categories and normally
distributed Person abilities. In the crossed design, each of 100 Persons was
rated by all three raters on all three criteria. In the rotating design,
each of 40 Persons was rated by two of the three raters on all three criteria.
The latter changes both sample size and assignment density, so it does not
isolate the effect of sparsity. Rater and criterion locations varied across
replications; the slope null was always true. Both fitted models estimated
the population mean and variance on the same 61-point integration grid.

| Slope facet | Rating design | Rejections at 5%, out of 100 | 95% Monte Carlo interval |
|---|---|---:|---:|
| Criterion | Crossed, 100 Persons | 5 | 1.6%--11.3% |
| Rater | Crossed, 100 Persons | 7 | 2.9%--13.9% |
| Criterion | Rotating, 40 Persons | 6 | 2.2%--12.6% |
| Rater | Rotating, 40 Persons | 7 | 2.9%--13.9% |

All 400 pairs supplied a test in these conditions. The intervals above are
exact binomial intervals describing uncertainty in the simulated rejection
rates, not parameter confidence intervals. The point estimates do not show a
large departure from 5%, but 100 replications per condition give limited
precision. They do not guarantee calibration under other sample sizes,
category support, dependence, assignment or population distributions. In the
first replication of each condition, increasing both fits to 121 integration
points changed the likelihood-ratio statistic by less than \(10^{-6}\); this
illustrates a sensitivity check, not a universal integration-error bound.

These explanations separate missing mfrmr functionality from failures of a
particular fit. They do not imply that every model needs every downstream
workflow, or that JML inherits the MML reference distribution.

## Relationship to generalized many-facet models

Ordinary GPCM adds a positive discrimination parameter to the partial-credit
kernel; it does not replace the item position or step parameters. In the
current many-facet notation,

\[
\log\frac{P(Y_o=k)}{P(Y_o=k-1)}
  = \alpha_{g(o)}\{\eta_o-\tau_{g(o),k}\},
\qquad \prod_g\alpha_g=1,
\]

where the additive predictor \(\eta_o\) contains person ability and the signed
locations of the observed facets. Thus the selected slope multiplies the
complete adjacent-category predictor: person ability, every facet location or
fitted interaction inside \(\eta_o\), and the owned step. It is not a
loading-only formulation in which discrimination multiplies ability but leaves
rater severity and other intercept effects unscaled. Unit slopes reduce this
expression to the equal-discrimination PCM kernel.

The phrase *generalized many-facet Rasch model* is broader and is not a unique
synonym for this implementation. For example, Uto and Ueno (2020, equation 9)
use multiplicative task and rater slope blocks,
\(\alpha_i\alpha_r\), together with rater severity and rater-specific steps.
The current mfrmr GPCM estimates one slope family. MML can separate the
slope owner from the step owner; JML requires `slope_facet == step_facet`.
It does not jointly estimate task and rater slopes or introduce a
multidimensional ability vector. Selecting criterion slopes and rater steps
therefore remains a restricted many-facet GPCM, with the entire adjacent-category
predictor multiplied by the selected criterion slope.

A rater-owned fit is close to a restricted Uto--Ueno form after suppressing the
task-slope block, provided the remaining step and identification contracts are
also aligned. A criterion-owned fit instead has criterion-owned slopes and
steps and should be described as an analogous restricted many-facet GPCM, not
as an exact fit of Uto and Ueno's equation. Because free discrimination relaxes
the defining equal-discrimination Rasch restriction, "generalized MFRM" means
a model generalized from the MFRM baseline rather than a strict Rasch model.

### Which facet receives different slopes?

The slope is not one common number for the entire fitted GPCM. The selected
`slope_facet` contributes one positive slope for each of its levels. If it has
\(G\) levels, the returned table contains \(G\) slopes and the model estimates
\(G-1\) free relative log-slope contrasts because their geometric mean is fixed
to one.

| Configuration | Meaning | Current status |
|---|---|---|
| `step_facet = "Criterion"`, `slope_facet = "Criterion"` | Each criterion has its own relative discrimination and criterion-specific steps. | Supported; inference availability is listed above. |
| `step_facet = "Rater"`, `slope_facet = "Rater"` | Each rater has its own model-conditional relative discrimination and rater-specific steps. | Code-supported with rater-interpretation caveats; a slope is not automatically evidence of rater consistency. |
| Criterion steps with rater slopes, or conversely | Slope owner and step owner differ. | Supported with MML; one slope family and existing output-specific inference checks. JML, weighting reviews and simulation/design workflows are unavailable for this structure. |
| Criterion slopes and rater slopes together | Effective discrimination could involve two slope blocks. | Unsupported in the current package; this is closer to a multiplicative generalized-MFRM extension. |

Every facet that is not selected remains an additive location term inside
\(\eta\). For example, a criterion-owned GPCM can still estimate rater severity,
but it does not give each rater a separate discrimination parameter. The
criterion slope nevertheless scales that rater-severity contribution because
the contribution is inside \(\eta\). “No separate rater slope” and “rater
severity is unscaled” are therefore not the same statement.

### Criterion discrimination with rater-specific category use

Suppose several raters score each performance on several rubric criteria.
Use `step_facet` for **whose category transitions may differ** and `slope_facet`
for **whose responses may distinguish ability more sharply**. These are distinct
questions from overall rater severity, which remains a location effect.

```{r separate-owners, eval=FALSE}
fit <- fit_mfrm(ratings, "Person", c("Rater", "Criterion"), "Score",
  model = "GPCM", method = "MML",
  step_facet = "Rater", slope_facet = "Criterion")
slopes <- confint(fit)
grid <- expand.grid(Theta = seq(-2, 2, length.out = 21),
                    Rater = "R01", Criterion = "C01") # use your fitted labels
curves <- mfrm_curve_intervals(fit, grid)
plot(curves, title = NULL, subtitle = NULL)
results <- mfrm_results(fit, intervals = list(slopes = slopes, curves = curves),
                        include = c("fit", "plots"), compute = "never")
saveRDS(results, "criterion-slopes-rater-steps.rds")
mfrm_report(readRDS("criterion-slopes-rater-steps.rds"))
```

Relative slopes have geometric mean one. Step contrasts are centered within
each rater, separately from severity. Adequate crossing and category use are
needed to distinguish these effects; a sparse or confounded design can leave
intervals unavailable. A large slope or unusual step is not, by itself, a
rater-quality or training diagnosis. The numerical workflow checks do not
establish finite-sample coverage for this new structure.

For a PCM comparison, keep the same data, rater steps, constraints and population
model; the GPCM adds one fewer slope contrast than there are criteria. The default
GPCM estimates a population mean and variance, whereas default PCM fixes its
population, so defaults alone do not form that matched comparison. Follow the
matched population example earlier in this vignette. Existing likelihood and
local-information checks still apply. `bootstrap_mfrm_gpcm()` preserves both
owners and the observed assignment for parametric resampling; this is not a
new-design simulation. `build_weighting_review()`, `extract_mfrm_sim_spec()`
and the simulation/design workflows still require a shared owner. Portable
GPCM calibration and simultaneous rater/criterion slope families remain unavailable.

### An explanation for a high-school reader

Imagine that several teachers grade presentations on several criteria.

- **PCM** gives everyone rulers with the same magnification. One criterion may
  be harder and one teacher may be stricter, but a one-unit ability difference
  changes the rating probabilities at the same basic sharpness everywhere.
- **The current mfrmr GPCM** lets you choose one box of rulers. If the box is
  `Criterion`, each criterion can have a different magnification. If the box is
  `Rater`, each rater can have a different magnification. You may choose only
  one box for magnification. With MML, another box may supply the category
  steps: for example, criteria can differ in magnification while teachers
  differ in how they use the score categories. With JML, the same box must
  supply both magnification and category steps.
- **A broader generalized MFRM** can give both the task/criterion and the rater
  their own magnification knobs. In the Uto--Ueno example, the effective
  sharpness is the product of the task and rater settings. It can also model
  rater severity and rater-specific use of rating categories.

Thus the current GPCM is a smaller branch inside the broader family: it adds a
carefully controlled slope vector to a many-facet PCM, but it is not the full
two-knob generalized MFRM. Also, once discrimination is allowed to vary, the
model is generalized *from* the Rasch baseline; it no longer retains the strict
equal-discrimination Rasch restriction.

## Before fitting: model-choice triage

Do not choose `GPCM` only because it is the most flexible model in the menu.
Start with the score interpretation.

| Model | Use when | Main risk if over-used |
|---|---|---|
| `RSM` | The rating scale is intended to share one category-threshold structure across the step facet. | Real threshold differences can be hidden in residual diagnostics. |
| `PCM` | Thresholds may differ by item, criterion, task, or another designated step facet, but rating events should still contribute equally after conditioning on the modeled facets. | It can absorb threshold heterogeneity without asking whether some levels are more discriminating. |
| `GPCM` | The analysis explicitly allows discrimination-based reweighting and treats slopes as part of the substantive sensitivity question. | Better statistical fit can be mistaken for a better operational scoring rule. |

This ordering matters for reporting. `RSM` and `PCM` are the package's
equal-weighting reference route; `GPCM` is a slope-aware extension.
If equal contribution of items, criteria, or raters is part of the validity
argument, a better-fitting `GPCM` should be reported as sensitivity
evidence rather than as an automatic replacement.

Category avoidance is a separate decision. For example, a rater who uses only
3--8 on a declared 1--10 scale shows local restriction of range if the omitted
categories are used elsewhere. Start with `data_quality_report()` and inspect
`category_usage_by_facet` and `category_usage_summary`; also examine assignment
and case mix, fit, and unexpected responses. Do not move to `GPCM` merely
because a rater avoids extreme categories. A rater-owned PCM/GPCM may expose
owner-specific threshold or slope support problems, but it changes the model
being estimated and does not by itself explain or repair the scoring practice.

### Comparing PCM and GPCM

The package provides three complementary comparison layers:

1. `compare_mfrm()` returns AIC, Person-based BIC and SABIC comparisons when
   `comparison$table$ICComparable` is true. Its local-solution check is
   independent of slope-interval availability. Add `nested = TRUE` to test
   equal relative slopes under the matched PCM/GPCM specifications below;
   inspect `$lrt` and `$comparison_basis$lrt_reason`.
2. `build_weighting_review()` shows what the GPCM slopes changed: the slope
   profile, facet-measure shifts, and redistribution of information across
   levels.
3. `build_model_choice_review(..., run_weighting_review = TRUE)` bundles the
   statistical comparison, operational model roles, support boundaries, and
   the weighting review.

For example, `dat` is a long-format rating table with `Person`, `Rater`,
`Criterion` and `Score` columns. Fit the same estimated-normal population
in both models; `~1` estimates its mean and variance without covariates.

```r
persons <- data.frame(Person = unique(dat$Person))
fit_pcm <- fit_mfrm(
  dat, "Person", c("Rater", "Criterion"), "Score",
  method = "MML", model = "PCM",
  step_facet = "Criterion", quad_points = 61,
  population_formula = ~1, person_data = persons, person_id = "Person"
)

fit_gpcm <- fit_mfrm(
  dat, "Person", c("Rater", "Criterion"), "Score",
  method = "MML", model = "GPCM",
  step_facet = "Criterion", slope_facet = "Criterion",
  quad_points = 61,
  population_formula = ~1, person_data = persons, person_id = "Person"
)

comparison <- compare_mfrm(fit_pcm, fit_gpcm,
                           labels = c("PCM", "GPCM"), nested = TRUE)
comparison$lrt
comparison$comparison_basis$lrt_reason
comparison$table[, c(
  "Label", "LogLik", "AIC", "BIC", "SABIC", "ICComparable"
)]

weighting <- build_weighting_review(fit_pcm, fit_gpcm, nested = TRUE)
summary(weighting)
weighting$comparison_contract

review <- build_model_choice_review(
  PCM = fit_pcm,
  GPCM = fit_gpcm,
  run_weighting_review = TRUE
)
summary(review)
review$model_roles[, c(
  "Model", "StepStructure", "StepCoordinates", "FreeStepParameters",
  "SlopeStructure", "SlopeCoordinates", "FreeSlopeParameters",
  "FitReadiness", "FormalInference", "Interpretation"
)]
```

`StepCoordinates` and `SlopeCoordinates` count the values shown in fitted
tables. `FreeStepParameters` and `FreeSlopeParameters` count the independent
coordinates optimized after identification constraints. For example, with
four score categories and four levels of the step facet, PCM and GPCM display
12 step coordinates but optimize 8 free step parameters because each
three-step ladder is sum-zero. GPCM additionally displays four relative slopes
but optimizes three free log slopes because their geometric mean is one.

The `comparison_contract` describes the evidence available from the two fits.
Its `FormalModelSelectionAvailable` field follows the verified IC comparison,
not numerical convergence alone:

| Evidence tier | What may be concluded? |
|---|---|
| `mml_information_criterion_comparison` | The IC and solution checks pass. Compare candidate AIC/BIC and their differences; choose an operational weighting policy on separate substantive grounds. Use `confint(fit, parm = "slopes")` for separately checked relative-slope intervals and `nested = TRUE` for the equal-slope test. |
| `mml_descriptive_not_inference_ready` | Both fits passed numerical checks, but inferential qualification is incomplete. Available raw fit values and weighting changes remain descriptive. More quadrature points alone do not qualify GPCM selection. |
| `mml_screening_or_review_grid_only` or `mml_noncomparable_descriptive_only` | Other MML comparison restrictions remain; neither tier enables automatic model selection. |
| `jml_descriptive_reweighting_only` | The fitted slopes, measure shifts, and information redistribution may be reviewed, but the unpenalized JML likelihood gain does not select GPCM. |
| `mml_numerical_review_only` or `jml_numerical_review_only` | At least one fit still needs numerical review. A difference between retained optimizer traces is not inferential model evidence. |
| `cross_method_or_data_not_comparable` | The methods or prepared data do not supply the matched comparison required here. |

For a JML question, follow that table with recovery conditions rather than an
IC or LRT claim. `evaluate_mfrm_recovery()` already distinguishes
`unit_slopes`, `near_flat`, `moderate`, and `high_dispersion` generator
conditions. The unit-slope condition asks how often harmless sampling and
optimizer variation can create an apparent slope spread; the non-unit
conditions ask whether the declared heterogeneity can be recovered. FACETS
can support the aligned PCM/JML side of this review, but its Table 7
discrimination remains a post-fit diagnostic and not a free-slope GPCM fit.

PCM is obtained from the aligned GPCM response kernel by setting all slopes to
one. Calling `compare_mfrm(..., nested = TRUE)` records `PCM_in_GPCM` when the
structural restrictions are verified. The result in `$lrt` is returned only
when the numerical and likelihood checks also pass; otherwise read
`$comparison_basis$lrt_reason`. Older saved comparisons must be rebuilt to
obtain the new decision. Weighting consequences still need substantive review.

## Response-family boundary

Every current `fit_mfrm()` likelihood is ordered categorical. The distinction
between response values and observation frequencies is essential:

| Input or estimand | Current mfrmr treatment |
|---|---|
| Ordered binary response | Supported as a two-category ordered response. |
| Ordered polytomous response | Supported through RSM, PCM, or GPCM. |
| Unordered nominal/multinomial response | Unsupported; category order enters the current likelihood. |
| Poisson, negative-binomial, or grouped binomial-trial count | Unsupported as a response family. An integer `Score` is only an ordered category code. |
| Row-frequency/replication weight for an ordered rating | A positive numeric `weight` can weight that row's conditional ordered-category likelihood contribution. |

Thus, the fact that the fitted category probabilities sum to one does not make
the model a nominal-response or multinomial-logit model. Likewise, a frequency
weight does not turn the score into a Poisson count or account for dependence
among replicated ratings. Nor is it a general collapsed-person frequency
table: under MML, powering responses inside one Person's conditional pattern
is not equivalent to replicating a complete Person pattern after
marginalization. FACETS provides separate `Bn` and `P` response models; those
are outside the current mfrmr estimator even though FACETS-style coverage
documentation describes the handoff boundary.

Within that supported scope, an explicit identification contract is still
required. With default MML,
`gpcm_mml_identification = "free_population"` estimates an intercept-only
person distribution \(N(\beta_0,\sigma^2)\), while the level-specific log
slopes are constrained to sum to zero (geometric mean 1). The fitted slopes are
therefore relative discriminations with \(G-1\) free contrasts, and
\(\sigma\alpha_g\) gives the equivalent slope on a fixed-latent-standard-
deviation scale. This retains the conventional common-discrimination degree
of freedom. The explicit legacy mode
`gpcm_mml_identification = "fixed_standard_normal"` instead fixes both
\(\sigma=1\) and the slope geometric mean to one and is a narrower model.

## GPCM is not one estimator

Muraki (1992) supplies the GPCM response model and an MML-EM estimation route;
it is not a source for the present JML implementation. The present mfrmr JML
route is an identified fixed-effects, unpenalized joint likelihood with
post-fit boundary/recession auditing. This differs from penalized JML for GPCM
(Wijayanto et al. 2021), whose quadratic regularization changes the objective,
and from software routes that impose finite parameter boxes. A penalized,
adjusted, or box-constrained point estimate can be a useful comparator, but it
is not the maximizer of mfrmr's original unpenalized JML objective.

This distinction matters even when all methods return finite numbers. With a
fixed number of observations per person, jointly estimated person coordinates
are incidental parameters; structural JML estimates can retain finite-test-
length bias as the number of persons grows (Hessen 2025). An extreme response
pattern can additionally make the ordinary finite JML maximum unattained.
mfrmr therefore keeps recovery/bias evidence, boundary handling, numerical
convergence, and cross-software agreement as separate checks.
Under JML, the geometric-mean constraint resolves the scale of the jointly
estimated person coordinates and slopes.

## Report wording templates

Use wording that matches the model actually fitted:

- `RSM`: "We fit a many-facet rating-scale Rasch model, treating category
  thresholds as common across the step facet."
- `PCM`: "We fit a many-facet partial-credit Rasch model, allowing thresholds
  to vary by the designated step facet while retaining equal discrimination."
- `GPCM`: "We fit a generalized partial-credit many-facet model
  as a slope-aware sensitivity analysis; interpretation focused on whether
  discrimination-based reweighting changed the substantive conclusions."

Avoid wording that says `GPCM` "improves the score" solely because it
improves log-likelihood, `AIC`, or `BIC`. The model can fit better while changing
the scoring contract.

## Checking the support boundary

`gpcm_capability_matrix()` is the canonical reference. It returns one row
per helper family with a `Status` column drawn from `supported`,
`supported_with_caveat`, `blocked`, and `deferred`. Read `Boundary` for the
interpretive limit and `RecommendedRoute` for the route to use next. The
default print is deliberately compact; subset by status to inspect a focused
set of rows.

```{r capability-supported}
library(mfrmr)
gpcm_capability_matrix("supported")[, c("Area", "Status")]
```

```{r capability-with-caveat}
gpcm_capability_matrix("supported_with_caveat")[, c("Area", "Status")]
```

```{r capability-blocked}
gpcm_capability_matrix("blocked")[, c("Area", "Status", "RecommendedRoute")]
```

```{r capability-deferred}
gpcm_capability_matrix("deferred")[, c("Area", "Status", "Boundary", "RecommendedRoute")]
```

The matrix is intentionally conservative. A row stays in `blocked` or
`deferred` even when some individual computation is already available, because the
scope statement includes the interpretation needed for a complete public
workflow rather than only checking whether code executes.

## Slope values, boundaries, and uncertainty

The finite `LogEstimate` and `Estimate` columns in `fit$slopes` are optimizer
outputs, not automatic evidence of a finite estimand. Use `ParameterStatus` and
the `Primary*` columns first. Under JML, mfrmr can certify a monotone
sum-to-zero log-slope path while holding the retained additive coordinates
fixed. Such a result yields `unbounded_low`, `unbounded_high`, or
`unbounded_both`. If no such path is certified, the joint Person--facet--step--
slope problem remains unevaluated; the finite optimizer output is retained only
as a numerical trace. A second bounded check lets additive coordinates move
with an ordered positive/negative log-slope pair. A competitive path from this
family is recorded as a candidate, but `ParameterStatus` remains
`not_evaluated` and the primary value remains unavailable: this sufficient
path result does not determine the global maximum of the non-concave GPCM
likelihood. Failure to find such a path is equally limited to that family.
Under MML, Persons are integrated out, so neither conditional JML result is
reused as marginal-likelihood evidence. The MML branch records a separate
sufficient-path check for the fitted finite-quadrature objective, but that
check is not a complete analysis of the continuous marginal likelihood.
That global boundary question is separate from the local MML approximation.
`confint()` or `diagnose_mfrm()` qualifies the relative slopes using the current
likelihood, gradient, unregularized information and the other checks described
above. Qualified diagnostic tables use `ParameterStatus = "finite_local_solution"`
and `PrimaryEstimateBasis = "verified_local_mml_solution"`; this does not claim
a global maximum. The unmodified fit may still carry its original unevaluated
parameter audit. Attaching diagnostics synchronizes its slope parameter rows,
while the global fit-readiness record remains unchanged.

Local `Optimizer*SE` and `Optimizer*CI` calculations remain available for
numerical review where possible, even if ordinary intervals are withheld.
`SEEligible`, `CIEligible`, `CIUse` and `InferenceReview` identify the separate
uncertainty decision. A positive information matrix alone is insufficient,
and a regularized inverse cannot qualify ordinary intervals.

## What can and cannot be compared across programs

A column labelled "discrimination" is not sufficient evidence that two
programs fitted the same GPCM. FACETS reports an element-discrimination
diagnostic after estimating its Rasch measures and explicitly does not allow
that diagnostic to alter the other estimates. It is therefore not a matched
counterpart of a jointly estimated free slope in `mfrmr`. FACETS remains useful
for unit- or explicitly fixed-discrimination reductions, category/threshold
and facet-measure comparisons within a common Rasch specification, and as a
deliberately different-model diagnostic benchmark.

### Model structure: which effect does discrimination change?

TAM and ConQuest both support more flexible slope designs than the current
mfrmr interface. That does not make every many-facet GPCM in these programs
the same model.

| Route | Where slopes are assigned | Relationship to current mfrmr |
|---|---|---|
| `mfrmr::fit_mfrm(model = "GPCM")` | One selected facet supplies positive relative slopes. MML permits a different step owner; JML requires the same owner. | Each slope multiplies the complete adjacent-category predictor. |
| TAM `tam.mml.2pl()` | Item slopes, shared slope groups or a design `E`; Example 14c combines this with a facet intercept design `A`. | The example has slopes on ability and separately additive facet intercepts. `tam.mml.mfr()` alone does not estimate slopes. |
| ConQuest `scoresfree` | By default, one slope for each generalized item, meaning a combination of facets; a `C` design can constrain sharing. | Sharing slopes by criterion still does not impose their products with estimated rater effects in the intercept design. |

The TAM construction is documented in
[Example 14, Model 14c](https://alexanderrobitzsch.github.io/TAM/reference/tam.mml.html).
ConQuest's [Note 8, pages 1--5](https://www.acer.org/files/Note_8--The_ConQuest_4_Model.pdf)
describes its scoring design, population model and identification. It updates
the earlier ConQuest 3 note; limitations described only in that older note
should not be assigned to current ConQuest.

For a rubric criterion \(c\), rater \(r\) and category transition \(k\),
compare these adjacent log odds, using minus signs for difficulty and severity:

\[
\begin{aligned}
\text{Current mfrmr:}\quad &\alpha_c(\theta_p-d_c-\rho_r-\tau_{ck}),\\
\text{Additive-intercept design:}\quad &a_c z_p-b_c-u_r-v_{ck}.
\end{aligned}
\]

The second expression represents the TAM example and a ConQuest design with
criterion-shared scores and additive rater intercepts. It is not every model
those programs can fit. The comparison is an algebraic consequence of their
documented designs: the first expression contains the product
\(\alpha_c\rho_r\); independent linear intercept and slope designs do not
impose that product of two estimated parameters.

For example, on the mfrmr scale, two raters differing by 0.5 severity units
produce an absolute log-odds difference of 0.4 for a criterion with slope 0.8 and 0.625
for one with slope 1.25. In the additive-intercept model, the shared rater
contrast has the same log-odds effect at every criterion. These are different
assumptions about how rater severity acts. Changing which facet supplies
slopes and steps does not by itself change this assumption.

`sirt::rm.facets()` provides a useful second MML
reference: it estimates item and/or rater slopes under product-one centering and
an estimated normal trait spread. Its item-only GPCM and equal-discrimination
many-facet reductions are candidate matched comparisons after category, threshold,
quadrature, and centering conventions are aligned. Its general free-slope rater
kernel is not the current mfrmr kernel: sirt uses the product of item and rater
slopes on the trait term while rater severity remains a separate location term,
whereas mfrmr assigns one slope owner and multiplies the complete adjacent-
category predictor. Such results are different-model sensitivity evidence, not
automatic numerical validation. `immer` provides PCM-design JML/CML/CCML
routes and a hierarchical rater model, not a direct free-GPCM many-facet match.
Its role is consequently a unit-slope reduction or alternative-model
sensitivity analysis rather than free-slope numerical validation.

### Exact overlap and what has been checked

With items only, free item-specific steps can absorb the slope multiplication.
For an estimated normal ability population, write
\(\theta=\mu+X\beta+\sigma\epsilon\), with standard-normal \(\epsilon\), and
\(\prod_i\alpha_i=1\). Set
\(z=(\theta-\mu)/\sigma\), \(a_i=\sigma\alpha_i\),
\(b_i=\alpha_i(d_i-\mu)\), \(v_{ik}=\alpha_i\tau_{ik}\) and
\(\beta_z=\beta/\sigma\). Then
\(\alpha_i(\theta-d_i-\tau_{ik})=a_i z-b_i-v_{ik}\).
Here \(\sigma\) is the residual population SD given any covariates \(X\);
without covariates it is the population SD. This derives an identical response
model on a unit residual-variance scale;
the intercept/step signs and centering must still match each program.
Fixing both population variance and the geometric mean of slopes to one
instead imposes an additional restriction, so it is not the same conversion.
The overlap is for positive slopes; ConQuest also exposes a `positivescores`
setting, so the permitted parameter space must be checked.

A retained validation comparison exercises the exact item-only TAM
overlap on TAM's `data.gpcm` fixture at q=31 and q=41. TAM fixes the latent
variance and estimates absolute slopes; mfrmr fixes the geometric mean of its
relative slopes and estimates the latent variance. After applying the analytic
scale and location transformation, maximum q-specific differences were about
`1.8e-5` for relative slopes, `1.3e-5` for transition thresholds, `6e-6` for
fitted category probabilities, and `1.4e-6` for deviance. The transformed TAM
probability formula itself agreed to binary64 precision. This is external
evidence for the item-only GPCM kernel, coordinate map, and continuous MML
target; it is not a comparison of the rater-plus-slope model, a cross-engine
SE validation, or permission to override mfrmr readiness. These results use
TAM 4.3-25. A separate q=31 many-facet re-expression was retained as a
different-model comparison after the algebraic distinction above was found.

An earlier local ConQuest 5.47.5 MML run checked the item-only overlap with
one population covariate, 120 Persons, five items and four categories. The
deviances differed by about `1.3e-5` after matched identification. Its exact
coordinate transformation reproduces probabilities within `1e-14`. This is
historical, version-specific evidence for that model and mapping, not a fresh
check of the current source, interval coverage or a multifacet slope model.
The public ConQuest overlap bundle does not automate this GPCM comparison.

### Inference: same estimates do not imply the same intervals

The saved TAM 4.3-25 fits retain marginal slope and intercept SEs, but not the
joint slope covariance or
slope--intercept cross-covariance required by the nonlinear transformation to
mfrmr's relative slopes, thresholds, and free population scale. Two positive-
definite covariance matrices can preserve exactly those reported marginal SEs
while producing different transformed SEs. This non-identification means the
comparison must be withheld rather than completed by assuming zero covariance.
TAM's documented [`tam.se()`](https://alexanderrobitzsch.github.io/TAM/reference/tam.se.html)
also omits parameter covariances and labels loading SEs for `GPCM.design`
experimental. This describes that route and the saved objects; it does not
rule out obtaining a joint matrix through additional methods.

ConQuest's [command reference](https://conquestmanual.acer.org/s4-00.html)
provides `estimatecovariances` with `stderr=empirical` and `matrixout`, and
`show` table 10 reports parameter-estimate error covariances. In contrast,
`export covariance` exports the **latent-population** covariance. It is not
the sampling covariance needed for transformed calibration SEs. Parameter
ordering, constraints and the inclusion of free score parameters must be
checked in the actual exported matrix before using it. No such current-source
joint-covariance comparison is claimed here. ConQuest's estimated-score route
is an MML comparator; its JML route does not estimate free scores.

For example, let \(s=\log\sigma\) and \(\ell_i=\log\alpha_i\). A
population-SD-standardized slope has \(\log a_i=s+\ell_i\), so its delta-method
variance requires

\[
\operatorname{Var}(\log a_i)=\operatorname{Var}(s)+
\operatorname{Var}(\ell_i)+2\operatorname{Cov}(s,\ell_i).
\]

Multiplying the current relative-slope interval endpoints by the estimated
\(\sigma\) omits two of these terms. Use
`confint(fit, parm = "slopes", scale = "standardized")` to include all three;
the default still reports relative slopes. Conversely, converting absolute
slopes to geometric-mean-one slopes requires their joint covariance.

TAM's [`anova()` method](https://alexanderrobitzsch.github.io/TAM/reference/anova.tam.html)
reports information criteria and likelihood-ratio comparisons. These methods
and mfrmr's current IC/LRT operations still require comparable likelihoods.
Replicating another program's fit requires the same model. IC selection can
instead compare different, nonnested models when their likelihoods refer to
the same observed data with compatible likelihood definitions and constants.
Verify category coding, population specification, each model's free-parameter
count and integration accuracy. An LRT additionally requires a nested null;
different many-facet slope actions are not automatically nested.
The simple PCM/GPCM null with \(G-1\) relative-slope
restrictions does not establish a reference distribution for arbitrary
cross-program design changes.

These distinctions also govern diagnostics. FACETS' discrimination, `mfrmr`
residual PCA or bias screens, and hierarchical-rater variance components answer
different questions. Agreement among them can be supporting evidence; a
difference is not a software error until the fitted likelihood, estimand,
identification, data rows, and output transformation have first been shown to
match.

A population-level validation also asked whether the current
complete-predictor slope and a loading-only slope are merely two coordinate
systems. They reduce exactly when slopes are all one or when the non-owner
Rater contrast is zero. In a fully crossed four-Criterion/four-Rater design,
however, their identifying adjacent-logit difference in differences is zero
for loading-only and
\(-\{\alpha_c-\alpha_d\}\{\rho_r-\rho_s\}\) for the current kernel. After all
available slopes, severities, and transition boundaries were re-estimated,
moderate and strong fixed conditions retained population KL losses of about
0.0016 and 0.0085--0.0088 per response, with maximum common-grid probability
differences of about 0.047--0.184. q=31 and q=41 conclusions agreed closely.
These are fixed model-form demonstrations, not practical cutoffs or
finite-sample operating characteristics. They support treating a loading-only
kernel as a separate possible future family rather than as an interpretation
switch for an existing `GPCM` fit.

Sparsity changes this distinction structurally, not merely by reducing sample
size. On an observed Rater-by-Criterion graph with no cycles, the products
\(\alpha_c\rho_r\) can be represented exactly by new additive Rater and
Criterion coordinates. Thus even a connected tree cannot distinguish the two
slope actions. A known-ability validation compared four connected
graphs. At 250 Persons, truth-selection rates were about 0.96 under complete
crossing, 0.76--0.78 for a balanced eight-edge cycle, and only 0.58 for an
eight-edge cycle localized to two Raters and two Criteria; the connected tree
remained observationally equivalent. Each Person was observed on every edge,
and abilities and both parameter vectors were supplied to the comparison, so
these are optimistic fixed-parameter results rather than JML or MML model-
selection performance. The design lesson is to inspect which slope and
severity contrasts are traversed by cycles, not to infer identifiability from
connectedness, density, or an anchor percentage alone.

A subsequent finite-sample validation removed the known-ability advantage and
refitted both slope actions by direct MML. The common identification estimated
geometric-mean-one relative slopes, sum-zero Rater severities, mean-zero
transition boundaries, and a normal population mean and standard deviation.
Across 12 replications per truth and design at 250 training Persons, all 96
candidate fits converged and all full 19-coordinate Hessians were positive
definite. Two starts differed by at most about `7.9e-6` log-likelihood units,
and q=31 versus q=41 changed none of the 48 paired family decisions. Complete
crossing selected the true family in every training and independent holdout
dataset. Under the balanced cycle, training selected the truth in 7--9 of 12
datasets and median relative-slope RMSE increased from about 0.06 to as much as
0.12. A stricter retained-solution polish reduced the largest gradient by
roughly an order of magnitude without materially changing likelihoods. These
small fixed counts clarify optimizer, curvature, and design contributions;
they do not calibrate a universal selection, standard-error, or readiness
rule.

## Source-grounded recovery interpretation

The `GPCM` route follows Muraki's generalized partial credit model
and its information-function extension. The package-specific `slope_regime`
labels are narrower than that model theory: they summarize the centered
log-slope spread of the simulation generator so recovery evidence can be
read against a declared stress condition. They are not model-fit tests and
they are not literature-derived adequacy cut points.

For simulation reporting, read direct recovery checks in an ADEMP-style order:
the data-generating mechanism first, then the estimands and performance
measures, and only then the row-level recovery diagnostics. In practice, this
means:

1. Build or extract an explicit `mfrm_sim_spec`.
2. Run `evaluate_mfrm_recovery()` for the direct parameter-recovery question.
3. Run `assess_mfrm_recovery()` with practical RMSE/bias limits.
4. Read `summary(recovery_review)`, then
   `recovery_review$condition_reporting_notes` and
   `recovery_review$condition_review`, then
   `recovery_review$diagnostic_reporting_notes` and
   `recovery_review$diagnostic_review` when optional diagnostics were retained,
   then
   `plot(recovery_review, type = "status")`, then
   `plot(recovery_review, type = "metrics", metric = "rmse")`.

## Supported routes and verification evidence

The following `GPCM` routes are documented and verified within the
stated constraints:

- **Fitting and core summaries** via `fit_mfrm(model = "GPCM", step_facet =
  ...)`, with the direct `MML` engine. Omitting `slope_facet` keeps the step
  owner; naming another facet separates the two owners in MML.
- **Posterior scoring and information** via `predict_mfrm_units()`,
  `sample_mfrm_plausible_values()`, `compute_information()`, and
  `plot_information()`.
- **Curve and category views** via `plot(fit, type = c("wright",
  "pathway", "ccc", "ccc_surface"))`, `category_structure_report()`, and
  `category_curves_report()`.
- **Slope-aware simulation specifications** via `build_mfrm_sim_spec()`
  and `simulate_mfrm_data()`.
- **Direct recovery checks** via `evaluate_mfrm_recovery()` and
  `assess_mfrm_recovery()`, including fitted GPCM slope recovery on the
  log-slope scale.

## What works with caveats

The following are exposed for `GPCM` but should be read as exploratory
screens rather than as Rasch-style invariance evidence:

- `diagnose_mfrm()` and the residual and unexpected-response stack:
  `unexpected_response_table()`, `displacement_table()`,
  `measurable_summary_table()`, `rating_scale_table()`,
  `interrater_agreement_table()`, `facet_quality_dashboard()`,
  `plot_qc_dashboard()`, `plot_marginal_fit()`,
  `plot_marginal_pairwise()`.
- `reporting_checklist()` and `precision_review_report()` route to the
  supported direct tables and plots. The broader APA/QC/export family is
  available as caveated sensitivity-reporting output with explicit
  `gpcm_boundary` rows.
- `build_misfit_casebook()` inherits the exploratory screening framing of
  its underlying sources.
- `estimate_bias()` now provides GPCM conditional screening rows with
  slope-aware information and profile-likelihood columns. Treat these rows as
  screening evidence for follow-up, not as standalone confirmatory fairness
  tests. `unexpected_after_bias_table()` provides a descriptive in-sample
  before/after flag comparison; a lower flag count does not show that bias has
  been removed.
- `estimation_iteration_report()` provides a slope-aware reconstructed
  optimization trajectory. It is a diagnostic replay, not the exact optimizer
  history or an additional convergence test.
- `analyze_dff()`, `analyze_dif()`, `dif_interaction_table()`,
  `dif_report()`, `plot_dif_heatmap()`, and `plot_dif_summary()` provide
  GPCM DFF/DIF screening and reporting surfaces with explicit
  `gpcm_boundary` rows.
- `build_apa_outputs()`, `build_visual_summaries()`, `run_qc_pipeline()`,
  `build_mfrm_manifest()`, `build_mfrm_replay_script()`,
  `export_mfrm_bundle()`, package-native scorefile export, and
  `build_linking_review()` return caveated `GPCM` reporting or
  exploratory-review objects with explicit `gpcm_boundary` rows. The
  package-native scorefile can include native structural delta-method
  expected-score SEs and score-side delta SEs selected by `score_se_method`
  when the required MML diagnostics are available, but those SEs are not
  FACETS-equivalent score-side uncertainty.
- `evaluate_mfrm_design()`, `predict_mfrm_population()`,
  `evaluate_mfrm_diagnostic_screening()`, and
  `evaluate_mfrm_signal_detection()` are available as caveated role-based
  repeated simulation/refit routes. Treat their outputs as design-level or
  screening sensitivity evidence, not as operational scoring, calibrated
  inferential testing, or arbitrary-facet planning validation.

The dashboard marks the fair-average panel unavailable under `GPCM`; use
`fair_average_table()` directly for the slope-aware element-conditional table
and `fair_average_table(fit_gpcm, fair_se = TRUE)` to inspect structural
fair-average SEs for non-person rows. These condition on Person EAP/reference
means and remain diagnostic-only: `FairCIEligible = FALSE`, including when
the covariance is computable or regularized. Full-refit coverage remains
unverified. Historical measure-level SE columns are not fair-score SEs.

## What is intentionally restricted

The slope-aware `fair_average_table()` route and package-native scorefile route
are available under `GPCM`, including native expected-score uncertainty and
score-side delta SEs where the required MML diagnostics support them. Full
FACETS-style score-side compatibility remains restricted because free
discrimination changes the relationship between the latent measure and
operational score-side summaries. Specifically:

- `facets_output_contract_review()` still depends on FACETS-style
  compatibility semantics that are not generalized to free discrimination.
- posterior-predictive checks and MCMC estimation are not available in mfrmr;
  use external Bayesian software when those analyses are required.
- Caveated reporting, export, linking, design-forecast, and screening helpers
  must keep their `gpcm_boundary` wording visible and must not imply
  FACETS-equivalent score-side uncertainty, operational scoring, calibrated
  screening criteria, or arbitrary-facet planning validation.

## Recommended substitutes

When a restricted helper is needed for a `GPCM` report, the practical paths
are:

- Refit with `model = "PCM"` if the discrimination-free assumption is
  defensible for the data and a full FACETS score-side review is required;
  `compare_mfrm()` can describe available likelihood differences for matched
  data and model specifications. A shared quadrature grid and a denser-grid
  sensitivity check address numerical comparability. The PCM/GPCM test also
  requires matching population and constraint settings. Relative-slope intervals
  and IC comparison have their own solution and likelihood checks.
- Keep the report on the `GPCM` fit itself but draft the manuscript section
  manually around the supported tables: `summary(fit)` for parameters,
  `diagnose_mfrm()` for residual fit, `facet_quality_dashboard()` for the
  per-facet quality summary, and `compute_information()` for precision
  evidence.
- When a Rasch-family baseline comparison is substantively useful, optionally
  generate a second manifest from a parallel `RSM` or `PCM` fit. The two fits
  can be reported side by side, with the `GPCM` fit identified as the
  discrimination-aware counterpart. This is not required to use the caveated
  `GPCM` manifest/replay/export route.

Restricted helpers use the capability matrix at runtime. An unsupported
`GPCM` call stops with the relevant limitation and a supported
alternative instead of producing a partial score-side or backend result. Use
`gpcm_capability_matrix()` or `mfrmr_output_guide("gpcm")` before choosing a
downstream route.

## A worked example

The `example_core` dataset includes a small synthetic block that supports a
`GPCM` fit. This example uses compact quadrature and iteration settings
for a shorter runtime; for substantive analysis, rerun with the package default
or a higher quadrature setting and a larger recovery design.

```r
library(mfrmr)
toy <- load_mfrmr_data("example_core")

fit_gpcm <- fit_mfrm(
  data = toy,
  person = "Person",
  facets = c("Rater", "Criterion"),
  step_facet = "Criterion",
  slope_facet = "Criterion",
  score = "Score",
  model = "GPCM",
  method = "MML",
  quad_points = 7,
  maxit = 20
)

fit_gpcm
gpcm_summary <- summary(fit_gpcm, profile = "fit", detail = "brief")
gpcm_summary$decision
gpcm_summary$inference_evidence

diag_gpcm <- diagnose_mfrm(fit_gpcm)
summary(diag_gpcm)

info <- compute_information(fit_gpcm)
plot_information(info)

rec_gpcm <- evaluate_mfrm_recovery(
  sim_spec = build_mfrm_sim_spec(
    n_person = 30,
    n_rater = 3,
    n_criterion = 4,
    raters_per_person = 2,
    model = "GPCM",
    step_facet = "Criterion",
    slope_facet = "Criterion",
    slopes = c(0.8, 1.0, 1.15, 1.05)
  ),
  reps = 10,
  model = "GPCM",
  fit_method = "MML",
  quad_points = 7,
  maxit = 20,
  include_diagnostics = TRUE,
  diagnostic_fit_df_method = "both",
  seed = 1
)
review_gpcm <- assess_mfrm_recovery(
  rec_gpcm,
  max_rmse = c(facet = 0.5, step = 0.5, slope = 0.25),
  max_abs_bias = c(default = 0.25)
)

summary(review_gpcm)$overview
summary(review_gpcm)$reading_order
review_gpcm$condition_reporting_notes[, c(
  "ConditionArea", "ReportingAttention", "ConditionFinding"
)]
review_gpcm$condition_review[, c(
  "Model", "GPCMSlopeRegime", "StressLevel", "ScoreSupportStatus"
)]
review_gpcm$diagnostic_reporting_notes[, c(
  "Facet", "ReportingAttention", "DiagnosticFinding"
)]
summary(review_gpcm)$diagnostic_review
plot(review_gpcm, type = "status")
plot(review_gpcm, type = "metrics", metric = "rmse")

```

Use the same `Decision` fields as for RSM and PCM. In particular, optimizer
convergence does not override incomplete GPCM identifiability, curvature, or
slope-boundary evidence. If `FormalInference` is `"No"`, the diagnostic and
information calls above are investigative routes for resolving the recorded
reason; they do not promote the fit to an inferential result.

The fit, summary, residual diagnostics, information, recovery, fair-average, and
conditional bias-screening helpers all run under `GPCM` with the caveats listed
above. `build_apa_outputs(fit_gpcm)` returns a caveated sensitivity-reporting
object with a `gpcm_boundary`; full FACETS score-side review remains on the
`RSM` / `PCM` route.

## Estimator references

- Muraki, E. (1992). *A generalized partial credit model: Application of an EM
  algorithm*. Applied Psychological Measurement, 16, 159--176.
- Uto, M., & Ueno, M. (2020). *A generalized many-facet Rasch model and its
  Bayesian estimation using Hamiltonian Monte Carlo*. Behaviormetrika, 47,
  469--496.
- Wijayanto, F., Mul, K., Groot, P., van Engelen, B. G. M., & Heskes, T.
  (2021). *Semi-automated Rasch analysis using in-plus-out-of-questionnaire log
  likelihood*. British Journal of Mathematical and Statistical Psychology, 74,
  313--339.
- Hessen, D. J. (2025). *A convexity-constrained parameterization of the random
  effects generalized partial credit model*. British Journal of Mathematical
  and Statistical Psychology, 78, 401--419.

## Current boundary

Score-side semantics for free-discrimination polytomous models differ from the
Rasch-family route. Use the current matrix returned by
`gpcm_capability_matrix()` as the workflow contract and
`gpcm_score_side_contract()` for score-side alternatives.
