---
title: "Portable calibration and fresh-session scoring"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Portable calibration and fresh-session scoring}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
is_cran_check <- !isTRUE(as.logical(Sys.getenv("NOT_CRAN", "false")))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = !is_cran_check
)
```

This workflow creates a portable calibration from an eligible fitted model,
validates and freezes it, saves it, and then scores new Persons without the
source fit or training responses. The scoring phase uses only the saved
artifact and the new response rows.

The portable workflow is deliberately narrower than fitted-object scoring.
It supports one observed score scale and one latent dimension for `RSM` or
`PCM` fitted by `MML` under the fixed standard-normal scoring basis. It does
not currently create portable artifacts for estimated-population or latent-
regression fits, `JML`, or bounded `GPCM`.

```{r capabilities}
library(mfrmr)

mfrm_calibration_capabilities()[, c(
  "Model", "Estimator", "ScoringBasis", "PortableCalibration"
)]
```

## Calibration session

The bundled `example_core` data are synthetic. This example uses 18 Persons
to keep vignette runtime modest; a substantive analysis should use its planned
sample and design rather than copying that count.

```{r create-calibration}
synthetic <- load_mfrmr_data("example_core")
person_ids <- unique(as.character(synthetic$Person))
training_ids <- person_ids[seq_len(18L)]
training <- synthetic[
  as.character(synthetic$Person) %in% training_ids,
  ,
  drop = FALSE
]

fit <- suppressWarnings(fit_mfrm(
  training,
  person = "Person",
  facets = c("Rater", "Criterion"),
  score = "Score",
  model = "RSM",
  method = "MML",
  quad_points = 5,
  maxit = 20
))

draft <- extract_mfrm_calibration(
  fit,
  calibration_id = "portable-rsm-example",
  source_fit_id = "synthetic-rsm-fit",
  scoring_quad_points = 31
)
review_mfrm_calibration(draft)

validated <- validate_mfrm_calibration(draft)
frozen <- freeze_mfrm_calibration(validated)
frozen

artifact_file <- tempfile(fileext = ".rds")
save_mfrm_calibration(frozen, artifact_file)
```

The fit-time quadrature and the scoring-time quadrature are distinct. Here the
small fit-time grid keeps the example quick, while the artifact explicitly
records the 31-point grid used for later posterior scoring.

## Fresh scoring session

In operational use, start a new R process and supply only the artifact path and
new response rows. The following chunk removes the fit, draft, validation
object, and training data before loading the artifact, so none can influence
the score.

```{r score-from-artifact}
new_rows <- synthetic[
  as.character(synthetic$Person) %in% person_ids[19:20],
  c("Person", "Rater", "Criterion", "Score"),
  drop = FALSE
]
new_ids <- stats::setNames(
  c("NEW_PERSON_1", "NEW_PERSON_2"), person_ids[19:20]
)
new_rows$Person <- unname(new_ids[as.character(new_rows$Person)])
rownames(new_rows) <- NULL

rm(fit, draft, validated, frozen, training, synthetic)
invisible(gc())

calibration <- load_mfrm_calibration(artifact_file)
scores <- score_mfrm_calibration(calibration, new_rows)
scores$estimates
scores$row_review
scores$person_dispositions

unlink(artifact_file)
```

For a separate scoring script, the essential code is:

```r
library(mfrmr)

calibration <- load_mfrm_calibration("reviewed-calibration.rds")
new_rows <- read.csv("new-responses.csv", stringsAsFactors = FALSE)
scores <- score_mfrm_calibration(calibration, new_rows)
write.csv(scores$estimates, "person-scores.csv", row.names = FALSE)
```

Unknown facet levels, unknown score values, invalid weights, ambiguous
duplicates, and malformed artifacts fail closed. Missing responses require an
explicit `missing_response = "omit"` policy. Every returned estimate is a
posterior EAP conditional on the recorded fixed point calibration and prior;
its uncertainty interval excludes calibration-parameter uncertainty. Validate
the intended score use, population transport, and decision consequences
outside this numerical workflow.
