---
title: "Exploring person, rater, and task attributes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Exploring person, rater, and task attributes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
available <- requireNamespace("mice", quietly = TRUE) &&
  requireNamespace("cluster", quietly = TRUE)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = available)
```

An assessment team wants to describe its raters' backgrounds before planning
training. Do groups summarize different combinations of experience, workload,
and training? Would those descriptions change with the number of groups,
feature weights, or plausible replacements for unrecorded attributes?

This example follows that question from a feature table to a comparison of
partitions. It uses **120 fictional raters**, not observations from an assessment
program. No true groups are planted. Successful execution does not establish
that these features, imputation models, or groups are appropriate for your
raters. The groups describe backgrounds; they do not measure rater quality,
severity, or ability, and do not identify which training will be effective.
Later sections add separate person and task attributes, link all three
classifications to a sparse assignment roster, and explain the roles of
feature selection and PCA.

These APIs belong to development version 0.2.4.9001. Install the optional
`cluster` and `mice` packages to run the entire example.

## Choose attributes and their meaning

Use one row per rater, rather than repeating attributes for every rating.
Keep identifiers separate from features. Select variables for the question
before examining which choices produce attractive groups.

| Column | Meaning and coding | Role here |
| --- | --- | --- |
| `ExperienceYears` | Completed years of rating experience; numeric | Feature |
| `AnnualRatings` | Number of responses rated in the previous year; numeric | Feature |
| `Specialty` | Main subject area; unordered factor | Feature |
| `TrainingLevel` | Highest completed training stage; ordered factor | Feature |
| `Certified` | Holds the relevant certification; logical | Feature |
| `WorkshopHours` | Hours of workshops attended in the previous year; numeric | Auxiliary imputation predictor |
| `MentoringYears` | Years in a formal mentoring role; undefined without that role | Reviewed, then excluded from the all-rater feature set |

```{r attributes}
library(mfrmr)
set.seed(20260921)
n <- 120L
raters <- data.frame(
  Rater = sprintf("R%03d", seq_len(n)),
  ExperienceYears = sample(0:30, n, replace = TRUE),
  AnnualRatings = round(exp(rnorm(n, log(250), 0.6))),
  Specialty = factor(sample(c("Language", "Science", "Social science"),
                            n, replace = TRUE)),
  TrainingLevel = ordered(sample(c("Introductory", "Intermediate", "Advanced"),
                                 n, replace = TRUE),
                          levels = c("Introductory", "Intermediate", "Advanced")),
  Certified = sample(c(FALSE, TRUE), n, replace = TRUE),
  WorkshopHours = sample(0:80, n, replace = TRUE)
)
has_mentoring_role <- sample(c(FALSE, TRUE), n, replace = TRUE, prob = c(.7, .3))
raters$MentoringYears <- ifelse(has_mentoring_role,
  pmin(raters$ExperienceYears, sample(0:10, n, replace = TRUE)), NA_real_)

# Introduce unrecorded values only after generating the fictional attributes.
feature_names <- c("ExperienceYears", "AnnualRatings", "Specialty",
                   "TrainingLevel", "Certified")
missing_counts <- c(12L, 18L, 6L, 10L, 10L)
for (j in seq_along(feature_names)) {
  raters[[feature_names[j]]][sample.int(n, missing_counts[j])] <- NA
}
head(raters)
```

The [Gower distance calculation](https://stat.ethz.ch/R-manual/R-devel/library/cluster/html/daisy.html)
accommodates these mixed types. Numeric features use their
observed ranges. Ordered levels use ordered integer codes, so the declared
order and the spacing implicit in that coding matter. Logical features treat
both matches (`FALSE/FALSE` and `TRUE/TRUE`) alike. Certification is therefore
not coded as a special presence-only attribute. Equal feature weights are a
starting choice; correlated or redundant attributes can still receive too much
combined influence. A highly unusual numeric value can also change its
feature's range and hence other raters' distances.

## Distinguish unrecorded and undefined values

Here a missing mentoring duration means that the attribute does not apply.
It is not a duration waiting to be estimated. A duration of zero would instead
mean an applicable mentoring role with no completed years.

```{r missing-reasons}
all_features <- c(feature_names, "MentoringYears")
initial <- mfrm_features(raters, "Rater", all_features)
reasons <- initial$missing
reasons$Reason <- ifelse(reasons$Feature == "MentoringYears",
                        "Not applicable: no mentoring role", "Not recorded")
review <- mfrm_features(raters, "Rater", all_features, reasons)
summary(review)
table(review$missing$Reason)
```

The target is a description of **all raters**. Requiring a mentoring duration
would exclude nonmentors. We therefore leave it out of the selected features
and the imputation predictors, retaining the original review. A separate
analysis confined to mentors would address a different population.

```{r selected-features}
selected_reasons <- subset(review$missing, Feature %in% feature_names)
features <- mfrm_features(raters, "Rater", feature_names, selected_reasons)
features$feature_summary
table(features$row_summary$Complete)
```

`mfrm_cluster(features, k = 3)` would stop on the remaining missing values.
Explicit `missing = "omit"` is available, but selects complete raters and
changes the population being described. It does not correct missing-data bias.

## Fit an explicit imputation model

Only cells reviewed as unrecorded are eligible here. The fictional omissions
were introduced at random; in an actual program, an explanation based on the
data collection process is needed. Missingness reasons alone do not identify
a statistical missingness mechanism. Unobserved-value-dependent nonresponse
requires additional sensitivity assumptions.

The [mice documentation](https://amices.org/mice/reference/mice.html) describes
the methods and predictor controls used below. `WorkshopHours` can help predict
unrecorded attributes without becoming a clustering feature. The identifier
is neither imputed nor used to predict other variables.

```{r imputation}
eligible <- subset(features$missing, Reason == "Not recorded")
imputation_data <- raters[c("Rater", feature_names, "WorkshopHours")]
where <- matrix(FALSE, nrow(imputation_data), ncol(imputation_data),
                dimnames = dimnames(imputation_data))
where[cbind(match(eligible$ID, imputation_data$Rater),
            match(eligible$Feature, names(imputation_data)))] <- TRUE
methods <- mice::make.method(imputation_data, where = where)
methods[feature_names] <- c("pmm", "pmm", "polyreg", "polr", "logreg")
methods["Rater"] <- ""
predictors <- mice::make.predictorMatrix(imputation_data)
predictors[, "Rater"] <- 0
predictors["Rater", ] <- 0
model <- mice::mice(imputation_data, m = 5, maxit = 5,
  method = methods, predictorMatrix = predictors, where = where,
  seed = 724, printFlag = FALSE)
model$loggedEvents
head(mice::convergence(model))
```

Five imputations and five iterations keep this executable example short; they
are not an adequacy recommendation. Inspect logged events, chain behavior,
observed versus imputed distributions, and substantive plausibility. A lack
of logged events does not establish convergence or a suitable model.
`mfrm_cluster_imputed()` checks consistency of the data, not these assumptions.

## Describe groups and compare choices

Reuse the **same fitted imputation model** for every setting. Changing the
imputations as well as the weights would change two things at once.

```{r grouping}
experience_weights <- setNames(rep(1, length(feature_names)), feature_names)
experience_weights["ExperienceYears"] <- 3
analyses <- list(
  TwoGroups = mfrm_cluster_imputed(features, model, eligible, k = 2),
  ThreeGroups = mfrm_cluster_imputed(features, model, eligible, k = 3),
  ExperienceWeighted = mfrm_cluster_imputed(features, model, eligible, k = 3,
                                          weights = experience_weights)
)
comparison <- mfrm_cluster_compare(analyses)
comparison$analysis_summary
summary(comparison)
weight_effect <- subset(comparison$comparison_summary,
  First == "ThreeGroups" & Second == "ExperienceWeighted")
```

`MeanChangedFraction` describes the fraction of rater pairs whose together/apart
status changes between settings, averaged over the supplied imputations.
The minimum and maximum show its range. The individual rows in
`comparison$comparisons` retain every imputation, including split/join counts
and adjusted Rand indices. An adjusted Rand index of one indicates identical
partitions, allowing arbitrary group numbering. Neither high agreement nor a
large silhouette proves that the groups are meaningful. Silhouettes under
different weights use different distances and should not automatically select
the weights.

In this fictional example, keeping three groups and raising the experience
weight from one to three changes the together/apart status of
`r if (available) sprintf("%.1f%%", 100 * weight_effect$MeanChangedFraction) else "the fraction shown in the table"`
of rater pairs on average across imputations. Thus, the selected weights
materially affect this particular description; presenting a single partition
would conceal that dependence. This percentage concerns pairs, not the
percentage of raters assigned to a different numbered group. It is not evidence
about an actual assessment program.

Also inspect `MinGroupSize`. A rare category can produce a very small group,
and repeated profiles can give high separation values. A high overall
silhouette can coexist with these conditions.

## Compare feature selections

Does including annual rating workload change the background groups? Hold the
raters, group count, method, and imputation model fixed, and remove
`AnnualRatings` from the clustering features. Select the corresponding
eligible cells rather than passing cells for an unselected feature.

```{r feature-selection}
background_names <- setdiff(feature_names, "AnnualRatings")
background_features <- mfrm_features(raters, "Rater", background_names,
  subset(selected_reasons, Feature %in% background_names))
background_only <- mfrm_cluster_imputed(background_features, model,
  subset(eligible, Feature %in% background_names), k = 3)
feature_comparison <- mfrm_cluster_compare(list(
  WithWorkload = analyses$ThreeGroups, WithoutWorkload = background_only))
feature_comparison$analysis_summary
feature_comparison$weights
summary(feature_comparison)
```

Here, removing workload changes the together/apart status of
`r if (available) sprintf("%.1f%%", 100 * feature_comparison$comparison_summary$MeanChangedFraction) else "the fraction shown in the table"`
of rater pairs on average across the five imputations. This measures the
description's dependence on including workload, not whether workload should
be excluded. The `Features` column records each analysis's feature count;
`weights` lists only its selected features. Silhouettes now use different
distances and cannot automatically choose a feature set.

Workload remains in the imputation model, so this does not examine removing
it as a predictor of missing attributes. Different feature selections must
reuse the same fitted `mids` object, and stored completed values are checked
against the corresponding imputation. A selection containing only complete
attributes can use its empty `missing` table as `impute`; it retains a
partition for every completion for comparison purposes.

The included entities must match across all results. With `missing = "omit"`,
removing an incomplete feature may admit additional raters, in which case the
comparison stops. Define and report a common cohort before fitting if that
is the intended question; the API does not silently intersect samples.

## Interpret the retained profiles

Inspect the actual profiles before attaching interpretations to the groups:

```{r interpretation}
one_completion <- analyses$ThreeGroups$analyses[[1]]
one_completion$profiles$numeric
subset(one_completion$profiles$categorical, Feature == "TrainingLevel")
head(one_completion$membership)
analyses$ThreeGroups$co_membership[1:5, 1:5]
```

These profiles describe **one completion**, not an average cluster or a final
assignment. Group numbers can change across completions. The `co_membership`
matrix instead records how often each pair is grouped together across all
imputations. For example, 0.8 means four of the five completions, conditional
on this model and these settings. It is not an 80% probability of a true shared
class. Numeric ranges are recalculated within each completion; their changes
also contribute to sensitivity.

The workflow answers whether proposed background descriptions depend strongly
on the supplied choices. It supplies no automatic preferred group count,
consensus assignment, resampling stability, Rubin-pooled inference, or
imputation of rating responses. Before using groups to make decisions about
raters, establish that the attributes and proposed use address the intended
question and assess the consequences with appropriate evidence.

## Read separation, feature profiles, and imputation sensitivity

The silhouette plot shows which raters fit their assigned group less clearly
under the selected distances. Negative widths warrant examining the rater's
profile and nearby groups; high widths do not establish useful or stable types.
The dashed line is the overall mean. Raters excluded from clustering have no
width; their IDs remain in the returned plot data.

```{r silhouette, fig.width=7, fig.height=5}
plot(one_completion)
```

Inspect one feature at a time so that its units and categories remain clear.
Numeric profiles show means and medians in original units, without confidence
intervals. Categorical profiles show proportions within each group, preserving
the order of training levels. These are descriptions of this completion only.

```{r profiles, fig.width=7, fig.height=5}
plot(one_completion, type = "profile", feature = "ExperienceYears")
plot(one_completion, type = "profile", feature = "TrainingLevel")
```

The next heatmap asks which pairs remain together across imputations. For
readability, this view explicitly selects the first 20 IDs in the input table;
it does not recluster them, change the denominator, or represent the remaining
pairs. Omit `ids` to display all raters. By default, ID labels are hidden above
50 displayed entities; the data are never sampled automatically.

```{r co-membership, fig.width=7, fig.height=6}
plot(analyses$ThreeGroups, ids = raters$Rater[1:20])
```

The colour scale always runs from zero to one. Grey denotes unavailable pairs
involving excluded raters, not pairs that never share a group. Rows retain
the requested ID order. The map does not infer a consensus partition or a
hierarchy, and it does not measure stability under new samples of raters.

For a custom figure, extract the same values without drawing or refitting:

```{r plot-values}
view <- plot(analyses$ThreeGroups, ids = raters$Rater[1:5], draw = FALSE)
plot_data(view)$matrix
```

## Compare PAM with a hierarchy

A hierarchy asks which groups merge as dissimilarity increases. Fit it
separately from PAM: `mfrm_cluster_hierarchical()` uses average linkage
(the mean of all cross-group Gower dissimilarities) or complete linkage
(their maximum). The same feature types, weights, and missingness rules apply.
Choose the group count explicitly; a tree does not automatically identify it.

For multiply imputed features, reuse the same model and eligible cells. This
holds the completed tables fixed while changing the grouping method.

```{r hierarchical-comparison}
average <- mfrm_cluster_imputed(features, model, eligible, k = 3,
  method = "hierarchical", linkage = "average")
complete <- mfrm_cluster_imputed(features, model, eligible, k = 3,
  method = "hierarchical", linkage = "complete")
method_comparison <- mfrm_cluster_compare(list(
  PAM = analyses$ThreeGroups, Average = average, Complete = complete))
method_comparison$analysis_summary
summary(method_comparison)
```

The comparison identifies methods and linkages, alongside group sizes and
separation. As before, pair-change summaries describe sensitivity to the
supplied choices; they do not select a preferred algorithm or establish true
rater types. Substantively different profiles or very small groups deserve
inspection even when the overall pair agreement is high.

```{r dendrogram, fig.width=7, fig.height=5}
plot(average$analyses[[1]])
```

This is the full tree for **one completion**, with boxes marking its three
groups. The 120 raters are all included; ID labels are hidden by default above
50 entities. For a smaller cohort, labels appear automatically. Use
`plot_data(plot(average$analyses[[1]], draw = FALSE))$table` to inspect IDs and
memberships in leaf order. Excluded raters retain unavailable memberships but
have no leaves. No entities are sampled to draw the tree.

Merge heights are linkage dissimilarities, not significance levels or branch
support. Tied distances can produce alternative trees; the group cut uses
merge order, so a unique horizontal cutting height may not exist. Inspect
each completion's profiles with `type = "profile"` or silhouettes with
`type = "silhouette"`. `plot(average)` displays pairwise co-membership across
imputations; it does not combine the trees into a consensus hierarchy.

## How this relates to MFRM rater-type research

Clustering has been used alongside MFRM, with both hierarchical and
nonhierarchical methods. In writing assessment,
[Eckes (2012), *Operational Rater Types in Writing Assessment*](https://doi.org/10.1080/15434303.2011.649381),
clustered MFRM rater-by-criterion bias estimates using hierarchical Ward
clustering, also checking complete and average linkage solutions. In music
performance assessment,
[Wesolowski (2019), *Predicting Operational Rater-Type Classifications Using
Rasch Measurement Theory and Random Forests*](https://doi.org/10.1111/jedm.12227),
reported nonhierarchical k-means clustering of 29 differential rater functioning
indices, followed by prediction of the resulting labels.

Those analyses ask about patterns of rating behaviour. This tutorial instead
groups external backgrounds such as experience and specialty. Its Gower
distances accommodate the selected mix of variable types, and PAM produces
a partition rather than a hierarchy. The separate hierarchical API uses
average or complete linkage. Ward's variance criterion requires appropriate
Euclidean geometry; this Gower interface does not define a Euclidean
conversion and does not offer Ward clustering (see
[Murtagh and Legendre, 2014](https://doi.org/10.1007/s00357-014-9161-z)). The cited
studies do not validate the present attributes, weights, imputation model, or
PAM groups, and do not establish that one method is preferable for all uses.
Grouping estimated MFRM effects would additionally require consideration of
their measurement uncertainty and comparability. The present external-feature
API does not propagate that uncertainty.

## Link separately grouped persons, raters, and tasks

An assessment team may also ask whether different kinds of persons and tasks
are represented across its rater groups. Keep three feature tables, with one
row per person, rater, or task, and a separate rating table containing their
IDs. The tables can have different numbers and types of features. Clustering
the joined rating rows would repeat attributes according to assignment
frequency and answer a different question about rating events.

| Table | Example features | Unit classified |
| --- | --- | --- |
| Persons | Years of language study, preparation hours, instruction mode | One person |
| Raters | Experience, workload, specialty, training, certification | One rater |
| Tasks | Time limit, word limit, response format | One task |

The following extends the fictional example with 240 persons and 12 tasks.
It uses the existing rater partition from **one completed table** only to
illustrate joining by ID. The other two tables have no missing attributes.
The group counts are illustrative choices, not estimated numbers of types.

```{r other-facet-features}
set.seed(7241)
persons <- data.frame(
  Person = sprintf("P%03d", 1:240),
  LanguageStudyYears = sample(0:15, 240, replace = TRUE),
  PreparationHours = round(exp(rnorm(240, log(20), 0.5))),
  InstructionMode = factor(sample(c("Classroom", "Online", "Hybrid"),
                                  240, replace = TRUE)))
tasks <- data.frame(
  Task = sprintf("T%02d", 1:12),
  TimeLimitMinutes = sample(c(10, 20, 30), 12, replace = TRUE),
  WordLimit = sample(c(100, 200, 400), 12, replace = TRUE),
  ResponseFormat = factor(rep(c("Summary", "Argument", "Explanation"), 4)))
person_groups <- mfrm_cluster(mfrm_features(persons, "Person",
  c("LanguageStudyYears", "PreparationHours", "InstructionMode")), k = 3)
task_groups <- mfrm_cluster(mfrm_features(tasks, "Task",
  c("TimeLimitMinutes", "WordLimit", "ResponseFormat")), k = 3)
```

Record the planned assignments before collecting scores. Here each person
receives three distinct tasks and each performance is assigned two distinct
raters. The full Person-by-Rater-by-Task cross-product is not the assignment
roster. Random scores below are placeholders for demonstrating coverage;
they provide no evidence about achievement, rater effects, or model fit.

```{r sparse-rating-roster}
assignments <- do.call(rbind, lapply(persons$Person, function(id) {
  do.call(rbind, lapply(sample(tasks$Task, 3), function(task) {
    data.frame(Person = id, Rater = sample(raters$Rater, 2), Task = task)
  }))
}))
ratings <- assignments
ratings$Score <- sample(1:4, nrow(ratings), replace = TRUE)
ratings$Score[sample(nrow(ratings), 72)] <- NA_integer_
rating_review <- describe_mfrm_data(ratings, person = "Person",
  facets = c("Rater", "Task"), score = "Score",
  rating_min = 1, rating_max = 4, keep_original = TRUE,
  include_agreement = FALSE, expected_design = assignments)
rating_review$structural_missingness$summary
rating_review$design_connectivity
```

There are 1,440 assigned cells out of 345,600 possible combinations, and
72 assigned scores are absent. Assignment coverage is therefore 95%, even
though fewer than 0.5% of all combinations were assigned. Neither a zero
score nor an imputed score is inserted into unassigned cells. With actual
data, supply the original roster: reconstructing it from observed scores
would conceal missing assigned ratings.

Start from the assignment roster so that even an entirely absent rating row
remains in the denominator. This example has one score per Person-Rater-Task
key; check uniqueness before joining. For repeated occasions, include the
occasion in the key. Attach the three independently obtained group labels
by ID with `match()`; row position in a feature table is not an ID.
The check stops this illustration if an ID is unknown or an entity has no
classification. In an analysis allowing omitted entities, report their
unclassified ratings separately rather than silently dropping them.

```{r link-facet-groups}
rating_key <- c("Person", "Rater", "Task")
stopifnot(!anyDuplicated(assignments[rating_key]),
          !anyDuplicated(ratings[rating_key]))
labeled <- merge(assignments, ratings, by = rating_key,
                 all.x = TRUE, sort = FALSE)
labeled$PersonGroup <- person_groups$membership$Cluster[
  match(labeled$Person, person_groups$membership$ID)]
labeled$RaterGroup <- one_completion$membership$Cluster[
  match(labeled$Rater, one_completion$membership$ID)]
labeled$TaskGroup <- task_groups$membership$Cluster[
  match(labeled$Task, task_groups$membership$ID)]
group_columns <- c("PersonGroup", "RaterGroup", "TaskGroup")
stopifnot(!anyNA(labeled[group_columns]))
for (column in group_columns) labeled[[column]] <- factor(labeled[[column]], 1:3)
labeled$Assigned <- 1L
labeled$Observed <- as.integer(!is.na(labeled$Score))
planned_counts <- xtabs(Assigned ~ PersonGroup + RaterGroup + TaskGroup,
                        data = labeled)
observed_counts <- xtabs(Observed ~ PersonGroup + RaterGroup + TaskGroup,
                         data = labeled)
group_coverage <- as.data.frame(planned_counts, responseName = "Assigned")
group_coverage$Observed <- as.vector(observed_counts)
group_coverage$Coverage <- ifelse(group_coverage$Assigned > 0,
  group_coverage$Observed / group_coverage$Assigned, NA_real_)
group_coverage
```

This table answers which group combinations were assigned and how many of
their scores were recorded. A combination with no assignments has unavailable
coverage, not zero coverage. Counts reflect the assignment design; they do
not demonstrate associations between traits or differences in rating quality.
Ratings share persons, raters, and tasks, so an ordinary independent-row
test of group differences would not account for this dependence. Raw score
means also confound person composition, rater severity, and task difficulty.
Such comparisons need an appropriate measurement model and sufficient
overlap, rather than treating the group labels as established latent classes.

If attributes are imputed for several facets, each imputation model must
respect its unit of analysis and relevant dependencies. The current API
analyzes one entity table at a time; it does not fit a joint cross-facet
imputation model or pool group-by-group rating effects. Pairing the first
completion of independently fitted person and rater models does not establish
a valid joint imputation. Nor can numeric group labels be matched across
completions without addressing label changes and membership uncertainty.

## Distinguish sparse assignment from missing attributes

Four situations require different decisions:

| Situation | Example | What to retain or review |
| --- | --- | --- |
| Unassigned rating | A rater was never scheduled to score a performance | Original assignment roster; no invented observation |
| Missing assigned rating | A scheduled score was not submitted | Assigned denominator, absence reason, observed overlap |
| Unrecorded external attribute | Training hours were not recorded | Eligible cells, predictors, imputation assumptions |
| Undefined external attribute | Mentoring years for someone with no mentoring role | Meaning of the attribute; not an ordinary imputation target |

Sparseness describes how much of a design is observed and how those
observations connect. MCAR means that missingness does not depend on the
observed or missing values. MAR allows dependence on observed values, but
not on missing values after conditioning on the observed information. MNAR
allows additional dependence on the missing values. Planned allocation
does not automatically imply MCAR: assignment may depend on recorded
proficiency or on information absent from the analysis. A missingness rate
or diagnostic test alone cannot establish MAR. See
[Rubin (1976), *Inference and Missing Data*](https://doi.org/10.1093/biomet/63.3.581)
for the conditions for ignoring a missingness process in inference.

Inspect both planned and observed connectivity, per-level support, and
coverage within relevant subgroups. Connected Person-facet graphs do not
by themselves establish identification of a complete model, especially with
interactions, or adequate precision. In a simulation of sparse MFRM designs,
[Wind and Ge (2021)](https://doi.org/10.1177/0013164420988108)
found that detecting differential rater functioning depended on the linking
design, not merely on having incomplete ratings. Their findings do not give
a universal acceptable missingness percentage. Imputing external attributes
does not add observed links or resolve confounding between disconnected
rating blocks. Grouping noisy estimated effects from sparse data may also
reflect differences in precision, which this feature API does not model.

## Many features, k-means, and PCA

More columns are not necessarily more useful information. Repeated measures
of one attribute can dominate a distance through their combined weight;
unrelated columns can obscure the differences of interest.
[Steinley and Brusco (2008)](https://doi.org/10.1007/s11336-007-9019-y)
examined variable selection for clustering and found that non-informative
variables and their distributions affected performance. Choose features and
their substantive domains before choosing weights. For example, equal total
weight for a domain containing ten features and one containing two requires
different per-feature weights. This is a substantive choice, not an automatic
correction for correlations. The comparison API supports both positive
feature weights and different selections on the same entities, as illustrated
above. Choosing features to maximize an attractive grouping still requires
separate validation; the comparison does not automate feature selection.

PCA and k-means answer different questions. PCA represents variation in
numeric features with orthogonal continuous components. K-means partitions
entities to minimize within-group squared Euclidean distances to centroids.
[Ding and He (2004), *Principal Component Analysis and Effective K-means
Clustering*](https://doi.org/10.1137/1.9781611972740.54)
describe their relationship through a continuous relaxation of discrete
cluster membership. This does not make ordinary PCA and discrete clustering
equivalent, or establish that the first two PCs recover useful groups.

For a fixed numeric input, retaining all nonzero PC scores without further
rescaling preserves its Euclidean distances. Truncating components changes
those distances and may remove low-variance separation as well as noise.
Standardizing features before PCA, or whitening the scores afterwards,
also changes the chosen geometry. An explained-variance threshold alone is
not evidence that a clustering structure was preserved. A two-component
plot is a projection, not a test for the existence of clusters.

Ordinary PCA and k-means should not receive arbitrary numeric codes for
nominal categories such as specialty. The present mixed-feature API uses
Gower distances with PAM or average/complete linkage; it does not implement
PCA, k-means, or automatic feature selection. Base R provides
[`prcomp()`](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/prcomp.html)
and [`kmeans()`](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/kmeans.html)
for separately specified numeric analyses. A future integrated route needs
explicit scaling, retained dimensions, initialization, and uncertainty
handling. Simply sending PC scores to the current Gower interface would
range-scale them again and would not reproduce Euclidean k-means.

## Plan for the size of the analysis

Pairwise distances require memory that grows quadratically with the number
of raters. Ordinary clustering permits up to 5,000 included entities; the
imputation workflow permits up to 5,000 total entities because its comparison
matrix includes excluded IDs. These are input limits, not a guarantee of low
memory use or acceptable run time. Multiple imputations also retain every
completed analysis. Save the fitted imputation model and clustering results;
`mfrm_cluster_compare()` reuses them without repeating the fits.

Plan for temporary distance and silhouette calculations as well as the saved
result. For scale, a run with 5,000 entities and eight mixed-type features,
fitting two hierarchical configurations and comparing them, used about
1.5 GiB of peak process memory on macOS arm64 with R 4.6.1. The returned
comparison occupied about 3.3 MiB. Average and complete linkage gave similar
memory requirements in this example. These single-run measurements include
R startup and data preparation; they are not a memory bound or a prediction
for other machines, features, or imputation counts. The 5,000-entity imputation
workflow was not evaluated in that example.

Increasing feature count also increases distance calculations and stored
profiles. A separate run with 1,000 entities and 1,000 mixed-type features,
again fitting and comparing two configurations, took about six seconds per
method and used approximately 400--430 MiB of peak process memory on the
same platform, for PAM, average linkage, and complete linkage. This was a
complete-feature workload without planted groups. It does not establish
useful classification, joint performance at the largest entity and feature
counts, or feasibility of a 1,000-feature imputation model. A large imputation
model requires its own predictor selection and checks for information,
dependence, and model fit.
