---
title: "mfrmr Visual Diagnostics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{mfrmr Visual Diagnostics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
is_cran_check <- !isTRUE(as.logical(Sys.getenv("NOT_CRAN", "false")))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  eval = !is_cran_check
)
```

This vignette connects figures to questions about fitted ratings, external
attributes and observed-score planning. It is organized around six practical
questions:

- How well do persons, facet levels, and categories target each other?
- Which observations or levels look locally unstable?
- Is the design linked well enough across subsets or forms?
- Where do residual structure and interaction screens point next?
- How can external attributes describe groups of raters?
- How could changing the number of tasks affect an observed-score composite?

Model examples use packaged data; the feature hierarchy uses eight fictional
raters. The `"publication"` and `"monochrome"` presets illustrate reusable
styles, including views that remain interpretable without colour.

If you are selecting figures for a report, use `reporting_checklist()` before
or alongside this vignette. Its `"Visual Displays"` rows now mirror the
public plotting family shown here.

## Find a figure by purpose

Use the plot guide before choosing a specialized function. It includes rater
feedback, category curves, external-feature groups and G/D-study planning.

Open a preview for the full figure, interpretation and the values used to draw
it. These are different questions: a severe rater need not misfit; feature
groups describe backgrounds; a D-study plans an observed-score measurement.

<style>
.mfrmr-gallery { display: grid; grid-template-columns: repeat(auto-fit, minmax(min(100%, 16rem), 1fr)); gap: 1rem; margin: 1.5rem 0; }
.mfrmr-gallery-card { border: 1px solid #737373; border-radius: .4rem; padding: .85rem; background: #fff; color: #202020; min-width: 0; }
.mfrmr-gallery-card a { display: block; color: #164f78; text-decoration: underline; text-underline-offset: .15em; }
.mfrmr-gallery-card a:focus-visible { outline: 3px solid #164f78; outline-offset: 4px; }
.mfrmr-gallery-card img { display: block; width: 100%; height: auto; margin: 0 0 .65rem; }
.mfrmr-gallery-card p:last-child { margin-bottom: 0; }
@media print { .mfrmr-gallery { display: block; } .mfrmr-gallery-card { break-inside: avoid; margin-bottom: 1rem; } }
</style>

```{r plot-gallery, eval = TRUE, echo = FALSE, results = 'asis'}
figures <- mfrmr::mfrmr_output_guide("plots")
# Only presentation locations live here; questions and support come from the guide.
gallery <- data.frame(
  PlotName = c("wright_map", "fit_pathway", "category_characteristic_curves",
    "subset_connectivity", "cluster_dendrogram", "multivariate_d_study"),
  Chunk = c("wright", "fit-pathway", "category-curves", "linking",
    "feature-hierarchy", "d-study-planning"),
  Preview = c("Wright map", "Infit versus location", "Category probability curves",
    "Subset coverage", "Feature dendrogram", "D-study coefficient curves"))
rows <- figures[match(gallery$PlotName, figures$PlotName), ]
stopifnot(!anyNA(rows$Question))
support <- c(dedicated = "Base R and dedicated ggplot conversion",
  native = "Returns ggplot directly",
  generic = "Base R; ggplot offers a different, table-based view",
  unavailable = "Base R; custom graphics from the saved data")
cat('::: {.mfrmr-gallery}

')
for (i in seq_len(nrow(gallery))) {
  preview <- if (!is_cran_check) paste0("!", "[", gallery$Preview[i],
    " preview]", "(", knitr::fig_chunk(gallery$Chunk[i], "png", number = 1),
    "){loading=\"lazy\"} ") else ""
  cat(sprintf('::: {.mfrmr-gallery-card}

[%s**%s**](#figure-%s)

%s

:::

',
    preview, rows$Question[i], gallery$Chunk[i], support[rows$GGPlot[i]]))
}
cat(':::
')
```

<details>
<summary>Browse all selected plot routes</summary>

```{r plot-guide, eval = TRUE}
figures <- mfrmr::mfrmr_output_guide("plots")
knitr::kable(figures[, c("Question", "ResultFunction", "GGPlot")])
```

</details>

Open the help for `ResultFunction` if you have not created the required result.
Choose a row and read `PlotCall`, `GGPlotCall`, `DataComponent` and `Notes`.
Replace `x` in `PlotCall` with the indicated result, and save its return value
as `p`. The call uses `draw = FALSE` to prepare the figure without opening a
window. Change it to `TRUE` to display it, or follow `GGPlotCall` to customize
it with ggplot2. A data component is a table to inspect; passing that component
to `as_ggplot()` may produce a different view from the full figure.

`dedicated` means there is a converter for the figure; `native` means the plot
already returns ggplot. `generic` describes a table-based graphic that need
not reproduce the original layout or annotations. `unavailable` still allows
the original plot and custom graphics from `plot_data(p)`. This is a selected
map, not a claim that every plot or component has been covered. Graphical
support does not extend the model's statistical assumptions.

## Minimal setup

```{r setup}
library(mfrmr)

toy <- load_mfrmr_data("example_operational")

fit <- fit_mfrm(
  toy,
  person = "Person",
  facets = c("Rater", "Criterion"),
  score = "Score",
  method = "MML",
  model = "RSM"
)

diag <- diagnose_mfrm(fit, residual_pca = "none")
checklist <- reporting_checklist(fit, diagnostics = diag)
subset(
  checklist$checklist,
  Section == "Visual Displays",
  c("Item", "Available", "NextAction")
)
```

## 1. Targeting and scale structure

### Wright map {#figure-wright}

Use the Wright map first when you want one shared logit view of persons,
facet levels, and step thresholds.

```{r wright, dev="png", fig.alt = "Wright map of person abilities, facet levels and category steps on the fitted logit scale, with requested uncertainty intervals. Higher facet locations indicate stricter scoring or greater difficulty."}
wright <- plot(fit, type = "wright", preset = "publication", show_ci = TRUE)
```

Interpretation:

- Compare person density on the left to facet and step locations on the right.
- Large gaps suggest weaker targeting in that logit region.
- Wide overlap in marginal confidence whiskers suggests imprecision; estimate
  the relevant pairwise contrast directly before claiming that two levels are
  separated or indistinguishable.

The locations used in the figure remain available as a table:

```{r wright-values}
head(plot_data(wright, component = "locations"))
```

The native view above remains the recommended analytic figure because it keeps
facet uncertainty and fitted step locations visible. For a closer FACETS-facing handoff, switch the
renderer and leave `show_ci` at its default `FALSE`. The fitted coordinates do
not change.

Every retained native facet level is labelled. Only label coordinates are
displaced for collision avoidance, with leader lines back to the fitted
points. Step thresholds form one vertical ladder and their labels include the
fitted transition logits.

```{r wright-facets-style, fig.alt = "Text-style Wright display with rubric labels Beginning, Developing, Secure and Advanced. Locations share a fitted logit scale; labels describe ordered categories rather than equal observed-score intervals."}
plot(
  fit,
  type = "wright",
  renderer = "facets",
  category_labels = c(
    `1` = "Beginning", `2` = "Developing", `3` = "Secure", `4` = "Advanced"
  )
)
```

This renderer uses one common logit ruler, a `*` person-frequency column,
signed facet headers, every facet level, and short labeled score-transition
lines. It is FACETS-style visual correspondence, not a claim that `mfrmr` and FACETS
produce numerically identical estimates. A numerical comparison requires output
from a documented FACETS version and aligned estimator, identification, score,
and orientation settings. Set `show_ci = TRUE` only when
you deliberately want a hybrid FACETS-style ruler with the mfrmr uncertainty
extension. Use `draw = FALSE` and inspect the `facets_style` tables when
rebuilding the display with ggplot2 or another graphics system.

Next, use the pathway map when you want to see how expected scores progress
across theta.

```{r pathway, fig.alt = "Pathway display of fitted response probabilities across ability, separated by display layer. It shows model predictions rather than tests of rater bias."}
plot(fit, type = "pathway", preset = "publication")
```

Interpretation:

- Steeper rises indicate stronger score progression.
- Dominant-category strips show where each category is most likely to govern the score.
- Flat or compressed regions suggest weaker category separation.

### Fit pathway {#figure-fit-pathway}

The expected-score pathway is not a fit pathway. To review measure against
Infit, place Infit on the horizontal axis and include person rows explicitly:

```{r fit-pathway, dev="png", fig.alt = "Infit mean squares against fitted locations for persons and facet levels, with requested location intervals. Mean-square bands are descriptive review thresholds, not automatic decisions."}
fit_pathway <- plot(
  fit,
  type = "fit_pathway",
  diagnostics = diag,
  fit_stat = "Infit",
  fit_scale = "mnsq",
  include_person = TRUE,
  show_ci = TRUE,
  preset = "publication"
)
```

Interpretation:

- The vertical axis remains the fitted measure in logits.
- The horizontal axis is Infit MnSq; the 1.0 line is the model-expectation reference.
- Vertical whiskers show measure uncertainty; person and non-person rows have distinct uncertainty bases, recorded in the draw-free payload metadata.
- Treat displaced or flagged rows as review prompts, not automatic exclusions.

```{r fit-pathway-values}
head(plot_data(fit_pathway, component = "table"))
```

### Category probabilities {#figure-category-curves}

Category curves ask which ordered score is most probable at each ability.
They complement the expected-score pathway: a single expected score can hide
which categories contribute to it.

```{r category-curves, dev="png", fig.alt="Predicted category probabilities across ability for the fitted rating scale. Line patterns and colours identify categories; the curves use a reference profile with other additive facet effects fixed at zero."}
category_curves <- plot(fit, type = "ccc", preset = "monochrome")
```

These are reference-profile probabilities, not observed category frequencies or
uncertainty bands. For PCM/GPCM, select and interpret the relevant step/slope
owner. A rarely dominant category is a reason to inspect counts and rubric
meaning, not automatic evidence that categories should be merged.

```{r category-values}
head(plot_data(category_curves, component = "probabilities"))
# Optional customization without refitting:
# as_ggplot(category_curves)
```

## 2. Local response and level issues

Unexpected-response screening is useful for case-level review.

```{r unexpected, fig.alt = "Observed ratings screened by standardized residual size and fitted response probability. Highlighted observations merit contextual review rather than automatic deletion."}
plot_unexpected(
  fit,
  diagnostics = diag,
  abs_z_min = 1.5,
  prob_max = 0.4,
  plot_type = "scatter",
  preset = "publication"
)
```

Interpretation:

- Upper corners combine large residual mismatch with low model probability.
- Repeated appearances of the same persons or levels are more informative than a single extreme point.

Displacement focuses on level movement rather than individual responses.

```{r displacement, fig.alt = "Level-specific displacement summaries for the fitted facets, including unanchored levels. The displayed shifts are diagnostic summaries, not evidence that a level changed over time."}
plot_displacement(
  fit,
  diagnostics = diag,
  anchored_only = FALSE,
  plot_type = "lollipop",
  preset = "publication"
)
```

Interpretation:

- Large absolute displacement indicates stronger tension between observed data and current calibration.
- For anchored runs, this is especially useful as an anchor-robustness screen.

### Strict marginal follow-up

When you need the package's latent-integrated follow-up path, switch to `MML`
and request `diagnostic_mode = "both"` so the legacy and strict branches stay
visible side by side. The chunk below uses compact quadrature for a shorter
runtime; final reporting should be based on a refit with the package default or
a higher quadrature setting.

```{r strict-marginal, fig.alt = "Marginal fit summaries using posterior-averaged response probabilities. These diagnostics have a different probability basis from conditional residual screens."}
fit_strict <- fit_mfrm(
  toy,
  person = "Person",
  facets = c("Rater", "Criterion"),
  score = "Score",
  method = "MML",
  model = "RSM",
  quad_points = 7,
  maxit = 40
)

diag_strict <- diagnose_mfrm(
  fit_strict,
  residual_pca = "none",
  diagnostic_mode = "both"
)

strict_checklist <- reporting_checklist(fit_strict, diagnostics = diag_strict)
subset(
  strict_checklist$checklist,
  Section == "Visual Displays" &
    Item %in% c("QC / facet dashboard", "Strict marginal visuals"),
  c("Item", "Available", "NextAction")
)

plot_marginal_fit(
  diag_strict,
  top_n = 12,
  preset = "publication"
)
```

Interpretation:

- Treat strict marginal plots as exploratory corroboration screens, not as standalone inferential tests.
- Use the checklist rows to confirm that the current run actually supports the strict branch before routing figures into a report.
- When pairwise follow-up is needed, continue with `plot_marginal_pairwise(diag_strict, preset = "publication")`.

## 3. Linking and coverage {#figure-linking}

When the design may be incomplete or spread across subsets, inspect the
coverage matrix before interpreting cross-subset contrasts.

```{r linking, dev="png", fig.alt = "Observation shares above a matrix of relative facet-level coverage across observed subsets. Matrix cells compare each subset with the largest observed count for that facet; they are not completion rates for planned assignments."}
sc <- subset_connectivity_report(fit, diagnostics = diag)
coverage <- plot(sc, type = "design_matrix", preset = "publication")
```

The cells divide each subset's observed number of facet levels by the largest
such count for that facet among the observed subsets. They do not show the
proportion of planned ratings that were completed. A single connected subset
can therefore have coverage values of one even with a sparse rating design.
Low relative counts invite a closer assignment review; they do not alone
establish weak linking, informative missingness, or random missingness.

```{r coverage-values}
plot_data(coverage, component = "matrix")
head(plot_data(coverage, component = "table"))
# Only the matrix is converted; the observation-share panel is not included.
# as_ggplot(coverage, component = "matrix")
```

Compare an explicit assignment roster with observed ratings when distinguishing
unassigned cells from missing assigned scores. Neither should be coded as zero.

Keep three uses of *network* separate. `mfrm_network_analysis()` analyzes the
assignment/co-observation graph and can expose disconnected measurement
subsets. `rater_network_analysis()` analyzes pairwise score relations, and
`rater_halo_network_analysis()` analyzes rater-by-criterion score-profile
correlations. The latter two do not establish assignment connectedness or a
common scale. Their centrality statistics are graph-theoretic quantities, not
rating-scale central tendency, MFRM severity logits, or causal halo evidence.

If you are working across administrations, follow up with anchor-drift plots:

```{r eval = FALSE}
drift <- detect_anchor_drift(current_fit, baseline = baseline_anchors)
plot_anchor_drift(drift, type = "heatmap", preset = "publication")
```

## 4. Residual structure and interaction screens

Residual PCA is a follow-up layer after the main fit screen.

```{r residual-pca, fig.alt = "Scree display of residual principal components when the requested decomposition is available. Residual components are diagnostic summaries, not automatically substantive dimensions."}
diag_pca <- diagnose_mfrm(fit, residual_pca = "both", pca_max_factors = 4)
pca <- analyze_residual_pca(diag_pca, mode = "both")
summary(pca)
if (nrow(pca$overall_table) > 0L) {
  plot_residual_pca(pca, mode = "overall", plot_type = "scree", preset = "publication")
}
```

The incomplete rating design in this example leaves some pairs without
enough shared Persons to compute their residual correlation, so the overall
PCA is unavailable. The summary retains the reason and any available
facet-specific results. Inspect shared-Person counts and residual variation;
do not replace missing correlations with zero or interpret the absence of a
scree plot as evidence for one dimension. A facet-specific PCA describes its
own residual matrix and cannot replace the unavailable overall analysis.

Interpretation:

- Early components with noticeably larger eigenvalues deserve follow-up.
- Scree review should usually be paired with loading review for the component of interest.

For interaction screening, use the packaged bias example.

```{r bias, fig.alt = "Rater-by-criterion interaction profiles from the fitted bias analysis. Differences identify combinations for review and do not establish a causal explanation."}
bias_df <- load_mfrmr_data("example_bias")

fit_bias <- fit_mfrm(
  bias_df,
  person = "Person",
  facets = c("Rater", "Criterion"),
  score = "Score",
  method = "MML",
  model = "RSM",
  quad_points = 7
)

diag_bias <- diagnose_mfrm(fit_bias, residual_pca = "none")
bias <- estimate_bias(fit_bias, diag_bias, facet_a = "Rater", facet_b = "Criterion")

plot_bias_interaction(
  bias,
  plot = "facet_profile",
  preset = "publication"
)
```

Interpretation:

- Facet profiles are useful for seeing whether a small number of levels drives most flagged interaction cells.
- Treat these plots as screening evidence; confirm with the corresponding tables and narrative reports.

## 5. Custom figures without losing the evidence boundary

The built-in plots are intended as safe defaults. Use
`preset = "monochrome"` when a journal, accessibility review, or print workflow
needs grayscale output. For journal figures, teaching material, dashboards, or
lab-specific styles, use `draw = FALSE` and the plot-data accessors instead of
editing screenshots.

```{r custom-plot-data, fig.alt = "Monochrome Wright map of person, facet and step locations on the common logit scale. The accompanying code extracts the plotted locations for custom displays."}
plot(fit, type = "wright", preset = "monochrome")

wright_payload <- plot(fit, type = "wright", draw = FALSE, preset = "publication")
plot_data_components(wright_payload)

locations <- plot_data(wright_payload, component = "locations")
head(locations)

pathway_long <- plot_data(
  fit,
  type = "pathway",
  component = "pathway_long",
  preset = "publication"
)
head(pathway_long[, c("Layer", "CurveGroup", "Theta", "Value")])
```

When you build a custom figure, keep the helper's guidance tables with the
plot data:

```{r custom-guidance}
names(wright_payload$data)
wright_payload$data$reference_lines
```

Those metadata are the guardrails for captions and interpretation. They let
you change colors, labels, panels, or rendering technology while preserving the
same measurement scale, reference lines, caveats, and reporting role used by
the package-native plot.

### Set an appearance once for several plots

Use `options(mfrmr.plot_preset = "monochrome")` to set a common default in the
development version. This example restores the previous option when it ends:

```{r session-preset}
local({
  old <- options(mfrmr.plot_preset = "monochrome")
  on.exit(options(old))
  saved <- plot(fit, type = "ccc", draw = FALSE)
  special <- plot(fit, type = "ccc", preset = "publication", draw = FALSE)
  c(session_default = saved$data$preset, explicit_override = special$data$preset)
})
```

The choices are `"standard"`, `"publication"`, `"compact"` and `"monochrome"`.
Explicit `preset` wins over the option; omitting both retains `"standard"`.
Explicit `preset = NULL` also selects the package default. Remove the session
setting with `options(mfrmr.plot_preset = NULL)`.

A supported `as_ggplot(saved)` conversion uses the saved preset, whereas a
new `plot(fit, ...)` call uses the current option. Set the option in a script
or use explicit arguments to reproduce its appearance in another session.
This applies to the common preset controls, including category, coverage and
network report plots. Extended-model plots with separate `palette` controls
retain their own settings. It does not change estimates, confidence levels,
screening thresholds or the user's global ggplot theme.

### Use a common title argument

In the development version, eight existing helpers accept `title`:
`plot_marginal_fit()`, `plot_marginal_pairwise()`, `plot_unexpected()`,
`plot_interrater_agreement()`, `plot_facets_chisq()`, `plot_bubble()`,
`plot_bias_interaction()` and `plot_facet_quality_dashboard()`.

```{r title-controls, fig.alt="Infit and Outfit mean squares for fitted facet levels, with the title omitted. Bubble sizes reflect rating counts; reference lines and the saved descriptive review band remain unchanged."}
plain <- plot_bubble(fit, diagnostics = diag, view = "infit_outfit",
                     title = NULL, preset = "monochrome", draw = FALSE)
as_ggplot(plain)
# Reuse the saved values, screening references and interval settings.
plain$data$reference_lines
```

Omit `title` for the default heading, supply `title = "Scoring patterns"` to
replace it, or use `title = NULL` (or `""`) to hide it. This does not remove
subtitles, legends, screening settings or interpretation notes. The dashboard
method also accepts `plot(dashboard, title = NULL)`. Keep review-only
limitations in the figure caption or accompanying report if its heading is
hidden.

Existing `main` calls and positional arguments still work without deprecation
warnings. `main = NULL` retains its earlier meaning of "use the default
heading". Do not supply both `main` and `title`, even with the same value.
These aliases do not imply that every plot has an `as_ggplot()` converter;
consult `mfrmr_output_guide("plots")` for the supported route.

### Fair Scores and annotations outside the figure

Use `plot_type = "measure"` to relate measures to fair scores, `"scatter"`
to compare observed and fair scores, or `"difference"` to rank their gaps.
FairM uses mean reference measures; FairZ uses zero reference measures.
FairZ is an expected score on the internal score scale, not a z-score.
Observed-minus-fair gaps also reflect assignment and person mix, so they do
not by themselves establish rater bias.

```{r fair-score-data, eval=FALSE}
p_fair <- plot_fair_average(
  fit, diagnostics = diag, facet = "Rater", metric = "FairZ",
  plot_type = "measure", show_ci = TRUE, preset = "monochrome",
  show_title = FALSE, show_notes = FALSE, draw = FALSE
)
p_fair$data$notes
p_fair$data$plot_data
p_fair$data$excluded
# Requires ggplot2:
g_fair <- as_ggplot(p_fair)
g_fair
attr(g_fair, "mfrmr_notes")
```

Use `draw = TRUE` to draw the base-R version. Titles and notes remain in the
returned object when hidden in the figure. The same annotation controls apply
to Wright, expected-score pathway, and CCC plots; they do not change readiness
or interval eligibility. Keep the applicable notes in the surrounding caption
or report. Other plotting helpers have their own documented options.

Fair-score intervals remain diagnostic-only (`CI_Eligible = FALSE`), with
full-refit coverage unverified. RSM/PCM plots propagate only the focal measure
SE, holding thresholds and reference measures fixed; they require a fitted
model, not just a stored fair-average bundle. GPCM-MML plots use available
structural delta-method SEs for non-Person rows, conditioning on Person
EAP/reference means. The table option `fair_se = TRUE` supplies the GPCM route,
not RSM/PCM fair-score SEs. Historical table `SE` columns describe measures.
Difference-view whiskers hold observed means fixed and are not confidence
intervals for the observed-minus-fair gap.

Expected-score and CCC curves use six line types: solid, dashed, dotted,
dotdash, longdash, and twodash. Fair-score plots use six point shapes. These
encodings repeat for larger sets; select facets or use panels to keep a
grayscale figure readable. Check labels and line separation at the intended
output size rather than relying on color or gray levels alone.

## 6. Secondary visual layer

The package ships a complementary visual layer for teaching and
diagnostic follow-up. These helpers are not default reporting figures;
use them after the main screens above.

- `plot_guttman_scalogram(fit, diagnostics)` renders a person x
  facet-level response matrix with an unexpected-response overlay, for
  teaching-oriented scalogram intuition and local triage.
- `plot_residual_qq(fit, diagnostics)` plots a Normal Q-Q of
  person-level standardized residual aggregates as exploratory
  follow-up on residual tail behavior.
- `plot_rater_trajectory(list(T1 = fit_a, T2 = fit_b))` tracks rater
  severity across named waves. The helper does not perform linking;
  supply waves that have already been placed on a common anchored
  scale (see `vignette("mfrmr-linking-and-dff")`) before interpreting
  movement as rater drift.
- `plot_rater_agreement_heatmap(fit, diagnostics)` renders a
  compact pairwise rater x rater agreement matrix; pass
  `metric = "correlation"` to colour by the Pearson-style `Corr`
  column instead of exact agreement.
- `response_time_review(data, person, facets, time)` summarizes
  response-time metadata by person, facet, and score category. Pair it with
  `plot_response_time_review()` for distribution and grouped timing plots.
  This is a descriptive QC layer, not a joint speed-accuracy model.
- `plot_shrinkage_funnel(fit_eb, show_ci = TRUE)` draws raw and
  empirical-Bayes shrunken facet estimates on the same row, with
  optional confidence whiskers for both estimates. Use this only after
  `apply_empirical_bayes_shrinkage()` or `fit_mfrm(..., facet_shrinkage =
  "empirical_bayes")`.

### Response-time QC context

If your rating-event data include response times, review them separately from
the MFRM likelihood. Rapid and slow response-time flags are descriptive
quality-control prompts; they do not change measures and should not be treated
as proof of disengagement, cheating, or speededness.

```{r response-time-review, fig.alt = c("Distribution of synthetic response times with selected lower and upper quantile screens. These thresholds describe this example and do not diagnose careless responding.", "Person-level summaries of synthetic response times. Compare flagged patterns with their rating counts and context before drawing substantive conclusions.")}
toy_rt <- toy
toy_rt$ResponseTime <- 12 + (seq_len(nrow(toy_rt)) %% 7) +
  as.numeric(toy_rt$Score)
toy_rt$ResponseTime[1] <- 2
toy_rt$ResponseTime[2] <- 38

rt <- response_time_review(
  toy_rt,
  person = "Person",
  facets = c("Rater", "Criterion"),
  score = "Score",
  time = "ResponseTime",
  rapid_quantile = 0.10,
  slow_quantile = 0.90
)

summary(rt)
plot_response_time_review(rt, type = "distribution", preset = "publication")
plot_response_time_review(rt, type = "person", preset = "publication")
```

Interpretation:

- Start with the distribution plot to see whether the rapid/slow thresholds are sensible for this administration.
- Inspect person and facet summaries for concentrated rapid or slow rates rather than isolated events.
- Keep timing flags separate from fit, bias, and validity claims unless the study design explicitly supports stronger speed-accuracy modeling.

### Small-N shrinkage with uncertainty

When a non-person facet has few levels or sparse observations, a large raw
severity estimate can be a noisy estimate rather than a stable facet signal.
The shrinkage funnel shows how far empirical-Bayes pooling moved each level
toward the facet mean and whether the uncertainty remains wide after pooling.

```{r shrinkage-funnel, fig.alt = "Raw and empirically shrunken facet estimates with their displayed uncertainty. Movement toward a facet center reflects shrinkage, not a change in observed ratings."}
fit_eb <- apply_empirical_bayes_shrinkage(fit)

shrink <- plot_shrinkage_funnel(
  fit_eb,
  show_ci = TRUE,
  ci_level = 0.95,
  preset = "publication",
  draw = FALSE
)

head(shrink$data$table[, c(
  "Facet", "Level", "RawEstimate", "RawCI_Lower", "RawCI_Upper",
  "ShrunkEstimate", "ShrunkCI_Lower", "ShrunkCI_Upper",
  "ShrinkageFactor"
)])

plot_shrinkage_funnel(
  fit_eb,
  show_ci = TRUE,
  ci_level = 0.95,
  preset = "publication"
)
```

Interpretation:

- Long raw-to-shrunken segments identify levels most affected by the partial-pooling prior.
- Wide raw whiskers that narrow after pooling indicate estimation instability, not automatic rater-quality failure.
- Report the shrinkage method and keep this display separate from bias, fit, or validity claims.

## 7. Attributes and assessment planning

### External-feature hierarchy {#figure-feature-hierarchy}

A hierarchy can summarize raters' backgrounds before discussing training.
These eight fictional raters illustrate the mechanics; their groups are not
estimates of severity, ability or rater quality. Choose attributes and their
coding for the substantive question before interpreting a tree.

```{r feature-input}
backgrounds <- data.frame(
  Rater = paste0("R", 1:8),
  Years = c(1, 2, 4, 6, 8, 10, 12, 15),
  Training = ordered(c("Basic", "Basic", "Advanced", "Basic",
    "Advanced", "Advanced", "Specialist", "Specialist"),
    levels = c("Basic", "Advanced", "Specialist")))
features <- mfrm_features(backgrounds, "Rater", c("Years", "Training"))
hierarchy <- mfrm_cluster_hierarchical(features, k = 3, linkage = "average")
```

```{r feature-hierarchy, dev="png", fig.alt="Average-linkage dendrogram for eight fictional raters, based on Gower dissimilarities in experience and ordered training level. Labelled leaves identify raters and boxes show the requested three groups; heights are dissimilarities, not significance levels."}
hierarchy_plot <- plot(hierarchy, preset = "monochrome")
```

Three groups were requested; the diagram does not select that number or
establish that the groups are real. Feature weights, coding and linkage can
change the branches. The full tree and memberships in leaf order are available
for inspection. The development version supports both the base plot and a
dedicated ggplot conversion of this saved tree.

```{r hierarchy-values}
plot_data(hierarchy_plot, component = "table")
tree <- plot_data(hierarchy_plot, component = "tree")
```

```{r hierarchy-ggplot, fig.width=8, fig.height=6, fig.alt="The same saved hierarchy of eight fictional raters rendered with ggplot. Leaf order, Gower merge heights and the three group boxes are unchanged. The caption explains that heights are not branch support."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  hierarchy_figure <- as_ggplot(hierarchy_plot, component = "tree") +
    ggplot2::labs(title = NULL, subtitle = NULL)
  print(hierarchy_figure)
}
```

The conversion keeps the saved label setting and preset. Use `labels = FALSE`
when creating a crowded tree to hide IDs without removing leaves. Where
heights tie, boxes follow the stored merge-order groups; they need not
correspond to a unique horizontal height cut. Removing titles or captions
with `ggplot2::labs()` leaves the tree and memberships in `plot_data()`.

For imputed features, group comparisons and PCA, continue with
[Exploring person, rater, and task attributes](mfrmr-external-features.html).

### D-study planning {#figure-d-study-planning}

A D-study asks how a specified measurement design would perform if we changed
facet counts. The packaged MGENOVA manual example has two scores, V and W,
observed for Persons crossed with Tasks. Here the target is their equal-weight
composite. There is no rater facet in this example.

```{r d-study-input}
tasks <- read.csv(system.file("extdata", "mgenova-table12.csv", package = "mfrmr"))
gstudy <- mfrm_multivariate_gstudy(tasks, c("V", "W"), rater = NULL)
planned <- mfrm_multivariate_d_study(gstudy,
  data.frame(Tasks = c(3, 6, 12)), weights = c(V = 0.5, W = 0.5))
```

```{r d-study-planning, dev="png", fig.alt="G and Phi coefficients for an equally weighted V and W composite at three, six and twelve planned Tasks. Separate panels distinguish relative and absolute decisions; points identify requested designs, and lines guide the eye."}
planning_plot <- plot(planned, type = "coefficients", preset = "monochrome")
```

G concerns relative comparisons among Persons; Phi concerns absolute score
interpretation. More tasks reduce the modeled task-related error while holding
the estimated covariance components and composite weights fixed. These are
observed-score projections, not MFRM ability reliability or a guarantee for a
new task population. Uncertainty in the estimated components is omitted.

```{r planning-values}
plot_data(planning_plot, component = "table")
# Customizable conversion of this selected score/composite:
# as_ggplot(planning_plot)
# Error in score units is a different view:
# plot(planned, type = "sem")
```

Use `?mfrm_multivariate_d_study` for score/composite selection and supported
crossed or nested designs. An ordinary `mfrm_d_study` object has its own base
plot; its automatic ggplot conversion is currently unavailable.

## Portable score-batch review

A batch returned by `score_mfrm_calibration()` is not a fitted-model diagnostic
object. Review it through its own methods:

```{r eval = FALSE}
summary(scores)
plot(scores, type = "interval", preset = "publication")
plot(scores, type = "precision", preset = "publication")
plot(scores, type = "edge_mass", preset = "publication")
```

The interval view highlights `scored_review` Persons. The precision view shows
valid response rows against posterior SD, and the edge-mass view compares
outer-node posterior mass with the artifact's recorded review threshold.
Persons with no valid responses remain in the summary and draw-free plot
payload rather than being assigned an artificial coordinate. These displays
do not replace source-fit diagnostics, and their uncertainty is conditional on
the frozen point calibration. See
`vignette("mfrmr-portable-calibration", package = "mfrmr")` for the complete
fresh-session workflow.

## Recommended sequence

For a compact visual workflow:

1. `reporting_checklist()` when you want the package to route which figures are already supported.
2. `plot_qc_dashboard()` for one-page triage.
3. `plot_unexpected()`, `plot_displacement()`, `plot_marginal_fit()`, and `plot_interrater_agreement()` for local follow-up.
4. `plot(fit, type = "wright")` and `plot(fit, type = "pathway")` for targeting and scale interpretation.
5. `plot_residual_pca()`, `plot_bias_interaction()`, and `plot_information()` for deeper structural review.
6. `response_time_review()` and `plot_response_time_review()` when response-time metadata are available.
7. `plot_shrinkage_funnel(show_ci = TRUE)` when empirical-Bayes shrinkage was applied.
8. `summary(scores)` and `plot(scores)` for a separately created portable
   score batch; do not use these as source-fit diagnostics.
9. `plot_guttman_scalogram()`, `plot_residual_qq()`,
   `plot_rater_trajectory()`, and `plot_rater_agreement_heatmap()` as
   the teaching / drift / agreement-heatmap follow-up layer.

## Related help

- `help("mfrmr_visual_diagnostics", package = "mfrmr")`
- `help("mfrmr_workflow_methods", package = "mfrmr")`
- `mfrmr_interval_guide("shrinkage")`
