---
title: "Portable calibration and fresh-session scoring"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Portable calibration and fresh-session scoring}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
is_cran_check <- !isTRUE(as.logical(Sys.getenv("NOT_CRAN", "false")))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  eval = !is_cran_check
)
options(width = 78)
```

This workflow creates a portable calibration from an eligible fitted model,
validates and freezes it, saves it, and then scores new Persons without the
source fit or training responses. The scoring phase uses only the saved
artifact and the new response rows.

The portable workflow is deliberately narrower than fitted-object scoring.
It supports one observed score scale and one latent dimension for `RSM` or
`PCM` fitted by `MML` under the fixed standard-normal scoring basis. It does
not currently create portable artifacts for estimated-population or latent-
regression fits, `JML`, or bounded `GPCM`.

Supported direct and group facet anchors and two-way facet interactions are
retained in the calibration. New responses must use its recorded non-Person
facet levels. Preserving an interaction for scoring does not establish its
statistical significance or the model's suitability for another population.

```{r capabilities}
library(mfrmr)

mfrm_calibration_capabilities()[, c(
  "Model", "Estimator", "ScoringBasis", "PortableCalibration"
)]
```

## Calibration session

The bundled `example_core` data are synthetic. This example uses 18 Persons
to keep vignette runtime modest; a substantive analysis should use its planned
sample and design rather than copying that count.

```{r create-calibration}
synthetic <- load_mfrmr_data("example_core")
person_ids <- unique(as.character(synthetic$Person))
training_ids <- person_ids[seq_len(18L)]
training <- synthetic[
  as.character(synthetic$Person) %in% training_ids,
  ,
  drop = FALSE
]

fit <- suppressWarnings(fit_mfrm(
  training,
  person = "Person",
  facets = c("Rater", "Criterion"),
  score = "Score",
  model = "RSM",
  method = "MML",
  quad_points = 5,
  maxit = 20
))

quadrature_review <- suppressWarnings(mml_quadrature_sensitivity(
  fit,
  training,
  quad_points = c(5, 7),
  theta_points = 41
))
summary(quadrature_review)
fit <- quadrature_review$fits$q7

draft <- extract_mfrm_calibration(
  fit,
  calibration_id = "portable-rsm-example",
  source_fit_id = "synthetic-rsm-fit",
  scoring_quad_points = 31,
  quadrature_review = quadrature_review
)
review_mfrm_calibration(draft)

validated <- validate_mfrm_calibration(draft)
frozen <- freeze_mfrm_calibration(validated)
frozen

artifact_file <- tempfile(fileext = ".rds")
save_mfrm_calibration(frozen, artifact_file)
```

The fit-time quadrature and the scoring-time quadrature are distinct. The two
small fit grids above keep this example quick; they are not recommendations
for substantive work. Inspect the continuous movement on grids chosen for the
application and extend them when needed. Public extraction accepts the exact
highest-grid fit in that review but does not decide whether its movement is
small enough. The artifact separately records the 31-point grid used for later
posterior scoring, and the response-linked review should be archived outside
the portable artifact when it is needed for an audit trail.

## Fresh scoring session

In operational use, start a new R process and supply only the artifact path and
new response rows. The following chunk removes the fit, draft, validation
object, and training data before loading the artifact, so none can influence
the score.

```{r score-from-artifact, fig.alt = "Posterior EAP intervals for two newly scored Persons. A blue circle marks a conditionally ready score, an orange triangle marks a score requiring review, and a dashed vertical line marks the zero-logit scale origin."}
new_rows <- synthetic[
  as.character(synthetic$Person) %in% person_ids[19:20],
  c("Person", "Rater", "Criterion", "Score"),
  drop = FALSE
]
new_ids <- stats::setNames(
  c("NEW_PERSON_1", "NEW_PERSON_2"), person_ids[19:20]
)
new_rows$Person <- unname(new_ids[as.character(new_rows$Person)])
rownames(new_rows) <- NULL

# Create one explicit review case for the tutorial figure. In operational
# scoring, never alter observed responses to manufacture a disposition.
new_rows$Score[new_rows$Person == "NEW_PERSON_2"] <- max(training$Score)

rm(fit, quadrature_review, draft, validated, frozen, training, synthetic)
invisible(gc())

calibration <- load_mfrm_calibration(artifact_file)
scores <- score_mfrm_calibration(
  calibration, new_rows, adaptive_quad_points = c(31, 61)
)
summary(scores)

plot(
  scores,
  type = "interval",
  preset = "publication"
)

unlink(artifact_file)
```

The base and optional ggplot2 renderers distinguish scored and review states
by shape as well as colour. The complete score and review tables remain in the
summary object even though its console printout shows only a compact preview.

The interval figure is the first score-batch review. Orange triangles identify
Persons whose `Disposition` is `scored_review`; inspect their `ReasonCodes`
before using the estimate. A Person with no valid response has no score
coordinate and therefore appears in `summary(scores)$review` and in the
draw-free plot payload's `unplotted_dispositions`, not as an artificial point.

Two focused follow-up views use the same score object without consulting the
source fit:

```{r score-review-payload, eval = FALSE}
# Valid response rows versus posterior SD.
plot(scores, type = "precision", preset = "publication")

# Posterior mass on the outer quadrature nodes versus the recorded threshold.
edge_review <- plot(scores, type = "edge_mass", draw = FALSE)
edge_review$data$selection_summary
edge_review$data$unplotted_dispositions
```

These displays review a returned score batch. They do not establish source-fit
quality, reliability, calibration transport, or validity. Every interval and
posterior SD remains conditional on the frozen point calibration and excludes
calibration-parameter uncertainty.

A small edge mass does not guarantee accurate integration: a narrow posterior
can fall between nodes near the middle of the grid. The optional
`adaptive_quad_points` above adds a separate numerical comparison, placing
nodes around each scored Person's posterior mode and scaling their spacing
to its local width. It holds the calibration parameters and prior fixed.

```{r score-integration-review}
summary(scores)$quadrature_overview
# Keep full precision when inspecting individual differences.
head(scores$quadrature_review[c(
  "Person", "AdaptiveNodes", "EAPChange", "PosteriorSDChange",
  "AdaptiveEAPChangeFromPrevious", "AdaptiveSDChangeFromPrevious", "Status"
)])
```

Check fixed-versus-adaptive differences and changes between adaptive orders,
including log-marginal likelihood contributions in the full table. Observation
weights, when supplied, enter the likelihood as in the original scoring call.
The first adaptive order
has no previous order, so its change columns are `NA`. A `computed` status
means the calculation finished; it does not certify accuracy. An `unavailable`
row retains the reason in `Detail`. Persons with no scored responses remain
in the ordinary disposition review and have no numerical integration row.
Archive the full table with the score batch when reporting numerical checks.
The comparison does not replace returned scores, intervals, or the artifact's
stored grid, and it does not change readiness. Use the same option in
`predict_mfrm_units()` or `mml_quadrature_sensitivity()` to review a fitted
object; the latter also retains its separate comparison of refitted models.

To use adapted grids during fitting and subsequent scoring, set
`mml_integration = "adaptive"` in the `fit_mfrm()` call above. The direct MML
engine then updates the grids while estimating parameters. The refit review
retains this mode, and the frozen artifact records
`scoring_algorithm = "adaptive_quadrature_eap_v2"`. Its stored nodes and weights
define the base Hermite rule; scoring places that rule around each new Person's
posterior. A small edge mass on an adapted grid is still not an accuracy
certificate. Keep the order comparison and report the integration mode.
Ordinary fitting defaults to `mml_integration = "fixed"`, and existing fixed-grid
artifacts retain their scoring algorithm.

New v2 scoring uses continuous equal-tail posterior intervals: a 95% interval
leaves 2.5% of the continuous posterior below each lower endpoint and above
each upper endpoint. EAP and SD still use the stored quadrature rule. Older
v1 artifacts retain their discrete grid endpoints for reproducibility and
carry a note that their continuous posterior mass may differ from the requested
level. A frozen point calibration's intervals exclude calibration-parameter
uncertainty; they do not guarantee 95% coverage at every fixed true ability.

Printed results and interval plots state the requested level and identify
saved grid-based intervals. Both `scores$estimates` and
`summary(scores)$estimates` retain `ScoringAlgorithm` and `IntervalLevel` for
CSV export, alongside the calibration and uncertainty basis. `IntervalLevel`
is the requested posterior probability, not measured frequentist coverage,
and is not rounded by `summary()`. For older saved score results, re-running
`summary()` recovers these fields from the result's stored settings without
recomputing scores.

The scoring prior is also an assumption about the target population. For
example, a new population with a wider ability distribution than the recorded
standard normal prior can have less than 95% of its true abilities inside
these intervals, even when the numerical integration is accurate. Increasing
the quadrature order checks approximation under the same prior; it does not
establish that this prior fits the new population. Report the scoring prior
and explain its relevance to the population receiving scores, alongside the
calibration and integration settings.

An intercept-only normal population fitted to training data can adjust the
scoring mean and variance, but it does not learn a skewed population shape.
Its score intervals still condition on the estimated calibration and population
parameters and exclude uncertainty from estimating them. Training and scoring
populations also need not match. Such fitted-object scoring currently requires
`readiness_policy = "review"` in `predict_mfrm_units()` and cannot be exported
as a portable calibration; see the latent-regression workflow in `?fit_mfrm`.

For example, a calibration may come from last year's cohort while this year's
cohort has a different ability distribution. With an intercept-only population
model, scoring the new response batch retains the fitted mean and variance;
it does not estimate a new batch population. Source readiness alone therefore
does not establish that the prior represents this year's people. Identify the
training and scoring cohorts in the report, explain why the scoring prior can
be applied to the new cohort, and examine sensitivity when that assumption is
uncertain. A larger training sample can estimate last year's population more
precisely without resolving a difference between the two populations.

For a separate scoring script, the essential code is:

```r
library(mfrmr)

calibration <- load_mfrm_calibration("reviewed-calibration.rds")
new_rows <- read.csv(
  "new-responses.csv",
  colClasses = "character",
  na.strings = "",
  check.names = FALSE,
  fileEncoding = "UTF-8-BOM"
)
scores <- score_mfrm_calibration(calibration, new_rows)
print(scores)
plot(scores, type = "interval", preset = "publication")
write.csv(scores$estimates, "person-scores.csv", row.names = FALSE)
```

Read identifiers as text: `001` and `1` can identify different Persons, and
`NA` can be a literal identifier. The example treats empty cells as missing;
if your response file uses another missing-score token, convert it only in
the score column before scoring. Do not let automatic CSV type conversion
merge people or change the frozen facet labels. Numeric score strings are
validated against the frozen score map by the scorer.

Keep `scores` with `saveRDS()` when the full review and numerical settings are
needed. The CSV above preserves per-score identities and uncertainty labels,
but does not contain the complete row review or optional integration review.
When reading that CSV back, also preserve identifier columns as text.

Unknown facet levels, unknown score values, invalid weights, ambiguous
duplicates, and malformed artifacts fail closed. Missing responses require an
explicit `missing_response = "omit"` policy.

## If scoring stops or asks for review

An input error stops the scoring call. A returned `scored_review` result has
an estimate that needs review; `not_scored` means no estimate is available.
Inspect `summary(scores)$review`, including Persons absent from the interval
plot. The printed review translates the reason codes into descriptions.

| What you see | What to check or change |
| --- | --- |
| An unseen rater, task, or other facet level | Check spelling and preserve IDs as text. A genuinely new level has no parameter in this calibration: return to calibration with an appropriate linking design. Do not rename it as an existing level or edit the frozen artifact. |
| A score outside the frozen map | Check the original rubric and missing-score codes. Correct an entry error or explicitly recode a documented missing marker. A changed rubric requires calibration review; widening a score range cannot update the saved map. |
| Missing scores or IDs | Correct unintended missing values. Explicit omission applies to missing scores, not missing Person or facet IDs, and does not correct selective nonresponse. Review omitted counts after rescoring. |
| Duplicate response cells | Check whether rows are accidental copies or distinct rating events. Correct copies; use the documented `event_id` argument only for genuine distinct events under the intended model. New event IDs do not model dependence between repeated performances. |
| A missing, nonfinite, or nonpositive weight on a scored row | Check the weight column and its intended meaning. Correct erroneous values; do not replace them with arbitrary constants just to obtain scores. |
| A malformed or incompatible artifact | Restore the original file and use a compatible package version. If it must be recreated, return to the source fit and the documented extraction/review workflow. Do not edit schema or validation fields to bypass the refusal. |
| Review because responses are very few, omitted, or all at one endpoint | Check the original ratings and valid/omitted counts. Interpret precision and dependence on the prior before using the estimate; further ratings may be needed. These flags do not by themselves diagnose an invalid Person or justify changing responses. |
| Review because posterior mass is near integration limits | Examine the numerical comparisons described above. Increasing optimizer iterations cannot alter a frozen calibration. If scoring settings need revision, return to the calibration workflow and retain a new artifact and its review. |
| No valid responses (`not_scored`) | Correct missing-score coding if erroneous, or obtain valid ratings. Retain the unavailable result; do not substitute a zero-logit score. |

A completed numerical comparison does not automatically clear these review
states. Record the reason, the action taken, and the calibration used with the
score batch.

Every returned estimate is a posterior EAP conditional on the recorded fixed
point calibration and prior;
its uncertainty interval excludes calibration-parameter uncertainty. Validate
the intended score use, population transport, and decision consequences
outside this numerical workflow.
