---
title: "Migrating from Facets to mfrmr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Migrating from Facets to mfrmr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

This vignette walks Facets users through the equivalent `mfrmr` workflow:
preparing data, fitting a many-facet Rasch model with Facets-compatible
defaults, generating the diagnostic and reporting tables that the canonical
Facets output stack provides, and verifying numeric parity between the two
systems.

## Mental model

The two stacks share the same psychometric framework but differ in
operating model.

| Concept | Facets (Linacre 2026) | mfrmr |
|---|---|---|
| Input | Specification file plus data file | `data.frame` in long format |
| Estimation | JMLE by default | `JML` (legacy default) or `MML` (recommended for new analyses) |
| Models | Rating-scale, partial-credit, polytomous step models | `RSM`, `PCM`, bounded `GPCM` |
| Output | Tables 0-30 plus graphic files | Returned R objects with `summary()` and `plot()` methods |
| Anchoring | `D=`, `A=` fields in the specification | `anchors` and `group_anchors` arguments to `fit_mfrm()` |
| Bias / interaction | Table 14 | `estimate_bias()` and `bias_interaction_report()` |
| Fair average | Table 7 fair-M average | `fair_average_table()` |
| Reproducibility | Specification file is the manifest | `build_mfrm_manifest()` plus `build_mfrm_replay_script()` |

## A one-shot legacy-compatible call

If the goal is to reproduce a Facets-style script with minimal R-side
plumbing, use `run_mfrm_facets()` (alias `mfrmRFacets()`):

```{r facets-mode}
library(mfrmr)
data("ej2021_study1", package = "mfrmr")

run <- run_mfrm_facets(
  data = ej2021_study1,
  person = "Person",
  facets = c("Rater", "Criterion"),
  score = "Score",
  model = "RSM",
  method = "JML"
)

names(run)
```

The wrapper returns the same `fit_mfrm()` and `diagnose_mfrm()` objects that
a step-by-step pipeline produces, plus the iteration log, fair-average
table, and rating-scale table:

```{r facets-mode-summary}
summary(run$fit)
head(run$fair_average)
```

For new analysis scripts, prefer `fit_mfrm(method = "MML")` directly. MML
integrates over the person distribution under an N(0, 1) prior and exposes
per-person posterior SEs that JML cannot produce.

## Translating the specification file

The mapping below covers the most common Facets specification keywords.

### Facets and labels

```
Facets = 3
Models = ?,?,?,R5
Labels =
  1, Examinee
    1 = P01
    ...
  2, Rater
    1 = R1
    ...
  3, Criterion
    1 = Content
    ...
```

translates to:

```r
fit_mfrm(
  data = examinee_long,
  person = "Examinee",
  facets = c("Rater", "Criterion"),
  score = "Score",
  rating_min = 1,
  rating_max = 5,
  model = "RSM"
)
```

`Models = ?,?,?,R5` becomes `model = "RSM"` and the `R5` rating-scale
declaration becomes `rating_min = 1, rating_max = 5`. For a partial-credit
specification, pass `model = "PCM"` and identify the facet that carries the
step thresholds with `step_facet = "Rater"` (or the appropriate facet
name).

### Anchoring

A Facets `D = 2, A =` block:

```
D = 2
A = 1, 0.0
    2, 0.5
```

becomes an `anchors` data frame:

```r
anchors <- data.frame(
  facet = "Rater",
  level = c("R1", "R2"),
  estimate = c(0.0, 0.5),
  stringsAsFactors = FALSE
)
fit <- fit_mfrm(..., anchors = anchors)
```

`audit_mfrm_anchors()` validates and reports on the anchor block before the
fit runs, surfacing connectivity, overlap, and minimum-sample issues.

### Bias and interaction

Facets Table 14 bias output between Rater and Criterion has a direct
equivalent:

```r
diag <- diagnose_mfrm(fit)
bias <- estimate_bias(fit, diag,
                      facet_a = "Rater", facet_b = "Criterion")
summary(bias)
```

`estimate_all_bias()` enumerates every non-person facet pair in one call.

### Group anchoring and DFF

Facets `D = ..., G =` group-anchor blocks for differential facet
functioning translate to the `group_anchors` argument and the
`analyze_dff()` follow-up:

```r
group_anchors <- data.frame(
  facet = "Criterion",
  level = "Content",
  group = c("Native", "Non-native"),
  estimate = c(0.0, 0.0),
  stringsAsFactors = FALSE
)
fit_g <- fit_mfrm(..., group_anchors = group_anchors)
dff <- analyze_dff(fit_g, diag, facet = "Criterion",
                   group = "FirstLanguage", method = "refit")
```

## Verifying numeric parity

When migrating an existing study, `facets_parity_report()` compares
the parameter estimates and fit statistics produced by the two systems:

```r
parity <- facets_parity_report(
  fit,
  facets_estimates = facets_table_csv,
  tolerance = 0.05
)
summary(parity)
```

The resulting report lists per-element differences and flags rows that
exceed the supplied tolerance. Use it as the sign-off step in a migration:
once `parity$flagged` is empty under a tolerance you accept, the two stacks
agree on the substantive output.

## Producing Facets-style output files

For audit trails or downstream tools that expect Facets output files,
`facets_output_file_bundle()` writes a parallel set of fixed-width or CSV
exports:

```r
files <- facets_output_file_bundle(
  fit,
  diagnostics = diag,
  out_dir = tempdir(),
  include = c("graph", "score")
)
```

For RSM and PCM the score-side helpers are available. Under bounded `GPCM`
the score-side bundle is intentionally restricted; see
`?gpcm_capability_matrix` and the `mfrmr-gpcm-scope` vignette for the
binding contract.

## Recommended next steps

After a Facets-equivalent fit is in hand, the canonical mfrmr reporting
route extends the analysis with:

- `diagnose_mfrm(diagnostic_mode = "both")` for the strict marginal screen
  alongside the residual stack.
- `reporting_checklist()` for a manuscript-readiness summary.
- `build_apa_outputs()` for Method and Results paragraphs and APA tables.
- `build_mfrm_manifest()` and `build_mfrm_replay_script()` for the
  reproducibility bundle that Facets specifications cannot produce
  out of the box.

The `mfrmr-workflow` vignette covers the full sequence end to end; the
`mfrmr-reporting-and-apa` vignette focuses on the manuscript surface; the
`mfrmr-linking-and-dff` vignette covers anchoring, drift, and DFF in
detail.
