---
title: "Missing scores on assigned ratings"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Missing scores on assigned ratings}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
is_cran_check <- !isTRUE(as.logical(Sys.getenv("NOT_CRAN", "false")))
knitr::opts_chunk$set(eval = !is_cran_check, collapse = TRUE, comment = "#>")
```

An assessment team has missing scores from ratings that were assigned. It
wants to compare two raters' severity while allowing for uncertainty about
those scores. Other performances were never assigned to one rater. Those
unassigned combinations are part of the design, not scores to invent.

This example uses fictional assessment data, a specified joint rating-scale
imputation model, separate MFRM analyses and a pooled rater difference. It
compares the result with direct inference from observed scores and changes
the missing-score assumption in a sensitivity analysis. The saved completions
are supplied with the package, so the analysis runs without installing a
Bayesian sampler. An accompanying R/Stan script regenerates them if required.
These functions are available in version 0.2.4.

## Keep assignment separate from missing scores

The original data have 48 persons, four raters and four criteria. Every rating
has an event ID. Six R04 Content events are unassigned. Among the remaining
R04 Content and Language events, scores are masked with probability
`plogis(-0.5 + (2.5 - mean_R01_score))`. All R01 scores remain observed.
The fixed random seed selects 30 missing assigned scores; five persons have
two missing scores each. Repeated ratings would need different event IDs even
if their person, rater and criterion were identical.

This missingness mechanism depends only on observed scores and is MAR by
construction. It illustrates ignorable missingness, not a test showing that
missingness in a real assessment is MAR. Hidden scores are saved solely to
check predictions; they are never supplied to the imputation model.

```{r roster}
library(mfrmr)
example <- readRDS(system.file("examples", "response-imputation.rds", package = "mfrmr"))
ratings <- example$ratings
missing_assigned <- example$missing
unassigned <- example$unassigned
held_out_scores <- example$held_out
impute_ids <- ratings$Event[missing_assigned]
m <- ncol(example$scores)
model <- example$imputation_model
with(ratings, table(Assigned, MissingScore = is.na(Score)))
```

Absent roster rows remain absent. If the input contains assigned events only,
omit the `assigned` argument. To deliberately exclude other assigned
nonresponse from imputation, retain it as missing and explicitly request
`missing = "omit"`. Review omission counts and their substantive consequences;
exclusion can still bias an analysis.

## Use a joint score model

Within each completion, the imputation model draws:

1. One joint set of rater severities, criterion difficulties and shared steps,
   allowing for their posterior uncertainty and covariance.
2. One latent ability for each person, informed by that person's observed
   ratings. All missing ratings of that person share this ability draw.
3. A category for each selected missing rating, conditional on those draws.

This retains uncertainty in both calibration and missing scores, as well as
dependence between missing ratings. Plugging in an EAP or drawing a separate
ability independently for each missing rating would be a different procedure.
Observed scores and unassigned events never change.

For a category coded internally as \(k=0,1,2,3\), the model is

\[
P(Y_{prc}=k\mid\theta_p,\rho_r,\beta_c,\tau)
\propto \exp\left\{k(\theta_p-\rho_r-\beta_c)
                 -\sum_{h=1}^{k}\tau_h\right\}.
\]

Persons are independent draws from a **known standard-normal population**.
The rater, criterion and step vectors each sum to zero, matching the default
completed-data analysis below. The model is an adjacent-category RSM;
its steps are not required to be ordered cumulative-logit cutpoints.
Orthonormal sum-zero coordinates for calibration have independent
`normal(0, 2.5)` priors (2.5 is the SD). This gives a proper, level-symmetric
prior on each centered vector. It does not turn the fixed raters into a
population of exchangeable random-rater effects with an estimated variance.

These are assumptions of this example, not automatic defaults for arbitrary
assessments. Review the population distribution, category model and dependence
structure. Covariates or additional effects can be needed when nonresponse
relates to information outside this model. Several tasks, shared random
raters or local dependence may require a different joint imputation model.

### Inspect sampling and preserve provenance

The saved result contains 40 completions selected at random without replacement
from 8,000 retained draws in four well-mixing chains. The generation script
requires Rhat below 1.01, bulk/tail effective sample sizes of at least 400,
no divergences or maximum-tree-depth hits and E-BFMI of at least 0.3. It saves
all chains before reviewing these diagnostics and stops if they fail.

```{r sampling-diagnostics}
diagnostics <- model$diagnostics
c(MaxRhat = max(diagnostics$rhat),
  MinBulkESS = min(diagnostics$ess_bulk),
  MinTailESS = min(diagnostics$ess_tail))
model$sampler
head(model$draw_ids)
```

Good sampling diagnostics do not prove that the response model describes the
assessment. Random selection approximates independent posterior draws here;
it cannot repair a poorly mixing chain. Forty completions are an example
setting: examine Monte Carlo error relative to the decision's required
precision. More completions reduce simulation noise, not model bias.

To regenerate the example, use the accompanying script with an existing
[CmdStanR/Stan installation](https://mc-stan.org/cmdstanr/). This optional step
compiles and samples the model; importing the supplied completions and running
the analyses below does not require that toolchain. The script is specific to
this example, not a general imputation API.

```{r regenerate, eval=FALSE}
source(system.file("examples", "response-imputation.R", package = "mfrmr"))
example <- response_imputation_example("mi-example-output")
```

The output directory retains the model code, data, sampler settings, all
chains and diagnostics. Use a new directory for a new run. Draw identities,
selected scores and model specification are also retained in the compact
example file. The full chains can be used for further posterior checks.

## Return completions to the original roster

Only eligible event IDs receive scores. A long-format `mids` object from
another suitable imputation model can instead be supplied directly; the API
checks its original data and `where` mask. Common category labels or a
successful import do not establish compatibility with the analysis model.

```{r completions}
completed <- lapply(seq_len(m), function(i) {
  result <- ratings
  result$Score[missing_assigned] <- example$scores[, i]
  result
})
review <- review_mfrm_imputations(
  ratings, completed, person = "Person", facets = c("Rater", "Criterion"),
  score = "Score", event_id = "Event", impute_ids = impute_ids, categories = 1:4,
  assigned = "Assigned", imputation_model = model
)
review
subset(summary(review), Facet == "Rater")
stopifnot(all(vapply(review$completed, function(d) {
  all(is.na(d$Score[unassigned])) &&
    identical(d$Score[!is.na(ratings$Score)], ratings$Score[!is.na(ratings$Score)])
}, logical(1))))
```

The support table separates observed and imputed information. A person or
rater with no observed scores depends especially strongly on the imputation
assumptions. Imputed links are not evidence of empirical overlap. All selected
scores must be completed in every imputation; no invalid completion is dropped.

## Check the missing-score predictions

The saved category probabilities average all 8,000 posterior draws. They are
more precise simulation summaries than the category frequencies from just 40
completions. Labels show percentages as well as color, so the figure can be
read without distinguishing colors. Each row is a marginal prediction;
dependence between rows is preserved in the joint completions.

```{r imputation-diagnostics, fig.width=6.5, fig.height=8, fig.alt="Category probabilities for thirty missing assigned ratings; each cell shows its percentage."}
probabilities <- example$probability_mean
n <- nrow(probabilities)
image(1:4, seq_len(n), t(probabilities[n:1, , drop = FALSE]),
      zlim = c(0, 1), col = colorRampPalette(c("#ffffd9", "#7fcdbb", "#41b6c4"))(64), axes = FALSE,
      xlab = "Score", ylab = "", main = "Predicted category probabilities")
axis(1, at = 1:4)
axis(2, at = seq_len(n), labels = rev(impute_ids), las = 1, cex.axis = .7)
for (j in 1:4) text(j, seq_len(n),
  labels = sprintf("%.0f%%", probabilities[n:1, j] * 100), cex = .65,
  col = "black")
```

Rounded zero percentages do not mean impossible categories. In addition to
sampling diagnostics, examine whether predictions describe the assessment.
Because we deliberately masked scores, we can compare the predictions with
held-out values. The Brier score compares probabilities with the observed
category; smaller is better.

```{r held-out-check}
prediction_check <- data.frame(
  Event = impute_ids, HeldOutScore = held_out_scores,
  PredictedMean = drop(probabilities %*% (1:4)),
  Brier = rowSums((probabilities - outer(held_out_scores, 1:4, "=="))^2)
)
head(prediction_check)
c(HeldOutMean = mean(held_out_scores),
  PredictedMean = mean(prediction_check$PredictedMean),
  Brier = mean(prediction_check$Brier))
```

The held-out mean is 2.10 and the predicted mean about 2.24. This small check
makes prediction discrepancies visible, but cannot establish population bias,
probability calibration or coverage of a rater interval. The RSM's population
and dependence assumptions still require substantive review.

## Analyze every completion on the same scale

The declared 1--4 category ladder and identical facet constraints are retained
in every fit. This workflow currently supports fixed-standard-normal RSM/PCM
MML analysis. PCM requires `step_facet = "Criterion"` explicitly. Shared anchors
can be supplied, but are treated as known: their uncertainty is not included.

```{r analyses}
analyses <- fit_mfrm_imputed(review, model = "RSM", quad_points = 61)
summary(analyses)[, c("Imputation", "Status", "Error", "Warnings")]
```

Before interpretation, review model fit, observed support and quadrature
sensitivity using the retained fits. The default quadrature grid is a starting
point, not an adequacy certificate. A fit can complete computation and still
be ineligible for inference. `analysis_summary` retains errors and warnings;
pooling requires every fit to be eligible and to have unregularized observed
information. Increasing the number of imputations does not cure a flawed
imputation model or a failed measurement analysis.

## Pool the rater contrast, including its covariance

For feedback, a prespecified difference is often more interpretable than an
isolated centered severity. Here a positive R04-minus-R01 value means that R04
is more severe under the default facet orientation. Differences are on the
common logit scale. The covariance between the two rater estimates contributes
to their difference's variance in every completed analysis.

```{r pooled-result}
levels <- analyses$fits[[1]]$config$facet_specs$Rater$levels
contrast <- matrix(0, 1, length(levels),
                   dimnames = list("R04 minus R01", levels))
contrast[1, c("R04", "R01")] <- c(1, -1)
pooled <- pool_mfrm_imputed(analyses, facet = "Rater", contrasts = contrast)
summary(pooled)
```

```{r pooled-plot, fig.width=7, fig.height=3, fig.alt="Pooled R04-minus-R01 severity contrast with a pointwise Rubin t interval incorporating within- and between-imputation covariance. The result is conditional on the supplied imputations and analysis assumptions."}
plot(pooled)
```

The development version also supports a ggplot made from these saved intervals:

```{r pooled-ggplot, fig.width=7, fig.height=4, fig.alt="The saved R04-minus-R01 contrast and its pointwise multiple-imputation t interval, rendered with ggplot. Changing the title does not recalculate the interval or change its confidence level."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  pooled_figure <- as_ggplot(pooled, title = NULL, preset = "monochrome")
  print(pooled_figure)
  plot_data(pooled_figure)$table[, c("Target", "Estimate", "Lower", "Upper", "DF")]
}
```

Default and `component = "table"` conversion keep the complete interval view.
The t interval is not replaced by a normal interval from `SE`; changing its
level requires a new, explicit pooling call. Fixed targets are open diamonds
without inferential intervals. Missing intervals have crosses, and infinite
endpoints have arrows whose tips are drawing limits rather than finite bounds.
Full contrasts, settings, degrees of freedom and information cautions remain in
`plot_data()`, even after titles or captions are hidden. These display symbols
do not establish interval coverage or classify rater quality.

`Estimate` averages the completed-data estimates. `WithinVariance` averages
their model-based variances; `BetweenVariance` captures their variation across
imputations. `TotalVariance` includes both, with the finite-imputation factor
`1 + 1/m`. `SE` and the pointwise t interval use that total.
`MonteCarloSE` instead quantifies simulation error in the pooled point estimate.
`MissingVarianceFraction` is not the percentage of missing scores.

The default `df_complete = Inf` uses large-sample complete-data inference.
There is no automatic conversion of rating-row counts into independent sample
size. A finite `df_complete` is available when the target has a justified
complete-data degrees-of-freedom approximation; it does not create one.
Fixed constraints are reported as fixed, without confidence intervals.

These are model-based MI intervals, treating any supplied anchors as known.
They do not guarantee
finite-sample coverage, robustness to misspecification, or simultaneous control
over a selected collection of rater flags. A severity difference is not a
measure of rater quality. Feedback should use content, observed behavior and
practical importance along with the uncertainty.

Person EAPs and posterior SDs remain in the separate fitted objects. They are
not ordinary parameter estimates and standard errors for the pooling operation
above. Neither averaged imputed scores nor Rubin-pooled EAPs substitute for an
appropriately specified person-scoring target.

## Compare with inference from the observed scores

If only scores are missing, the observed-score likelihood can already estimate
this contrast under the same RSM and ignorable missingness assumptions.
Multiple imputation does not add observed information. It is useful when
completed rosters are needed for a downstream analysis or an explicit
missing-score sensitivity assumption. For this single contrast, direct
observed-score MML is a simpler starting point.

Keep the full roster and its missingness review. Here we explicitly select its
732 observed assigned scores after the known masking procedure. This selection
is justified by the example's construction; it is not a remedy for unreviewed
nonresponse, missing identifiers or a disconnected design. Passing the roster
with `NA` scores directly to `fit_mfrm()` retains an input-review state, which
the interval helper will ask you to resolve.

```{r observed-comparison}
observed_ratings <- subset(ratings, Assigned & !is.na(Score))
observed_fit <- fit_mfrm(observed_ratings, "Person", c("Rater", "Criterion"),
  "Score", model = "RSM", method = "MML", rating_min = 1, rating_max = 4,
  quad_points = 121, keep_original = TRUE, attach_diagnostics = FALSE)
observed_interval <- mfrm_facet_intervals(observed_fit, "Rater", contrasts = contrast)
comparison <- rbind(
  data.frame(Method = "Observed-score MML",
             summary(observed_interval)[c("Estimate", "SE", "Lower", "Upper")]),
  example$reference,
  data.frame(Method = "Joint-model MI",
             summary(pooled)[c("Estimate", "SE", "Lower", "Upper")])
)
comparison
```

The saved Bayesian reference comes from the same observed-score posterior
used for imputation. Its `SE` column is a **posterior SD**, and its bounds are
an equal-tailed credible interval; the other rows use frequentist interval
approximations. In this example, the three estimates are about 0.46--0.47 logits
with uncertainty scales about 0.14--0.15 logits. The MI estimate's Monte Carlo
SE is about 0.008 logits. Increasing from 61 to 121 quadrature points changes
the direct MML estimate by less than 0.000002 logits.

This agreement is reassuring for this calculation, but does not prove
congeniality or repeated-sampling coverage. The imputer has proper calibration
priors; the completed-data analyses use MML estimates and inverse observed
information without those priors. Treating the latter as Bayesian posterior
moments is a large-sample approximation. Prior influence, sparse support and
model misspecification can make that approximation inadequate. Merely sharing
the response likelihood is insufficient; see Bartlett and Hughes (2020).

## What repeated sampling showed

A separate simulation examined this joint-RSM completion workflow for a fixed
rater contrast. It used 200 independent datasets, each with 80 persons, three
raters, two criteria and scores 0--2. Twenty ratings were structurally
unassigned. Missing scores were concentrated in the third rater's ratings;
forty completions were used per incomplete roster. The true R1-minus-R3
contrast was 0.5 logits, and the known standard-normal ability distribution
and RSM were correctly specified.

Each dataset was analyzed under two missingness mechanisms. In the MAR
condition, missingness depended on the first rater's observed scores. In the
MNAR condition, it additionally depended on the missing rating's own value,
with lower scores more likely to be missing. Both conditions were analyzed
using the same MAR imputer. This second condition deliberately violates the
missingness assumption; it does not evaluate an MNAR imputation model.

| Missingness | Procedure | Available / 200 | Coverage, % (95% MC interval) | Mean bias, logits | Mean interval width, logits |
| --- | --- | ---: | ---: | ---: | ---: |
| MAR | Observed-score MML | 200 | 97.0 (93.6--98.9) | -0.007 | 0.813 |
| MAR | Observed-score Bayes | 198 | 97.0 (93.5--98.9) | -0.004 | 0.814 |
| MAR | Joint-model MI | 198 | 96.5 (92.9--98.6) | -0.004 | 0.817 |
| MNAR | Observed-score MML | 200 | 25.0 (19.2--31.6) | +0.574 | 0.859 |
| MNAR | Observed-score Bayes | 200 | 24.5 (18.7--31.1) | +0.578 | 0.859 |
| MNAR | Joint-model MI | 200 | 26.0 (20.1--32.7) | +0.574 | 0.864 |

Coverage is the fraction of available nominal 95% intervals containing the
true contrast. The parentheses describe **Monte Carlo uncertainty in that
coverage rate**, not an interval for a rater's severity. The Bayesian row uses
credible intervals. Two MAR posteriors missed the prespecified Rhat <1.01
criterion (1.0106 and 1.0104); their Bayesian and MI results were unavailable,
and neither case was replaced. Counting these as unsuccessful results, MI
returned an interval containing the truth in 191/200 trials (95.5%), compared
with 191/198 (96.5%) among available intervals. All 15,920 completed-data
MML fits from the retained posteriors were eligible.

In this MAR design, all three methods met the prespecified bounded coverage,
availability and bias criteria. For MI, the bias MC interval was
[-0.033, 0.025] logits. The mean point-estimate Monte Carlo SE was about 8%
of the inferential SE. This supports this specific workflow under the tested
assumptions, not a universal nominal-coverage guarantee, arbitrary supplied
imputations or a validation of other populations, sparse designs or model
families.

The two conditions had almost identical average missing fractions: **14.14%
and 14.19% of assigned ratings**. Nevertheless, the MNAR condition produced
large positive bias and very poor coverage for all three methods. MI and
direct inference can agree closely while both miss the target. Missingness
concentrated in one rater also makes the overall missing fraction an
insufficient description. More completions cannot correct the wrong
missing-score distribution. Review the recording process, relevant auxiliary
information and justified sensitivity assumptions before substantive feedback.

## Challenge the missingness assumption

The baseline uses the same conditional score model for observed and missing
assigned ratings, given the observed ratings and common latent ability. To examine a specific departure,
suppose an unrecorded score would be **one category lower** than its baseline
imputation, with the lowest category retained at 1. This is a deliberately
strong, discrete pattern-mixture sensitivity scenario. It is not an estimated
MNAR mechanism, a universal correction or an instruction to subtract a
category from real ratings. Its size and direction need subject-matter
justification in an actual assessment.

Use the same baseline draws to isolate this change. Only the selected
scores change; observed and unassigned events stay as they were. Analyze and
pool each assumption separately, rather than treating the assumptions as
additional imputations of one model.

```{r sensitivity}
lower_scores <- lapply(completed, function(d) {
  d$Score[missing_assigned] <- pmax(1L, d$Score[missing_assigned] - 1L)
  d
})
lower_review <- review_mfrm_imputations(
  ratings, lower_scores, person = "Person", facets = c("Rater", "Criterion"),
  score = "Score", event_id = "Event", impute_ids = impute_ids, categories = 1:4,
  assigned = "Assigned", imputation_model = list(
    baseline = model, assumption = "Missing scores one category lower, floored at 1")
)
lower_analyses <- fit_mfrm_imputed(lower_review, model = "RSM", quad_points = 61)
lower_pooled <- pool_mfrm_imputed(lower_analyses, facet = "Rater", contrasts = contrast)
comparison <- rbind(
  cbind(Assumption = "Joint RSM baseline", summary(pooled)),
  cbind(Assumption = "Missing scores one category lower", summary(lower_pooled))
)
comparison[, c("Assumption", "Estimate", "SE", "Lower", "Upper", "MonteCarloSE")]
paired_change <- drop(lower_pooled$estimates - pooled$estimates)
c(ContrastChange = mean(paired_change),
  MonteCarloSEOfChange = sd(paired_change) / sqrt(m))
```

Read the change in the rater contrast together with its uncertainty and your
prespecified practically important difference. A modest change here would
concern only this 30-event scenario; it would not establish robustness to
other missing scores or mechanisms. A large change would identify dependence
on the assumed missing scores, not prove which assumption is true. Inspect
every analysis's warnings and eligibility before interpreting either row.
Neither an interval overlapping zero nor agreement between two scenarios
validates the imputation model.
The paired Monte Carlo SE quantifies simulation noise in the change between
these two assumptions; it is not a sampling SE for that change.

## Save enough to review and reproduce the result

```{r save, eval=FALSE}
saveRDS(pooled, "rater-imputation-result.rds")
saved <- readRDS("rater-imputation-result.rds")
summary(saved)
write.csv(summary(saved), "rater-imputation-table.csv", row.names = FALSE)
saved$analyses$imputations$imputation_model$diagnostics
saveRDS(list(baseline = pooled, lower_scores = lower_pooled),
        "rater-imputation-sensitivity.rds")
```

The saved result retains the original roster, completions, imputation model,
all fitted analyses and pooling settings. The CSV contains the selected summary,
so retain the RDS alongside it. Use `mfrmr_output_guide("imputation")` for these
dedicated output routes; a pooled result is not an `mfrm_results()` input.
Its plot preserves the stored MI bounds. For custom graphics, extract
`plot_data(saved)$table` and specify the estimate and both endpoints explicitly;
`as_ggplot()` has dedicated conversion for the saved pooled interval view,
including `component = "table"`; use `plot_data()` for other components.
Changing the contrast or confidence
level reuses the fits; changing the imputation model requires new completions
and analyses. Saved objects contain response data and identifiers: choose an
appropriate storage location when using real assessments.

## References

- Andrich, D. (1978). A rating formulation for ordered response categories.
  *Psychometrika*, 43, 561--573. <https://doi.org/10.1007/BF02293814>.
- Rubin, D. B. (1987). *Multiple Imputation for Nonresponse in Surveys*. Wiley.
- Barnard, J., and Rubin, D. B. (1999). Small-sample degrees of freedom with
  multiple imputation. *Biometrika*, 86, 948--955.
  <https://doi.org/10.1093/biomet/86.4.948>.
- Bartlett, J. W., and Hughes, R. A. (2020). Bootstrap inference for multiple
  imputation under uncongeniality and misspecification.
  <https://doi.org/10.1177/0962280220932189>.
- Stan User's Guide: [posterior predictive sampling](https://mc-stan.org/docs/stan-users-guide/posterior-prediction.html)
  and [missing data](https://mc-stan.org/docs/stan-users-guide/missing-data.html).
